---
title: "Adolescent QC"
output: html_notebook
---

```{r setup, include = FALSE, echo = FALSE} 
require(knitr) 
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(phyloseq)
library(tidyverse)
library(vegan) # For rarecurve()
```

# Read in data
```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

```{r}
# Rarefied richness estimates NEED TO REPLACE
# rare.df <- 
#      here('data', 
#           'processed', 
#           'rarefaction',
#           'Rarefied trnL counts (1,000 reads).rds') %>% 
#      readRDS
# 
# rare.df
```
# Pre-process

```{r}
# Subset to Adolescent samples
ps <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0 , .)

ps
```

36 samples here had fewer than 1,000 reads, not included above.

```{r}
alpha <- 
     rare.df %>% 
     filter(study == 'Adolescent') %>% 
     select(name, study, subj, R24_id:o, starts_with('obs'))
```

```{r}
# Add in full set of observations, including samples with fewer than read count threshold
asvtab <- data.frame(ps@otu_table@.Data)
obs <- apply(asvtab, 1, function(x){sum(x > 0)})

add <- 
     data.frame(obs = obs) %>% 
     rownames_to_column(var = 'name')

# Populate with sample data
add <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, R24_id, visit, o) %>% 
     right_join(add, by = c('row' = 'name')) %>% 
     rename(name = row)

alpha <- 
     add %>% 
     filter(!(name %in% alpha$name)) %>% 
     bind_rows(alpha)
```

## Format variables

Convert types for better plotting and downstream analyses

### Age

```{r}
sample_data(ps)$age_years <- 
     ((sample_data(ps)$age_days)/365) %>% 
     floor() %>% 
     factor()

head(sample_data(ps)$age_years)
```

### SES

```{r}
sample_data(ps)$ses <- factor(sample_data(ps)$ses,
                              levels = c('<24,999',
                                         '25,000-49,999',
                                         '50,000-74,999',
                                         '75,000-99,999',
                                         '>100,000'))
```

```{r}
# Make additional variable for over/under 100k
sample_data(ps)$over100k <- ifelse(sample_data(ps)$ses == '>100,000',
                                   yes = '>$100K',
                                   no = '<$100K')
```

```{r}
sample_data(ps)$worry <- factor(sample_data(ps)$worry,
                                levels = c('Never true',
                                           'Sometimes true',
                                           'Often true'))

sample_data(ps)$actual <- factor(sample_data(ps)$actual,
                                 levels = c('Never true',
                                            'Sometimes true',
                                            'Often true'))
```

```{r}
# Make "any" category for worry/actual insecurity
sample_data(ps) <- 
     sample_data(ps) %>% 
     data.frame() %>% 
     mutate(
          across(.cols = c(worry, actual),
                 .fns = ~ ifelse(.x != 'Never true',
                                 yes = 'Yes',
                                 no = 'No'),
                 .names = '{.col}_ever'
                 )
     )
```

## Bind to demographic data

```{r}
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row:reads_tot, R24_id:actual_ever)  # Adolescent metadata
```

```{r}
# Pull from sample data, and join to results
alpha <- 
     samdf %>% 
     select(name = row, 
            age_days, 
            age_years,
            ses, 
            over100k,
            starts_with('worry'),
            starts_with('actual')) %>% 
     right_join(alpha)
```

# Analyze

## Sample demographics

```{r}
samdf <- data.frame(ps@sam_data)
```

```{r}
# How many unique subjects?
n_distinct(samdf$subj)
```

```{r}
samdf %>% 
     select(subj, casecontrol) %>% 
     distinct() %>% 
     count(casecontrol) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, ethnicity) %>% 
     distinct() %>% 
     count(ethnicity) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, race) %>% 
     distinct() %>% 
     count(race) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, ses) %>% 
     distinct() %>% 
     count(ses) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```


## Read depth

## trnL taxa

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(subj) %>% 
     summarize(mean = mean(obs, na.rm = TRUE),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.mean, aes(x = mean, y = mean_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Mean *trnL* richness<br>\n(All samples)', 
          y = 'Mean rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adolescent rarefied vs observed trnL richness (mean).pdf')),
       height = 4, width = 4)
```

### Range

```{r}
adolescent.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(across(c(obs, obs_rare),
                      list(med = ~median(.x, na.rm = TRUE),
                           sd = ~sd(.x, na.rm = TRUE))
                      )
               )

adolescent.stats
```

```{r}
adolescent.stats %>% 
     select(-subj) %>% 
     apply(2, range, na.rm = TRUE)
```

## Ordination

### Jaccard

```{r}
ord <- 
     ordinate(ps,
              method = 'PCoA',
              distance = 'jaccard',
              binary = TRUE)
```

```{r}
p <-  
     plot_ordination(ps,
                     ord,
                     type = 'samples',
                     color = 'subj')

p + theme(legend.position = 'none')
```

```{r}
# Customize plot 
data <- data.frame(ord$vectors)
samdf <- data.frame(ps@sam_data)

data <- bind_cols(data, samdf)

ggplot(data, aes(x = Axis.1, y = Axis.2, color = worry_ever)) +
     geom_point(alpha = 0.8,
                size = 3) +
     labs(x = 'PCo1 (9.1%)',
          y = 'PCo2 (7.9%)') +
     coord_equal() +
     theme_bw() +
     theme(legend.title = element_blank()) +
     theme(axis.line = element_line(size = 1,
                                    color = 'gray80'),
           axis.text = element_text(size = 12,
                                    face = 'bold',
                                    color = 'gray80'), 
           axis.title = element_text(size = 14,
                                     face = 'bold',
                                     color = 'gray80'), 
           axis.ticks = element_line(size = 1, 
                                     color = 'gray80'),
           legend.text = element_text(size = 10,
                                      face = 'bold',
                                      color = 'gray40'),
           panel.border = element_blank(),
           panel.grid.major = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_Metaproteomic ordination (Bray-Curtis).pdf')))
```

```{r}
vegan::adonis(distance(ps, 
                       method = 'jaccard',
                       binary = TRUE) ~ reads_tot,
              data = samdf)
```

## Age

```{r}
# Visualize on the ordinated data
```

```{r}
# Statistical test
# Need to first remove participants with no age data
dist.age <- 
     ps %>% 
     subset_samples(!is.na(age_days)) %>% 
     distance(., 
              method = 'jaccard', 
              binary = TRUE)

samdf.age <- 
     ps %>% 
     subset_samples(!is.na(age_days)) %>% 
     sample_data() %>% 
     data.frame()
```

```{r}
set.seed('1116')
vegan::adonis(dist.age ~ age_days,
                   data = samdf.age)
```

```{r}
set.seed('1116')
vegan::adonis(dist.age ~ age_years,
                   data = samdf.age)
```

### In days

```{r}
alpha %>% 
     filter(visit == 1) %>% 
     ggplot(aes(x = age_days/365, y = obs)) +
     geom_point(alpha = 0.5) +
     geom_smooth(method = 'lm') +
     labs(x = 'Age', y = '*trnL* richness') +
     scale_x_continuous(breaks = seq(10,18, 2)) +
     theme_bw() +
     theme(axis.title = element_text(size = 14,
                                     face = 'bold'),
           axis.title.y = ggtext::element_markdown(),
           axis.text = element_text(size = 12))
     
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '4',
            paste0(plotID, '_A_Adolescent raw trnL richness vs. age.pdf')),
       height = 4, width = 4)

# ggsave(here('..',
#             'development',
#             'docs',
#             'r01',
#             '2021_11',
#             paste0(plotID, 'Adolescent trnL richness vs. age (continuous).pdf')),
#        height = 4, width = 3)
```

```{r}
summary(lm(age_days/365 ~ obs, alpha))
```


```{r}
ggplot(aes(x = , y = obs_rare)) +
     geom_point(alpha = 0.5) +
     # geom_smooth(method = 'lm') +
     labs(x = 'Age', y = '*trnL* richness') +
     theme_bw() +
     theme(axis.title = element_text(size = 14),
           axis.title.y = ggtext::element_markdown())
```


```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_A_Adolescent rarefied trnL richness vs. age.pdf')),
       height = 4, width = 4)
```


```{r}
alpha %>% 
     filter(visit == 1) %>% 
     filter(!is.na(obs) & !is.na(age_days)) %>% 
     cor.test(~ obs + age_days,
              data = ., 
              method = 'spearman',
              alternative = 'less')
```

```{r}
alpha %>% 
     filter(visit == 1) %>% 
     filter(!is.na(obs_rare) & !is.na(age_days)) %>% 
     cor.test(~ obs_rare + age_days,
              data = ., 
              method = 'spearman',
              alternative = 'less')
```

```{r}
alpha %>% 
     filter(visit == 1) %>% 
     lm(obs ~ age_days, data = .) %>% 
     summary()
```

### In years

```{r}
alpha %>% 
     filter(visit == 1) %>% 
     group_by(age_years) %>% 
     count()
```


```{r}
alpha %>% 
     filter(visit == 1) %>% 
     ggplot(aes(x = age_years, y = obs, group = age_years)) +
     geom_boxplot() +
     labs(x = 'Age', y = '*trnL* richness') +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())

```

```{r}
alpha %>% 
     filter(visit == 1) %>% 
     mutate(age_group = cut(age_years, 
                            breaks = c(9, 11, 13, 15, 18),
                            labels = c('10-11',
                                       '12-13', 
                                       '14-15',
                                       '16-18'))
            ) %>% 
     ggplot(aes(x = age_group, y = obs, group = age_group)) +
     geom_boxplot() +
     labs(x = 'Age', y = '*trnL* richness') +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown(),
           axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..',
            'development',
            'docs',
            'r01',
            '2021_11',
            paste0(plotID, 'Adolescent trnL richness vs. age (summarized categories).pdf')),
       height = 3, width = 2)
```

## SES

```{r}
# Completeness of each measure
alpha %>% 
     group_by(ses) %>% 
     count()
```
```{r}
alpha %>% 
     group_by(worry) %>% 
     count()
```

```{r}
alpha %>% 
     group_by(actual) %>% 
     count()
```

### Household income

```{r}
# Statistical test
# Need to first remove participants with no age data
dist.ses <- 
     ps %>% 
     subset_samples(!is.na(ses)) %>% 
     distance(., 
              method = 'jaccard', 
              binary = TRUE)

samdf.ses <- 
     ps %>% 
     subset_samples(!is.na(ses)) %>% 
     sample_data() %>% 
     data.frame()
```

```{r}
set.seed('1116')
vegan::adonis(dist.ses ~ over100k,
                   data = samdf.ses)
```

```{r}
alpha.stats <- 
     alpha %>% 
     filter(!is.na(ses)) %>% 
     group_by(subj, ses, over100k) %>% 
     summarize(obs_mean = mean(obs),
               obs_rare_mean = mean(obs_rare))
```

```{r}
# Non-rarefied data
# Remove NAs, make one observation per participant
ggplot(alpha.stats, aes(x = ses, y = obs_mean)) +
     geom_boxplot() +
     labs(x = 'Household income', y = '*trnL* richness') +
     theme_classic() +
     theme(axis.title = element_text(size = 14),
           axis.title.y = ggtext::element_markdown())
```
```{r}
kruskal.test(alpha.stats$obs_mean, 
             alpha.stats$ses)
```

```{r}
summary(aov(obs_mean ~ ses, alpha.stats))
```


```{r}
# Thresholded and rarefied data
# Remove NAs, make one observation per participant
alpha %>% 
     filter(!is.na(ses)) %>% 
     group_by(subj, ses) %>% 
     summarize(obs_rare_mean = mean(obs_rare)) %>% 
     ggplot(aes(x = ses, y = obs_rare_mean)) +
     geom_boxplot() +
     labs(x = 'Household income', y = '*trnL* richness') +
     theme_classic() +
     theme(axis.title = element_text(size = 14),
           axis.title.y = ggtext::element_markdown())
```
```{r}
kruskal.test(alpha.stats$obs_rare_mean, 
             alpha.stats$ses)
```

```{r}
summary(aov(obs_rare_mean ~ ses, alpha.stats))
```

#### Over/under $100K

```{r}
ggplot(alpha.stats, aes(x = over100k, y = obs_mean)) +
     geom_boxplot() +
     labs(x = 'Household\nincome\n', 
          y = '*trnL* richness') +
     ylim(0, 22) +
     theme_classic() +
     theme(axis.title = element_text(size = 14,
                                     face = 'bold'),
           axis.title.y = ggtext::element_markdown(),
           axis.text = element_text(size = 12))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Rarefied trnL richness x household income.pdf')),
       height = 4, width = 2)
```

```{r}
t.test(obs_mean ~ over100k,
       data = alpha.stats)
```

```{r}
ggplot(alpha.stats, aes(x = over100k, y = obs_rare_mean)) +
     geom_boxplot() +
     labs(x = 'Household income', 
          y = '*trnL* richness') +
     theme_classic() +
     theme(axis.title = element_text(size = 14),
           axis.title.y = ggtext::element_markdown())
```
```{r}
t.test(obs_rare_mean ~ over100k,
       data = alpha.stats)
```

### Food insecurity

```{r}
# Statistical test
# Need to first remove participants with no age data
dist.insecurity <- 
     ps %>% 
     subset_samples(!is.na(worry)) %>% 
     distance(., 
              method = 'jaccard', 
              binary = TRUE)

samdf.insecurity <- 
     ps %>% 
     subset_samples(!is.na(worry)) %>% 
     sample_data() %>% 
     data.frame()
```

```{r}
set.seed('1116')
vegan::adonis(dist.insecurity ~ worry_ever,
                   data = samdf.insecurity)
```

```{r}
set.seed('1116')
vegan::adonis(dist.insecurity ~ actual_ever,
                   data = samdf.insecurity)
```


```{r}
select(alpha, starts_with('worry'), starts_with('actual'))
```

```{r}
# Remove NA results, group multiple measures within each subject
alpha.stats <- 
     alpha %>% 
     filter(!is.na(worry) | !is.na(actual)) %>% 
     group_by(subj,
              worry, worry_ever,
              actual, actual_ever) %>% 
     summarize(obs_mean = mean(obs),
               obs_rare_mean = mean(obs_rare)) %>% 
     ungroup()
```

#### Worry 
##### Raw richness

```{r}
ggplot(alpha.stats, aes(x = worry, y = obs_mean)) +
     geom_boxplot() +
     theme_classic()
```
```{r}
kruskal.test(alpha.stats$obs_mean, 
             alpha.stats$worry)
```


```{r}
ggplot(alpha.stats, aes(x = worry_ever, y = obs_mean)) +
     geom_boxplot() +
     labs(x = 'Worried about\nfood running out\nin past year',
          y = '*trnL* richness') +
     ylim(0, 22) +
     theme_classic() +
     theme(axis.title = element_text(size = 14,
                                     face = 'bold'),
           axis.title.y = ggtext::element_markdown(),
           axis.text = element_text(size = 12))
     
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Raw trnL richness x food worry.pdf')),
       height = 4, width = 2)
```

```{r}
# Two-independent sample t-test
t.test(obs_mean ~ worry_ever,
       data = alpha.stats)
```

##### Threshold

```{r}
ggplot(alpha.stats, aes(x = worry, y = obs_rare_mean)) +
     geom_boxplot() +
     theme_classic()
```

```{r}
ggplot(alpha.stats, aes(x = worry_ever, y = obs_rare_mean)) +
     geom_boxplot() +
     labs(x = 'Worried about food\nrunning out\nin past year',
          y = '*trnL* richness') +
     theme_classic() +
     theme(axis.title = element_text(size = 14),
           axis.title.y = ggtext::element_markdown())
```
```{r}
# Two-independent sample t-test
t.test(obs_rare_mean ~ worry_ever,
       data = alpha.stats)
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Rarefied trnL richness x food worry.pdf')),
       height = 4, width = 2)
```

#### Actual lack of food

##### Raw richness
```{r}
ggplot(alpha.stats, aes(x = actual, y = obs_mean)) +
     geom_boxplot() +
     theme_classic()
```
```{r}
kruskal.test(alpha.stats$obs_mean, 
             alpha.stats$actual)
```


```{r}
ggplot(alpha.stats, aes(x = actual_ever, y = obs_mean)) +
     geom_boxplot() +
     labs(x = 'Food ran\nout in\npast year',
          y = '*trnL* richness') +
     ylim(0, 22) +
     theme_classic() +
     theme(axis.title = element_text(size = 14,
                                     face = 'bold'),
           axis.title.y = ggtext::element_markdown(),
           axis.text = element_text(size = 12))
     
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Raw trnL richness x food ran out.pdf')),
       height = 4, width = 2)
```

```{r}
# Two-independent sample t-test
t.test(obs_mean ~ actual_ever,
       data = alpha.stats)
```