---
title: "Adolescent summary"
output: html_notebook
---

# Setup

```{r setup, include = FALSE, echo = FALSE} 
require(knitr) 
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(eulerr)
library(here)
library(MButils) # For lowest_level() naming function
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data
```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

# Pre-process

Preserve full object, because I want to compare to other cohorts.

```{r}
ps.adol <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     subset_samples(reads_tot != 0) %>% 
     prune_taxa(taxa_sums(.) > 0, .) 

ps.adol
```

## Update sample data

# Analyze

## Sample demographics

```{r}
samdf <- 
     data.frame(ps.adol@sam_data) %>% 
     rownames_to_column(var = 'row')
```

```{r}
# How many unique subjects?
n_distinct(samdf$subj)
```

```{r}
samdf %>% 
     select(subj, casecontrol) %>% 
     distinct() %>% 
     count(casecontrol) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, ethnicity) %>% 
     distinct() %>% 
     count(ethnicity) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, race) %>% 
     distinct() %>% 
     count(race) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
```{r}
samdf %>% 
     select(subj, ses) %>% 
     distinct() %>% 
     count(ses) %>% 
     mutate(percentage = n/n_distinct(samdf$subj))
```
## trnL taxa

### Number

```{r}
ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .)
```

```{r}
# Distinct families, species, genera
ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     tax_table() %>% 
     data.frame() %>% 
     apply(2, function(x){n_distinct(x)})
```

### Prevalence

```{r}
melt <- psmelt(ps.adol)
```

```{r}
# Tally taxa by subject
taxa.counts <- 
     melt %>% 
     group_by(subj, OTU) %>% 
     summarize(detected = any(Abundance > 0)) %>% 
     ungroup() %>% 
     group_by(OTU) %>% 
     summarize(detected = sum(detected)/length(detected)) %>% 
     arrange(desc(detected)) 

taxa.counts
```

```{r}
# What's the distribution of these prevalences?
p.prev <- 
     ggplot(taxa.counts, aes(x = detected * 100)) +
     geom_histogram(binwidth = 5,
                    boundary = 0) +
     labs(x = 'Detected in\n[X%] of subjects', y = 'Frequency') +
     xlim(0, 100) 

p.prev
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_Subject prevalence histogram.pdf')),
       height = 3, width = 2)
```

```{r}
# Skewness
e1071::skewness(taxa.counts$detected)
```

Manually label names

```{r}
# top10 %>% 
#      pull(name) %>% 
#      unique() %>% 
#      sort() %>% 
#      data.frame(name = .) %>% 
#      write_csv(here('data',
#                     'processed',
#                     'miniseq',
#                     'taxa_names.csv'))
```

#### By family

```{r}
melt %>% 
     filter(study == 'Adolescent') %>% 
     group_by(subj, family) %>% 
     summarize(detected = any(Abundance > 0)) %>% 
     ungroup() %>% 
     group_by(family) %>% 
     summarize(detected = sum(detected)/length(detected)) %>% 
     arrange(desc(detected)) 
```

### Mean pMR

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(subj) %>% 
     summarize(mean = mean(obs, na.rm = TRUE),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.mean, aes(x = mean, y = mean_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Mean *trnL* richness<br>\n(All samples)', 
          y = 'Mean rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adolescent rarefied vs observed trnL richness (mean).pdf')),
       height = 4, width = 4)
```

### Range

```{r}
adolescent.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(across(c(obs, obs_rare),
                      list(med = ~median(.x, na.rm = TRUE),
                           sd = ~sd(.x, na.rm = TRUE))
                      )
               )

adolescent.stats
```

```{r}
adolescent.stats %>% 
     select(-subj) %>% 
     apply(2, range, na.rm = TRUE)
```

## Ordination

### PCA

```{r}
# Presence-absence data
asvtab <-
     ps.adol %>%
     # subset_samples(reads_tot > 10000) %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     otu_table() %>%
     .@.Data

# Convert from counts to binary presence-absence
asvtab[asvtab > 0] = 1
asvtab[asvtab = 0] = 0

asvtab %>% as.numeric() %>% unique()
```

```{r}
# # CLR-transformed relative abundance
# asvtab <- 
#      ps.adol %>% 
#      subset_samples(reads_tot > 10000) %>% 
#      transform_sample_counts(., function(x){x/sum(x)}) %>% 
#      otu_table() %>% 
#      .@.Data 
# 
# # Add pseudocount, transform
# asvtab.clr <- driver::clr(asvtab + 0.0001)
# 
# # Replace ASV name, though note not interpretable as strict ASV any longer
# colnames(asvtab.clr) <- colnames(asvtab)
```

```{r}
# PCA
pca <- prcomp(asvtab, 
              center = TRUE, # Try these a few different ways to see
              scale = FALSE)

pca.df <- 
        data.frame(pca$x) %>% 
        rownames_to_column(var = 'row')

# % variance explained
eigs <- pca$sdev^2
ve.pc1 <- as.character(100*round(eigs[1] / sum(eigs), 3))
ve.pc2 <- as.character(100*round(eigs[2] / sum(eigs), 3))

# Add back sample data
pca.df <- left_join(pca.df, samdf)

# Calculate plotting limits based on largest value observed in PC axes 1 and 2
limit <- max(abs(pca.df[, c('PC1', 'PC2')])) +
          0.05*(max(abs(pca.df[, c('PC1', 'PC2')])))
```

#### Plot

```{r}
# batch.colors <- brewer.pal(9, 'Set3')
# read.depth.colors <- brewer.pal(5, 'RdBu')[5:1] # Reverse
# read.depth.colors[3] <- 'gray80' # Darken

pca.plot <- 
     ggplot(pca.df, 
            aes(x = PC1, y = PC2)) + 
     geom_point(size = 2, alpha = 0.6) +
     # scale_color_manual(values = read.depth.colors) + 
     coord_equal() +
     labs(x = paste0(' PC1 (', ve.pc1, '%)'),
          y = paste0(' PC2 (', ve.pc2, '%)')) + 
     xlim(-limit, limit) + ylim(-limit, limit) 
     
pca.plot
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript',
            'supplemental', 
            'adolescent',
            paste0(plotID, '_PCA on CLR-transformed abundance.pdf')),
       device = cairo_pdf,
       height = 4, width = 4)
```

##### Biplot

```{r}
# Biplot

# Calculate loadings
V <- pca$rotation # Eigenvectors
L <- diag(pca$sdev) # Diag mtx w/sqrts of eigenvalues on diag.

loadings <- V %*% L
     
# Get loadings for first 2 PCs and format for plotting
pythag <- function(a, b){sqrt(a^2 + b^2)}
loadings.12 <- 
     data.frame(loadings[, 1:2]) %>%
     rename(PC1 = X1, PC2 = X2) %>% 
     mutate(variable = row.names(loadings)) %>% 
     mutate(length = pythag(PC1, PC2), 
            slope = PC2/PC1,
            ang = atan(slope)*(180/pi))

loadings.plot <- top_n(loadings.12, 10, wt = length) 
```

```{r}
# Rename loadings with lowest taxonomic level
loadings.taxtab <- 
     tax_table(ps.adol)[row.names(loadings.plot)] %>% 
     data.frame() %>% 
     MButils::lowest_level()

loadings.plot <- 
     loadings.taxtab %>% 
     rownames_to_column(var = 'asv') %>% 
     select(asv, name) %>% 
     right_join(loadings.plot, by = c('asv' = 'variable'))
```

```{r}
# What quadrant of the plot is the label in?
q1 <- filter(loadings.plot, PC1 > 0 & PC2 > 0)
q2 <- filter(loadings.plot, PC1 < 0 & PC2 > 0)
q3 <- filter(loadings.plot, PC1 < 0 & PC2 < 0)
q4 <- filter(loadings.plot, PC1 > 0 & PC2 < 0)
     
pca.biplot <- 
     pca.plot + 
     geom_segment(data = loadings.plot,
                  aes(x = 0, y = 0, 
                      xend = PC1, yend = PC2),
                  color = 'black',
                  arrow = arrow(angle = 15, 
                                length = unit(0.1, 'inches')))

# Then add geom_text quadrant-by-quadrant, aligning text accordingly
     if (dim(q1)[1] != 0) {
          pca.biplot <- pca.biplot +
               geom_text(data = q1, aes(x = PC1, y = PC2, hjust = 0, angle = ang,
                                        label=paste0('   ', name), 
                                        fontface = 'bold'), 
                         color = 'black', show.legend = FALSE)
     }
     if (dim(q2)[1] != 0) {
          pca.biplot <- pca.biplot +
               geom_text(data = q2, aes(x = PC1, y = PC2, hjust = 1, angle = ang,
                                        label=paste0(name, '   '),
                                        fontface = 'bold'),
                         color = 'black', show.legend = FALSE)
     }
     if (dim(q3)[1] != 0) {
          pca.biplot <- pca.biplot +
               geom_text(data = q3, aes(x = PC1, y = PC2, hjust = 1, angle = ang,
                                        label=paste0(name, '   '), 
                                        fontface = 'bold'), 
                         color = 'black', show.legend = FALSE)
     }
     if (dim(q4)[1] != 0) {
          pca.biplot <- pca.biplot +
               geom_text(data = q4, aes(x = PC1, y = PC2, hjust = 0, angle = ang,
                                        label=paste0('   ', name), 
                                        fontface = 'bold'), 
                         color = 'black', show.legend = FALSE)
     }
```

```{r}
pca.biplot
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 'ordination', 
            paste0(plotID, '_Overall trnL composition, all samples, biplot.pdf')),
       height = 6, width = 6)
```

### PCoA
#### Jaccard

```{r}
ord <- 
     ordinate(ps.adol,
              method = 'PCoA',
              distance = 'jaccard',
              binary = TRUE)
```

```{r}
p <-  
     plot_ordination(ps.adol,
                     ord,
                     type = 'samples')

p + 
     # coord_equal() + 
     theme(legend.position = 'none')
```

```{r}
# Customize plot 
data <- 
     data.frame(ord$vectors) %>% 
     rownames_to_column(var = 'row')

data <- full_join(data, samdf)

ggplot(data, aes(x = Axis.1, y = Axis.2, color = zbmi)) +
     geom_point(alpha = 0.8,
                size = 3) +
     labs(x = 'PCo1 (9.1%)',
          y = 'PCo2 (7.9%)') +
     coord_equal() +
     theme_bw() +
     theme(legend.title = element_blank()) +
     theme(axis.line = element_line(size = 1,
                                    color = 'gray80'),
           axis.text = element_text(size = 12,
                                    face = 'bold',
                                    color = 'gray80'), 
           axis.title = element_text(size = 14,
                                     face = 'bold',
                                     color = 'gray80'), 
           axis.ticks = element_line(size = 1, 
                                     color = 'gray80'),
           legend.text = element_text(size = 10,
                                      face = 'bold',
                                      color = 'gray40'),
           panel.border = element_blank(),
           panel.grid.major = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_Metaproteomic ordination (Bray-Curtis).pdf')))
```

```{r}
vegan::adonis(distance(ps.adol, 
                       method = 'jaccard',
                       binary = TRUE) ~ age,
              data = samdf)
```

## Set intersection

```{r}
# Update taxa names to their lowest specified level
taxtab <- data.frame(ps@tax_table@.Data)

all(taxa_names(ps) == row.names(taxtab)) 
taxa_names(ps) <- lowest_level(taxtab)$name
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

all <- taxa_names(ps)
```

```{r}
# Now, return taxon name to ASV sequence
taxa_names(ps) <- row.names(taxtab)
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
146 - length(weightloss.taxa)
146 - length(adult1.taxa)
146 - length(adult2.taxa)
146 - length(adolescent.taxa)
```

### Only found in Adolescent

```{r}
setdiff(adolescent.taxa,
        union(adult1.taxa, 
              union(adult2.taxa, weightloss.taxa))
)
```

For these taxa, how many samples were they detected in? With what kind of prevalence?

```{r}
adol.only <- 
     setdiff(adolescent.taxa,
             union(adult1.taxa, 
                   union(adult2.taxa, weightloss.taxa)))

# Get their ASV sequences
adol.only <- 
     taxtab %>% 
     filter(name %in% adol.only) %>% 
     rownames_to_column(var = 'asv')
```

```{r}
# Get sample prevalences
psmelt(ps.adol) %>% 
     filter(OTU %in% adol.only$asv) %>% 
     group_by(OTU) %>% 
     summarize(prevalence = sum(Abundance > 0),
               cumulative_reads = sum(Abundance)) %>% 
     left_join(adol.only, ., by = c('asv' = 'OTU'))
```

### Only found in others

```{r}
setdiff(union(adult1.taxa, 
              union(adult2.taxa, weightloss.taxa)),
        adolescent.taxa
)
```

### Euler plot

```{r}
euler.df <- 
     data.frame(taxon = all) %>% 
     mutate(`Weight Loss` = taxon %in% weightloss.taxa,
            `Adult-1` = taxon %in% adult1.taxa,
            `Adult-2` = taxon %in% adult2.taxa,
            `Adolescent` = taxon %in% adolescent.taxa) %>% 
     column_to_rownames(var = 'taxon')

euler.df
```
```{r}
fit <- euler(euler.df)
fit
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results',
         'manuscript',
         'supplemental',
         'adolescent',
         paste0(plotID,
                '_Intersection with other cohorts.pdf')),
    height = 4, width = 4)

plot(fit)
```

### UpSet plot

```{r}
list.input <- 
     list(`Weight Loss` = weightloss.taxa,
          `Adult-1` = adult1.taxa,
          `Adult-2` = adult2.taxa,
          `Adolescent` = adolescent.taxa)
```

```{r}
upset(fromList(list.input), 
      order.by = "freq",
      mb.ratio = c(0.65, 0.35))
```

## Prevalence differences

Calculate taxon prevalence in each cohort
```{r}
prev_by_study <- 
     psmelt(ps) %>% 
     group_by(study, OTU) %>% 
     summarize(prevalence = sum(Abundance > 0)/n_distinct(Sample)) 

prev_by_study
```
```{r}
# Pivot results to wide
prev_by_study <- 
     pivot_wider(prev_by_study,
                 names_from = study,
                 values_from = prevalence)

prev_by_study
```
Now for each we'd want an ANOVA-like test to see if there are differences

```{r}
results.df <- 
     data.frame(taxon = NULL,
                comparison = NULL,
                diff = NULL,
                lwr = NULL,
                upper = NULL,
                p.adj = NULL)
```


```{r}
melt <- psmelt(ps)

for (taxon in unique(melt$OTU)){
     melt.temp <- 
          melt %>% 
          filter(OTU == taxon) %>% 
          mutate(detected = ifelse(Abundance > 0,
                                   yes = 1, 
                                   no = 0)) %>% 
          select(study, detected)

     f <- detected ~ study

     m <- glm(f, family = binomial(link = 'logit'), data = melt.temp) 
     result <- 
          aov(m) %>% 
          TukeyHSD
     
     rows <- 
          data.frame(result$study) %>% 
          rownames_to_column(var = 'comparison') %>% 
          filter(grepl('Adolescent', comparison))
     
     rows$taxon <- taxon
     
     results.df <- bind_rows(results.df, rows)
}
```

```{r}
# Run overall p value adjustment
# Could also do a crude Bonferonni where I multiply by # of taxa (suspect I'm double-counting here)
results.df$p.adj_adj <- p.adjust(results.df$p.adj,
                                 method = 'BH')
```

```{r}
# Get histogram of p values
ggplot(results.df, aes(x = p.adj)) +
     geom_histogram(binwidth = 0.05,
                    boundary = 0)
```
```{r}
# Get histogram of p values
ggplot(results.df, aes(x = p.adj_adj)) +
     geom_histogram(binwidth = 0.05,
                    boundary = 0)
```

```{r}
results.df %>% 
     filter(p.adj_adj < 0.05) %>% 
     ggplot(aes(x = diff)) +
     geom_histogram(binwidth = 0.1, boundary = 0.05)
```
More things are reduced in this cohort.

```{r}
different <- 
     results.df %>% 
     filter(p.adj_adj < 0.05) %>% 
     group_by(taxon) %>% 
     count() %>% 
     filter(n == 3) %>% 
     select(OTU =taxon)
```

8 taxa differentially present in POMMS compared to *all* adult cohorts.

```{r}
left_join(different, select(melt, OTU, name)) %>% 
     distinct() %>% 
     left_join(results.df, c('OTU' = 'taxon')) %>% View()

```


Maybe we could present these with an interpretation of coefficients?