---
title: "Adolescent metabolites"
output: html_notebook
---

```{r setup, include = FALSE, echo = FALSE} 
require(knitr) 
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(fitdistrplus)
library(lme4) # For glmer
library(lmerTest)
library(patchwork)
library(phyloseq)
library(tidyverse)
library(vegan) # For rarecurve()

select <- dplyr::select
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)

## Below is trying to adapt to Nature Methods guidance, but hard to reproduce
## across figs.  Need to read more about how others do this. 
# theme_set(theme_bw() +
#                theme(
#                     axis.ticks = element_line(size = 0.25),
#                     axis.title = element_text(size = 7,
#                                               face = 'bold'),
#                     legend.key.size = unit(2.5, units = 'mm'),
#                     legend.title = element_text(size = 6, 
#                                                 face = 'bold'),
#                     text = element_text(size = 5),
#                     panel.grid.major = element_line(size = 0.25),
#                     panel.grid.minor = element_line(size = 0.125),
#                     strip.text = element_text(face = 'bold')
#                      )
# )
```


# Read in data
```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed',
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

```{r}
# Rarefied richness estimates
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

# Pre-process

## trnL

```{r}
# Filter to Adolescent cohort only
ps <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     # prune_samples(sample_sums(.) > 0, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'POMMS') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study) 

alpha <- 
    left_join(add, alpha,
              by = c('row' = 'name'))
```

## Demographic data

```{r}
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adolescent') %>% 
     select(row:reads_tot, R24_id:reads_food)  # Adolescent metadata
```

### Completeness

Check completeness of important model variables.  Note that `table` won't include these: have to do with `count`.

```{r}
# Do all subjects have an entry timepoint?
n_distinct(samdf$subj)

samdf %>% 
     filter(timepoint == 'Entry') %>% 
     nrow()
```

Almost, but not everyone.  So will group by subject below rather than just filtering to one timepoint. 

```{r}
# Check only one measure per subject
samdf %>%
     group_by(subj, ethnicity) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(ethnicity) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, sex) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(sex) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, race) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(race) %>% 
     count()
```

```{r}
# Recode these to reduce categories with small number
# Mixed-race responses as "Multiple"
samdf$race[grep(' ', samdf$race)] <- 'Multiple'

samdf %>% 
     group_by(subj, race) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(race) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, casecontrol) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(casecontrol) %>% 
     count()
```
```{r}
hist(samdf$zbmi)
```

May want to consider different breakpoints here to test nuance to case-control finding (e.g. adding Case (low); Case (med); Case (high)) based on breakpoints in the below plot.

```{r}
samdf %>% 
     filter(casecontrol == 'Case' & timepoint == 'Entry') %>%
     ggplot(aes(x = zbmi)) +
     geom_histogram()
```

```{r}
# Note that many BMI entries missing (all? at 6 months)
samdf %>% 
     group_by(timepoint) %>% 
     summarize(n_missing = sum(is.na(zbmi)))
```
```{r}
# Try this recoding
samdf <- 
     samdf %>% 
     mutate(zbmi_case = ifelse(casecontrol == 'Case',
                               yes = zbmi,
                               no = 0), 
            # Label controls as 0 for cut, below
            casecontrol_detail = 
                 cut(zbmi_case, 
                     breaks = c(-0.1, 1.5, 2.0, 2.5, 4.0), 
                     right = FALSE, 
                     labels = c('Control',
                                'Case (zBMI < 2.0)', 
                                'Case (2.0 <= zBMI < 2.5)', 
                                'Case (zBMI > 2.5)')) %>% 
                 factor(exclude = NULL))

samdf %>% select(subj, 
                 starts_with('case'), 
                 starts_with('zbmi'))
```
```{r}
samdf %>% 
     group_by(casecontrol_detail) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(tanner) %>% 
     count()
```

```{r}
# Only 5 missing. Recode as "Unknown"
samdf <- 
     samdf %>% 
     mutate(tanner = ifelse(is.na(tanner),
                            yes = 'Unknown',
                            no = tanner),
            tanner = factor(tanner,
                            levels = c('2', '3', '4', '5', 
                                       'Unknown')))

```

#### SES indicators

```{r}
samdf %>% 
     group_by(subj, ses) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(ses) %>% 
     count()
```
```{r}
# Make an over/under 100k category
# (The poverty threshold for a 2-person family would be closer to 50k)
samdf <- mutate(samdf, ses_over100k = (ses == '>100,000'))
```

```{r}
samdf %>% 
     group_by(ses_over100k) %>% 
     count()
```

```{r}
# Make "any" category for worry/actual insecurity
samdf <- 
     mutate(
          samdf,
          across(.cols = c(worry, actual),
                 .fns = ~ ifelse(.x != 'Never true',
                                 yes = 'Yes',
                                 no = 'No'),
                 .names = '{.col}_ever'
                 )
     )
```

```{r}
samdf %>% 
     group_by(subj, worry_ever) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(worry_ever) %>% 
     count()
```
```{r}
samdf %>% 
     group_by(subj, actual_ever) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(actual_ever) %>% 
     count()
```

```{r}
# Make a joint category for any positive response to a food insecurity question
samdf <- 
     samdf %>% 
     mutate(insecurity_any = 
                 ifelse(worry_ever == 'Yes' | actual_ever == 'Yes',
                        yes = 'Yes',
                        no = 'No'))
```

```{r}
# Recode NA as Unknown so these aren't discarded from the model
samdf$ses[is.na(samdf$ses)] <- 'Unknown'
samdf$ses_over100k[is.na(samdf$ses_over100k)] <- 'Unknown'
samdf$worry_ever[is.na(samdf$worry_ever)] <- 'Unknown'
samdf$actual_ever[is.na(samdf$actual_ever)] <- 'Unknown'
samdf$insecurity_any[is.na(samdf$insecurity_any)] <- 'Unknown'
```

#### Age

```{r}
# Age into quartiles by year, as in POMMS group analysis
samdf <- 
     samdf %>% 
     mutate(age_years = floor(age_days/365),
            age_q = cut(age_years,
                        breaks = 4))

samdf %>% 
     group_by(subj, age_q) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(age_q) %>% 
     count()
```

If age not available from one of the samples, take it from another:

```{r}
correction <- 
     samdf %>% 
     group_by(subj) %>% 
     summarize(n_age_q = n_distinct(age_q)) %>% 
     filter(n_age_q == 2)

samdf %>% 
     filter(subj %in% correction$subj) %>% 
     select(subj, starts_with('age')) %>% 
     arrange(subj)
```
Okay, so some of these are appropriately when the participant's age has changed.  We want to just keep those that are NA. 

```{r}
# Get a list of subjects for which we'll replace one value with the other
correction <- 
     samdf %>% 
     filter(subj %in% correction$subj) %>% 
     select(subj, starts_with('age')) %>% 
     group_by(subj) %>% 
     summarize(na.present = any(is.na(age))) %>% 
     filter(na.present == TRUE) %>% 
     pull(subj)

length(correction)
```

Awesome!! This is exactly the same as the number of missing values above.  

```{r}
samdf %>% 
     filter(subj %in% correction) %>% 
     select(subj, timepoint, starts_with('age')) %>% 
     arrange(subj, desc(timepoint))
```
Manually inspected and for all of these it's the 6 month timepoint missing. So add half a year (183 d) to their age_days variable, then convert.

```{r}
correction.df <- 
     samdf %>% 
     filter(subj %in% correction) %>% 
     filter(timepoint == 'Entry') %>% 
     select(subj, timepoint, age_days) %>% 
     mutate(timepoint = '6 months',
            age_days_corr = age_days + 183) %>% 
     select(subj, timepoint, ends_with('corr'))

correction.df
```
```{r}
# Bind to sample data and replace values if needed
samdf <- 
     samdf %>% 
     select(-age) %>% 
     left_join(correction.df) %>% 
     mutate(age_days = 
                 ifelse(is.na(age_days),
                        yes = age_days_corr,
                        no = age_days)) 

# Peek at results
samdf %>% 
     filter(subj %in% correction.df$subj) %>% 
     select(subj, starts_with('age')) %>% 
     arrange(subj)

# Now, can drop corrected column, recalculate 
samdf <- 
     samdf %>% 
     select(-age_days_corr) %>% 
          mutate(age_years = floor(age_days/365),
            age_q = cut(age_years,
                        breaks = 4))
```

```{r}
# Any missing now? 
samdf %>% 
     group_by(subj, age_q) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(age_q) %>% 
     count()
```

```{r}
# Convert to 10s of years for better visualization
samdf$age_decades <- samdf$age_days/365/10 
```

#### Blood pressure

```{r}
bp <- 
     samdf %>% 
     select(subj, visit, matches('_bp'))

head(bp)
```
```{r}
bp <- 
     pivot_longer(bp,
                  cols = matches('_bp'),
                  names_to = 'measure',
                  values_to = 'value')

bp
```
```{r}
bp <- 
     bp %>% 
     separate(col = measure,
              into = c('measure', 'type'),
              sep = '_',
              extra = 'merge') %>% 
     pivot_wider(names_from = 'type',
                 values_from = 'value') %>% 
     rename(percentile = bp_percentile)

bp
```
```{r}
# Plot distribution
ggplot(bp, aes(bp)) +
     geom_histogram(binwidth = 5, 
                    boundary = 0) +
     facet_wrap(~measure,
                nrow = 2) +
     labs(x = 'Blood pressure',
          y = 'Frequency') +
     theme(plot.margin = unit(c(0.125, 0.25, 0, 0), units = 'inches'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            'supplemental',
            'adolescent',
            paste0(plotID, '_BP histogram.pdf')),
       height = 4, width = 3)
```

```{r}
ggplot(bp, aes(percentile)) +
     geom_histogram(binwidth = 5,
                    boundary = 0) +
     facet_wrap(~measure,
                nrow = 2) +
     ylim(0, 105) +
     labs(x = 'BP percentile',
          y = 'Frequency') +
     theme(plot.margin = unit(c(0.125, 0.25, 0, 0), units = 'inches'))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            'supplemental',
            'adolescent',
            paste0(plotID, '_BP percentile histogram.pdf')),
       height = 4, width = 3)
```

#### Set reference

```{r}
class(samdf$sex)

samdf$sex <- factor(samdf$sex,
                          levels = c('M', 'F'))

levels(samdf$sex)
```

```{r}
class(samdf$race)

samdf$race <- factor(samdf$race,
                     levels = c('WH', 
                                'BL', 
                                'AS', 
                                'Multiple'))

levels(samdf$race)
```

```{r}
class(samdf$ethnicity)

samdf$ethnicity <- factor(samdf$ethnicity,
                          levels = c('NH', 'HI'))

levels(samdf$ethnicity)
```

```{r}
samdf$age_q <- as.ordered(samdf$age_q)
levels(samdf$age_q)
```

```{r}
class(samdf$casecontrol)

samdf$casecontrol <- factor(samdf$casecontrol,
                          levels = c('Control', 'Case'))

levels(samdf$casecontrol)
```

```{r}
levels(samdf$casecontrol_detail)
```

```{r}
levels(samdf$tanner)
```


```{r}
class(samdf$ses)

samdf$ses <- 
     factor(samdf$ses,
            levels = c('<24,999',
                       '25,000-49,999',
                       '50,000-74,999',
                       '75,000-99,999',
                       '>100,000',
                       'Unknown')) 
levels(samdf$ses)
```

```{r}
class(samdf$worry_ever)

samdf$worry_ever <- 
     factor(samdf$worry_ever,
            levels = c('No',
                       'Yes',
                       'Unknown')) 

levels(samdf$worry_ever)
```

```{r}
class(samdf$actual_ever)

samdf$actual_ever <- 
     factor(samdf$actual_ever,
            levels = c('No',
                       'Yes',
                       'Unknown')) 

levels(samdf$actual_ever)
```

### Metabolites

From the POMMS interim report, metabolites that had differences between case and control were:

OB: higher HOMA-IR, HBA1c, CRP, ALT; lower HDL

Targeted: 
OB: higher valine, glycerol, insulin; lower KIC, KMV, glycine

```{r}
# Not sure what code for glycerol is, left off here
metabs <- c('alt', 
            'c_reactive_protein',
            'Gly',
            'hdl',
            'hba1c',
            'homa_ir',
            'hdl',
            'insulin',
            'KIC', # α-ketoisocaproate
            'KMV', # α-keto-β-methylvalerate
            'Val')

# Now, add AAs (apart from Gly, Val above)
metabs <- c(metabs, 
            # ESSENTIAL
            'Phe',
            # 'Thr', # Not collected 
            # 'Trp',# Not collected 
            'Met',
            'Leu.Ile', # Joint entry for leucine and isoleucine
            # 'Lys', # Not collected
            'His',
            # OTHER
            'Ala',
            'Ser',
            'Pro',
            'Tyr',
            'Asx',
            'Glx',
            'Orn',
            'Cit',
            'Arg')

# All targeted metabolites, to be used downstream
targeted.panel <- 
     samdf %>% 
     select(KIV:X3.HIB) %>% 
     names()
```

```{r}
# Pull from sample data, and join to results
alpha <- 
     samdf %>% 
     select(row, 
            subj,
            starts_with('casecontrol'), 
            timepoint,
            starts_with('age'), # Want flexibility in choosing here
            sex,
            tanner,
            race,
            ethnicity,
            bmi,
            zbmi, 
            matches('_bp'),
            starts_with('ses'),
            worry_ever,
            actual_ever,
            insecurity_any,
            one_of(metabs)) %>% 
     right_join(alpha)

alpha
```

```{r}
# Make a raw version to make this easier to work with below
alpha.raw <- 
     alpha %>% 
     select(-c(threshold, obs_rare, se)) %>% 
     distinct() 

dim(alpha.raw)
```

# Analyze

## Rough pass blood pressure

```{r}
bp <- 
     alpha %>% 
     select(subj,
            timepoint, 
            obs,
            matches('_bp')) %>% 
     distinct()

dim(bp)
bp <- filter(bp,
             if_any(matches('_bp'),
                    ~!is.na(.x)))
dim(bp)
```

```{r}
any(is.na(bp$systolic_bp))
any(is.na(bp$systolic_bp_percentile))
```
Strangely the above expression isn't catching some NAs in the percentile column, but let's just plot raw relationships first.

```{r}
bp <- 
     bp %>% 
     pivot_longer(cols = matches('_bp'),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     separate(measure,
              into = c('measure', 'type'),
              sep = '_',
              extra = 'merge') %>% 
     pivot_wider(names_from = 'type',
                 values_from = 'value') %>% 
     rename(percentile = bp_percentile)

bp
```
```{r}
ggplot(bp, aes(x = obs,
               y = bp)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~measure) +
     labs(x = 'pMR', y = 'Blood pressure')
```

```{r}
ggplot(bp, aes(x = obs,
               y = percentile)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~measure) +
     labs(x = 'pMR', y = 'BP percentile')
```

## Summarize per-subject

```{r}
alpha %>% 
     summarize(med = median(obs),
               mad = mad(obs))
```

## Model: pMR response

### Linear model

```{r}
alpha.subset <- 
     alpha %>% 
     filter(reads_food > 0) %>%
     # filter(threshold == 1000 & !is.na(obs_rare)) %>%
     select(-c(obs_rare, threshold, se)) %>% 
     distinct() %>% 
     # mutate(obs = ifelse(obs == 0,
     #                     yes = 1, no = obs)) %>% 
     filter(timepoint == 'Entry') 

dim(alpha.subset)
```

```{r}
descdist(alpha$obs, discrete = TRUE, boot = 500)
```


```{r}
# Check distribution
p.pmr.hist <- 
     ggplot(alpha.subset, aes(x = obs)) +
     geom_histogram(binwidth = 2,
                    boundary = 0) +
     labs(x = 'pMR\n', y = 'Frequency') +
     theme(panel.grid.minor = element_blank())

p.pmr.hist
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '4',
            paste0(plotID, '_pMR response variable histogram.pdf')),
       height = 3, width = 3)
```

```{r}
# QQ plot
qqnorm(alpha.subset$obs, pch = 1, frame = FALSE)
qqline(alpha.subset$obs, col = "steelblue", lwd = 2)
```

```{r}
# Mixed model
# f = "obs ~ age_years + sex + race + ethnicity + casecontrol + zbmi + ses + (1|subj)"
# model <- lmer(f,
#               data = alpha.subset)

f = "obs ~ age_decades + sex + race + ethnicity + casecontrol + ses + actual_ever"

model <- lm(f,
            data = alpha.subset)

summary(model)
```

###### Plot betas

```{r}
# Get confidence intervals
cis <- 
     confint(model) %>% 
     data.frame() %>% 
     rownames_to_column(var = 'covar') 

# Get beta hat and p from model
betas <- 
     summary(model)$coeff[, c(1, 4)] %>% 
     data.frame() %>% 
     rownames_to_column(var = 'covar') 

betas <- 
     full_join(betas, cis) %>% 
     rename(p = Pr...t..,
            CI2.5 = X2.5..,
            CI97.5 = X97.5..)

betas
rm(cis)
```

```{r}
# Assign covariate grouping
betas$group <- NULL
betas$group[grepl(pattern = 'age|race|sex|ethnicity',
                  betas$covar)] <- 'Demographic'
betas$group[grepl(pattern = 'case',
                  betas$covar)] <- 'Clinical'
betas$group[grepl(pattern = 'ses|ever',
                  betas$covar)] <- 'Socioeconomic'

betas$group <- factor(betas$group,
                      levels = c('Socioeconomic',
                                 'Clinical',
                                 'Demographic'))
```

```{r}
# Refactor by values of Estimate
betas <- 
     betas %>% 
     arrange(group, covar)
```

```{r}
# Recode names for more interpretable plot
betas$covar <- 
     factor(betas$covar,
            levels = c('actual_everUnknown',
                       'actual_everYes',
                       'sesUnknown',
                       'ses>100,000',
                       'ses75,000-99,999',
                       'ses50,000-74,999',
                       'ses25,000-49,999',
                       'casecontrolCase',
                       'ethnicityHI',
                       'raceMultiple',
                       'raceAS',
                       'raceBL',
                       'sexF',
                       'age_decades',
                       '(Intercept)'),
            labels = c('foodinsecUnknown',
                       'foodinsecYes',
                       'incomeUnknown',
                       'income>100K',
                       'income75-99K',
                       'income50-74K',
                       'income25-49K',
                       'casecontrolCase',
                       'ethnicityHispanic',
                       'raceMultiple',
                       'raceAsian',
                       'raceBlack',
                       'sexFemale',
                       'ageDecades',
                       '(Intercept)'))
```

```{r}
# Recode p value as significant, trend, n.s.
betas$p_cat <- 
     cut(betas$p,
         breaks = c(0, 0.05, 0.10, 1),
         labels = c('Significant',
                    'Trend',
                    'Not significant')) 
```

```{r}
# Plots in style of Yap et al
p.betas <- 
     betas %>% 
     filter(covar != "(Intercept)") %>% # Exclude intercept
     # Exclude Unknown covariate categories (missing data)
     filter(!grepl(pattern = 'Unknown', covar)) %>% 
     ggplot(aes(x = covar, y = Estimate, color = p<=0.05)) +
     geom_hline(yintercept = 0,
                linetype = 'dotted',
                color = '#E15759') +
     geom_point(size = 2) +
     geom_linerange(aes(ymin = CI2.5, 
                       ymax = CI97.5,
                       color = p<=0.05),
                    size = 1) +
     scale_color_manual(values = c('gray40', 'black')) +
     labs(x = 'Covariate', y = 'Regression coefficient') +
     coord_flip() +
     theme(legend.position = 'none')

p.betas
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '4',
            paste0(plotID, '_pMR ~ demographics betas.pdf')),
       height = 3, width = 5)
```

###### Fit
```{r}
fits <- 
     # data.frame(y = model@resp$y, # For lmer() output
     data.frame(y = model$model$obs,
                y_hat = fitted(model),
                resid = residuals(model))
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_point() +
     geom_abline(slope = 1,
                 intercept = 0,
                 color = 'gray80') +
     labs(x = 'pMR (actual)', y = 'pMR (fitted)') +
     scale_y_continuous(breaks = seq(0, 20, 5),
                        limits = c(0, 22)) +
     coord_equal()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR ~ demographics fitted v. actual, lmer with zeros and missing data removed.pdf')),
       height = 4, width = 4)
```

```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_Residuals versus pMR fits, lmer with zeros and missing data removed.pdf')),
       height = 4, width = 4)
```

Perhaps re: Justin's outliers point, it would be good to visually inspect the outliers in the dataset.  What ASVs make them up-- do they seem plausible?

```{r}
car::vif(model)
```

### GLM

Tried this:
- with all data as a negative binomial GLMM; didn't converge (think because so few measurements per subject): see here https://stackoverflow.com/questions/53034261/warning-lme4-model-failed-to-converge-with-maxgrad
- used only entry timepoints, and as a negative binomial GLM

#### Negative binomial

```{r}
# Using alpha.subset and f from above
model <- 
     glm.nb(f, 
            data = alpha.subset)

summary(model)
```

```{r}
# Get confidence intervals
cis <- 
     confint(model) %>% 
     data.frame() %>% 
     rownames_to_column(var = 'covar') 

# Get beta hat and p from model
betas <- 
     summary(model)$coeff[, c(1, 4)] %>% 
     data.frame() %>% 
     rownames_to_column(var = 'covar') 

betas <- 
     full_join(betas, cis) %>% 
     rename(p = Pr...z..,
            CI2.5 = X2.5..,
            CI97.5 = X97.5..)

betas
rm(cis)
```

```{r}
# Assign covariate grouping
betas$group <- NULL
betas$group[grepl(pattern = 'age|race|sex|ethnicity',
                  betas$covar)] <- 'Demographic'
betas$group[grepl(pattern = 'case',
                  betas$covar)] <- 'Clinical'
betas$group[grepl(pattern = 'ses|ever',
                  betas$covar)] <- 'Socioeconomic'

betas$group <- factor(betas$group,
                      levels = c('Socioeconomic',
                                 'Clinical',
                                 'Demographic'))
```

```{r}
# Refactor by values of Estimate
betas <- 
     betas %>% 
     arrange(group, Estimate)

betas$covar <- factor(betas$covar,
                         levels = betas$covar)
```

```{r}
# Plots in style of Yap et al
betas %>% 
     filter(covar != "(Intercept)") %>% # Exclude intercept
     # Exclude Unknown covariate categories (missing data)
     filter(!grepl(pattern = 'Unknown', covar)) %>% 
     ggplot(aes(x = covar, y = Estimate, color = p<=0.05)) +
     # geom_rect(xmin = ) + 
     geom_hline(yintercept = 0,
                linetype = 'dotted',
                color = '#E15759') +
     geom_point() +
     geom_linerange(aes(ymin = CI2.5, 
                       ymax = CI97.5,
                       color = p<=0.05)) +
     scale_color_manual(values = c('gray40', 'black')) +
     labs(x = 'Covariate', y = 'Regression coefficient') +
     coord_flip() +
     theme(legend.position = 'none')
```

```{r}
fits <- 
     data.frame(y = model$y,
                y_hat = fitted(model),
                resid = residuals(model))
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_abline(slope = 1, intercept = 0) +
     geom_point() +
     labs(x = 'pMR (actual)', y = 'pMR (fitted)') +
     ylim(0, 28) + xlim(0, 28)
```
```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```

Hmmm. This residual structure is better but the model is hardly predictive. 

### Compare beta estimates

```{r}
ggplot(results.df, aes(x = n, y = beta, color = p <= 0.05)) +
     geom_point() +
     facet_wrap(~coeff,
                scales = 'free_y') +
     scale_color_manual(values = c('gray80', '#59A14F'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'linear models', 
            paste0(plotID, '_betas, models 1-9.pdf')),
       height = 4, width = 10)
```

### Visualize raw relationships

#### Age

```{r}
p.age <- 
     ggplot(alpha.subset, aes(x = age_decades*10,
                         y = obs)) +
     geom_point(alpha = 0.2) +
     geom_smooth(method = 'lm',
                 color = '#e03531') +
     scale_x_continuous(breaks = seq(10, 18, by = 2)) +
     labs(title = 'Age (years)', y = 'pMR') + 
     ylim(0, 35) +
     theme(axis.title.x =  element_blank())

p.age
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_pMR x age raw relationship.pdf')),
       device = cairo_pdf,
       height = 3, width = 2.5)
```

#### Income categories

```{r}
# Clean up household income labels for plotting
levels(alpha.subset$ses)
```

```{r}
alpha.subset$ses <- 
     factor(alpha.subset$ses,
            labels = c('<$25K',
                       '$25-$49K',
                       '$50-$74K',
                       '$75K-$99K',
                       '>$100K',
                       'Unknown'))
```

```{r}
alpha.subset %>% 
     filter(ses != 'Unknown') %>% # Remove Unknown category
     ggplot(aes(x = ses, y = obs)) +
     geom_boxplot() +
     labs(x = 'Household income', y = 'pMR') +
     theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
     theme(axis.title.y = element_blank())
```

```{r}
# As dotplot
p.income <- 
     alpha.subset %>% 
     filter(ses != 'Unknown') %>% # Remove Unknown category
     ggplot(aes(x = ses, y = obs)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Income', y = 'pMR')  +
     theme(axis.text.x = element_text(angle = 30, hjust = 1),
           axis.title = element_blank()) 

p.income
```


```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_pMR x income raw relationship.pdf')),
       device = cairo_pdf,
       height = 3.5, width = 2.5)
```

#### Case status

```{r}
alpha.subset$casecontrol <- 
     factor(alpha.subset$casecontrol,
            labels = c('Healthy\n(control)',
                       'Obesity\n(case)'))
```

```{r}
# As barplot
alpha.subset %>% 
     ggplot(aes(x = casecontrol, y = obs)) +
     geom_boxplot() +
     labs(x = 'Health status', y = 'pMR') +
     theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
# As dotplot
p.case <- 
     alpha.subset %>% 
     ggplot(aes(x = casecontrol, y = obs)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Weight status', y = 'pMR') +
     theme(axis.title = element_blank())

p.case
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_pMR x case status raw relationship.pdf')),
       device = cairo_pdf,
       height = 4, width = 3)
```

#### Food insecurity

```{r}
p.insec <- 
     alpha.subset %>% 
     filter(actual_ever != 'Unknown') %>% 
     ggplot(aes(x = actual_ever, y = obs)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Food insecurity', y = 'pMR') +
     theme(axis.title = element_blank())

p.insec
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_pMR x food insecurity raw relationship.pdf')),
       device = cairo_pdf,
       height = 3, width = 2.5)
```

Now, combine all of the above with patchwork:

```{r}
((p.prev / p.pmr.hist ) | (p.betas)) /
(p.age | p.income | p.case | p.insec)
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Combined figure.pdf')),
       device = cairo_pdf,
       height = 120, width = 180, units = 'mm')
```


## Model: pMR predictor

```{r}
# Log transform metabolites
alpha.raw <- 
     alpha.raw %>% 
     mutate(across(one_of(metabs),
                   ~log(.x),
                   .names = '{.col}_log')) %>% 
     # Replace with NA if -Inf
     mutate(across(ends_with('_log'),
                   ~ifelse(.x == -Inf,
                           yes = NA,
                           no = .x)))
```

## Structure of metabolite data

```{r}
# Get in simple form: only metabolite variables
# No duplicated rows

dim(alpha.raw)
metab.df <- 
     alpha.raw %>% 
     select(name,
            one_of(metabs),
            ends_with('_log'))

dim(metab.df)
```

##### Distribution
```{r}
# Raw data
pivot_longer(metab.df,
             one_of(metabs),
             names_to = 'metabolite',
             values_to = 'value') %>% 
     ggplot(aes(x = value)) +
     geom_histogram() +
     # ggplot(aes(x = obs, y = value)) +
     # geom_point(alpha = 0.5) +
     labs(y = 'Count') +
     facet_wrap(~metabolite,
                scales = 'free') +
     theme(axis.title.x = element_blank(),
           axis.text = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_metabolite response variable histogram (with zeros).pdf')),
       height = 6, width = 6)
```


```{r}
# Logged
pivot_longer(metab.df,
             ends_with('log'),
             names_to = 'metabolite',
             values_to = 'value') %>% 
     ggplot(aes(x = value)) +
     geom_histogram() +
     # ggplot(aes(x = obs, y = value)) +
     # geom_point(alpha = 0.5) +
     facet_wrap(~metabolite,
                scales = 'free') +
     labs(y = 'Count') +
     facet_wrap(~metabolite,
                scales = 'free') +
     theme(axis.title.x = element_blank(),
           axis.text = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_log(metabolite) response variable histogram (with zeros).pdf')),
       height = 6, width = 6)
```

```{r}
# Do any have negative values?
metab.df %>% 
     select(one_of(metabs)) %>% 
     apply(2, function(x){range(x, na.rm = TRUE)})
```

```{r}
# How often do NAs occur?
metab.df %>% 
     select(one_of(metabs)) %>% 
     apply(2, function(x){sum(is.na(x))})
```

These differ by individual, so don't want to remove them en masse.

##### Cross-correlation

Can also make these plots using ggplot's geom_tile

```{r}
# 24 metabolites in all
length(metabs)
```

```{r}
# Make heatmap of correlations between them
cor.raw <- 
     metab.df %>% 
     select(all_of(metabs)) %>% 
     as.matrix() %>% 
     cor(use = 'na.or.complete',
         method = 'spearman')

data.frame(cor.raw)
```
```{r}
# Heatmap
corrplot(cor.raw, 
         order = "hclust", 
         method = "circle")
```

```{r}
# Make heatmap of correlations between them
cor.log <- 
     metab.df %>% 
     select(ends_with('log')) %>% 
     as.matrix() %>% 
     cor(use = 'na.or.complete',
         method = 'spearman')

data.frame(cor.log)
```
```{r}
# Heatmap

corrplot(cor.log, 
         order = "hclust", 
         method = "circle")
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results',
            'manuscript',
            'supplemental',
            'adolescent',
            paste0(plotID, '_Correlation between metabolites.pdf')),
       height = 4, width = 4)
```

```{r}
# Check serine, ornithine, proline based on results
cor.log[c('Ser_log', 'Orn_log', 'Pro_log'),
        c('Ser_log', 'Orn_log', 'Pro_log')]
```

```{r}
hist(cor.log)
```
So these are on the upper end of what's typical. 

##### Principal components (targeted metabolites)

Another idea is to see if pMR is predictive of PC1, 2, ... from an ordination based on the serum targeted metabolites.

Do an ordination:
- Pull complete cases of targeted metabolite data
- Center and scale
- Do a PCA
- Extract first n PCs and add them to dataframe for testing

```{r}
# Get all targeted metabolite columns
targeted.df <- 
     samdf %>% 
     select(name = row, 
            one_of(targeted.panel))

dim(targeted.df)
```

```{r}
# Get complete cases
targeted.df <- targeted.df[complete.cases(targeted.df), ]
dim(targeted.df)
```

```{r}
# PCA
pca <- 
     targeted.df %>% 
     remove_rownames() %>% 
     column_to_rownames(var = 'name') %>% 
     as.matrix() %>% 
     prcomp(center = TRUE,
            scale = TRUE)
```

```{r}
# Pull coordinates, join to pMR data
pca.df <- 
     pca$x %>% 
     data.frame() %>% 
     rownames_to_column(var = 'name') %>% 
     left_join(select(alpha.raw, name, obs)) %>% 
     select(name, obs, everything())
```

```{r}
# Percent of variance explained for plot labels
percentage <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 2)
percentage <- paste(colnames(pca$x), 
                    paste0('(', as.character(percentage), "%", ")", sep="") )
```

```{r}
color.scheme <- 
     RColorBrewer::brewer.pal(n = 8, name = 'RdBu') %>% 
     rev()

# Plot
ggplot(pca.df, aes(x = PC1, y = PC2, color = obs)) +
     geom_point(alpha = 0.9) +
     coord_equal() +
     scale_color_gradientn(colors = color.scheme) +
     labs(x = percentage[1], y = percentage[2], color = 'pMR') +
     theme(panel.grid = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'adolescent',
#             paste0(plotID, '_PCA on serum metabolites.pdf')),
#        device = cairo_pdf, 
#        height = 4, width = 4)
```

```{r}
set.seed(20220202)
# Try a PERMANOVA with pMR
vegan::adonis(dist(pca$x, method = 'euclidean') ~ obs, 
              data = pca.df)
```

```{r}
# Pull first PC for inclusion given large gap from % variance explained by
# remainder
head(percentage)
```

```{r}
alpha.raw <- 
     pca.df %>% 
     select(name, PC1) %>% 
     right_join(alpha.raw)
```

## Linear model

TODO:
get R2 and p for each nominally significant pMR-metabolite relationship to label plot

Check a few above to establish distribution and link function.
```{r}
alpha.filt <- filter(alpha.raw, !is.na(Ser_log))
```

```{r}
# Create a skewness-kurtosis plot to find the best distribution for data
descdist(alpha.filt$Ser_log, discrete = FALSE, boot = 500)
```
Ok, confirming what we saw above that things look normal. 

Following what I think is Nathan Bilmeyer's code from [POMMS github repository](https://github.com/POMMS/methods_baseline_insights/blob/master/2_statistical_analysis_and_figures/Fig3_and_Supplement_script.R):

Had initially tried this as using metabolite plus demographics to predict trnL-- Lawrence suggests doing this the other way around, using trnL to predict metabolite (as would be done in epi studies). 

```{r}
metabs <- paste(metabs,
                'log',
                sep = '_')

metabs
```

```{r}
# Add PC1 from above
metabs <- c(metabs, 'PC1')
```

```{r}
# Only consider entry timepoint for simplicity
alpha.subset <- 
     alpha.raw %>% 
     filter(timepoint == 'Entry')
```

```{r}
f1     = "`%s` ~ obs" # Unadjusted model
f2     = "`%s` ~ obs + age_years + sex + race + ethnicity + casecontrol" # Adjusted model

results     = data.frame(f = NA,
                         Metabolite=NA, 
                         Number=1, 
                         Beta=1, 
                         SE=1, 
                         Tval=1, 
                         Pval=1)

for(metab in metabs) {
     for (formula in c(f1, f2)){
          f         = sprintf(formula, metab)
          model     = tryCatch(glm(f, data=alpha.subset, 
                                       family = gaussian()),
                                   error=function(e) { NA })
          n         = tryCatch(nobs(model),                      
                                   error=function(e) { NA })
          s         = tryCatch(summary(model)$coef[2,],          
                                   error=function(e) { rep(NA, 4) })
          results   = rbind(results, c(formula, metab, n, s))
     }
}

results          = results[-1,]
results$f  = factor(results$f,
                          levels = c(f1, f2),
                          labels = c('Unadjusted', 'Adjusted'))
results$Number   = as.numeric(results$Number)
results$Beta     = as.numeric(results$Beta)
results$SE       = as.numeric(results$SE)
results$Tval     = as.numeric(results$Tval)
results$Pval     = as.numeric(results$Pval)
results$adj.Pval = p.adjust(results$Pval, method="fdr")
results          = results[order(results$adj.Pval),]
```

```{r}
results
```
```{r}
# Get model fit for PC1 (last iteration above)
summary(
     lm(PC1 ~ obs + age_years + sex + race + ethnicity + casecontrol,
        data = alpha.subset)
)
```

Follow up on serine, ornithine, proline

```{r}
# Plot
alpha.raw %>% 
     select(obs, Ser_log, Orn_log, Pro_log) %>% 
     pivot_longer(ends_with('log'),
                  names_to = 'metabolite',
                  values_to = 'value'
     ) %>% 
     ggplot(aes(x = obs, y = value)) + 
     geom_point() +
     geom_smooth(method = 'lm') +
     facet_wrap(~metabolite,
                scales = 'free_y')
```

```{r}
# Can I simply show this as the raw value (not logged)?
alpha.raw %>% 
     select(obs, Ser, Orn, Pro, Arg) %>% 
     pivot_longer(-obs,
                  names_to = 'metabolite',
                  values_to = 'value'
     ) %>% 
     mutate(metabolite = factor(metabolite,
                                levels = c('Arg',
                                           'Ser',
                                           'Orn',
                                           'Pro')),
            significant = ifelse(metabolite == 'Arg',
                                 yes = FALSE,
                                 no = TRUE)) %>%  
     ggplot(aes(x = obs, y = value)) + 
     geom_point(alpha = 0.5) +
     geom_smooth(aes(color = significant),
                 method = 'lm') +
     facet_wrap(~metabolite,
                scales = 'free_y',
                nrow = 1) +
     labs(y = 'Metabolite concentration',
          x = 'pMR') +
     scale_color_manual(values = c('gray40', '#E15759')) +
     theme(legend.position = 'none')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '5',
            paste0(plotID, '_serum metabolites.pdf')),
       device = cairo_pdf, 
       height = 3, width = 8)
```

```{r}
# Model fits: serine

f1     = "Ser_log ~ obs" # Unadjusted model
f2     = "Ser_log ~ obs + age_years + sex + race + ethnicity + casecontrol" # Adjusted model

m1 <- lm(f1, data = alpha.subset)
summary(m1)
```
```{r}
m2 <- lm(f2, data = alpha.subset)
summary(m2)
```

```{r}
# Is adjusted model a better fit?
anova(m1, m2)
```

