---
title: "Adolescent metabolites"
output: html_notebook
---

```{r setup, include = FALSE, echo = FALSE} 
require(knitr) 
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(fitdistrplus)
library(lme4) # For glmer
library(phyloseq)
library(tidyverse)
library(vegan) # For rarecurve()

select <- dplyr::select
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```


# Read in data
```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed',
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

```{r}
# Rarefied richness estimates
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (1,000 reads).rds') %>% 
     readRDS

rare.df
```

# Pre-process

```{r}
# Intentionally *not* removing samples with 0 reads here
ps <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0 , .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(study == 'Adolescent') %>% 
     select(name, study, subj, R24_id:o, starts_with('obs'))
```

```{r}
# Add in full set of observations, including samples with fewer than read count threshold
asvtab <- data.frame(ps@otu_table@.Data)
obs <- apply(asvtab, 1, function(x){sum(x > 0)})

add <- 
     data.frame(obs = obs) %>% 
     rownames_to_column(var = 'name')

# Populate with sample data
add <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, R24_id, visit, o) %>% 
     right_join(add, by = c('row' = 'name')) %>% 
     rename(name = row)

alpha <- 
     add %>% 
     filter(!(name %in% alpha$name)) %>% 
     bind_rows(alpha)
```

POMMS-70 doesn't have a rarefied observation, should figure out why.

```{r}
# Recode NAs in rarefied data as 0 observations
alpha$obs_rare[is.na(alpha$obs_rare)] <- 0 
```


## Bind to demographic data

```{r}
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row:reads_tot, R24_id:reads_food)  # Adolescent metadata
```

### Completeness

Check completeness of important model variables.  Note that `table` won't include these: have to do with `count`.

```{r}
# Do all subjects have an entry timepoint?
n_distinct(samdf$subj)

samdf %>% 
     filter(timepoint == 'Entry') %>% 
     nrow()
```

Almost, but not everyone.  So will group by subject below rather than just filtering to one timepoint. 

```{r}
# Check only one measure per subject
samdf %>%
     group_by(subj, ethnicity) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(ethnicity) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, sex) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(sex) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, race) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(race) %>% 
     count()
```

```{r}
# Recode these to reduce categories with small number
# Mixed-race responses as "Multiple"
samdf$race[grep(' ', samdf$race)] <- 'Multiple'

samdf %>% 
     group_by(subj, race) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(race) %>% 
     count()
```

```{r}
samdf %>% 
     group_by(subj, casecontrol) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(casecontrol) %>% 
     count()
```
```{r}
hist(samdf$zbmi)
```


```{r}
samdf %>% 
     group_by(subj, ses) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(ses) %>% 
     count()
```
```{r}
# Age into quartiles by year, as in POMMS group analysis
samdf <- 
     samdf %>% 
     mutate(age_years = floor(age_days/365),
            age_q = cut(age_years,
                        breaks = 4))

samdf %>% 
     group_by(subj, age_q) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(age_q) %>% 
     count()
```
How come some of these are incomplete? Will need to address.

#### Set reference

```{r}
class(samdf$sex)

samdf$sex <- factor(samdf$sex,
                          levels = c('M', 'F'))

levels(samdf$sex)
```

```{r}
class(samdf$race)

samdf$race <- factor(samdf$race,
                     levels = c('WH', 
                                'BL', 
                                'AS', 
                                'Multiple'))

levels(samdf$race)
```

```{r}
class(samdf$ethnicity)

samdf$ethnicity <- factor(samdf$ethnicity,
                          levels = c('NH', 'HI'))

levels(samdf$ethnicity)
```

```{r}
levels(samdf$age_q)
```

```{r}
class(samdf$casecontrol)

samdf$casecontrol <- factor(samdf$casecontrol,
                          levels = c('Case', 'Control'))

levels(samdf$casecontrol)
```

```{r}
class(samdf$ses)

samdf$ses <- factor(samdf$ses,
                          levels = c('<24,999',
                                     '25,000-49,999',
                                     '50,000-74,999',
                                     '75,000-99,999',
                                     '>100,000'))


levels(samdf$ses)
```

### Metabolites

From the POMMS interim report, metabolites that had differences between case and control were:

OB: higher HOMA-IR, HBA1c, CRP, ALT; lower HDL

Targeted: 
OB: higher valine, glycerol, insulin; lower KIC, KMV, glycine

```{r}
# Not sure what code for glycerol is, left off here
metabs <- c('alt', 
            'c_reactive_protein',
            'Gly',
            'hdl',
            'hba1c',
            'homa_ir',
            'hdl',
            'insulin',
            'KIC', # α-ketoisocaproate
            'KMV', # α-keto-β-methylvalerate
            'Val')

# Now, add AAs (apart from Gly, Val above)
metabs <- c(metabs, 
            # ESSENTIAL
            'Phe',
            # 'Thr', # Not collected 
            # 'Trp',# Not collected 
            'Met',
            'Leu.Ile', # Joint entry for leucine and isoleucine
            # 'Lys', # Not collected
            'His',
            # OTHER
            'Ala',
            'Ser',
            'Pro',
            'Tyr',
            'Asx',
            'Glx',
            'Orn',
            'Cit',
            'Arg')
```

```{r}
# Pull from sample data, and join to results
alpha <- 
     samdf %>% 
     select(name = row, 
            subj,
            casecontrol, 
            timepoint,
            starts_with('age'), # Want flexibility in choosing here
            sex,
            race,
            ethnicity,
            zbmi, 
            ses,
            one_of(metabs)) %>% 
     right_join(alpha)

alpha
```
# Analyze

Need to think about-- should I be handling two timepoints?  I think best to just start with 1 here. 

I could include a subject effect, which would handle, right?

```{r}
dim(alpha)
# alpha <-
#      alpha %>%
#      filter(timepoint == 'Entry')

dim(alpha)
```

## Response variable distribution

How about just visualizing data to start?

```{r}
# Log transform metabolites
alpha <- 
     alpha %>% 
     mutate(across(one_of(metabs),
                   ~log(.x),
                   .names = '{.col}_log')) %>% 
     # Replace with NA if -Inf
     mutate(across(ends_with('_log'),
                   ~ifelse(.x == -Inf,
                           yes = NA,
                           no = .x)))
```

```{r}
# Raw data
pivot_longer(alpha,
             one_of(metabs),
             names_to = 'metabolite',
             values_to = 'value') %>% 
     ggplot(aes(x = value)) +
     geom_histogram() +
     # ggplot(aes(x = obs, y = value)) +
     # geom_point(alpha = 0.5) +
     facet_wrap(~metabolite,
                scales = 'free')
```

```{r}
# Logged
pivot_longer(alpha,
                  ends_with('log'),
                  names_to = 'metabolite',
                  values_to = 'value') %>% 
     ggplot(aes(x = value)) +
     geom_histogram() +
     # ggplot(aes(x = obs, y = value)) +
     # geom_point(alpha = 0.5) +
     facet_wrap(~metabolite,
                scales = 'free')
```

```{r}
# Do any have negative values?
alpha %>% 
     select(one_of(metabs)) %>% 
     apply(2, function(x){range(x, na.rm = TRUE)})
```

```{r}
# How often do NAs occur?
alpha %>% 
     select(one_of(metabs)) %>% 
     apply(2, function(x){sum(is.na(x))})
```

These differ by individual, so don't want to remove them en masse.

## GLM parameters

Check a few above to establish distribution and link function.
```{r}
alpha.filt <- filter(alpha, !is.na(alt))
```

```{r}
# Create a skewness-kurtosis plot to find the best distribution for data
descdist(alpha.filt$alt, discrete = FALSE, boot = 500)
```
```{r}
# Fit  data to a variety of distributions to double check above results. 
fit.norm <- fitdist(alpha.filt$alt, "norm")
fit.lognorm <- fitdist(alpha.filt$alt, "lnorm")
fit.gamma <- fitdist(alpha.filt$alt, "gamma")
```

```{r}
plot(fit.norm)
```

```{r}
plot(fit.gamma)
```

Following what I think is Nathan Bilmeyer's code from [POMMS github repository](https://github.com/POMMS/methods_baseline_insights/blob/master/2_statistical_analysis_and_figures/Fig3_and_Supplement_script.R):

Had initially tried this as using metabolite plus demographics to predict trnL-- Lawrence suggests doing this the other way around, using trnL to predict metabolite (as would be done in epi studies). 

```{r}
metabs <- paste(metabs,
                'log',
                sep = '_')

metabs
```

```{r}
formula     = "`%s` ~ obs + age_q + sex + race + ethnicity + casecontrol"

results     = data.frame(Metabolite=NA, Number=1, Beta=1, SE=1, Tval=1, Pval=1)

for(metab in metabs) {
	f             = sprintf(formula, metab)
	model         = tryCatch(glm(f, data=alpha, 
	                             family = gaussian(link = 'log')),
	                             # family = Gamma()),
	                         error=function(e) { NA })
	n             = tryCatch(nobs(model),                           
	                         error=function(e) { NA })
	s             = tryCatch(summary(model)$coef[2,],               
	                         error=function(e) { rep(NA, 4) })
	results       = rbind(results, c(metab, n, s))
}

results          = results[-1,]
results$Number   = as.numeric(results$Number)
results$Beta     = as.numeric(results$Beta)
results$SE       = as.numeric(results$SE)
results$Tval     = as.numeric(results$Tval)
results$Pval     = as.numeric(results$Pval)
results$adj.Pval = p.adjust(results$Pval, method="fdr")
results          = results[order(results$adj.Pval),]
```

```{r}
results
```
```{r}
formula     = "`%s` ~ obs_rare + age_q + sex + race + ethnicity + casecontrol"

results     = data.frame(Metabolite=NA, Number=1, Beta=1, SE=1, Tval=1, Pval=1)

for(metab in metabs) {
	f             = sprintf(formula, metab)
	model         = tryCatch(glm(f, data=alpha, 
	                             family=gaussian()),
	                             # family = Gamma()),
	                         error=function(e) { NA })
	n             = tryCatch(nobs(model),                           
	                         error=function(e) { NA })
	s             = tryCatch(summary(model)$coef[2,],               
	                         error=function(e) { rep(NA, 4) })
	results       = rbind(results, c(metab, n, s))
}

results          = results[-1,]
results$Number   = as.numeric(results$Number)
results$Beta     = as.numeric(results$Beta)
results$SE       = as.numeric(results$SE)
results$Tval     = as.numeric(results$Tval)
results$Pval     = as.numeric(results$Pval)
results$adj.Pval = p.adjust(results$Pval, method="fdr")
results          = results[order(results$adj.Pval),]
```

```{r}
results
```

## Richness response variable

```{r}
# Check distribution
alpha %>% 
     select(-obs_rare_se) %>% 
     pivot_longer(cols = starts_with('obs'),
                  names_to = 'metric',
                  values_to = 'value') %>% 
     mutate(metric = factor(metric,
                            levels = c('obs', 'obs_rare'),
                            labels = c('pMR', 'Rarefied pMR'))) %>% 
     ggplot(aes(x = value)) +
     geom_histogram(binwidth = 2,
                    boundary = 0) +
     facet_wrap(~metric, nrow = 2) +
     labs(x = 'Value', y = 'Count') +
     theme(panel.grid.minor = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR response variable histogram (with zeros).pdf')),
       height = 4, width = 2)
```

Sayan's steps:
1. Try generic linear model
     - without zeros
     - turn zeros into ones
     - exclude missing data
     - impute missing data
seeing if weirdness happens here also
2. Try then as a GLM with appropriate fmaily and link.

### Linear model

```{r}
sum(alpha$obs == 0)
```

##### Remove 0s, exclude missing data

```{r}
alpha.subset <- 
     alpha %>% 
     filter(obs != 0) %>% 
     filter(!is.na(ses)) %>% 
     filter(!is.na(age_years))

dim(alpha.subset)
```

```{r}
f1 = "obs ~ age_years + sex + race + ethnicity + casecontrol + zbmi + (1|subj)"
f2 = "obs ~ age_years + sex + race + ethnicity + casecontrol + zbmi + ses + (1|subj)"

model1 <- lmer(f1, 
             data = alpha.subset) 

model2 <- lmer(f2, 
             data = alpha.subset) 

summary(model1)
```

```{r}
summary(model2)
```


###### Fit
```{r}
fits <- 
     data.frame(model = 1,
                y = model1@resp$y,
                y_hat = fitted(model1),
                resid = residuals(model1))

fits <- 
     data.frame(model = 2,
                y = model2@resp$y,
                y_hat = fitted(model2),
                resid = residuals(model2)) %>% 
     bind_rows(fits, .)
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_point() +
     geom_abline(slope = 1,
                 intercept = 0,
                 color = 'gray80') +
     facet_wrap(~model) +
     labs(x = 'pMR (actual)', y = 'pMR (fitted)') +
     scale_y_continuous(breaks = seq(0, 20, 5),
                        limits = c(0, 22)) +
     coord_equal()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR ~ demographics fitted v. actual, lmer with zeros and missing data removed.pdf')),
       height = 4, width = 4)
```

```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     facet_wrap(~model) +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```

#### Replace 0 with 1s, remove missing data

```{r}
alpha.subset <- 
     alpha %>% 
     mutate(obs = ifelse(obs == 0,
                         yes = 1,
                         no = obs)) %>% 
     filter(!is.na(ses)) %>% 
     filter(!is.na(age_years))

dim(alpha.subset)
```

```{r}
f1 = "obs ~ age_years + sex + race + ethnicity + casecontrol + zbmi + (1|subj)"
f2 = "obs ~ age_years + sex + race + ethnicity + casecontrol + zbmi + ses + (1|subj)"

model1 <- lmer(f1, 
             data = alpha.subset) 

model2 <- lmer(f2, 
             data = alpha.subset) 

summary(model1)
```

```{r}
summary(model2)
```

###### Fit
```{r}
fits <- 
     data.frame(model = 1,
                y = model1@resp$y,
                y_hat = fitted(model1),
                resid = residuals(model1))

fits <- 
     data.frame(model = 2,
                y = model2@resp$y,
                y_hat = fitted(model2),
                resid = residuals(model2)) %>% 
     bind_rows(fits, .)
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_point() +
     geom_abline(slope = 1,
                 intercept = 0,
                 color = 'gray80') +
     facet_wrap(~model) +
     labs(x = 'pMR (actual)', y = 'pMR (fitted)') +
     scale_y_continuous(breaks = seq(0, 20, 5),
                        limits = c(0, 22)) +
     coord_equal()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR ~ demographics fitted v. actual, lmer with zeros and missing data removed.pdf')),
       height = 4, width = 4)
```

```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     facet_wrap(~model) +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```

Perhaps re: Justin's outliers point, it would be good to visually inspect the outliers in the dataset.  What ASVs make them up-- do they seem plausible?

#### Multicollinearity

```{r}
car::vif(model2)
```


#### Fit to data

Plot fitted values against actual data-- how good is the model?

```{r}
fits <- 
     data.frame(y = model2$y,
                y_hat = model2$fitted.values,
                resid = model2$residuals)
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_point() +
     labs(x = 'pMR (actual)', y = 'pMR (fitted)') +
     coord_equal()
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR ~ demographics fitted v. actual.pdf')),
       height = 4, width = 4)
```


#### Residuals

```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```
```{r}
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'adolescent',
            paste0(plotID, '_pMR ~ demographics residuals.pdf')),
       height = 4, width = 8)
```


### Rarefied
```{r}
f = "obs_rare ~ age_q + sex + race + ethnicity + casecontrol + zbmi + ses"
model3 <- 
     glm(f, 
         data = alpha, 
         family = Gamma())

summary(model3)
```

```{r}
fits <- 
     data.frame(y = model3$y,
                y_hat = model3$fitted.values,
                resid = model3$residuals)
```

```{r}
ggplot(fits, aes(x = y, y = y_hat)) +
     geom_point() +
     labs(x = 'Rarefied pMR (actual)', y = 'Rarefied pMR (fitted)') +
     ylim(0, 21) + xlim(0, 21)
```
```{r}
ggplot(fits, aes(x = y_hat, y = resid)) +
     geom_point() +
     geom_hline(yintercept = 0,
                linetype = 'dashed',
                color = 'red') +
     labs(x = 'pMR (fitted)', y = 'Residuals') +
     coord_equal()
```

