---
title: "Adult-1, Adult-2 diet survey data"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## DHQ3

### Summary responses

```{r}
# Adult-1
summary.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3_summary.csv') %>% 
     read_csv()

head(summary.adult1)
```

```{r}
# Adult-2
summary.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3_summary.csv') %>% 
     read_csv()

head(summary.adult2)
```

### Detailed responses

```{r}
# Adult-1
detail.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3_detail.csv') %>% 
     read_csv()

head(detail.adult1)
```

```{r}
# Adult-2
detail.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3_detail.csv') %>% 
     read_csv()

head(detail.adult2)
```

# Pre-process

```{r}
# Add study variable

summary.adult1$study <- 'Adult-1'
summary.adult2$study <- 'Adult-2'

detail.adult1$study <- 'Adult-1'
detail.adult2$study <- 'Adult-2'
```


```{r}
# Join together, arrange columns
summary <- 
     bind_rows(summary.adult1, summary.adult2) %>% 
     select(study, everything())

detail <- 
     bind_rows(detail.adult1, detail.adult2) %>% 
     select(study, everything())
```

```{r}
# Remove raw data files
rm(summary.adult1, 
   summary.adult2,
   detail.adult1, 
   detail.adult2)
```

## Tidy

```{r}
# Remove un-needed columns (metadata, supplements)
summary <- select(summary, 
                  -c(`Record Number`,
                     starts_with('SUPP'))
)
```


```{r}
# Re-code sex
summary <- 
     summary %>% 
     rename(sex = `Sex (1=male; 2=female)`) %>% 
     mutate(sex = factor(sex,
                         labels = c('Male', 'Female')))

detail <- 
     detail %>% 
     rename(sex = `Sex (1=male; 2=female)`) %>% 
     mutate(sex = factor(sex,
                         labels = c('Male', 'Female')))
```

## QC and filter

From DHQ3 website:
There are no data cleaning or exclusion programs built into the software. Researchers can decide, based on break offs, energy, or other considerations, which respondents to exclude from analyses. Different investigators clean data in different ways: some like calorie exclusions, some like exclusions based on percentile cutpoints of nutrients or missing data, some like to transform data and then exclude, others want to exclude those respondents who did not take the task seriously (however that is defined). By looking at output files, one can write a program to check issues related to whether an individual always checks the same frequency or portion size. It is our experience that it is difficult to discern these types of errors from energy exclusions. FFQ data can be messy for some respondents and prone to measurement error for all.

Filtering criteria from Asnicar:
"Submitted FFQs were excluded if more than 10 food items were left unanswered or if the total energy intake estimate derived from the FFQ as a ratio of the participant’s estimated basal metabolic rate (determined by the Harris–Benedict equation) was more than 2 s.d. outside the mean of this ratio (<0.52 or >2.58)."
This is for a Harvard FFQ.

A minimum 10 kcal filter removes 3 participants who it appears minimally completed the survey. (Is there a DHQ3-reported parameter that reads this out??)

Other ideas for more data cleaning-- check for survey completeness, change sample ID to CHOMP participant ID, quality/sanity checks on data, filter incomplete responses...

```{r}
range(summary$`Energy (kcal)`,
      na.rm = TRUE)
```

## Normalize to kcal intake

```{r}
# Get columns to normalize
names(ffq)
```

```{r}
norm.cols <- 
     ffq %>% 
     select(`*Gluten (g)`:`Alcohol drinks`) %>% 
     names()
```

```{r}
# Calculate estimate of amount per 1000 kcal
ffq <- 
     ffq %>% 
     mutate(across(one_of(norm.cols),
                   ~ .x/`Energy (kcal)`*1000,
                   .names = '{.col}_per_1000kcal'
                   )
     )
```

# Calculate indices

### FVS

In order to calculate, need to do a few things:
- Remove entries that shouldn't be counted (i.e. water)
- Combine food codes duplicated across seasons or settings
- Filter to only those consumed with a frequency greater than once weekly

```{r}
# Simplify columns as we don't need nutrients
fvs <- select(detail, 
                 study,
                 subj, 
                 q_id,
                 food_id,
                 food_name,
                 daily_freq)

fvs
```

```{r}
# Remove water 

# Entries this term shows up in:
filter(fvs,
       grepl(pattern = '[Ww]ater',
             food_name)) %>% 
     pull(food_name) %>% 
     unique()

# Remove
fvs <- filter(fvs,
              !grepl(pattern = '[Ww]ater',
                     food_name))

dim(fvs)
```

```{r}
# How often might a code be repeated?
fvs %>% 
     group_by(subj, food_id, food_name) %>% 
     count() %>% 
     arrange(desc(n))
```
Spot-checked these and even values of up to 6 are reasonable: this would happen if someone added milk to every hot drink, for example. 

What's tricky here is for seasonality, I think I'd want to average, but for different types of "occasions", I'd want to sum. 

```{r}
# Sum
# This is most permissive, not accurate for all cases, but just try for now

fvs <- 
     fvs %>% 
     group_by(study, subj, food_id, food_name) %>% 
     summarize(daily_freq = sum(daily_freq))

fvs
```
Just for fun-- what's the range of daily frequencies?

```{r}
ggplot(fvs, aes(x = daily_freq)) +
     geom_histogram(binwidth = 0.01,
                    boundary = 0)
```
```{r}
# Filter
dim(fvs)
fvs <- filter(fvs,
              daily_freq >= 1/7) # Greater than once weekly
dim(fvs)
```

```{r}
# Calculate FVS
fvs <- 
     fvs %>% 
     ungroup() %>% 
     group_by(study, subj) %>% 
     count(fvs = .)

dim(fvs)
```

5 participants with missing data.  Think it's these 5 with a FVS of 0

```{r}
ggplot(fvs, aes(x = study, y = n)) +
     geom_boxplot() +
     labs(y = 'FVS',
          x = 'Study') +
     theme(axis.title.x = element_blank())
```

```{r}
fvs %>% 
     group_by(study) %>% 
     summarize(range(n))
```
Mann-Whitney U (Wilcoxon): This “does not estimate the difference in medians (a common misconception) but rather the median of the difference between a sample from x and a sample from y.”

```{r}
wilcox.test(n ~ study,
            data = fvs,
            conf.int = TRUE)
```

Would be good to inspect a few of these entries and see if they hold.

### DDS

### HEI-2015

### PDI

# Save output

```{r}
# Join together results from above
output <- fvs
```

```{r}
write_csv(
     output,
     here('data', 
          'processed',
          'DHQ3', 
          'Adult-1 and Adult-2 dietary indices.csv'
          )
)
```

