---
title: "Adult-1, Adult-2 FVS"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 5),
                    axis.title = element_text(size = 7,
                                              face = 'bold'),
                    strip.text = element_text(size = 7,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### DHQ3-based FVS

Calculated in Adult-1, Adult-2 dietary indices notebook.

```{r}
fvs <- 
     here('data',
          'processed',
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

# Pre-process

## trnL

### Filter phyloseq

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(alpha, add,
              by = c('name' = 'row'))
```

Then below, can just filter by desired threshold in order to calculate measures.

### Mean richness

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj, threshold) %>% 
     summarize(mean = mean(obs),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

# Analyze

## FVS

```{r}
# Join mean richness to PDI data
fvs <- 
     alpha.mean %>% 
     left_join(fvs)
```

### Raw data

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_tot, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'FVS\n(all items)',
          y = 'Mean pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x total FVS.pdf')),
       height = 2.5, width = 4)
```


```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_tot + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_tot + mean,
                            method = 'spearman')$p.value) 
            
```

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_plant, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'FVS\n(plant items)',
          y = 'Mean pMR')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x plant FVS.pdf')),
       height = 2.5, width = 4)
```

```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_plant + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_plant + mean,
                            method = 'spearman')$p.value) 
            
```

### Residuals

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_tot_residual, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'Food Variety Score residual\n(all items)',
          y = 'Mean pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x total FVS.pdf')),
       height = 2.5, width = 4)
```


```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_tot_residual + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_tot_residual + mean,
                            method = 'spearman')$p.value) 
            
```

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_plant_residual, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study,
                nrow = 2) +
     labs(x = 'FVS\n(plant items)',
          y = 'Mean pMR')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_raw trnL richness x plant FVS.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_plant_residual + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_plant_residual + mean,
                            method = 'spearman')$p.value) 
            
```

### Rarefied data

```{r}
fvs %>% 
     ggplot(aes(x = fvs_plant, y = mean_rare)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_grid(~study + threshold) +
     labs(x = 'Food Variety Score',
          y = 'Mean pMR')
```

## Sampling scheme

First, want to enumerate sampling schemes.
- 1: One sample, anytime
- 2: Two consecutive samples (adjacent days)-- ADULT-1 ONLY
- 3: NHANES spacing: two samples spaced between 3-10 days apart
- 4: 3 weeks of weekly samples
- 5: 4 " " " "-- ADULT-1 ONLY
- 6: 5 " " " "-- ADULT-1 ONLY
- 7: 6 " " " " -- ADULT-1 ONLY
- 5: All available samples

Maybe: create a sample distance (date) matrix? 

### Adult-1

```{r}
# Get sample data for Adult-1
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adult-1')

# Get observed richness
alpha.1 <- 
     alpha %>% 
     select(-c(threshold, obs_rare, se)) %>% 
     filter(study == 'Adult-1') %>% 
     distinct()
```

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

#### Scheme 1

Note that this will cover duplicated sample sets as i gets closer to 18, but can detect and remove them downstream.
```{r}
# Do scheme 1 first: 1 sample from each participant
set.seed(1222022)

for (iter in (1:100)){
     # Make subsampled, merged phyloseq
     samples <- 
          alpha.1 %>% 
          group_by(subj) %>% 
          slice_sample(n = 1) %>% 
          pull(name)
     
     # Join to FVS
     alpha.temp <- 
          alpha.1 %>% 
          filter(name %in% samples) %>% 
          left_join(fvs, by = 'subj')
     
     # Calculate correlation
     # Common warning: cannot compute exact p-value with ties
     suppressWarnings(spearman <- cor.test(alpha.temp$obs, 
                                           alpha.temp$fvs_plant_residual,
                                           method = 'spearman'))
     
     # Update results dataframe
     row <- c(1, 
              'Adult-1', 
              spearman$estimate, 
              spearman$p.value)
     names(row) <- c('scheme', 'study', 'rho', 'p')
     results.df <- bind_rows(results.df, row)
}
```

#### 2

Scheme 2: Two consecutive daily samples
My idea here would be first to subset to only the samples that I already know have one adjacent to them.  Then, at each iteration, could randomly pick one sample per participant, and then randomly pick one of its adjoining samples.

```{r}
# Make a sample "distance matrix"
d <- 
     dist(samdf$CH_day) %>% 
     as.matrix() 

colnames(d) <- rownames(d) <- samdf$row
```

```{r}
adjacent <- list()
for (s in unique(samdf$subj)){
     # For each subject, make a list of samples that are adjacent to at least 
     # one other sample
     
     # Get all the samples for that subject
     samples <- 
          filter(samdf,
                 subj == s) %>% 
          pull(row)
     
     # Subset the distance matrix
     d.temp <- d[samples, samples]
     
     # Pull any column that has a 1 (indicating a consecutive day from at least
     # one other subject)
     adjacent[[s]] <- 
          apply(d.temp, 2, function(x){any(x == 1)}) %>% 
          which() %>% 
          names()
}
```

```{r}
# How many subjects have at least one consecutive sample?
lapply(adjacent, length) %>% 
     unlist()
```

```{r}
# Drop those without
adjacent <- 
     adjacent[lapply(adjacent, 
                     function(x){length(x) > 0}) %>% 
                   unlist()]

length(adjacent)
```

Okay, so now want to calculate correlation from here. 
```{r}
set.seed(1232022)

for (iter in (1:100)){
     adjacent.df <- data.frame()
     for (s in names(adjacent)){
          # Randomly sample one of that subject's samples
          s1 <- sample(adjacent[[s]],
                       size = 1)
          
          # Subset the distance matrix to only that subject's samples
          d.temp <- d[s1, adjacent[[s]]]
          s2s <- 
               which(d.temp == 1) %>% 
               names()
               
          # Find another sample that's 1 day away in distance
          s2 <- sample(s2s, size = 1)
          
          # Pull observed values, calculate mean, join to FVS
          alpha.temp <- 
               alpha.1 %>% 
               filter(name %in% c(s1, s2)) %>% 
               group_by(subj) %>% 
               summarize(mean_obs = mean(obs)) %>% 
               left_join(fvs, by = c('subj'))
          
          # Add to the running list of values for this iteration
          adjacent.df <- bind_rows(adjacent.df,
                                   alpha.temp)
     }
          
     # Calculate correlation
     # Common warning: cannot compute exact p-value with ties
     suppressWarnings(spearman <- cor.test(adjacent.df$mean_obs, 
                                           adjacent.df$fvs_plant_residual,
                                           method = 'spearman'))
     
     # Update results dataframe
     row <- c(2, 
              'Adult-1', 
              spearman$estimate, 
              spearman$p.value)
     names(row) <- c('scheme', 'study', 'rho', 'p')
     results.df <- bind_rows(results.df, row)
}
```

#### 3

```{r}
nhanes <- list()
for (s in unique(samdf$subj)){
     # For each subject, make a list of samples that are 3-10 days from least 
     # one other sample
     
     # Get all the samples for that subject
     samples <- 
          filter(samdf,
                 subj == s) %>% 
          pull(row)
     
     # Subset the distance matrix
     d.temp <- d[samples, samples]
     
     # Pull any column that has a 1 (indicating a consecutive day from at least
     # one other subject)
     nhanes[[s]] <- 
          apply(d.temp, 2, function(x){any(x %in% 3:10)}) %>% 
          which() %>% 
          names()
}
```

```{r}
# Unlike adjacent sampling, this includes all subjects
length(nhanes)
```

Okay, so now want to calculate correlation from here. 
```{r}
set.seed(1232022)

for (iter in (1:100)){
     nhanes.df <- data.frame()
     for (s in names(nhanes)){
          # Randomly sample one of that subject's samples
          s1 <- sample(nhanes[[s]],
                       size = 1)
          
          # Subset the distance matrix to only that subject's samples
          d.temp <- d[s1, nhanes[[s]]]
          s2s <- names(d.temp)[d.temp %in% 3:10]
               
          # Find another sample that's 1 day away in distance
          s2 <- sample(s2s, size = 1)
          
          # Pull observed values, calculate mean, join to FVS
          alpha.temp <- 
               alpha.1 %>% 
               filter(name %in% c(s1, s2)) %>% 
               group_by(subj) %>% 
               summarize(mean_obs = mean(obs)) %>% 
               left_join(fvs, by = c('subj'))
          
          # Add to the running list of values for this iteration
          nhanes.df <- bind_rows(nhanes.df,
                                 alpha.temp)
     }
          
     # Calculate correlation
     # Common warning: cannot compute exact p-value with ties
     suppressWarnings(spearman <- cor.test(nhanes.df$mean_obs, 
                                           nhanes.df$fvs_plant_residual,
                                           method = 'spearman'))
     
     # Update results dataframe
     row <- c(3, 
              'Adult-1', 
              spearman$estimate, 
              spearman$p.value)
     names(row) <- c('scheme', 'study', 'rho', 'p')
     results.df <- bind_rows(results.df, row)
}
```

#### 4 through 7 (weekly)

```{r}
# Schemes 4-7 (3 to 6 weeks of weekly sampling)
set.seed(1232021)

for (i in 3:6){ # Number of samples (only from separate weeks)
     for (iter in (1:100)){
          # Subsample
          samples <- 
               alpha.1 %>% 
               group_by(subj, CH_week) %>% 
               slice_sample(n = 1) %>% # Pick one sample from each week at random
               ungroup() # 6 random weekly samples

          # Now, based on # samples needed, randomly sample a starting week
          # that will provide the right number of consecutive samples
          # 3: start_week from 1 to 4
          # 4: start_week from 1 to 3
          # 5: start_week from 1 to 2
          # 6: start_week is 1
          
          start_week_max <- 6 - i + 1
          samples_start <- 
               samples %>% 
               filter(CH_week <= start_week_max) %>% 
               group_by(subj) %>% 
               arrange(subj, CH_week) %>% 
               slice_sample(n = 1) %>% 
               select(subj, 
                      CH_week_start = CH_week)
          
          # Now join back to randomly selected samples
          samples <- 
               samples %>% 
               left_join(samples_start, by = 'subj') %>% 
               # Choose the right number of weeks
               filter(CH_week >= CH_week_start & 
                           CH_week < CH_week_start + i) %>% 
               # Pull sample names
               pull(name)
          
          # Pull observed values, calculate mean, join to FVS
          alpha.temp <- 
               alpha.1 %>% 
               filter(name %in% samples) %>% 
               group_by(subj) %>% 
               summarize(mean_obs = mean(obs)) %>% 
               left_join(fvs, by = c('subj'))
          
          # Calculate correlation
          # Common warning: cannot compute exact p-value with ties
          suppressWarnings(spearman <- cor.test(alpha.temp$mean_obs, 
                                                alpha.temp$fvs_plant_residual,
                                                method = 'spearman'))
          
          # Update results dataframe
          row <- c(i + 1, 
                   'Adult-1', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

#### Summarize and plot

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
# Intercept: get from full dataset
rho_full = 0.5214177
p_full = 0.008975764	
```

```{r}
# Sampling scheme labels:
schemes <- 
     c('Single sample\n',
       'Two samples\n(1d apart)', 
       'Two samples\n(3-10d apart)',
       'Three samples\n(1/wk)',
       'Four samples\n(1/wk)',
       'Five samples\n(1/wk)',
       'Six samples\n(1/wk)')
```


```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = rho_full,
                linetype = 'dotted',
                color = '#8CD17D',
                size = 1.5) +
     scale_x_continuous(breaks = 1:7,
                        labels = schemes) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme(axis.title.y = ggtext::element_markdown(),
           axis.text.x = element_text(angle = 40, hjust = 1),
           panel.grid.minor.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Adult-1 trnL x FVS by sampling scheme, rho.pdf')),
       height = 60, width = 80, units = 'mm')
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05,
                # yintercept = p_full,
                linetype = 'dotted',
                color = '#E15759',
                size = 1.5) +
     labs(y = 'Spearman *p*', x = 'Samples (n)') +
     scale_x_continuous(breaks = 1:7,
                        labels = schemes) +
     # scale_y_log10() +
     theme(axis.title.y = ggtext::element_markdown(),
           axis.text.x = element_text(angle = 40, hjust = 1),
           panel.grid.minor.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '4',
            paste0(plotID, '_Adult-1 trnL x FVS correlations by sampling scheme, p.pdf')),
       height = 60, width = 80, units = 'mm')
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```
 
```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.52, # Hard-coded from prior result 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x hPDI correlations restricted to 1 random sample per week, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x hPDI correlations restricted to 1 random sample per week, p.pdf')),
       height = 4, width = 3)
```