---
title: "Adult-1, Adult-2 diversity indices"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseqs
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### ASA24

```{r}
# CHOMP-only food counts
recall <- 
     here('..',
          'chomp',
          'data', 
          'processed',
          'ASA24', 
          'Food_Tree',
          'richness_ASA24.rds') %>% 
     readRDS()
```

### DHQ3

#### Adult-1

```{r}
ffq.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3.csv') %>% 
     read_csv()

head(ffq.adult1)
```

#### Adult-2

```{r}
ffq.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3.csv') %>% 
     read_csv()

head(ffq.adult2)
```

```{r}
# Join together
ffq <- bind_rows(ffq.adult1, ffq.adult2)

rm(ffq.adult1, ffq.adult2)
```

# Pre-process

## trnL

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(alpha, add,
              by = c('name' = 'row'))
```

Then below, can just filter by desired threshold in order to calculate measures.

## Diet surveys

### ASA24

```{r}
recall
```

### DHQ3

```{r}
range(ffq$`Energy (kcal)`, na.rm = TRUE)
```


```{r}
# Crude quality filter
ffq <- filter(ffq, 
              !(`Energy (kcal)` < 500 | `Energy (kcal)` > 4000))
```

```{r}
ffq <- select(ffq, 
              subj,
              sex = `Sex (1=male; 2=female)`, 
              matches('HEI'))
names(ffq)
```

```{r}
ffq <- 
     ffq %>% 
     rename(hei_tot = 'Total HEI-2015 Score',
            hei_veg = 'HEI-2015 - Total Vegetables - Component Score',
            hei_greensbeans = 'HEI-2015 - Greens and Beans - Component Score',
            hei_fruit = 'HEI-2015 - Total Fruits - Component Score',
            hei_wholefruits = 'HEI-2015 - Whole Fruits - Component Score',
            hei_grain = 'HEI-2015 - Whole Grains - Component Score') %>% 
     # Make custom plant column
     mutate(hei_plant_all = hei_veg + hei_greensbeans + hei_fruit + 
                 hei_wholefruits + hei_grain,
            hei_non_plant = hei_tot - hei_plant_all)

ffq
```
# Analyze

```{r}
samdf <- data.frame(ps@sam_data)
```

## Richness by participant

```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     # summarize(med = median(obs_rare, na.rm = TRUE), 
     #           sd = sd(obs_rare, na.rm = TRUE)) %>% 
     summarize(med = median(obs, na.rm = TRUE),
               sd = sd(obs, na.rm = TRUE)) %>%
     arrange(desc(med))

alpha$subj <- factor(alpha$subj, levels = alpha.stats$subj)
```

```{r}
alpha %>% 
     ggplot(aes(x = subj, y = obs, color = study)) +
     facet_wrap(~study, scales = 'free_x') +
     geom_boxplot() +
     labs(x = 'Participant', y = '*trnL* richness') +
     scale_color_manual(values = c('black', 'gray60')) +
     # scale_fill_manual(values = c('black', 'gray60')) +
     theme_classic() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title = element_text(size = 14,
                                    face = 'bold'),
          axis.title.y = ggtext::element_markdown(),
          strip.text = element_text(size = 12,
                                    face = 'bold'),
          legend.position = 'None'
     )
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'manuscript', '2',
            paste0(plotID, '_A_raw trnL richness by participant.pdf')),
       height = 3, width = 9)
```

```{r}
adult.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(obs_rare,
                            na.rm = TRUE),
               mean = mean(obs_rare,
                           na.rm = TRUE),
               sd = sd(obs_rare,
                       na.rm = TRUE),
               var = var(obs_rare,
                         na.rm = TRUE))

range(adult.stats$med, na.rm = T)
range(adult.stats$sd, na.rm = T)
```

### Within-subject vs. between-subject

#### ANOVA (R01 ONLY)

```{r}
# Adult-1 
mod.1 <- 
     alpha %>% 
     filter(study == 'Adult-1') %>% 
     aov(obs ~ subj, data = . )

summary(mod.1)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod.1)[[1]]['Mean Sq'][1,]
var.within <- summary(mod.1)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     filter(study == 'Adult-1') %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```


```{r}
# Adult-2 
mod.2 <- 
     alpha %>% 
     filter(study == 'Adult-2') %>% 
     aov(obs ~ subj, data = . )

summary(mod.2)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod.2)[[1]]['Mean Sq'][1,]
var.within <- summary(mod.2)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     filter(study == 'Adult-2') %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```


### Overall

```{r}
mod <- 
     aov(obs ~ subj, data = alpha)

summary(mod)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod)[[1]]['Mean Sq'][1,]
var.within <- summary(mod)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```

#### Permutation-based strategy

##### Raw data

Aggregating within-subject CVs using info from (this site) [https://www-users.york.ac.uk/~mb55/meas/cv.htm]: "We find the coefficient of variation for each subject separately, square these, find their mean, and take the square root of this mean. We can call this the root mean square approach."

```{r}
within.between <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(sigma = sd(obs, na.rm = T),
               mu = mean(obs, na.rm = T)) %>% 
     mutate(cv_within = sigma/mu)
```

```{r}
# Now summarize as above
within.between %>% 
     ungroup() %>% 
     mutate(cv_within_sq = cv_within^2) %>% # Square them
     # group_by(study) %>% 
     summarize(mean_sq = mean(cv_within_sq)) %>% # Find the mean
     mutate(root_mean_sq = sqrt(mean_sq)) # Take the square root
```
```{r}
# Now calculate for overall data
alpha %>% 
     group_by(study, subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     # group_by(study) %>% 
     summarize(
          mu_pop = mean(mu_subj, 
                        na.rm = T), # Population mean of mean per-subject obs.
          sigma_pop = sd(mu_subj, 
                         na.rm = T)) %>% # Population sd
     mutate(cv_between = sigma_pop/mu_pop) 
```

```{r}
# Calculate standard deviation for overall cohort and join to data
within.between <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     group_by(study) %>% 
     summarize(
          mu_pop = mean(mu_subj, 
                        na.rm = T), # Population mean of mean per-subject richness
          sigma_pop = sd(mu_subj, 
                         na.rm = T)) %>% # Population sd
     mutate(cv_between = sigma_pop/mu_pop) %>% 
     select(study, cv_between) %>% 
     right_join(within.between)
```

```{r}
# Calculate difference
within.between <- 
     mutate(within.between,
            delta = cv_within - cv_between)
```

```{r}
# Plot raw data
ggplot(within.between, aes(x = study, y = cv_within)) +
     geom_boxplot() +
     ylab('Within-subject CV') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 
            'development', 
            'docs',
            'r01', 
            '2021_11',
            'within-between',
            paste0(plotID, '_CV on rarefied richness.pdf')),
       height = 3, width = 3)
```

```{r}
# Plot delta
ggplot(within.between, aes(x = study, y = delta)) +
     geom_boxplot() +
     ylab('Within-subject CV - between-subject CV') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 
            'development', 
            'docs',
            'r01', 
            '2021_11',
            'within-between',
            paste0(plotID, '_delta CV on raw richness.pdf')),
       height = 3, width = 3)
```

##### Shuffled data

```{r}
set.seed(10222021)

# Shuffle
alpha.rand <- alpha 
alpha.rand$obs <- sample(alpha.rand$obs, length(alpha.rand$obs))
```

```{r}
within.between.rand <- 
     alpha.rand %>% 
     group_by(study, subj) %>% 
     summarize(sd_within = sd(obs))
```

```{r}
# Calculate standard deviation for overall cohort ad join to data
within.between.rand <- 
     alpha.rand %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs)) %>% 
     ungroup() %>% 
     group_by(study) %>% 
     summarize(sd_between = sd(mean)) %>% 
     right_join(within.between.rand)
```

```{r}
# Calculate difference
within.between.rand <- 
     mutate(within.between.rand,
            delta = sd_within - sd_between)
```

```{r}
# Bind together
within.between$data <- 'Raw'
within.between.rand$data <- 'Shuffled'
```

```{r}
within.between <- bind_rows(within.between,
                            within.between.rand)
```

```{r}
ggplot(within.between, aes(x = data, y = delta)) +
     geom_boxplot() + 
     facet_wrap(~study) + 
     labs(y = 'sd (within) - sd (between)') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```


#### KS test

```{r}
# Get vector for between-subject distribution: the mean trnL richness of each subject

between <- 
     adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     pull(mean)

subjects <- 
     adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     pull(subj) %>% 
     sort()

results.df <- 
     data.frame(
          study = NULL,
          subj = NULL,
          D = NULL,
          P = NULL
     )

for (s in subjects){
     # Compare between-subjects distribution to within-subject distribution
     # for this particular subject
     result <- 
          alpha %>% 
          filter(subj == s) %>% 
          pull(obs_rare) %>% 
          ks.test(., between)
     
     row = c('Adult-1', s, result$statistic, result$p.value)
     names(row) <- c('study', 'subj', 'D', 'p')
     
     results.df <- bind_rows(results.df, row)
}
```

```{r}
# Plot results as p-value histogram
results.df$p <- as.numeric(results.df$p)

hist(results.df$p, breaks = 20)
```


```{r}
fit.within <- 
     aov(obs ~ subj,
         data = alpha)
```

```{r}
summary(fit.within)
```

Think I need an additional layer of summarizing here??

```{r}
fit.between <- 
     aov(obs ~ study,
         data = alpha)
```

```{r}
summary(fit.between)
```


```{r}
aov(obs ~ study + subj,
         data = alpha) %>% 
     summary()
```


### Mean vs. variance

```{r}
# Add cohort information in order to calculate separately
adult.stats <- 
     samdf %>% 
     select(subj, study) %>% 
     right_join(adult.stats) %>% 
     distinct()
```

```{r}
# Adult-1
adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + var,
              data = ., 
              method = 'spearman')
```

```{r}
# Adult-2
adult.stats %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + var,
              data = ., 
              method = 'spearman')
```

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs),
               mean_shannon = mean(Shannon),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.mean, aes(x = mean, y = mean_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Mean *trnL* richness<br>\n(All samples)', 
          y = 'Mean rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             'supplemental',
#             paste0(plotID, '_rarefied vs observed trnL richness (mean).pdf')),
#        height = 4, width = 4)
```

```{r}
# Average trnL richness within each subject
alpha.median <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(median = median(obs, na.rm = TRUE),
               median_rare = median(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.median, aes(x = median, y = median_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Median *trnL* richness<br>\n(All samples)', 
          y = 'Median rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             'supplemental',
#             paste0(plotID, '_rarefied vs observed trnL richness (median).pdf')),
#        height = 4, width = 4)
```

Based on above plots, move forward with mean

## HEI

### Mean per-participant pMR

Think I want to pivot the data wider, then calculate means, then pivot long again.

```{r}
head(alpha)
```
```{r}
alpha.wide <- 
     alpha %>% 
     select(-c(se, reads_food)) %>% 
     pivot_wider(names_from = 'threshold',
                 values_from = 'obs_rare',
                 names_glue = '{.value}_{threshold}') %>% 
     select(name, starts_with('obs'), everything())

alpha.wide
```
```{r}
# Now, calculate means

# Ignoring NAs
alpha.mean <- 
     alpha.wide %>% 
     group_by(study, subj) %>% 
     summarise(across(starts_with('obs'),
                      ~mean(.x, na.rm = TRUE))) %>% 
     ungroup()

# Replacing NAs with 0s (strict interpretation as *no* plant matter)
alpha.mean.0 <- 
     alpha.wide %>% 
     mutate(across(starts_with('obs_'),
                   ~ifelse(is.na(.x),
                           yes = 0,
                           no = .x))) %>% 
     group_by(study, subj) %>% 
     summarise(across(starts_with('obs'),
                      ~mean(.x, na.rm = TRUE))) %>% 
     ungroup()

rm(alpha.wide)
```

### Join to FFQ

```{r}
# Join to FFQ data
ffq <- 
     alpha.mean.0 %>%
     # alpha.mean %>% 
     left_join(ffq)
```

```{r}
# How many subjects are missing DHQ3 data?
ffq %>% 
     select(starts_with(c('HEI', 'hei'))) %>% 
     apply(1, function(x){all(is.na(x))}) %>% 
     sum()
```

This is 5 at baseline, and 9 with crude calorie filter.

### Visualizations (joint)

```{r}
# Make long for easy vis at all rarefaction thresholds
ffq.long <- 
     ffq %>%
     rename(obs_Raw = obs) %>% 
     pivot_longer(starts_with('obs'),
                  names_to = 'rarefaction',
                  values_to = 'pMR') %>% 
     select(study, subj, rarefaction, pMR, everything()) %>% 
     mutate(rarefaction = gsub(rarefaction, 
                               pattern = 'obs_|obs_rare_',
                               replacement = '')) %>% 
     mutate(rarefaction = factor(rarefaction,
                                 levels = c('Raw',
                                            '10000',
                                            '5000',
                                            '1000')))

ffq.long
```


```{r}
ggplot(ffq.long, aes(x = hei_plant_all, y = pMR)) +
     geom_point(alpha = 0.7, size = 2) +
     facet_grid(vars(study), vars(rarefaction),
                scales = 'free') +
     labs(y = 'Mean pMR', 
          x = 'HEI-2015\nplant components score')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '3',
            paste0(plotID, 
                   '_rarefied trnL richness (below threshold as 0) x HEI plant.pdf')),
       device = cairo_pdf(),
       height = 3, width = 7)
```

```{r}
# Manuscript plot
ffq.long %>% 
     filter(rarefaction == 1000) %>% 
     ggplot(aes(x = hei_plant_all, y = pMR)) +
     geom_point(alpha = 0.7, size = 2) +
     facet_wrap(~study) +
     labs(y = 'Mean pMR', 
          x = 'HEI-2015\nplant components score')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '3',
            paste0(plotID, 
                   '_rarefied trnL richness (below threshold as 0) x HEI plant.pdf')),
       device = cairo_pdf(),
       height = 3, width = 4)
```

```{r}
# Plant HEI
ffq.long %>% 
     group_by(study, rarefaction) %>% 
     summarize(rho = cor.test(~pMR + hei_plant_all, 
                               method = 'spearman')$estimate,
               p = cor.test(~pMR + hei_plant_all,
                             method = 'spearman')$p.value)
```
```{r}
# Total HEI
ffq.long %>% 
     group_by(study, rarefaction) %>% 
     summarize(rho = cor.test(~pMR + hei_tot, 
                               method = 'spearman')$estimate,
               p = cor.test(~pMR + hei_tot,
                             method = 'spearman')$p.value)
```


```{r}
# Non-plant HEI
ffq.long %>% 
     group_by(study, rarefaction) %>% 
     summarize(rho = cor.test(~pMR + hei_non_plant, 
                               method = 'spearman')$estimate,
               p = cor.test(~pMR + hei_non_plant,
                             method = 'spearman')$p.value)
```
## Sampling scheme

### Adult-1

```{r}
# Get sample data for Adult-2
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adult-1')
```

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

Note that this will cover duplicated sample sets as i gets closer to 6, but can detect and remove them downstream.
```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(6302021)

for (i in 1:18){ # Number of samples
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj) %>% 
               slice_sample(n = i) %>% 
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(ffq)
          
          # Calculate correlation by study
          spearman <- cor.test(ffq.temp$obs_mean, ffq.temp$hei_plant_all,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-1', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
# Intercept: get from full dataset
y <- results.df$rho[results.df$scheme == 18]
```

```{r}
# Plot
results.df %>% 
     filter(scheme %in% seq(3, 18, 3)) %>% 
     ggplot(aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = y, 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = seq(3, 18, 3)) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-1 trnL x HEI correlations by sampling scheme, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
results.df %>% 
     filter(scheme %in% seq(3, 18, 3)) %>% 
     ggplot(aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = seq(3, 18, 3)) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-1 trnL x HEI correlations by sampling scheme, p.pdf')),
       height = 4, width = 3)
```

#### Restrict to separate weeks

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(6202021)

for (i in 1:6){ # Number of samples (only from separate weeks)
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj, CH_week) %>% 
               slice_sample(n = 1) %>% # Pick one sample from each week at random
               ungroup() %>% 
               group_by(subj) %>% # 6 random weekly samples
               slice_sample(n = i) %>% # Pick i weeks' worth of samples
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(ffq)
          
          # Calculate correlation by study
          spearman <- cor.test(ffq.temp$obs_mean, ffq.temp$hei_plant_all,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-1', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```
Note that 1,000 iterations saturates this measure, and gives a similar result to 1000. 
```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.4, # Hard-coded from prior result 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x HEI correlations restricted to 1 random sample per week, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x HEI correlations restricted to 1 random sample per week, p.pdf')),
       height = 4, width = 3)
```

### Adult-2

```{r}
# Get sample data for Adult-2
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adult-2')
```

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

Note that this will cover duplicated sample sets as i gets closer to 6, but can detect and remove them downstream.
```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(6302021)

for (i in 1:6){ # Number of samples
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj) %>% 
               slice_sample(n = i) %>% 
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(ffq)
          
          # Calculate correlation by study
          spearman <- cor.test(ffq.temp$obs_mean, ffq.temp$hei_plant_all,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-2', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# Intercept: get from full dataset
y <- results.df$rho[results.df$scheme == 6]
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = y, 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-2 trnL x HEI correlations by sampling scheme, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-2 trnL x HEI correlations by sampling scheme, p.pdf')),
       height = 4, width = 3)
```

#### Restrict to separate weeks

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

```{r}
# Update Adult-2 sample data to have a "week" entry
samdf <- 
     mutate(samdf,
            ONR_week = gsub('T|F', '', ONR_day),
            ONR_week = as.numeric(ONR_week))
```

```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(6202021)

for (i in 1:3){ # Number of samples (only from separate weeks)
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj, ONR_week) %>% 
               slice_sample(n = 1) %>% # Pick one sample from each week at random
               ungroup() %>% 
               group_by(subj) %>% # 3 random weekly samples
               slice_sample(n = i) %>% # Pick i weeks' worth of samples
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(ffq)
          
          # Calculate correlation by study
          spearman <- cor.test(ffq.temp$obs_mean, ffq.temp$hei_plant_all,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-2', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.38, # Hard-coded from prior result 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:3) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-2 trnL x HEI correlations restricted to 1 random sample per week, rho.pdf')),
       height = 4, width = 2)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:3) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-2 trnL x HEI correlations restricted to 1 random sample per week, p.pdf')),
       height = 4, width = 2)
```


```{r}
for (scheme in c('s2', 's3', 's4', 's5')){ # For each sampling scheme
     scheme.var <- get(scheme)
     for (i in seq_along(scheme.var)){ # For each set of samples within that scheme
          # Make subsampled phyloseq
          ps.temp <- subset_samples(ps.foods,
                                    ONR_day %in% scheme.var[[i]])
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
               estimate_richness(ps.temp, 
                                 measures = c('Observed', 
                                              'Shannon')) %>% 
               rownames_to_column(var = 'name') %>% 
               mutate(subj = gsub('_.*$', '', name))) # Get subject ID
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(mean_Obs = mean(Observed),
                         mean_Shannon = mean(Shannon)) %>% 
               left_join(ffq)
          
          # Calculate Pearson correlation
          pearson <- cor.test(ffq.temp$mean_Shannon, ffq.temp$hei_plant_all,
                              alternative = 'greater',
                              method = 'pearson')
          
          # Update results dataframe
          row <- c(scheme, i, pearson$estimate, pearson$p.value)
          names(row) <- c('scheme', 'i' , 'Pearson_r', 'Pearson_p')
          results.df <- bind_rows(results.df, row)
          }
}

```

Plot results:

```{r}
# Pivot longer for side-by-side plotting
results.df <- 
     pivot_longer(results.df,
                  cols = starts_with('Pearson'),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     mutate(value = as.numeric(value))
```

Outliers are throwing off plot limits, and I can't adjust across facets. Just save plots separately and display side-by side.

```{r}
labels <- c('1 sample',
            '2 samples,\nconsecutive',
            '2 samples,\nalternate week',
            '3 samples,\nconsecutive',
            '3 samples,\nweekly')

results.df %>% 
     filter(measure == 'Pearson_p') %>% 
     ggplot(aes(x = scheme, y = value)) +
     geom_boxplot(outlier.shape = NA, size = 1.5) +
     # geom_hline(yintercept = 0.43, size = 1.5, alpha = 0.5,
     #            color = 'springgreen4', linetype = 'dashed') +
     geom_hline(yintercept = 0.05, size = 1.5, alpha = 0.5,
                color = 'red', linetype = 'dashed') +
     scale_x_discrete(labels = labels) +
     labs(y = 'Pearson *p*-value') +
     # labs(y = 'Pearson *r*') +
     ylim(0, 0.25) +
     theme_bw() +
     theme(axis.line = element_line(size = 1.5),
           axis.ticks = element_line(size = 1.5),
           axis.text.x = element_text(angle = 30, hjust = 1,
                                      size = 14, face = 'bold', color = 'gray40'),
           axis.text.y = element_text(size = 14, face = 'bold', color = 'gray40'),
           axis.title.x = element_blank(), 
           axis.title.y = ggtext::element_markdown(size = 18, face = 'bold', 
                                                   color = 'gray40'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 'development', 'results', '2021 R01',
            paste0(plotID, '_trnL x HEI correlations by sampling scheme.png')),
       height = 5, width = 5)
```


## vs. ASA24

```{r}
# Concatenate richness measures
recall.rows <- 
     select(recall, subj = ID, starts_with('richness')) %>% 
     pivot_longer(cols = starts_with('richness'),
                  names_to = 'category',
                  values_to = 'value') %>% 
     mutate(category = gsub('richness_', '', category))

trnL.rows <- 
     rare.df %>% 
     filter(study == 'Adult-1') %>% 
     select(subj, value = obs_rare) %>% 
     mutate(category = 'plant')

recall.rows$measure = '24-hour recall'
trnL.rows$measure = 'Metabarcoding'

trnL.v.asa <- 
     bind_rows(recall.rows, trnL.rows)
```

```{r}
trnL.v.asa %>% 
     filter(category == 'plant') %>% 
     ggplot(aes(x = subj, y = value, color = measure)) +
     # geom_point()
     geom_boxplot() +
     theme_bw() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
# Try averaging within each participant to get one value
trnL.v.asa.avg <- 
     trnL.v.asa %>% 
     filter(category == 'plant') %>% 
     group_by(measure, subj) %>% 
     summarize(mean = mean(value)) %>% 
     pivot_wider(names_from = 'measure',
                 values_from = 'mean')
```

```{r}
ggplot(trnL.v.asa.avg, aes(x = `24-hour recall`, y = Metabarcoding)) +
     geom_point() +
     xlim(0, 20) + ylim(0, 20) +
     coord_equal()
```
Consider: haven't filtered data for completeness?

```{r}
cor.test(trnL.v.asa.avg$`24-hour recall`, 
         trnL.v.asa.avg$Metabarcoding,
         method = 'spearman')
```

