---
title: "Adult-1, Adult-2 FVS"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### DHQ3-based FVS

Calculated in Adult-1, Adult-2 dietary indices notebook.

```{r}
fvs <- 
     here('data',
          'processed',
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

# Pre-process

## trnL

### Filter phyloseq

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(alpha, add,
              by = c('name' = 'row'))
```

Then below, can just filter by desired threshold in order to calculate measures.

### Mean richness

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj, threshold) %>% 
     summarize(mean = mean(obs),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

# Analyze

## FVS

```{r}
# Join mean richness to PDI data
fvs <- 
     alpha.mean %>% 
     left_join(fvs)
```

### Raw data

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_tot, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'FVS\n(all items)',
          y = 'Mean pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x total FVS.pdf')),
       height = 2.5, width = 4)
```


```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_tot + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_tot + mean,
                            method = 'spearman')$p.value) 
            
```

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_plant, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'FVS\n(plant items)',
          y = 'Mean pMR')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x plant FVS.pdf')),
       height = 2.5, width = 4)
```

```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_plant + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_plant + mean,
                            method = 'spearman')$p.value) 
            
```

### Residuals

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_tot_residual, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'Food Variety Score residual\n(all items)',
          y = 'Mean pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x total FVS.pdf')),
       height = 2.5, width = 4)
```


```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_tot_residual + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_tot_residual + mean,
                            method = 'spearman')$p.value) 
            
```

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs_plant_residual, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study,
                nrow = 2) +
     labs(x = 'FVS\n(plant items)',
          y = 'Mean pMR')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_raw trnL richness x plant FVS.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs_plant_residual + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_plant_residual + mean,
                            method = 'spearman')$p.value) 
            
```

### Rarefied data

```{r}
fvs %>% 
     ggplot(aes(x = fvs_plant, y = mean_rare)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_grid(~study + threshold) +
     labs(x = 'Food Variety Score',
          y = 'Mean pMR')
```

```{r}
fvs %>% 
     group_by(study, threshold) %>% 
     summarize(rho = cor.test(~ fvs_plant_residual + mean_rare,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs_plant_residual + mean_rare,
                            method = 'spearman')$p.value) 
            
```

Interesting. Only significant at the lowest level of rarefaction.  Why is this?

One could be we're tossing out samples.
Another could be that small amounts shouldn't be contributing to our richness estimates.
Would expect this to stabilize if I tried something like Shannon, which would minimize the contribution.  This might be a nice place to explore.