---
title: "Adult 1, Adult-2 food accumulation curves"
output: html_notebook
---

See what unique foods accumulate over consecutive daily samples.
From Ruel: "An interesting analysis by Drewnowski and colleagues measured cumulative dietary variety (based on individual foods) in American adults over a period of 15 d (9). The individual curves show that, as expected, the number of different foods consumed increases with time and eventually plateaus at a point that defines a person’s entire "food repertoire" over this period."

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    legend.text = element_text(size = 11))
)
```

# Setup

```{r setup, include = FALSE, echo = FALSE}
require(knitr)
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(phyloseq)
library(tidyverse)
```

# Read in data

```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed', 
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

```{r}
# Rarefied counts
alpha <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS()

dim(alpha)
```

# Pre-process

```{r}
# Subset to Adult-1, Adult-2 samples
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     subset_samples(reads_food > 0) %>% 
     prune_taxa(taxa_sums(.) > 0 , .)

ps
```

```{r}
# Melt
melt <- psmelt(ps)
```

```{r}
# Simplify sample data
melt <- 
     select(melt,
            OTU:CH_weekday, ONR_day,
            superkingdom:name)
```

```{r}
# Sequence samples
# Adult-1: 
melt <- 
     melt %>% 
     # Create combination week-weekday label
     mutate(CH_day = paste(melt$CH_week, melt$CH_weekday)) %>% 
     # Recode NA
     mutate(CH_day = ifelse(CH_day == 'NA NA',
                            yes = NA,
                            no = CH_day))

unique(melt$CH_day) %>% sort()
```

```{r}
unique(melt$ONR_day)
```

```{r}
# Collapse this ordering into a single variable, so it can be iterated over below
melt <- 
     melt %>% 
     unite('study_day', c(CH_day, ONR_day), na.rm = TRUE) %>% 
     # Simplify columns
     select(Sample, 
            Abundance,
            study,
            subj, 
            study_day,
            superkingdom:name)
```

```{r}
# Make a factor to structure ordering
melt$study_day <- factor(melt$study_day,
                      levels = c(
                           # Adult-1 levels
                           '1 Wednesday', 
                           '1 Thursday', 
                           '1 Friday', 
                           '2 Wednesday', 
                           '2 Thursday', 
                           '2 Friday', 
                           '3 Wednesday', 
                           '3 Thursday', 
                           '3 Friday', 
                           '4 Wednesday', 
                           '4 Thursday', 
                           '4 Friday', 
                           '5 Wednesday', 
                           '5 Thursday', 
                           '5 Friday', 
                           '6 Wednesday', 
                           '6 Thursday', 
                           '6 Friday',
                           '6 Saturday',
                           # Adult-2 levels
                           '2', # T1
                           '5', # F1
                           '9', # T2
                           '12', # F2
                           '16', # T3
                           '19')) # F3
```

## Richness

```{r}
# Filter and simplify rarefied richness
alpha <- 
     alpha %>% 
     filter(grepl(name,
                  pattern = 'CHOMP|ONR') & reads_food > 0)
```

```{r}
# Add full obs, not just those passing filter
asvtab <- ps@otu_table@.Data
add <- apply(asvtab, 1, function(x){sum(x > 0)})

add <- 
     data.frame(obs = add) %>% 
     rownames_to_column(var = 'name')
     
# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) %>% 
     right_join(add, by = c('row' = 'name')) %>% 
     rename(name = row)

alpha <- 
     full_join(alpha,
               add)

rm(add)
```

# Analyze

## Calculate food accumulation

```{r}
# Make dataframe for storing results
results.df <- 
     data.frame(
          study = NULL,
          subj = NULL,
          sample_n = NULL,
          detected_so_far = NULL
     )
```

```{r}
# Check if any study day is NA
sum(is.na(melt$study_day))
```

```{r}
# Iterate over each subject
for (s in unique(melt$subj)){
     
     # Filter to only that subject's samples, and taxa > 0
     melt.s <- 
          filter(melt, subj == s & Abundance > 0)
     
     # Get number of samples
     n <- n_distinct(melt.s$study_day)
     n
     
     # Arrange in order
     melt.s <- arrange(melt.s, study_day)

     # Get study day of each
     days <- unique(melt.s$study_day)
     
     # Initial entries
     detected_names = NULL # No plants detected yet 
     
     # Add origin to data
     row <- 
          data.frame(study = unique(melt.s$study),
                     subj = unique(melt.s$subj),
                     sample_n = 0,
                     detected_so_far = 0) 
     
     # Append to results
     results.df <- bind_rows(results.df,
                             row)

     # Then, iterate over each one of that subject's samples in order, counting
     # cumulative trnL taxa as we go
     for (i in seq(n)){
          current <- 
               filter(melt.s, 
                      study_day == days[i]) %>% 
               pull(name) %>% 
               sort()
          
          current <- union(detected_names, current)
          
          row <- 
               data.frame(study = unique(melt.s$study),
                          subj = unique(melt.s$subj),
                          sample_n = i,
                          detected_so_far = length(current)) 
          
          # Append to results
          results.df <- bind_rows(results.df,
                                  row)
          
          # Update taxa detected so far
          detected_names <- current
     }
     
}
```

```{r}
# Do dimensions check out?
# We have number of unique subj-sample pairs, plus each subject again (0 row)

melt %>% 
     group_by(subj, study_day) %>% 
     count() %>% 
     dim()

n_distinct(melt$subj)
```

```{r}
# 423 + 58 # Rarefied data
566 + 60 # All data
dim(results.df)
```

## Plot

```{r}
ggplot(results.df, aes(x = sample_n, y = detected_so_far, group = subj)) +
     geom_line(alpha = 0.5) +
     labs(x = 'Number of samples', y = 'Cumulative pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
         'manuscript',
         'supplemental',
         paste0(plotID,
                '_Adult-1, Adult-2 taxon accumulation curves, all data.pdf')),
       height = 4,
       width = 5)
```

How does this correlate with mean number of observations?

```{r}
# Get the cumulative amount for each subject-study day pairing
endpoint <- 
     results.df %>% 
     group_by(subj) %>% 
     arrange(sample_n) %>% 
     summarize(obs_cumulative = last(detected_so_far))
```

```{r}
# Save accumulation data for subsequent use
# write_csv(
#      results.df,
#      here(
#           'data', 
#           'processed',
#           'accumulation',
#           'Adult-1 and Adult-2 cumulative taxon counts.csv'
#      )
# )
```

```{r}
# Average taxon counts within each participant
alpha.mean <- 
     alpha %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     group_by(subj) %>% 
     summarize(obs_mean = mean(obs),
               obs_rare_mean = mean(obs_rare, na.rm = TRUE))
```

```{r}
# Join the two together
endpoint <- full_join(endpoint, alpha.mean)
```

```{r}
ggplot(endpoint, aes(x = obs_mean, y = obs_cumulative)) +
     geom_point(size = 2, alpha = 0.6) +
     xlim(0, 70) + ylim(0, 70) +
     coord_equal() +
     labs(x = 'Mean pMR', y = 'Cumulative pMR') 
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
         'manuscript',
         'supplemental',
         paste0(plotID,
                '_Adult-1, Adult-2 cumulative taxa vs mean taxa.pdf')),
       device = cairo_pdf,
       height = 4,
       width = 4)
```

```{r}
# Calculate correlation
cor.test(endpoint$obs_cumulative, endpoint$obs_mean,
         method = 'spearman',
         alternative = 'greater')
```

## Fit curves

Species accumulation curves modeled with either an exponential or linear function. 

```{r}
# Make a numeric subject ID by cohort (1...n_1 in Adult 1, and 1...n_2 in Adult-2).
results.df <- 
     arrange(results.df, study, subj)

results.df <- 
     results.df %>% 
     group_by(study) %>% 
     mutate(subj_numeric = factor(subj, 
                                  labels = 1:n_distinct(subj))) %>% 
     ungroup()
```


```{r}
# Make overview plot of fits (Adult-1)
results.df %>% 
     filter(sample_n > 0) %>% # Remove intercept (for log below)
     filter(study == 'Adult-1') %>% 
     ggplot(aes(x = sample_n, y = detected_so_far)) +
     geom_point(size = 2, alpha = 0.6) +
     geom_smooth(aes(color = study),
                 method = 'lm', 
                 formula = y ~ log(x)) +
     facet_wrap(~ subj_numeric, 
                nrow = 7, ncol = 4) +
     ylim(0, 70) + xlim(0,18) +
     labs(y = 'Cumulative pMR', 
          x = 'Stool samples (*n*)') +
     theme(legend.position = 'none',
           strip.text = element_blank(),
           axis.title.x = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
         'manuscript',
         'supplemental',
         paste0(plotID,
                '_Adult-1 taxon accumulation curves.pdf')),
       device = cairo_pdf, 
       height = 10,
       width = 5)
```

```{r}
# Make overview plot of fits (Adult-2)
results.df %>% 
     filter(sample_n > 0) %>% # Remove intercept (for log below)
     filter(study == 'Adult-2') %>% 
     ggplot(aes(x = sample_n, y = detected_so_far)) +
     geom_point(size = 2, alpha = 0.6) +
     geom_smooth(aes(color = study),
                 method = 'lm', 
                 formula = y ~ log(x)) +
     facet_wrap(~ subj_numeric, 
                nrow = 8, ncol = 4) +
     ylim(0, 70) + xlim(0,18) +
     labs(y = 'Cumulative pMR', 
          x = 'Stool samples (*n*)') +
     theme(legend.position = 'none',
           axis.title.x = ggtext::element_markdown(),
           strip.text = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
         'manuscript',
         'supplemental',
         paste0(plotID,
                '_Adult-2 taxon accumulation curves.pdf')),
       device = cairo_pdf, 
       height = 10,
       width = 5)
```

```{r}
# Make dataframe for holding results
fits.df <- 
     data.frame(
          study = NULL,
          subj = NULL,
          intercept = NULL,
          intercept_p = NULL,
          beta = NULL,
          beta_p = NULL,
          r_squared = NULL
     )

# Add columns for predicted values: samples 1:20?
```

```{r}
subjects <- unique(results.df$subj)
for (s in subjects){
# Remove intercept from data
data <- 
     filter(results.df, subj == 'FLU18') %>% 
     filter(sample_n != 0)
}
```


```{r}
fit <- lm(detected_so_far ~ log(sample_n), 
         data)

summary(fit)
```

```{r}
predict(fit, newdata = data.frame(sample_n = 1:20))
```

