---
title: "Adult-1, Adult-2 PDI"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS

rare.df
```
## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### DHQ3-based PDI

```{r}
# Adult-1
pdi.adult1 <-
     here('data', 
          'processed', 
          'Adult-1', 
          'PDI scores by subject.csv') %>% 
     read_csv()

head(pdi.adult1)
```

```{r}
# Adult-2
pdi.adult2 <-
     here('data', 
          'processed', 
          'Adult-2', 
          'PDI scores by subject.csv') %>% 
     read_csv()

head(pdi.adult2)
```

```{r}
# Join together
pdi <- bind_rows(pdi.adult1, pdi.adult2)

rm(pdi.adult1, pdi.adult2)
```

# Pre-process

## trnL

### Filter phyloseq

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

### Full alpha diversity measures

```{r}
alpha <- 
     rare.df %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     select(name, study, subj, CH_week, ONR_day, starts_with('obs'))
```

```{r}
# Add in full set of observations and Shannon diversity, including samples with fewer than read count threshold
add <- 
     estimate_richness(ps, 
                       measures = c('Observed',
                                    'Shannon')) %>% 
     rownames_to_column(var = 'name') %>% 
     mutate(name = gsub('\\.', '-', name)) %>%  # Correct '.' to '-' substitution
     rename(obs = Observed)

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) %>% 
     right_join(add, by = c('row' = 'name')) %>% 
     rename(name = row)

alpha <- 
     full_join(alpha,
               add)
```

### Mean per participant

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs),
               mean_shannon = mean(Shannon),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

# Analyze

## PDI

```{r}
# Join mean richness to PDI data
pdi <- 
     alpha.mean %>% 
     left_join(pdi)
```

```{r}
# Make labels plot-friendly
names(pdi)[names(pdi) == 'pdi'] <- 'PDI'
names(pdi)[names(pdi) == 'hpdi'] <- 'hPDI'
names(pdi)[names(pdi) == 'updi'] <- 'uPDI'
```

```{r}
# Select relevant columns, get in long form
pdi.long <- 
     pdi %>% 
     select(study:uPDI) %>% 
     pivot_longer(cols = matches('pdi'),
                  names_to = 'index', 
                  values_to = 'value')

pdi.long$index <- factor(pdi.long$index,
                         levels = c('PDI', 'hPDI', 'uPDI'))
```

### Adult-1

```{r}
pdi.long %>% 
     filter(study == 'Adult-1') %>% 
     ggplot(aes(x = mean, y = value)) +
     geom_point() +
     facet_wrap(~index) +
     labs(x = 'Mean *trnL* richness',
          y = 'Dietary index score') +
     theme_bw() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title = element_text(size = 14))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             '2',
#             paste0(plotID, '_C trnL richness x PDI indices (Adult-1).pdf')),
#        height = 3, width = 7)
```

```{r}
pdi %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + PDI,
              data = ., 
              method = 'spearman')
```

```{r}
pdi %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + hPDI,
              data = ., 
              method = 'spearman')
```

```{r}
pdi %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + uPDI,
              data = ., 
              method = 'spearman')
```


### Adult-2

```{r}
pdi.long %>% 
     filter(study == 'Adult-2') %>% 
     ggplot(aes(x = mean, y = value)) +
     geom_point() +
     facet_wrap(~index) +
     labs(x = 'Mean *trnL* richness',
          y = 'Dietary index score') +
     theme_bw() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title = element_text(size = 14))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             '2',
#             paste0(plotID, '_C trnL richness x PDI indices (Adult-2).pdf')),
#        height = 3, width = 7)
```


```{r}
pdi %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + PDI,
              data = ., 
              method = 'spearman')
```

```{r}
pdi %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + hPDI,
              data = ., 
              method = 'spearman')
```

```{r}
pdi %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + uPDI,
              data = ., 
              method = 'spearman')
```

## Backwards selection

Organize as a function, so that can be run on both Adult-1 and Adult-2 cohorts, with the results compared.

```{r}
# Helper function to calculate mean richness from participants, with optional subsetting of taxa
mean_richness <- 
     function(ps, taxa = NULL){
          # ps: phyloseq object
          # taxa: list of taxa to EXCLUDE from richness calculation
          
          # Get all data
          melt <- 
               psmelt(ps) %>% 
               suppressWarnings() # For phyloseq objects where melting leads to variable name redundancy 
          # Calculate mean per-participant trnL richness
          richness <- 
               melt %>% 
               filter(Abundance > 0) %>% 
               filter(!(OTU %in% taxa)) %>% 
               group_by(Sample, subj) %>% 
               count() %>% # Richness
               ungroup() %>% 
               group_by(subj) %>% 
               summarize(mean_obs = mean(n)) # Within-subject average richness
          
          richness
     }
```

```{r}
backwards_selection <- 
     function(pdi,
              phyloseq){
          
          # INITIALIZATION
          # First, calculate trnL x PDI without limiting taxa, as above:
          # Get simplified richness dataframe
          richness <- mean_richness(phyloseq, taxa = NULL)
          
          # Subset PDI to relevant columns: subject and hPDI
          pdi <- select(pdi, subj, hpdi)
          
          # Join together
          richness <- left_join(richness, pdi,
                                by = 'subj')
          
          # Calculate correlation
          # Get warning: "Cannot compute exact p-value with ties"
          spearman <- 
               cor.test(richness$mean_obs, 
                        richness$hpdi,
                        method = 'spearman') %>% 
               suppressWarnings()
          
          # Initialize variables for while loop
          max_rho <- -Inf
          new_rho <- spearman$estimate
          dropped_features <- c()
          all_features <- taxa_names(phyloseq)
          
          # BACKWARDS SELECTION
          while (new_rho >= max_rho){
               # Update rho to this value, then begin the process again
               max_rho <- new_rho
               
               # Book-keeping dataframe for iterating over features
               results <- 
                    data.frame(feature = NULL,
                               rho = NULL,
                               p = NULL)
               
               for (f in all_features){
                    # Calculate a new richness with that feature dropped
                    # (in addition to all prior drops)
                    richness <- 
                         mean_richness(phyloseq,
                                       taxa = c(dropped_features,
                                                f))
                    
                    # Join to hPDI
                    richness <- left_join(richness, pdi,
                                          by = 'subj')
                    
                    # Calculate correlation
                    spearman <- 
                         cor.test(richness$mean_obs, 
                                  richness$hpdi,
                                  method = 'spearman') %>% 
                         suppressWarnings()
                    
                    
                    # Bind to results dataframe
                    row <- c(f,
                             spearman$estimate, 
                             spearman$p.value)
                    
                    names(row) <- c('feature', 'rho', 'p')
                    
                    results <- bind_rows(results, row)
               }
               
               # Do type conversion on results
               results <- 
                    results %>% 
                    mutate(across(-feature, as.numeric)) 
               
               # Remove p values that aren't significant
               results <- filter(results, p < 0.05)
               
               # Find the maximum correlation coefficient
               new_rho <- max(results$rho) 
               
               # If we're going to loop again, preserve the feature to drop
               # and remove it from the all_features list
               if (new_rho >= max_rho){
                    # Find the feature to remove
                    # Note: there might be more than one: If so, choose at random
                    drop <- 
                         results %>% 
                         filter(rho == new_rho) %>% 
                         pull(feature) %>% 
                         sample(size = 1) 
                    
                    dropped_features <- c(dropped_features, drop)
                    all_features <- 
                         setdiff(taxa_names(phyloseq),
                                  dropped_features)
               }
               
               # Print where we are
               cat('Features dropped:', length(dropped_features), '\n')
          }
          
          # Return the list of dropped features
          dropped_features
     }
```

