---
title: "Adult-1, Adult-2 QC"
output: html_notebook
---

# Setup

```{r setup, include = FALSE, echo = FALSE} 
require(knitr) 
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(fitdistrplus) # For response variable distribution
library(here)
library(lme4) # For glmer
library(phyloseq)
library(scales)
library(tidyverse)
library(vegan) # For rarecurve()

select <- dplyr::select # Otherwise masked by MASS
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```


# Read in data

## trnL

```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (1,000 reads).rds') %>% 
     readRDS

rare.df
```

### Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

# Pre-process

```{r}
# Subset to Adult-1, Adult-2 samples
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0 , .)

ps
```

## Combine richness measures

```{r}
# Begin with richness in this dataset
alpha <- 
     estimate_richness(ps, 
                       measures = 'Observed') %>% 
     rownames_to_column(var = 'name') %>% 
     mutate(name = gsub('\\.', '-', name)) %>%  # Correct '.' to '-'
     rename(obs = Observed)
```

```{r}
# Populate with sample data
alpha <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) %>% 
     right_join(alpha, by = c('row' = 'name')) %>% 
     rename(name = row)
```

```{r}
# Add in full set of observations, including samples below read count threshold
add <- 
     rare.df %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     select(name, study, subj, CH_week, ONR_day, starts_with('obs'))

alpha <- 
     full_join(alpha,
               add)
```

# Analyze

```{r}
# Pull sample data
samdf <- data.frame(ps@sam_data)
```

## Compare between cohorts

Here the options for comparing two counts are 
1. A Mann-Whitney U test (non-parametric)
2. A Poisson GLM

#### Read depth

```{r}
ggplot(samdf, aes(x = study, y = reads_food)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Cohort', y = 'Sequencing depth') +
     scale_y_continuous(label = comma,
                        limits = c(0, 175000)) 
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Adult-1 and Adult-2 cohort sequencing depth.pdf')),
#        height = 3, width = 2.5)
```

```{r}
wilcox.test(reads_food ~ study,
            data = samdf,
            conf.int = TRUE)
```

#### Raw richness

```{r}
ggplot(alpha, aes(x = study, y = obs)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Cohort', y = 'pMR') +
     lims(y = c(0, 55))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Adult-1 and Adult-2 cohort richness.pdf')),
#        height = 3, width = 2)
```

Mann-Whitney U (Wilcoxon): This “does not estimate the difference in medians (a common misconception) but rather the median of the difference between a sample from x and a sample from y.”

```{r}
wilcox.test(obs ~ study,
            data = alpha,
            conf.int = TRUE)
```

```{r}
model <- glmer(obs ~ study + (1 | subj),
               data = alpha,
               family = poisson())

summary(model)
```

```{r}
# How well does the model fit the data?
yhat <- fitted(model)
y <- alpha$obs

ggplot(data.frame(yhat = yhat,
                  y = y),
       aes(x = y, y = yhat)) +
     geom_point()
```

```{r}
# Residual plots
plot(yhat, residuals(model))
```

#### Rarefied

```{r}
ggplot(alpha, aes(x = study, y = obs_rare)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Cohort', y = 'pMR') +
     lims(y = c(0, 55))
```
```{r}
wilcox.test(obs_rare ~ study,
            data = ,
            conf.int = TRUE)
```

### Depth v. richness

```{r}
alpha <- 
     samdf %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, reads_food) %>% 
     right_join(alpha, by = c('row' = 'name'))

alpha 
```

```{r}
ggplot(alpha, aes(x = reads_food, y = obs)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~study) +
     labs(x = 'Sequencing depth', y = 'pMR') +
     scale_x_continuous(label = comma) +
     theme(axis.text.x = element_text(angle = 20, hjust = 1))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_pMR x dequencing depth.pdf')),
#        height = 3, width = 6)
```

## Relate to variables

Pulling all the experimental and biological variables I think could be related:

```{r}
# Check extraction type
samdf %>% 
     group_by(extraction_type) %>% 
     count()
```

```{r}
# See if PCR date can uniquely identify batch
samdf %>% 
     group_by(pcr_date, pcr_batch) %>% 
     count()
```

```{r}
# It can't; make a unique variable and use that instead
samdf <- 
     samdf %>% 
     mutate(pcr_batch_overall = paste(pcr_date,
                                      pcr_batch, 
                                      sep = '_')) %>% 
     mutate(pcr_batch_overall = factor(pcr_batch_overall,
                                       levels = c("7/22/20_1",
                                                  "11/5/20_1",
                                                  "11/6/20_1",
                                                  "11/9/20_2",
                                                  "11/11/20_3",
                                                  "12/9/20_3",
                                                  "1/12/21_1",
                                                  "1/20/21_2",
                                                  "1/20/21_3"),
                                       labels = 1:9)) # Put in order by time

samdf %>% 
     group_by(pcr_date, pcr_batch_overall) %>% 
     count() %>% 
     arrange(pcr_batch_overall) 

sample_data(ps) <- samdf
```

```{r}
vars <- 
     c(
          # Biological
          # 'subj', # Leave this out: to be added as random effect
          # Either
          'qubit',
          'ng_added',
          # Experimental
          'extraction_type',
          'pcr_batch_overall',
          'seq_date',
          # 'primer', # Only one in these sequencing runs
          'input_volume',
          'pcrb_cycles'
          )
```

```{r}
# Check completeness
all(complete.cases(samdf[, vars]))
```

### Individual tests

#### Correlations

```{r}

```

#### Mann-Whitney U

Extraction type
```{r}
ggplot(samdf, aes(x = extraction_type, y = reads_food)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Extraction type', y = 'Sequencing depth') +
     scale_y_continuous(label = comma)
```
Most of the outliers are from the 2021/01/05 run (so thinking this is perhaps a sequencing yield effect, because 2021/02/04 also had 4 ul of a number of plate-extracted samples). 

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'experimental optimization',
#             paste0(plotID, '_Sequencing depth x extraction type.pdf')),
#        height = 3, width = 2.5)
```

```{r}
wilcox.test(reads_food ~ extraction_type,
            data = samdf,
            conf.int = TRUE)
```

PCR batch
```{r}
ggplot(samdf, aes(x = pcr_batch_overall, y = reads_food)) +
     facet_wrap(~study,
                scales = 'free_x') +
     geom_boxplot(width = 0.5) +
     labs(x = 'PCR batch', y = 'Sequencing depth') +
     scale_y_continuous(label = comma)
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'experimental optimization',
#             paste0(plotID, '_Sequencing depth x PCR batch.pdf')),
#        height = 3, width = 5)
```

```{r}
wilcox.test(reads_food ~ extraction_type,
            data = samdf,
            conf.int = TRUE)
```

Sequencing batch
```{r}
ggplot(samdf, aes(x = seq_date, y = reads_food)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Sequencing batch', y = 'Sequencing depth') +
     scale_y_continuous(label = comma) +
     theme(axis.text.x = element_text(angle = 20,
                                      hjust = 1))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            'supplemental',
            'experimental optimization',
            paste0(plotID, '_Sequencing depth x sequencing batch.pdf')),
       height = 3.5, width = 2.5)
```

Input volume
```{r}
ggplot(samdf, aes(x = input_volume, y = reads_food)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'Input volume', y = 'Sequencing depth') +
     scale_y_continuous(label = comma)
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'experimental optimization',
#             paste0(plotID, '_Sequencing depth x input volume.pdf')),
#        height = 3, width = 2.5)
```

PCRB cycles

```{r}
ggplot(samdf, aes(x = factor(pcrb_cycles), y = reads_food)) +
     geom_boxplot(width = 0.5) +
     labs(x = 'PCRB cycles', y = 'Sequencing depth') +
     scale_y_continuous(label = comma)
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            'supplemental',
            'experimental optimization',
            paste0(plotID, '_Sequencing depth x PCRB cycles.pdf')),
       height = 3, width = 2.5)
```

### PERMANOVA

Idea here would be PERMANOVA on Jaccard for each variable above, then extract R2 and p value. 

### Linear model

```{r}
# Response variable: read depth
ggplot(samdf, aes(x = reads_food)) +
     geom_histogram(boundary = 0,
                    binwidth = 10000) +
     labs(y = 'Frequency', x = 'Sequencing depth')
```

```{r}
# Create a skewness-kurtosis plot to find the best distribution for data
descdist(samdf$reads_food, discrete = FALSE, boot = 500)
```
```{r}
# Fit  data to a variety of distributions to double check above results. 
fit.norm <- fitdist(samdf$reads_food, "norm")
fit.gamma <- fitdist(samdf$reads_food, "gamma", method = 'mme') # MLE error
```

```{r}
plot(fit.norm)
```

```{r}
plot(fit.gamma)
```

Note that non-positive values aren't allowed for Gamma family.  Would need to turn zeros into ones. Try lm for now

```{r}
# Try a model with everything
f <- paste('reads_food ~ ',
           paste(vars, collapse = ' + '))

f
```

```{r}
m <- lm(f, data = samdf)
```

```{r}
summary(m)
```


### Per-participant

## Richness by participant

```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(obs_rare, na.rm = TRUE),
               sd = sd(obs_rare, na.rm = TRUE)) %>%
     # summarize(med = median(obs, na.rm = TRUE),
     #           sd = sd(obs, na.rm = TRUE)) %>%
     arrange(desc(med), desc(sd))

alpha$subj <- factor(alpha$subj, levels = alpha.stats$subj)
```

```{r}
alpha %>% 
     ggplot(aes(x = subj, y = obs_rare, fill = study)) +
     facet_wrap(~study, scales = 'free_x') +
     geom_boxplot() +
     labs(x = 'Participant', y = 'Rarefied *trnL* richness') +
     # scale_color_manual(values = c('black', 'gray60')) +
     scale_fill_manual(values = c('gray40', 'gray80')) +
     theme_classic() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title = element_text(size = 14,
                                    face = 'bold'),
          axis.title.y = ggtext::element_markdown(),
          strip.text = element_text(size = 12,
                                    face = 'bold'),
          legend.position = 'None'
     )
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            '2',
            paste0(plotID, '_A_rarefied (1,000 reads) trnL richness by participant.pdf')),
       height = 3, width = 9)
```

```{r}
# Cohort-level
alpha %>% 
     ggplot(aes(x = study, y = obs, fill = study)) +
     geom_boxplot() +
     labs(x = 'Study', y = '*trnL* richness') +
     # scale_color_manual(values = c('black', 'gray60')) +
     scale_fill_manual(values = c('gray40', 'gray80')) +
     theme_classic() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title = element_text(size = 14,
                                    face = 'bold'),
          axis.title.y = ggtext::element_markdown(),
          strip.text = element_text(size = 12,
                                    face = 'bold'),
          legend.position = 'None'
     )
```

```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(obs_rare,
                            na.rm = TRUE),
               mean = mean(obs_rare,
                           na.rm = TRUE),
               sd = sd(obs_rare,
                       na.rm = TRUE))

range(alpha.stats$med, na.rm = T)
range(alpha.stats$sd, na.rm = T)
```

```{r}
alpha %>% 
     ggplot(aes(x = subj, y = obs, color = study)) +
     facet_wrap(~study, scales = 'free_x') +
     geom_boxplot() +
     labs(x = 'Participant', y = '*trnL* richness') +
     scale_color_manual(values = c('black', 'gray60')) +
     # scale_fill_manual(values = c('black', 'gray60')) +
     theme_classic() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title = element_text(size = 14,
                                    face = 'bold'),
          axis.title.y = ggtext::element_markdown(),
          strip.text = element_text(size = 12,
                                    face = 'bold'),
          legend.position = 'None'
     )
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'manuscript', '2',
            paste0(plotID, '_A_raw trnL richness by participant.pdf')),
       height = 3, width = 9)
```

### Within-subject vs. between-subject

#### ANOVA (R01 ONLY)

```{r}
# Adult-1 
mod.1 <- 
     alpha %>% 
     filter(study == 'Adult-1') %>% 
     aov(obs ~ subj, data = . )

summary(mod.1)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod.1)[[1]]['Mean Sq'][1,]
var.within <- summary(mod.1)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     filter(study == 'Adult-1') %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```


```{r}
# Adult-2 
mod.2 <- 
     alpha %>% 
     filter(study == 'Adult-2') %>% 
     aov(obs ~ subj, data = . )

summary(mod.2)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod.2)[[1]]['Mean Sq'][1,]
var.within <- summary(mod.2)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     filter(study == 'Adult-2') %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```


### Overall

```{r}
mod <- 
     aov(obs ~ subj, data = alpha)

summary(mod)
```

```{r}
# Get partitioning of variance
var.between <- summary(mod)[[1]]['Mean Sq'][1,]
var.within <- summary(mod)[[1]]['Mean Sq'][2,]
```

Now need to get means. But it's not entirely clear to me *what* mean to use here. The mean of each subject's average richness? This is what it looks like in this assay development link: https://influentialpoints.com/Training/coefficient_of_variation.htm 
```{r}
u <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     summarize(mu_pop = mean(mu_subj, 
                             na.rm = T)) # Population mean of mean per-subject richness

u
```
```{r}
cat('Within', 100*sqrt(var.within)/u$mu_pop, '\n')
cat('Between', 100*sqrt(var.between)/u$mu_pop, '\n')
```

#### Permutation-based strategy

##### Raw data

Aggregating within-subject CVs using info from (this site) [https://www-users.york.ac.uk/~mb55/meas/cv.htm]: "We find the coefficient of variation for each subject separately, square these, find their mean, and take the square root of this mean. We can call this the root mean square approach."

```{r}
within.between <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(sigma = sd(obs, na.rm = T),
               mu = mean(obs, na.rm = T)) %>% 
     mutate(cv_within = sigma/mu)
```

```{r}
# Now summarize as above
within.between %>% 
     ungroup() %>% 
     mutate(cv_within_sq = cv_within^2) %>% # Square them
     # group_by(study) %>% 
     summarize(mean_sq = mean(cv_within_sq)) %>% # Find the mean
     mutate(root_mean_sq = sqrt(mean_sq)) # Take the square root
```
```{r}
# Now calculate for overall data
alpha %>% 
     group_by(study, subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     # group_by(study) %>% 
     summarize(
          mu_pop = mean(mu_subj, 
                        na.rm = T), # Population mean of mean per-subject obs.
          sigma_pop = sd(mu_subj, 
                         na.rm = T)) %>% # Population sd
     mutate(cv_between = sigma_pop/mu_pop) 
```

```{r}
# Calculate standard deviation for overall cohort and join to data
within.between <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mu_subj = mean(obs, na.rm = T)) %>% # Mean per-subject richness
     ungroup() %>% 
     group_by(study) %>% 
     summarize(
          mu_pop = mean(mu_subj, 
                        na.rm = T), # Population mean of mean per-subject richness
          sigma_pop = sd(mu_subj, 
                         na.rm = T)) %>% # Population sd
     mutate(cv_between = sigma_pop/mu_pop) %>% 
     select(study, cv_between) %>% 
     right_join(within.between)
```

```{r}
# Calculate difference
within.between <- 
     mutate(within.between,
            delta = cv_within - cv_between)
```

```{r}
# Plot raw data
ggplot(within.between, aes(x = study, y = cv_within)) +
     geom_boxplot() +
     ylab('Within-subject CV') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 
            'development', 
            'docs',
            'r01', 
            '2021_11',
            'within-between',
            paste0(plotID, '_CV on rarefied richness.pdf')),
       height = 3, width = 3)
```

```{r}
# Plot delta
ggplot(within.between, aes(x = study, y = delta)) +
     geom_boxplot() +
     ylab('Within-subject CV - between-subject CV') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 
            'development', 
            'docs',
            'r01', 
            '2021_11',
            'within-between',
            paste0(plotID, '_delta CV on raw richness.pdf')),
       height = 3, width = 3)
```

##### Shuffled data

```{r}
set.seed(10222021)

# Shuffle
alpha.rand <- alpha 
alpha.rand$obs <- sample(alpha.rand$obs, length(alpha.rand$obs))
```

```{r}
within.between.rand <- 
     alpha.rand %>% 
     group_by(study, subj) %>% 
     summarize(sd_within = sd(obs))
```

```{r}
# Calculate standard deviation for overall cohort ad join to data
within.between.rand <- 
     alpha.rand %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs)) %>% 
     ungroup() %>% 
     group_by(study) %>% 
     summarize(sd_between = sd(mean)) %>% 
     right_join(within.between.rand)
```

```{r}
# Calculate difference
within.between.rand <- 
     mutate(within.between.rand,
            delta = sd_within - sd_between)
```

```{r}
# Bind together
within.between$data <- 'Raw'
within.between.rand$data <- 'Shuffled'
```

```{r}
within.between <- bind_rows(within.between,
                            within.between.rand)
```

```{r}
ggplot(within.between, aes(x = data, y = delta)) +
     geom_boxplot() + 
     facet_wrap(~study) + 
     labs(y = 'sd (within) - sd (between)') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```


#### KS test

```{r}
# Get vector for between-subject distribution: the mean trnL richness of each subject

between <- 
     adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     pull(mean)

subjects <- 
     adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     pull(subj) %>% 
     sort()

results.df <- 
     data.frame(
          study = NULL,
          subj = NULL,
          D = NULL,
          P = NULL
     )

for (s in subjects){
     # Compare between-subjects distribution to within-subject distribution
     # for this particular subject
     result <- 
          alpha %>% 
          filter(subj == s) %>% 
          pull(obs_rare) %>% 
          ks.test(., between)
     
     row = c('Adult-1', s, result$statistic, result$p.value)
     names(row) <- c('study', 'subj', 'D', 'p')
     
     results.df <- bind_rows(results.df, row)
}
```

```{r}
# Plot results as p-value histogram
results.df$p <- as.numeric(results.df$p)

hist(results.df$p, breaks = 20)
```


```{r}
fit.within <- 
     aov(obs ~ subj,
         data = alpha)
```

```{r}
summary(fit.within)
```

Think I need an additional layer of summarizing here??

```{r}
fit.between <- 
     aov(obs ~ study,
         data = alpha)
```

```{r}
summary(fit.between)
```


```{r}
aov(obs ~ study + subj,
         data = alpha) %>% 
     summary()
```


### Mean vs. variance

```{r}
# Add cohort information in order to calculate separately
adult.stats <- 
     samdf %>% 
     select(subj, study) %>% 
     right_join(adult.stats) %>% 
     distinct()
```

```{r}
# Adult-1
adult.stats %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + var,
              data = ., 
              method = 'spearman')
```

```{r}
# Adult-2
adult.stats %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + var,
              data = ., 
              method = 'spearman')
```

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs),
               mean_shannon = mean(Shannon),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.mean, aes(x = mean, y = mean_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Mean *trnL* richness<br>\n(All samples)', 
          y = 'Mean rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             'supplemental',
#             paste0(plotID, '_rarefied vs observed trnL richness (mean).pdf')),
#        height = 4, width = 4)
```

```{r}
# Average trnL richness within each subject
alpha.median <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(median = median(obs, na.rm = TRUE),
               median_rare = median(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

```{r}
# How well do these numbers correlate, now that there are different numbers of observations?
ggplot(alpha.median, aes(x = median, y = median_rare)) +
     geom_abline(slope = 1,
                 intercept = 0, 
                 color = 'gray80',
                 linetype = 'dashed') +
     geom_point() +
     coord_equal() +
     labs(x = 'Median *trnL* richness<br>\n(All samples)', 
          y = 'Median rarefied *trnL* richness<br>(>10,500 reads)') +
     xlim(0, 40) + ylim(0, 40) +
     theme_classic() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             'supplemental',
#             paste0(plotID, '_rarefied vs observed trnL richness (median).pdf')),
#        height = 4, width = 4)
```

Based on above plots, move forward with mean

## Read depth

```{r}
# Faceted by study
ggplot(samdf, aes(x = reads_food, fill = study)) +
     geom_histogram(binwidth = 10500, boundary = 0) +
     geom_vline(xintercept = 10500, 
                color = 'red',
                linetype = 'dashed') +
     facet_wrap(~study, nrow = 2) +
     scale_fill_manual(values = c('black', 'gray60')) +
     labs(x = '*trnL* read count', y = 'Samples (n)') +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1),
           axis.title.x = ggtext::element_markdown(),
           legend.title = element_blank(),
           legend.position = 'none')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(
     here(
          'results', 
          'manuscript',
          'supplemental',
          paste0(plotID, 
                 '_Adult-1, Adult-2 read count histogram.pdf')),
     height = 4, width = 2)
```


## Rarefaction curves

Idea would be to randomly subsample 10 samples from each read count bin, then plot their curve.
```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- otu_table(ps)@.Data
dim(asvtab)
```

```{r}
set.seed(8181)

# Randomly choose 10 samples from each bin
samples <- 
     samdf %>% 
     rownames_to_column(var = 'row') %>% 
     group_by(reads_food_binned) %>% 
     sample_n(8) %>% 
     ungroup() %>% 
     select(row, reads_food_binned)
```

```{r}
for (level in levels(samples$reads_food_binned)){
     cat(level, '\n')
     rows <- 
          samples %>% 
          filter(reads_food_binned == level) %>% 
          pull(row)
     
     plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
     pdf(here('results', 
              'manuscript',
              'supplemental',
              paste0(plotID, 
                     '_Adult-1, Adult-2 rarefaction curves ',
                     level,
                     ' reads.pdf')),
       height = 3, 
       width = 5)
     plot <- rarecurve(asvtab[rows, ], col = "blue")
     dev.off()
}
```

