---
title: "Adult-1, Adult-2 diversity indices"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseqs
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### DHQ3-based HEI

Calculated in Adult-1, Adult-2 dietary indices notebook.

```{r}
hei <- 
     here('data',
          'processed',
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

# Pre-process

## trnL

### Filter

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     subset_samples(reads_food > 0) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(add, alpha,
              by = c('row' = 'name'))
```

### Mean richness

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj, threshold) %>% 
     summarize(mean = mean(obs),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

## Diet surveys

Subset to PDI values only

```{r}
hei <- 
     hei %>% 
     select(study, subj, kcal, pass_qc,
            matches('hei')) 
```

# Analyze

## HEI

```{r}
# Join mean richness to HEI data
hei <- left_join(alpha.mean, hei)
```

# Analyze

```{r}
# Select relevant columns, get in long form
hei.long <- 
     hei %>% 
     pivot_longer(cols = matches('hei'),
                  names_to = 'index', 
                  values_to = 'value')

hei.long$index <- factor(hei.long$index,
                         levels = c('hei_plant_all', 'hei_non_plant'),
                         labels = c('HEI (plant)',
                                    'HEI (non-plant)'))
```

#### Raw
```{r}
hei.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     filter(index == 'HEI (plant)' & pass_qc) %>% 
     distinct() %>% 
     ggplot(aes(x = value, y = mean)) +
     geom_point(size = 2, alpha = 0.6) +
     facet_wrap(~study, nrow = 2) +
     ylim(0, 26) +
     labs(y = 'Mean pMR',
          x = 'HEI-2015\n(plant component)') 
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x HEI.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
hei.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     filter(index == 'HEI (non-plant)' & pass_qc) %>% 
     distinct() %>% 
     ggplot(aes(x = value, y = mean)) +
     geom_point(size = 2, alpha = 0.6) +
     facet_wrap(~study, nrow = 2) +
     ylim(0, 26) +
     labs(y = 'Mean pMR',
          x = 'HEI-2015 (non-plant\ncomponents)') 
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x HEI (animal).pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
hei.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     filter(pass_qc) %>%
     distinct() %>% 
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean,
                            method = 'spearman')$p.value) %>% 
     mutate(p_adj = p.adjust(p, method = 'BH'))
```

#### Rarefied

```{r}
hei.long %>% 
     filter(index == 'HEI (plant)') %>% 
     ggplot(aes(x = value, y = mean_rare)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_grid(vars(study),
                vars(threshold)) +
     labs(y = 'Mean pMR',
          x = 'HEI component') 
```
```{r}
hei.long %>% 
     filter(pass_qc & threshold == 1000) %>% 
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean_rare,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean_rare,
                            method = 'spearman')$p.value) %>% 
     mutate(p_adj = p.adjust(p, method = 'BH'))
```
