---
title: "Adult-1, Adult-2 FVS"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### DHQ3-based FVS

Calculated in Adult-1, Adult-2 dietary indices notebook.

```{r}
fvs <- 
     here('data',
          'processed',
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

# Pre-process

## trnL

### Filter phyloseq

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(alpha, add,
              by = c('name' = 'row'))
```

Then below, can just filter by desired threshold in order to calculate measures.

### Mean richness

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj, threshold) %>% 
     summarize(mean = mean(obs),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

# Analyze

## FVS

```{r}
# Join mean richness to PDI data
fvs <- 
     alpha.mean %>% 
     left_join(fvs)
```

### Raw data

```{r}
# Raw data
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = fvs, y = mean)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study) +
     labs(x = 'Food Variety Score',
          y = 'Mean pMR')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             '2',
#             paste0(plotID, '_raw trnL richness x FVS.pdf')),
#        height = 7, width = 5)
```

```{r}
fvs %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     group_by(study) %>% 
     summarize(rho = cor.test(~ fvs + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ fvs + mean,
                            method = 'spearman')$p.value) 
            
```

### Rarefied data

```{r}
fvs %>% 
     ggplot(aes(x = fvs, y = mean_rare)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_grid(~study + threshold) +
     labs(x = 'Food Variety Score',
          y = 'Mean pMR')
```

## Sampling scheme

```{r}
# First remove overall observations from PDI data frame
pdi <- select(pdi, -starts_with('obs'))

pdi
```

### Adult-1

```{r}
# Get sample data for Adult-1
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adult-1')
```

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

Note that this will cover duplicated sample sets as i gets closer to 18, but can detect and remove them downstream.
```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(11232021)

for (i in 1:18){ # Number of samples
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj) %>% 
               slice_sample(n = i) %>% 
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          pdi.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(pdi)
          
          # Calculate correlation by study
          # Common warning: cannot compute exact p-value with ties
          spearman <- cor.test(pdi.temp$obs_mean, 
                               pdi.temp$hPDI,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-1', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
# Intercept: get from full dataset
y <- results.df$rho[results.df$scheme == 18]
```

```{r}
# Plot
results.df %>% 
     filter(scheme %in% seq(3, 18, 3)) %>% 
     ggplot(aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = y, 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = seq(3, 18, 3)) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-1 trnL x PDI correlations by sampling scheme, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
results.df %>% 
     filter(scheme %in% seq(3, 18, 3)) %>% 
     ggplot(aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = seq(3, 18, 3)) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-1 trnL x PDI correlations by sampling scheme, p.pdf')),
       height = 4, width = 3)
```

#### Restrict to separate weeks

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(11232021)

for (i in 1:6){ # Number of samples (only from separate weeks)
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj, CH_week) %>% 
               slice_sample(n = 1) %>% # Pick one sample from each week at random
               ungroup() %>% 
               group_by(subj) %>% # 6 random weekly samples
               slice_sample(n = i) %>% # Pick i weeks' worth of samples
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          pdi.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(pdi)
          
          # Calculate correlation
          spearman <- cor.test(pdi.temp$obs_mean, 
                               pdi.temp$hPDI,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-1', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```
 
```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.52, # Hard-coded from prior result 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x hPDI correlations restricted to 1 random sample per week, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-1 trnL x hPDI correlations restricted to 1 random sample per week, p.pdf')),
       height = 4, width = 3)
```

### Adult-2

```{r}
# Get sample data for Adult-2
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     filter(study == 'Adult-2')
```

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

Note that this will cover duplicated sample sets as i gets closer to 6, but can detect and remove them downstream.
```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(11232021)

for (i in 1:6){ # Number of samples
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj) %>% 
               slice_sample(n = i) %>% 
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          pdi.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(pdi)
          
          # Calculate correlation by study
          spearman <- cor.test(pdi.temp$obs_mean, pdi.temp$hPDI,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-2', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# Intercept: get from full dataset
y <- results.df$rho[results.df$scheme == 6]
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = y, 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-2 trnL x hPDI correlations by sampling scheme, rho.pdf')),
       height = 4, width = 3)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:6) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            '3',
            paste0(plotID, '_Adult-2 trnL x hPDI correlations by sampling scheme, p.pdf')),
       height = 4, width = 3)
```

#### Restrict to separate weeks

```{r}
# For storing results
results.df <- data.frame(
     scheme = NULL, 
     study = NULL,
     rho = NULL,
     p = NULL)
```

```{r}
# Update Adult-2 sample data to have a "week" entry
samdf <- 
     mutate(samdf,
            ONR_week = gsub('T|F', '', ONR_day),
            ONR_week = as.numeric(ONR_week))
```

```{r}
# Do scheme 1 first: Randomly sample i samples from each participant
set.seed(11232021)

for (i in 1:3){ # Number of samples (only from separate weeks)
     for (iter in (1:100)){
          # Make subsampled, merged phyloseq
          samples <- 
               samdf %>% 
               group_by(subj, ONR_week) %>% 
               slice_sample(n = 1) %>% # Pick one sample from each week at random
               ungroup() %>% 
               group_by(subj) %>% # 3 random weekly samples
               slice_sample(n = i) %>% # Pick i weeks' worth of samples
               pull(row)
          
          ps.temp <- prune_samples(samples, ps)
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
                                estimate_richness(ps.temp, 
                                                  measures = c('Observed')) %>% 
                                rownames_to_column(var = 'row') %>% 
                                # Correct '.' to '-' in sample name
                                mutate(row = gsub('\\.', '-', row)) %>% 
                                # Join to full sample data
                                left_join(samdf) %>% 
                                select(row, study, subj, obs = Observed)) 
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          pdi.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(obs_mean = mean(obs)) %>% 
               left_join(pdi)
          
          # Calculate correlation by study
          spearman <- cor.test(pdi.temp$obs_mean, 
                               pdi.temp$hPDI,
                               alternative = 'greater',
                               method = 'spearman')
          
          # Update results dataframe
          row <- c(i, 
                   'Adult-2', 
                   spearman$estimate, 
                   spearman$p.value)
          names(row) <- c('scheme', 'study', 'rho', 'p')
          results.df <- bind_rows(results.df, row)
     }
}
```

```{r}
# Remove duplicates
dim(results.df)
results.df <- distinct(results.df)
dim(results.df) 
```

```{r}
cols = c('scheme', 'rho', 'p')
results.df <- 
     mutate(results.df,
            across(one_of(cols),
            ~ as.numeric(.x)))
```

```{r}
# How many in each category?
table(results.df$scheme)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = rho, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.46, # Hard-coded from prior result 
                linetype = 'dotted',
                color = 'palegreen3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:3) +
     labs(y = 'Spearman *rho*', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14), 
          axis.title.y = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-2 trnL x hPDI correlations restricted to 1 random sample per week, rho.pdf')),
       height = 4, width = 2)
```

```{r}
# Plot
ggplot(results.df, aes(x = scheme, y = p, group = scheme)) +
     geom_boxplot() +
     geom_hline(yintercept = 0.05, 
                linetype = 'dotted',
                color = 'tomato3',
                size = 1.5) +
     scale_x_continuous(breaks = 1:3) +
     labs(y = 'Spearman p-value', x = 'Samples (n)') +
     theme_classic() +
     theme(
          axis.title = element_text(size = 14),
          axis.title.y = ggtext::element_markdown())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            paste0(plotID, '_Adult-2 trnL x hPDI correlations restricted to 1 random sample per week, p.pdf')),
       height = 4, width = 2)
```


```{r}
for (scheme in c('s2', 's3', 's4', 's5')){ # For each sampling scheme
     scheme.var <- get(scheme)
     for (i in seq_along(scheme.var)){ # For each set of samples within that scheme
          # Make subsampled phyloseq
          ps.temp <- subset_samples(ps.foods,
                                    ONR_day %in% scheme.var[[i]])
          
          # Calculate alpha diversity
          # Common warning here when there are few singletons in the data (only
          # a handful of samples)
          suppressWarnings(alpha <- 
               estimate_richness(ps.temp, 
                                 measures = c('Observed', 
                                              'Shannon')) %>% 
               rownames_to_column(var = 'name') %>% 
               mutate(subj = gsub('_.*$', '', name))) # Get subject ID
          
          # Average alpha diversity results within each participant, and join to 
          # FFQ data
          ffq.temp <- 
               alpha %>% 
               group_by(subj) %>% 
               summarize(mean_Obs = mean(Observed),
                         mean_Shannon = mean(Shannon)) %>% 
               left_join(ffq)
          
          # Calculate Pearson correlation
          pearson <- cor.test(ffq.temp$mean_Shannon, ffq.temp$hei_plant_all,
                              alternative = 'greater',
                              method = 'pearson')
          
          # Update results dataframe
          row <- c(scheme, i, pearson$estimate, pearson$p.value)
          names(row) <- c('scheme', 'i' , 'Pearson_r', 'Pearson_p')
          results.df <- bind_rows(results.df, row)
          }
}

```

Plot results:

```{r}
# Pivot longer for side-by-side plotting
results.df <- 
     pivot_longer(results.df,
                  cols = starts_with('Pearson'),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     mutate(value = as.numeric(value))
```

Outliers are throwing off plot limits, and I can't adjust across facets. Just save plots separately and display side-by side.

```{r}
labels <- c('1 sample',
            '2 samples,\nconsecutive',
            '2 samples,\nalternate week',
            '3 samples,\nconsecutive',
            '3 samples,\nweekly')

results.df %>% 
     filter(measure == 'Pearson_p') %>% 
     ggplot(aes(x = scheme, y = value)) +
     geom_boxplot(outlier.shape = NA, size = 1.5) +
     # geom_hline(yintercept = 0.43, size = 1.5, alpha = 0.5,
     #            color = 'springgreen4', linetype = 'dashed') +
     geom_hline(yintercept = 0.05, size = 1.5, alpha = 0.5,
                color = 'red', linetype = 'dashed') +
     scale_x_discrete(labels = labels) +
     labs(y = 'Pearson *p*-value') +
     # labs(y = 'Pearson *r*') +
     ylim(0, 0.25) +
     theme_bw() +
     theme(axis.line = element_line(size = 1.5),
           axis.ticks = element_line(size = 1.5),
           axis.text.x = element_text(angle = 30, hjust = 1,
                                      size = 14, face = 'bold', color = 'gray40'),
           axis.text.y = element_text(size = 14, face = 'bold', color = 'gray40'),
           axis.title.x = element_blank(), 
           axis.title.y = ggtext::element_markdown(size = 18, face = 'bold', 
                                                   color = 'gray40'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..', 'development', 'results', '2021 R01',
            paste0(plotID, '_trnL x HEI correlations by sampling scheme.png')),
       height = 5, width = 5)
```

