---
title: "Adult-1, Adult-2 diversity indices"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseqs
ps <- 
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS

rare.df
```
## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

### ASA24

```{r}
# CHOMP-only food counts
recall <- 
     here('..',
          'chomp',
          'data', 
          'processed',
          'ASA24', 
          'Food_Tree',
          'richness_ASA24.rds') %>% 
     readRDS()
```

# Pre-process

## trnL

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     select(name, study, subj, CH_week, ONR_day, starts_with('obs'))
```

```{r}
# Add in full set of observations, including samples below read count threshold
add <- 
     estimate_richness(ps, 
                       measures = 'Observed') %>% 
     rownames_to_column(var = 'name') %>% 
     mutate(name = gsub('\\.', '-', name)) %>%  # Correct '.' to '-' substitution
     rename(obs = Observed)

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) %>% 
     right_join(add, by = c('row' = 'name')) %>% 
     rename(name = row)

alpha <- 
     full_join(alpha,
               add)
```

## Diet surveys

### ASA24

```{r}
recall
```

# Analyze

```{r}
samdf <- data.frame(ps@sam_data)
```

## Richness by participant

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(mean = mean(obs, na.rm = TRUE),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

## Diversity indices

## DDS


### Visualizations (joint)

```{r}
ggplot(ffq, aes(x = mean, y = hei_plant_all)) +
     geom_point(alpha = 0.8, size = 2) +
     facet_wrap(~study) +
     labs(x = 'Mean *trnL* richness', 
          y = 'HEI-2015\nplant components score') +
     theme_bw() +
     theme(axis.title = element_text(size = 14,
                                     face = 'bold'),
           axis.title.x = ggtext::element_markdown(),
           axis.text = element_text(size = 12),
           strip.text = element_text(size = 12,
                                      face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_B trnL richness x HEI plant.pdf')),
       height = 3, width = 7)
```

```{r}
ggplot(ffq, aes(x = mean_rare, y = hei_plant_all)) +
     geom_point() +
     facet_wrap(~study) +
     labs(x = 'Mean rarefied *trnL* richness', 
          y = 'HEI-2015 plant components score') +
     theme_bw() +
     theme(axis.title = element_text(size = 14),
           axis.title.x = ggtext::element_markdown())
```

```{r}
ggplot(ffq, aes(x = mean_shannon, y = hei_plant_all)) +
     geom_point(alpha = 0.8, size = 2) +
     facet_wrap(~study) +
     labs(x = 'Mean *trnL* Shannon diversity', 
          y = 'HEI-2015\nplant components score') +
     theme_bw() +
     theme(axis.title = element_text(size = 14),
           axis.title.x = ggtext::element_markdown())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            'supplemental',
            paste0(plotID, '_trnL Shannon div x HEI plant.pdf')),
       height = 3, width = 7)
```

### Adult-1

#### Total

```{r}
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean_shannon + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean_rare + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

#### Plant only

```{r}
# Stats
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean + hei_plant_all,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
# Stats
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean_shannon + hei_plant_all,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
# Stats
ffq %>% 
     filter(study == 'Adult-1') %>% 
     cor.test(~ mean_rare + hei_plant_all,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

#### Linear fit

```{r}
fit.1 <- 
     ffq %>% 
     filter(study == 'Adult-1') %>% 
     lm(hei_plant_all ~ mean, data = .)

summary(fit.1)
```

```{r}
ffq %>% 
     filter(study == 'Adult-1') %>% 
     ggplot(aes(x = mean, y = hei_plant_all)) +
     # geom_abline(slope = fit.1$coefficients[[2]],
     #             intercept = fit.1$coefficients[[1]],
     #             color = 'gray',
     #             linetype = 'dashed') +
     geom_point() +
     annotate('text', x = 20, y = 5,
     label = 'atop(rho == 0.40, p == 0.03)',
              parse = TRUE,
              size = 5,
              color = 'blue') +
     xlim(0, 25) + ylim(0, 30) +
     labs(x = 'Mean *trnL* richness',
          y = 'HEI plant components') +
     theme_bw() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title = element_text(size = 14, 
                                     face = 'bold'),
           axis.text = element_text(size = 12, 
                                    face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..',
            'development',
            'docs',
            'r01',
            '2021_11',
            paste0(plotID, '_Adult-1 HEI plant x trnL.pdf')),
       height = 3, width = 3)
```


### Adult-2

#### Total

```{r}
ffq %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
ffq %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean_shannon + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

```{r}
ffq %>% 
     filter(study == 'Adult-2') %>% 
     cor.test(~ mean_rare + hei_tot,
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

## vs. ASA24

```{r}
# Concatenate richness measures
recall.rows <- 
     select(recall, subj = ID, starts_with('richness')) %>% 
     pivot_longer(cols = starts_with('richness'),
                  names_to = 'category',
                  values_to = 'value') %>% 
     mutate(category = gsub('richness_', '', category))

trnL.rows <- 
     rare.df %>% 
     filter(study == 'Adult-1') %>% 
     select(subj, value = obs_rare) %>% 
     mutate(category = 'plant')

recall.rows$measure = '24-hour recall'
trnL.rows$measure = 'Metabarcoding'

trnL.v.asa <- 
     bind_rows(recall.rows, trnL.rows)
```

```{r}
trnL.v.asa %>% 
     filter(category == 'plant') %>% 
     ggplot(aes(x = subj, y = value, color = measure)) +
     # geom_point()
     geom_boxplot() +
     theme_bw() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
# Try averaging within each participant to get one value
trnL.v.asa.avg <- 
     trnL.v.asa %>% 
     filter(category == 'plant') %>% 
     group_by(measure, subj) %>% 
     summarize(mean = mean(value)) %>% 
     pivot_wider(names_from = 'measure',
                 values_from = 'mean')
```

```{r}
ggplot(trnL.v.asa.avg, aes(x = `24-hour recall`, y = Metabarcoding)) +
     geom_point() +
     xlim(0, 20) + ylim(0, 20) +
     coord_equal()
```
Consider: haven't filtered data for completeness?

```{r}
cor.test(trnL.v.asa.avg$`24-hour recall`, 
         trnL.v.asa.avg$Metabarcoding,
         method = 'spearman')
```

