---
title: "Adult-1, Adult-2 PDI"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed',
          'phyloseq', 
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts.rds') %>% 
     readRDS

rare.df
```

## Cumulative richness estimates

```{r}
cumulative.df <- 
     here('data',
          'processed',
          'accumulation',
          'Adult-1 and Adult-2 cumulative taxon counts.csv') %>% 
     read_csv()
```

## Diet surveys

Calculated in Adult-1, Adult-2 dietary indices notebook.

```{r}
pdi <- 
     here('data',
          'processed',
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

# Pre-process

## trnL

### Filter phyloseq

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     subset_samples(reads_food > 0) %>%
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

```{r}
alpha <- 
     rare.df %>% 
     filter(grepl(name, pattern = 'CHOMP|ONR') )
```

```{r}
# Add in sample metadata

# Populate with sample data
add <- 
     ps@sam_data %>%
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, CH_week, ONR_day) 

alpha <- 
    left_join(add, alpha,
              by = c('row' = 'name'))
```

Then below, can just filter by desired threshold in order to calculate measures.

### Mean richness

```{r}
# Average trnL richness within each subject
alpha.mean <-
     alpha %>% 
     group_by(study, subj, threshold) %>% 
     summarize(mean = mean(obs),
               mean_rare = mean(obs_rare, na.rm = TRUE)) %>% 
     ungroup()
```

## Diet surveys

Subset to PDI values only

```{r}
pdi <- 
     select(pdi,
            study, subj, pass_qc,
            matches('PDI'))
```

# Analyze

## PDI

```{r}
# Join mean richness to PDI data
pdi <- 
     alpha.mean %>% 
     left_join(pdi)
```

### Totals

```{r}
# Select relevant columns, get in long form
pdi.long <- 
     pdi %>% 
     pivot_longer(cols = c(PDI, hPDI, uPDI),
                  names_to = 'index', 
                  values_to = 'value')

pdi.long$index <- factor(pdi.long$index,
                         levels = c('PDI', 'hPDI', 'uPDI'))
```

```{r}
# PDI visualization (need to rework given new data structure)
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = index, y = value)) + 
     geom_boxplot() +
     ylim(18, 90) +
     facet_wrap(~study) +
     labs(y = 'Score value', x = 'Score type') +
     theme_bw()
```

```{r}
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     filter(index == 'hPDI' & pass_qc) %>% 
     distinct() %>% 
     ggplot(aes(x = value, y = mean)) +
     geom_point(size = 2, alpha = 0.6) +
     facet_wrap(~study,
                nrow = 2) +
     ylim(0, 26) +
     labs(y = 'Mean pMR',
          x = 'hPDI\n') 
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x hPDI.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     filter(index == 'uPDI' & pass_qc) %>% 
     distinct() %>% 
     ggplot(aes(x = value, y = mean)) +
     geom_point(size = 2, alpha = 0.6) +
     facet_wrap(~study,
                nrow = 2) +
     ylim(0, 26) +
     labs(y = 'Mean pMR',
          x = 'uPDI\n') 
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '2',
            paste0(plotID, '_raw trnL richness x uPDI.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
# Raw
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     filter(pass_qc) %>%
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean,
                            method = 'spearman')$p.value) %>% 
     mutate(p_adj = p.adjust(p, method = 'BH'))
```

```{r}
# Rarefied
pdi.long %>% 
     filter(threshold == 1000) %>% 
     filter(pass_qc) %>% 
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean_rare,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean_rare,
                            method = 'spearman')$p.value) %>% 
     mutate(p_adj = p.adjust(p, method = 'BH'))
            
```

#### Linear fit

```{r}
fit.1 <- 
     pdi %>% 
     filter(study == 'Adult-1') %>% 
     lm(hPDI ~ mean, data = .)

summary(fit.1)
```

```{r}
# Mean, SD of raw data and fitted values
pdi %>% 
     filter(study == 'Adult-1', !is.na(hPDI)) %>% 
     pull(hPDI) %>% 
     data.frame(y = .,
                yhat = fitted(fit.1)) %>% 
     summarize(across(.fns = list(mean = mean, sd = sd)))
```

```{r}
pdi %>% 
     filter(study == 'Adult-1') %>% 
     ggplot(aes(x = mean, y = hPDI)) +
     geom_point() +
     # geom_abline(slope = fit.1$coefficients[[2]],
     #             intercept = fit.1$coefficients[[1]],
     #             color = 'blue',
     #             linetype = 'dashed') +
     annotate('text', x = 20, y = 30, 
              # label = 'atop(R^2 == 0.34, p == 0.002)',
              label = 'atop(rho == 0.52, p == 0.008)',
              parse = TRUE,
              size = 5,
              color = 'blue') +
     xlim(5, 25) + ylim(20, 76) +
     labs(x = 'Mean *trnL* richness',
          y = 'hPDI') +
     theme_bw() +
     theme(axis.title.x = ggtext::element_markdown(),
           axis.title = element_text(size = 14, 
                                     face = 'bold'),
           axis.text = element_text(size = 12, 
                                    face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('..',
            'development',
            'docs',
            'r01',
            '2021_11',
            paste0(plotID, '_Adult-1 hPDI x trnL.pdf')),
       height = 3, width = 3)
```


```{r}
fit.2 <- 
     pdi %>% 
     filter(study == 'Adult-2') %>% 
     lm(hPDI ~ mean, data = .)

summary(fit.2)
```

```{r}
# Mean, SD of raw data and fitted values
pdi %>% 
     filter(study == 'Adult-2', !is.na(hPDI)) %>% 
     pull(hPDI) %>% 
     data.frame(y = .,
                yhat = fitted(fit.2)) %>% 
     summarize(across(.fns = list(mean = mean, sd = sd)))
```

### Residuals

Have calculated residuals for hPDI and uPDI, but overall PDI does not have a relationship to calorie intake.

```{r}
# Select relevant columns, get in long form
pdi.long <- 
     pdi %>% 
     pivot_longer(cols = c(hPDI_residual, uPDI_residual),
                  names_to = 'index', 
                  values_to = 'value')

pdi.long$index <- factor(pdi.long$index)
```

```{r}
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>% 
     distinct() %>% 
     ggplot(aes(x = value, y = mean, color = pass_qc)) +
     geom_point(size = 2, alpha = 0.8) +
     facet_wrap(~study + index,
                nrow = 2) +
     labs(y = 'Mean pMR',
          x = 'hPDI\n') 
     
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '3',
            paste0(plotID, '_raw trnL richness x hPDI.pdf')),
       device = cairo_pdf,
       height = 4, width = 2.5)
```

```{r}
# Raw
pdi.long %>% 
     select(-c(threshold, mean_rare)) %>%
     filter(pass_qc) %>% 
     distinct() %>% 
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean,
                            method = 'spearman')$p.value) 
```

```{r}
# Rarefied
pdi.long %>%
     filter(pass_qc & threshold == 1000) %>% 
     group_by(study, index) %>% 
     summarize(rho = cor.test(~ value + mean_rare,
                           method = 'spearman')$estimate,
               p = cor.test(~ value + mean_rare,
                            method = 'spearman')$p.value) 
```
## Backwards selection

Organize as a function, so that can be run on both Adult-1 and Adult-2 cohorts, with the results compared.

```{r}
# Helper function to calculate mean richness from participants, with optional subsetting of taxa
mean_richness <- 
     function(ps, taxa = NULL){
          # ps: phyloseq object
          # taxa: list of taxa to EXCLUDE from richness calculation
          
          # Get all data
          melt <- 
               psmelt(ps) %>% 
               suppressWarnings() # For phyloseq objects where melting leads to variable name redundancy 
          # Calculate mean per-participant trnL richness
          richness <- 
               melt %>% 
               filter(Abundance > 0) %>% 
               filter(!(OTU %in% taxa)) %>% 
               group_by(Sample, subj) %>% 
               count() %>% # Richness
               ungroup() %>% 
               group_by(subj) %>% 
               summarize(mean_obs = mean(n)) # Within-subject average richness
          
          richness
     }
```

```{r}
backwards_selection <- 
     function(pdi,
              phyloseq){
          
          # INITIALIZATION
          # First, calculate trnL x PDI without limiting taxa, as above:
          # Get simplified richness dataframe
          richness <- mean_richness(phyloseq, taxa = NULL)
          
          # Subset PDI to relevant columns: subject and hPDI
          pdi <- select(pdi, subj, hPDI)
          
          # Join together
          richness <- left_join(richness, pdi,
                                by = 'subj')
          
          # Calculate correlation
          # Get warning: "Cannot compute exact p-value with ties"
          spearman <- 
               cor.test(richness$mean_obs, 
                        richness$hPDI,
                        method = 'spearman') %>% 
               suppressWarnings()
          
          # Initialize variables for while loop
          max_rho <- -Inf
          new_rho <- spearman$estimate
          dropped_features <- c()
          all_features <- taxa_names(phyloseq)
          
          # BACKWARDS SELECTION
          while (new_rho >= max_rho){
               # Update rho to this value, then begin the process again
               max_rho <- new_rho
               
               # Book-keeping dataframe for iterating over features
               results <- 
                    data.frame(feature = NULL,
                               rho = NULL,
                               p = NULL)
               
               for (f in all_features){
                    # Calculate a new richness with that feature dropped
                    # (in addition to all prior drops)
                    richness <- 
                         mean_richness(phyloseq,
                                       taxa = c(dropped_features,
                                                f))
                    
                    # Join to hPDI
                    richness <- left_join(richness, pdi,
                                          by = 'subj')
                    
                    # Calculate correlation
                    spearman <- 
                         cor.test(richness$mean_obs, 
                                  richness$hPDI,
                                  method = 'spearman') %>% 
                         suppressWarnings()
                    
                    
                    # Bind to results dataframe
                    row <- c(f,
                             spearman$estimate, 
                             spearman$p.value)
                    
                    names(row) <- c('feature', 'rho', 'p')
                    
                    results <- bind_rows(results, row)
               }
               
               # Do type conversion on results
               results <- 
                    results %>% 
                    mutate(across(-feature, as.numeric)) 
               
               # Remove p values that aren't significant
               results <- filter(results, p < 0.05)
               
               # Find the maximum correlation coefficient
               new_rho <- max(results$rho) 
               
               # If we're going to loop again, preserve the feature to drop
               # and remove it from the all_features list
               if (new_rho >= max_rho){
                    # Find the feature to remove
                    # Note: there might be more than one: If so, choose at random
                    drop <- 
                         results %>% 
                         filter(rho == new_rho) %>% 
                         pull(feature) %>% 
                         sample(size = 1) 
                    
                    dropped_features <- c(dropped_features, drop)
                    all_features <- 
                         setdiff(taxa_names(phyloseq),
                                  dropped_features)
               }
               
               # Print where we are
               cat('Features dropped:', length(dropped_features), '\n')
          }
          
          # Return the list of dropped features
          dropped_features
     }
```

### Adult-1

```{r}
ps.adult1 <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .)
```

#### With cross-validation

```{r}
samdf <- 
     data.frame(ps.adult1@sam_data) %>% 
     rownames_to_column(var = 'row')
```


```{r}
# Using group k-fold cross-validation data splitting function from caret
# "For Group k-fold cross-validation, the data are split such that no group is contained in both the modeling and holdout sets. One or more group could be left out, depending on the value of k."
set.seed(1019)
folds <- 
     groupKFold(samdf$subj, k = 3)

# Make list for storing results
results.1 <- list()

# For each fold:
for (f in folds){
     # Get sample names of training data
     samples.train <- samdf$row[f]
     
     # Subset the phyloseq object to the training set only
     ps.train <- 
          ps.adult1 %>% 
          prune_samples(samples.train, .) %>% 
          prune_taxa(taxa_sums(.) > 0, .)
     
     # Perform backward selection process, storing results
     result <- backwards_selection(pdi, 
                                   ps.train)
     
     results.1[[length(results.1) + 1]] <- result
}
```

```{r}
# saveRDS(results.1, 
#         here('data', 
#              'processed', 
#              'Adult-1', 
#              'k-fold CV',
#              'k = 3 results.rds'))
```

So now, the strategy is to
     - test each result on its corresponding test fold
          - how many improve/worsen?
          - if worsen, does "backing off" 5 iterations help?
     - of results that hold in test set, what are the core ASVs preserved in/removed from the dataset?

```{r}
summary.df <- 
     data.frame(
          fold = NULL, # Fold number
          set = NULL, # Training or test
          optimized = NULL, # 1: Features dropped; 0: Full dataset
          rho = NULL, # Spearman rho
          p = NULL # Spearman p    
)

for (f in seq_along(folds)){
     # Get training and test sets for this fold
     samples.train <- samdf$row[folds[[f]]]
     samples.test <- samdf$row[!(samdf$row %in% samples.train)]
     ps.train <- prune_samples(samples.train, ps.adult1)
     ps.test <- prune_samples(samples.test, ps.adult1)
     
     # Calculate hPDI correlation on training set
     richness <- 
          ps.train %>% 
          mean_richness() %>% 
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              'train',
              0,
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'set',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
     
     # Calculate hPDI correlation on test set
     richness <- 
          ps.test %>% 
          mean_richness() %>% 
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              'test',
              0,
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'set',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
     
     # Re-calculate the above with selected features only
     dropped <- results[[f]]
     
     # Training
     richness <- 
          ps.train %>% 
          # mean_richness(taxa = dropped) %>% # Exclude the dropped features
          mean_richness(taxa = dropped[1:(length(dropped)-10)]) %>% # iterations back
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              'train',
              1, # Only selected features
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'set',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
     
     # Calculate hPDI correlation on test set
     richness <- 
          ps.test %>% 
          # mean_richness(taxa = dropped) %>% # Only selected features 
          mean_richness(taxa = dropped[1:(length(dropped)-10)]) %>% # # iterations back
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              'test',
              1, # Only selected features
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'set',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
}
```

```{r}
summary.df$rho <- as.numeric(summary.df$rho)
summary.df$p <- as.numeric(summary.df$p)
summary.df$set <- factor(summary.df$set,
                         levels = c('train', 'test'))
```

```{r}
summary.df %>% 
     pivot_longer(cols = c(rho, p),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     ggplot(aes(x = set, y = value, color = fold)) +
     geom_point(alpha = 0.7, size = 3) +
     facet_grid(measure ~ optimized) +
     theme_bw()
```

Ok, so this appears to be *improving* the result in both the training set and the testing set, which is promising.

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'cross-validation',
#             paste0(plotID, '_Adult-1 3-fold cross-validation summary (-10 steps).pdf')),
#        height = 3, width = 5)
```

A few ideas that come to mind here:
- Step back 5 steps in each model, test again
     + Tried this: Improvements with 5, decrease with 10
- Flip flop these three onto Adult-2 and test performance
     + With dropped features: Decline in results, think still overfitting
     - With kept features (think these are conceptually different questions)
- Look for "conserved" features kept in the set
- Try forward selection instead?

#### Test on 2

```{r}
ps.adult2 <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0 , .)
```

##### Dropped 
```{r}
# Dataframe for storing results
summary.df <- 
     data.frame(
          fold = NULL, # Fold number
          optimized = NULL, # With (1) or without (0) feature selection
          rho = NULL, # Spearman rho
          p = NULL # Spearman p    
)

for (f in seq_along(results)){
     # Calculate hPDI correlation on Adult-2 cohort
     richness <- 
          ps.adult2 %>% 
          mean_richness() %>% 
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              0, 
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
     
     # Re-calculate the above with selected features only
     dropped <- results[[f]] # Dropped taxa for this fold

     # Training
     richness <- 
          ps.adult2 %>% 
          # mean_richness(taxa = dropped) %>% # Exclude the dropped features
          mean_richness(taxa = dropped[1:(length(dropped)-10)]) %>% # iterations back
          left_join(pdi, by = 'subj')
                    
     spearman <- 
          cor.test(richness$mean_obs, 
                   richness$hPDI,
                   method = 'spearman') %>% 
          suppressWarnings() # Can't compute exact p-value with ties

     row <- c(f,
              1, # Only selected features
              spearman$estimate, 
              spearman$p.value)
     
     names(row) <- c('fold',
                     'optimized',
                     'rho',
                     'p')
     
     summary.df <- bind_rows(summary.df, row)
}
```

```{r}
summary.df$rho <- as.numeric(summary.df$rho)
summary.df$p <- as.numeric(summary.df$p)
summary.df$fold <- factor(summary.df$fold)
summary.df$optimized <- factor(summary.df$optimized)
```

```{r}
summary.df %>% 
     pivot_longer(cols = c(rho, p),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     ggplot(aes(x = optimized, y = value, color = fold)) +
     geom_point(alpha = 0.7, size = 3) +
     facet_wrap(~ measure) +
     theme_bw()
```

Ok, so this is still overfitting-- declines in performance when tested on other dataset.  And compared to when I did this using the *full* dataset, it's seems worse on average (rho there was 0.36), and only one of the folds here exceeds that.

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'cross-validation',
            paste0(plotID, '_Adult-1 dropped on Adult-2 (-10 steps).pdf')),
       height = 3, width = 5)
```

What about, instead, intersection between dropped features??
Overfitting indicates that we are tossing more than we should.  So a more stringent requirement would be to only discard a feature if it were discarded in all the folds tested.

Find the core set of dropped taxa across the 3 folds.
```{r}
dropped <- Reduce(intersect, results)

dropped
```

Now, try repeating correlations:

```{r}
# Dataframe for storing results
summary.df <- 
     data.frame(
          optimized = NULL, # With (1) or without (0) feature selection
          rho = NULL, # Spearman rho
          p = NULL # Spearman p    
)

# Calculate hPDI correlation on Adult-2 cohort
richness <- 
     ps.adult2 %>% 
     mean_richness() %>% 
     left_join(pdi, by = 'subj')

spearman <- 
     cor.test(richness$mean_obs, 
              richness$hPDI,
              method = 'spearman') %>% 
     suppressWarnings() # Can't compute exact p-value with ties

row <- c(0, 
         spearman$estimate, 
         spearman$p.value)

names(row) <- c('optimized',
                'rho',
                'p')

summary.df <- bind_rows(summary.df, row)

# Re-calculate the above with selected features only

richness <- 
     ps.adult2 %>%
     mean_richness(taxa = dropped) %>% # Remove the intersection of all the dropped taxa 
     left_join(pdi, by = 'subj')

spearman <- 
     cor.test(richness$mean_obs, 
              richness$hPDI,
              method = 'spearman') %>% 
     suppressWarnings() # Can't compute exact p-value with ties

row <- c(1, # Only selected features
         spearman$estimate, 
         spearman$p.value)

names(row) <- c('optimized',
                'rho',
                'p')

summary.df <- bind_rows(summary.df, row)
```

```{r}
summary.df$rho <- as.numeric(summary.df$rho)
summary.df$p <- as.numeric(summary.df$p)
summary.df$optimized <- factor(summary.df$optimized)
```

```{r}
summary.df %>% 
     pivot_longer(cols = c(rho, p),
                  names_to = 'measure',
                  values_to = 'value') %>% 
     ggplot(aes(x = optimized, y = value, color = fold)) +
     geom_point(alpha = 0.7, size = 3) +
     facet_wrap(~ measure) +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'cross-validation',
            paste0(plotID, '_Adult-1 kept on Adult-2 (intersection).pdf')),
       height = 3, width = 5)
```

### Adult-2

#### With cross-validation

```{r}
samdf <- 
     data.frame(ps.adult2@sam_data) %>% 
     rownames_to_column(var = 'row')
```

```{r}
# Using group k-fold cross-validation data splitting function from caret
set.seed(1022)
folds <- 
     groupKFold(samdf$subj, k = 3)

# Make list for storing results
results.2 <- list()

# For each fold:
for (f in folds){
     # Get sample names of training data
     samples.train <- samdf$row[f]
     
     # Subset the phyloseq object to the training set only
     ps.train <- 
          ps.adult2 %>% 
          prune_samples(samples.train, .) %>% 
          prune_taxa(taxa_sums(.) > 0, .)
     
     # Perform backward selection process, storing results
     result <- backwards_selection(pdi, 
                                   ps.train)
     
     results.2[[length(results.2) + 1]] <- result
}
```

```{r}
# saveRDS(results.2,
#         here('data',
#              'processed',
#              'Adult-2',
#              'k-fold CV',
#              'k = 3 results.rds'))
```

#### 2 on 1

```{r}
# Calculate updated correlation coefficient
# Use helper function from above
# Get all data
melt.1 <- 
     psmelt(ps.adult1) 

# Calculate mean per-participant trnL richness
richness.1 <- 
     melt.1 %>% 
     filter(Abundance > 0) %>% 
     filter(OTU %in% features.2) %>% # Kept taxa from Adult-1 
     group_by(Sample, subj) %>% 
     count() %>% # Richness
     ungroup() %>% 
     group_by(subj) %>% 
     summarize(mean_obs = mean(n)) # Within-subject average richness

richness.1 <- left_join(richness.1, pdi, by = 'subj')

cor.test(richness.1$mean_obs,
         richness.1$hpdi,
         method = 'spearman')
```

### Dropped taxa

Want to examine the lowest-level dropped names across each fold and dataset.

```{r}
# Read in past results if needed
results.1 <-
     here('data',
          'processed',
          'Adult-1',
          'k-fold CV',
          'k = 3 results.rds') %>%
     readRDS()

results.2 <-
     here('data',
          'processed',
          'Adult-2',
          'k-fold CV',
          'k = 3 results.rds') %>%
     readRDS()
```

```{r}
# Convert list results to data frame
# Adult-1
results.1 <- lapply(results.1, 
                    function(x){data.frame(study = 'Adult-1',
                                           iter = 1:length(x),
                                           dropped = x)})

names(results.1) <- 1:length(results.1) # Name by fold number

results.1 <- bind_rows(results.1, .id = 'fold') # Bind all together
     
# Repeat for Adult-2
results.2 <- lapply(results.2, 
                    function(x){data.frame(study = 'Adult-2',
                                           iter = 1:length(x),
                                           dropped = x)})

names(results.2) <- 1:length(results.2) 

results.2 <- bind_rows(results.2, .id = 'fold')
     
# Now combine these two
results <- bind_rows(results.1, results.2)
rm(results.1, results.2)
```

```{r}
# Join with taxonomy
results <- 
     data.frame(ps@tax_table) %>% 
     rownames_to_column(var = 'asv') %>% 
     right_join(results, by = c('asv' = 'dropped')) %>% 
     select(study, fold, iter, asv, name, everything()) %>% 
     arrange(study, fold, iter)
```

```{r}
# Summarize most common within dataset
results %>% 
     group_by(asv, name) %>% 
     count() %>% 
     arrange(desc(n))
```
```{r}
# Do the same pulling back 5 iterations, since this gave better performance? 
results %>% 
     group_by(study, fold) %>% 
     filter(row_number() <= (n() - 5)) %>% 
     ungroup() %>% 
     group_by(asv, name) %>% 
     count() %>% 
     arrange(desc(n))
```
Would it also be reasonable to apply a weight? Assuming that earlier things discarded are weighted more heavily.  Could sum iteration numbers to get a score, and then rank by this score?

```{r}
results %>% 
     group_by(study, fold) %>% 
     filter(row_number() <= (n() - 5)) %>% 
     ungroup() %>% 
     group_by(asv, name) %>% 
     summarize(n = length(iter), 
               score = sum(iter)) %>% 
     arrange(desc(n), score)
```

Only those dropped in all 3 folds of cohort:

```{r}
results %>% 
     group_by(study, asv, name) %>%
     count() %>% 
     filter(n == 3)
```

The first 5 drops of each fold: 
```{r}
results %>% 
     group_by(study, fold) %>% 
     slice_head(n = 5) %>% 
     View()
```

Consider instead what's kept:
```{r}
# Remove any ASV that dropped in all 3 folds of either dataset
dropped.in.3 <- 
     results %>% 
     group_by(study, asv, name) %>%
     count() %>% 
     filter(n == 3) %>% 
     pull(asv) 

ps %>% 
     prune_taxa(!(taxa_names(.) %in% dropped.in.3), .) %>% 
     psmelt() %>% 
     pull(name) %>% 
     unique() %>% sort()
```

