---
title: "Alpha diversity"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
library(vegan)
```

# Read in data

## trnL

```{r}
# Read in last saved phyloseqs
outdir <- here('data', 'processed', 'miniseq', 'combined')
ps <- readRDS(file.path(outdir, '20210714_ps.rds'))

rm(outdir)
```

## Diet surveys

### DHQ3

```{r}
ffq <- read_csv(here('data', 'processed', 'DHQ3', 
                     'Full dataset with CHOMP IDs.csv'))
head(ffq)
```
Pull subject ID and any columns relevant to HEI.

```{r}
ffq <- select(ffq, 
              subj, 
              age = Age,
              sex = `Sex (1=male; 2=female)`, 
              matches('HEI'))
names(ffq)
```

I am assuming here that total vegetables, total fruits, and total whole grains are non-overlapping and include the other listed plant categories, but I am not certain. 
Tally these up independently, and also tally all plant categories for comparison.

```{r}
ffq <- 
     ffq %>% 
     rename(hei_tot = 'Total HEI-2015 Score',
            hei_veg = 'HEI-2015 - Total Vegetables - Component Score',
            hei_greensbeans = 'HEI-2015 - Greens and Beans - Component Score',
            hei_fruit = 'HEI-2015 - Total Fruits - Component Score',
            hei_wholefruits = 'HEI-2015 - Whole Fruits - Component Score',
            hei_grain = 'HEI-2015 - Whole Grains - Component Score') %>% 
     # Make custom plant column
     mutate(hei_plant_uniq = hei_veg + hei_fruit + hei_grain,
            hei_plant_all = hei_veg + hei_greensbeans + hei_fruit + 
                 hei_wholefruits + hei_grain)

ffq
```

## Filter

### Foods only

```{r}
# Make a foods-only object
ps
ps.foods <- subset_taxa(ps, !is.na(kingdom))
ps.foods
```


```{r}
# Update read counts in phyloseq object
sample_data(ps.foods)$reads <- sample_sums(ps.foods)
```

```{r}
# What percentage of reads does this retain (overall)?
sum(sample_sums(ps.foods))/sum(sample_sums(ps))
```

```{r}
# What percentage of reads does this retain (per sample)?
(sample_sums(ps.foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     ggplot(., aes(x = percentage)) +
     geom_histogram(binwidth = 0.05, boundary = 0) +
     theme_bw()
```

Follow up on samples that have <90% of their reads preserved after removing those not identified to plant.

```{r}
(sample_sums(ps.foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     filter(is.na(percentage) | percentage < 0.9)
```

```{r}
# Remove samples that now have no reads
ps.foods <- 
     ps.foods %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)
```

# Alpha diversity (trnL)

```{r}
alpha <- 
     estimate_richness(ps.foods, measures = c('Observed', 'Shannon')) %>% 
     # Clean up formatting
     rownames_to_column(var = 'name')

alpha
```

```{r}
# Join to read counts
samdf <- 
     data.frame(ps.foods@sam_data)

alpha <- 
     samdf %>% 
     select(name, reads, subj) %>% 
     right_join(alpha)

alpha
```

### vs. read counts

```{r}
alpha.long <- pivot_longer(alpha, cols = c(Observed, Shannon), 
                           names_to = 'measure', values_to = 'value')
```

```{r}
ggplot(alpha.long, aes(x = reads, y = value)) +
     geom_point() +
     labs(x = 'Read depth', y = 'Alpha diversity') +
     facet_wrap(~measure, scales = 'free_y') +
     theme_bw()
```

## Rarefaction

```{r}
# Pick a read count threshold
thresh <- 10000
```

```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- 
     ps.foods %>% 
     subset_samples(reads >= thresh) %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     otu_table(.)

asvtab <- asvtab@.Data
```

```{r}
raremin <- 
     asvtab %>% 
     rowSums() %>% 
     min()

raremin # The lowest read depth in the filtered data
```

```{r}
asvtab.rare <- rarefy(asvtab, raremin)
head(asvtab.rare) # The expected number of species if only raremin reads
```

```{r}
# View curves
rarecurve(asvtab[1:20, ], col = "blue")
```

```{r}
# How does this compare to actual number passed to analysis below?
rarefaction.df <- 
     data.frame(obs_rare = asvtab.rare) %>% 
     rownames_to_column(var = 'name')

# Add actual # observed taxa
rarefaction.df <- 
     rowSums(asvtab > 0) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name') %>% 
     full_join(rarefaction.df)

# Add 
```

```{r}
ggplot(rarefaction.df, aes(x = obs, y = obs_rare)) +
     geom_point() +
     coord_equal()
```

## By subject 

```{r}
# How well do observed and Shannon correlate with one another?
ggplot(alpha, aes(Observed, Shannon)) +
     geom_point() +
     coord_fixed(ratio = max(alpha$Observed)/max(alpha$Shannon))
```

```{r}
cor.test(alpha$Shannon, alpha$Observed, 
         alternative = 'greater',
         method = 'pearson')
```


```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(Observed), sd = sd(Observed)) %>% 
     arrange(desc(med))

alpha$subj <- factor(alpha$subj, levels = alpha.stats$subj)

ggplot(alpha, aes(x = subj, y = Observed)) +
     geom_boxplot() +
     labs(x = 'Participant') +
     theme_bw() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(), 
          panel.grid.major.x = element_blank()
     )
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'miniseq', 'Combined', 'alpha diversity',
            paste0(plotID, '_Alpha diversity by subject, raw.pdf')),
       height = 3, width = 5)
```

```{r}
# Factor by largest standard deviation instead
```



How does this change with filtering out of low quality samples?
What about at different phylogenetic levels?

# Dietary diversity (FFQ)

Average alpha diversity results within each participant, and join to FFQ data.

```{r}
ffq <- 
     alpha.long %>% 
     group_by(subj, measure) %>% 
     summarize(mean = mean(value), sd = sd(value)) %>% 
     left_join(ffq)

ffq
```

## vs. total HEI score
```{r}
ggplot(ffq, aes(x = mean, y = hei_tot)) +
     geom_point() +
     facet_wrap(~measure, scales = 'free_x')
```

```{r}
ffq.wide <- 
     ffq %>% 
     select(-sd) %>% 
     pivot_wider(names_from = measure, values_from = mean)

ffq.wide
```
```{r}
cor.test(ffq.wide$Observed, ffq.wide$hei_tot,
         alternative = 'two.sided',
         method = 'pearson')
```

```{r}
cor.test(ffq.wide$Shannon, ffq.wide$hei_tot,
         alternative = 'two.sided',
         method = 'pearson')
```

## vs. plant-components of HEI score

### Non-redundant
```{r}
ggplot(ffq, aes(x = mean, y = hei_plant_uniq)) +
     geom_point() +
     facet_wrap(~measure, scales = 'free_x')
```

```{r}
cor.test(ffq.wide$Observed, ffq.wide$hei_plant_uniq,
         alternative = 'greater',
         method = 'spearman')
```

```{r}
cor.test(ffq.wide$Shannon, ffq.wide$hei_plant_uniq,
         alternative = 'greater',
         method = 'spearman')
```

### All

```{r}
ggplot(ffq, aes(x = mean, y = hei_plant_all)) +
     geom_point() +
     facet_wrap(~measure, scales = 'free_x') +
     labs(x = 'Alpha diversity value by metabarcoding',
          y = 'HEI-2015 plant components score') +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 'miniseq', 'Combined', 'alpha diversity',
#             paste0(plotID, '_HEI-2015, all plant categories.png')),
#        height = 3, width = 5)
```

```{r}
cor.test(ffq.wide$Observed, ffq.wide$hei_plant_all,
         alternative = 'greater',
         method = 'spearman')
```

```{r}
cor.test(ffq.wide$Shannon, ffq.wide$hei_plant_all,
         alternative = 'greater',
         method = 'spearman')
```

## vs. demographics

### Age
```{r}
ggplot(ffq, aes(x = age, y = mean)) +
     geom_point() +
     facet_wrap(~measure, scales = 'free_y')
```

```{r}
summary(lm(Observed ~ age, ffq.wide))
```

### Sex

```{r}
ggplot(ffq, aes(x = sex, y = mean, by = sex, group = sex)) +
     geom_boxplot() + 
     facet_wrap(~measure, scales = 'free_y')
```

```{r}
men <- ffq.wide$Shannon[ffq.wide$sex == 1]
women <- ffq.wide$Shannon[ffq.wide$sex == 2]

t.test(men, women)
```
