---
title: "Alpha diversity"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
library(vegan)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseqs
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

# Alpha diversity

```{r}
alpha <- 
     estimate_richness(ps, measures = c('Observed', 'Shannon')) %>% 
     # Clean up formatting
     rownames_to_column(var = 'name') %>% 
     mutate(name = gsub('\\.', '-', name))

alpha
```

```{r}
# Join to read counts
samdf <- 
     data.frame(ps@sam_data) %>% 
     select(-name) %>% # SUBJ_WEEK_DATE format, incomplete across cohorts
     rownames_to_column(var = 'name')

alpha <- 
     samdf %>% 
     select(name, study, subj, reads) %>% 
     right_join(alpha)

alpha
```
### vs. read counts

```{r}
alpha.long <- pivot_longer(alpha, cols = c(Observed, Shannon), 
                           names_to = 'measure', values_to = 'value')
```

```{r}
ggplot(alpha.long, aes(x = reads, y = value)) +
     geom_point() +
     labs(x = 'Read depth', y = 'Alpha diversity') +
     facet_wrap(~measure, scales = 'free_y') +
     theme_bw()
```

## Rarefaction

```{r}
# Pick a read count threshold
thresh <- 1
```

```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- 
     ps %>% 
     subset_samples(reads_food >= thresh) %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     otu_table(.)

asvtab <- asvtab@.Data
dim(asvtab)
```

```{r}
raremin <- 
     asvtab %>% 
     rowSums() %>% 
     min()

raremin # The lowest read depth in the filtered data
```

```{r}
asvtab.rare <- rarefy(asvtab, raremin)
head(asvtab.rare) # The expected number of species if only raremin reads
```

```{r}
# View curves
# Randomly choose 20 CHOMP rows
rarecurve(asvtab[1:20, ], col = "blue")
```

More important than the curves may be conclusions we can draw from the slope: `rareslope` calculates the slope of `rarecurve` at `sample` size (can be a vector).
```{r}
slopes <- rareslope(asvtab, seq(100, 20000, 100))
```

```{r}
slopes.long <- 
     slopes %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     pivot_longer(cols = -row,
                  names_to = 'sample',
                  values_to = 'slope') %>% 
     mutate(sample = as.numeric(gsub('N', '', sample)),
            study = gsub('-\\d+$', '', row))

slopes.long
```

```{r}
ggplot(slopes.long, aes(x = sample, y = slope, group = sample)) +
     geom_point() 
```

```{r}
means <- colSums(slopes)/dim(slopes)[1]
```

```{r}
# What does slope *mean*? For each additional read, we see that many additional species in the sample. Taking the inverse asks: How many reads do we need to see one additional species?
1/means
```

Ok, so let's consider then a 5000-read threshold, which would mean we'd need 14,626 additional reads to get one more species. Could perhaps go even lower based on inflection point of graph above.

```{r}
# How does this compare to actual number passed to analysis below?
rarefaction.df <- 
     data.frame(obs_rare = asvtab.rare) %>% 
     rownames_to_column(var = 'name')

# Add actual # observed taxa
rarefaction.df <- 
     rowSums(asvtab > 0) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name') %>% 
     full_join(rarefaction.df)

# Add 
```

```{r}
ggplot(rarefaction.df, aes(x = obs, y = obs_rare)) +
     geom_point() +
     coord_equal()
```

## By subject 

```{r}
# How well do observed and Shannon correlate with one another?
ggplot(alpha, aes(Observed, Shannon)) +
     geom_point() +
     coord_fixed(ratio = max(alpha$Observed)/max(alpha$Shannon))
```

```{r}
cor.test(alpha$Shannon, alpha$Observed, 
         alternative = 'greater',
         method = 'pearson')
```

```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(Observed), sd = sd(Observed)) %>% 
     arrange(desc(med))

alpha$subj <- factor(alpha$subj, levels = alpha.stats$subj)

ggplot(alpha, aes(x = subj, y = Observed)) +
     geom_boxplot() +
     facet_wrap(~study, scales = 'free_x') +
     labs(x = 'Participant') +
     theme_bw() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(), 
          panel.grid.major.x = element_blank()
     )
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'miniseq', 'Combined', 'alpha diversity',
            paste0(plotID, '_Alpha diversity by study, raw.pdf')),
       height = 3, width = 5)
```

```{r}
alpha %>% 
     filter(study %in% c('CHOMP', 'ONR')) %>% 
     # filter(reads > 10000) %>% 
     ggplot(aes(x = subj, y = Observed, fill = study)) +
     geom_boxplot() +
     # facet_wrap(~study, scales = 'free_x') +
     labs(x = 'Participant') +
     theme_bw() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(), 
          panel.grid.major.x = element_blank()
     )
```


```{r}
# Factor by largest standard deviation instead
```



How does this change with filtering out of low quality samples?
What about at different phylogenetic levels?
Adding more comprehensive demographic data?