---
title: "Figure 1"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
library(vegan)
```

# Load data

## trnL richness (+/- rarefaction)

```{r}
alpha <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS()
```

## Weight Loss menu data

```{r}
# Menu phyloseq object (taxa consumed)
ps.menu <- 
     here('data', 
          'processed', 
          'Weight Loss', 
          '20210220_ps_menu.rds') %>% 
     readRDS()
```

```{r}
# Menu summaries (nutrients consumed)
nutrients <- 
     here('data',
          'processed',
          'Weight Loss',
          'Nutrients per day.csv') %>% 
     read_csv()
```

# Pre-process

## Metabarcoding data

```{r}
# Subset to Weight Loss samples and metadata only
alpha <- 
     alpha %>% 
     filter(study == 'Weight Loss') %>% 
     # Drop columns that are all NA
     select(
          where(function(x){ 
               !all(is.na(x))
          })
     ) %>% 
      # Re-order remainder
     select(
          subj,
          date,
          obs, 
          threshold, 
          starts_with('obs_rare')
          ) 

alpha
```

```{r}
# Add link to menu day at varying lag

# Label each stool sample with a week of preceding days as SUBJ_DATE
for (delta in 1:7){
     varname <- paste0('delta', delta)
     alpha <- mutate(alpha, 
                     !!varname := date - delta,)
}

alpha
```

## Menu data

```{r}
head(sample_names(ps.menu))
```

```{r}
# Filter to plants only
ntaxa(ps.menu)
ps.menu <- subset_taxa(ps.menu,
                       kingdom == 'Viridiplantae')
ntaxa(ps.menu)
```

### Calculate richness

```{r}
# Get intake table
asvtab.menu <- ps.menu@otu_table@.Data 
# Calculate number of non-zero entries
alpha.menu <- apply(asvtab.menu,
                    1, 
                    function(x){sum(x > 0)})  

# Reformat as dataframe
alpha.menu <- 
     data.frame(obs_menu = alpha.menu) %>% 
     rownames_to_column(var = 'row') %>% 
     separate(row, 
              into = c('subj', 'date_menu'), 
              sep = '_')
```

```{r}
# Add weekday
alpha.menu <- 
     alpha.menu %>% 
     mutate(date_menu = as.Date(date_menu, format = '%Y-%m-%d'),
            weekday_menu = weekdays(date_menu)) %>% 
     select(subj, date_menu, weekday_menu, obs_menu)

alpha.menu
```
```{r}
# Join to nutrient intakes
alpha.menu <- 
     left_join(alpha.menu, 
               nutrients,
               by = c('subj', 
                      'date_menu' = 'date', 
                      'weekday_menu' = 'weekday'))

rm(nutrients)
```

# Analyze

## Range of taxa 

Overall dataset

```{r}
range(alpha.menu$obs_menu)
```

What if I filter out weekends (lower quality?)
```{r}
alpha.menu %>% 
     filter(!(weekday %in% c('Saturday', 'Sunday'))) %>% 
     pull(obs_menu) %>% 
     range()
```

Median within each subject

```{r}
# Including weekends
alpha.menu %>% 
     group_by(subj) %>% 
     summarize(median = median(obs_menu),
               sd = sd(obs_menu))
```

```{r}
# Excluding weekends
alpha.menu %>% 
     filter(!(weekday %in% c('Saturday', 'Sunday'))) %>% 
     group_by(subj) %>% 
     summarize(median = median(obs_menu),
               sd = sd(obs_menu))
```

## Correlation with trnL richness

Have two dataframes:

* `alpha`
* `alpha.menu`

First, might want to understand day-to-day variability in trnL richness compared to menu richness.


```{r}
ggplot(alpha.menu, aes(x = date_menu, y = obs_menu, color = subj)) +
     geom_point() +
     facet_wrap(~subj, scales = 'free_x') +
     theme_bw() +
     theme(legend.position = 'none') 
```
```{r}
ggplot(alpha, aes(x = date, y = obs_rare, color = subj)) +
     geom_point() +
     facet_wrap(~subj, scales = 'free_x') +
     theme_bw() +
     theme(legend.position = 'none') 
```

```{r}
# 1 day of lag
lag <- 'delta1'

# Make joint object
alpha.joint <- 
     alpha %>%
     select(subj,
            date,
            date_menu = all_of(lag), 
            starts_with('obs')) %>% 
     # mutate(delta = gsub('delta', '', lag)) %>% 
     left_join(alpha.menu) %>% 
     select(subj,
            starts_with('date'),
            weekday_menu,
            starts_with('obs'),
            everything())

head(alpha.joint)
```
```{r}
ggplot(alpha.joint, aes(x = obs_menu, y = obs_rare, color = subj)) +
     geom_abline(slope = 1, intercept = 0) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```
```{r}
cor.test(alpha.joint$obs_rare,
         alpha.joint$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

Should I test other days, or leave it at that?

What about excluding weekends?

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>%
     cor.test(~ obs_rare + obs_menu, 
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

How about Sundays, Mondays only
```{r}
alpha.joint %>% 
     filter(weekday_menu %in% c('Saturday', 'Sunday')) %>%
     cor.test(~ obs_rare + obs_menu, 
              data = .,
              method = 'spearman')
```

How about in comparison to the day after today?

```{r}
alpha.next <- 
     alpha %>% 
     select(subj, date, obs_rare) %>% 
     mutate(date_menu = date + 1) %>% 
     left_join(alpha.menu) %>% 
     select(subj,
            starts_with('date'),
            weekday_menu,
            starts_with('obs'),
            everything())

alpha.next
```

```{r}
ggplot(alpha.next, aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```
```{r}
cor.test(alpha.next$obs_rare,
         alpha.next$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

How about with a random day (within each subject)?

```{r}
# Make joint object
alpha.rand <- 
     alpha %>%
     select(subj, date, obs_rare)

# Randomly choose a comparator day (within each subject)
sample_subj <- function(s, n){
     alpha.menu %>% 
          filter(subj == s) %>% 
          pull(row) %>% 
          sample(size = n, replace = FALSE)
}
```

```{r}
# How many entries per subject do we need?
alpha.rand %>% 
     group_by(subj) %>% 
     count()
```


```{r}
set.seed(0816)

rand.menu <- 
     c(
          sample_subj('AK65', 9),
          sample_subj('JN55', 13),
          sample_subj('ML48', 6),
          sample_subj('PC74', 4)
     ) %>% 
     data.frame(comparator_rand = .) %>% 
     left_join(alpha.menu,
               by = c('comparator_rand' = 'row'))

alpha.rand <- 
     alpha.rand %>% 
     arrange(subj) %>%
     select(-subj) %>% 
     bind_cols(rand.menu)
```

```{r}
ggplot(alpha.rand, aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```

```{r}
cor.test(alpha.rand$obs_rare,
         alpha.rand$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

How about with a random day (any day)?

```{r}
alpha.rand <- 
     alpha %>%
     mutate(sample = paste(subj, date, sep = '_'))  %>% 
     select(subj, sample, obs_rare)

dim(alpha.rand)
```

```{r}
set.seed(1234)
alpha.rand <- 
     bind_cols(alpha.rand, 
               row = sample(alpha.menu$row, 32))

alpha.rand <- 
     alpha.rand %>% 
     select(-subj) %>% 
     left_join(alpha.menu)
```

```{r}
ggplot(alpha.rand, aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```
```{r}
cor.test(alpha.rand$obs_rare,
         alpha.rand$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

## Relationship to macronutrient intake

Start with just one day's comparison. 
```{r}

```

