---
title: "R Notebook"
output: html_notebook
---

---
title: "Figure 1"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(metacoder)
library(tidyverse)
library(vegan)
```

# Read in data

## trnL

```{r}
# Read in last saved phyloseqs
outdir <- here('data', 'processed', 'miniseq', 'combined')
ps <- readRDS(file.path(outdir, '20210714_ps.rds'))

rm(outdir)
```

## Filter

### Foods only

```{r}
# Make a foods-only object
ps
ps.foods <- subset_taxa(ps, !is.na(kingdom))
ps.foods
```


```{r}
# Update read counts in phyloseq object
sample_data(ps.foods)$reads <- sample_sums(ps.foods)
```

```{r}
# What percentage of reads does this retain (overall)?
sum(sample_sums(ps.foods))/sum(sample_sums(ps))
```

```{r}
# What percentage of reads does this retain (per sample)?
(sample_sums(ps.foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     ggplot(., aes(x = percentage)) +
     geom_histogram(binwidth = 0.05, boundary = 0) +
     theme_bw()
```

Follow up on samples that have <90% of their reads preserved after removing those not identified to plant.

```{r}
(sample_sums(ps.foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     filter(is.na(percentage) | percentage < 0.9)
```

```{r}
# Remove samples that now have no reads
ps.foods <- 
     ps.foods %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)
```

# Visualize taxa

## Metacoder tree

```{r}
# Add lowest level information
taxtab <- 
     data.frame(ps.foods@tax_table) %>% 
     lowest_level()
```

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps.foods)
```

```{r}
x
```

Helpful functions for accessing data
```{r}
taxon_names(x) %>% head() # Taxa names
taxon_ranks(x) %>% head() # Taxa ranks
```

```{r}
# Convert counts to presence-absence
x$data$otu_table <- 
     counts_to_presence(x,
                        data = 'otu_table', 
                        cols = x$data$sample_data$sample_id)

# Returned logical; convert to numeric
x$data$otu_table <- 
     x$data$otu_table %>% 
     mutate_at(vars(-taxon_id), as.numeric)

# Calculate taxon abundance in this presence-absence scheme
# Not sure how this is working
x$data$taxon_counts <- 
     x$data$otu_table %>% 
     column_to_rownames(var = 'taxon_id') %>% 
     rowSums()

x$data$taxon_counts$total <- 
     rowSums(x$data$taxon_counts[, -1]) # -1 = taxon_id column
```

```{r}
# Plot heat tree based on presence-absence data
x %>% 
     heat_tree(node_size = n_obs,
               node_size_axis_label = "Number of ASVs",
               node_color = total,
               node_label = taxon_names)
```

