---
title: "Figure 2 (Adult-1, Adult-2)"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

# Read in data

## trnL

Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseqs
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

### Rarefied richness estimates

```{r}
rare.df <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS

rare.df
```
## Diet surveys

### ASA24

### DHQ3

## Diet surveys

### DHQ3

#### Adult-1

```{r}
ffq.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3.csv') %>% 
     read_csv()

head(ffq.adult1)
```

#### Adult-2

```{r}
ffq.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3.csv') %>% 
     read_csv()

head(ffq.adult2)
```


# Pre-process

## trnL

```{r}
# Filter to Adult-1 and -2 cohorts only
ps <- 
     ps %>% 
     subset_samples(study %in% c('Adult-1', 'Adult-2')) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps
```

## Diet surveys

Pull subject ID and any columns relevant to HEI.

```{r}
ffq <- select(ffq, 
              subj, 
              age = Age,
              sex = `Sex (1=male; 2=female)`, 
              matches('HEI'))
names(ffq)
```

I am assuming here that total vegetables, total fruits, and total whole grains are non-overlapping and include the other listed plant categories, but I am not certain. 
Tally these up independently, and also tally all plant categories for comparison.

```{r}
ffq <- 
     ffq %>% 
     rename(hei_tot = 'Total HEI-2015 Score',
            hei_veg = 'HEI-2015 - Total Vegetables - Component Score',
            hei_greensbeans = 'HEI-2015 - Greens and Beans - Component Score',
            hei_fruit = 'HEI-2015 - Total Fruits - Component Score',
            hei_wholefruits = 'HEI-2015 - Whole Fruits - Component Score',
            hei_grain = 'HEI-2015 - Whole Grains - Component Score') %>% 
     # Make custom plant column
     mutate(hei_plant_uniq = hei_veg + hei_fruit + hei_grain,
            hei_plant_all = hei_veg + hei_greensbeans + hei_fruit + 
                 hei_wholefruits + hei_grain)

ffq
```


# Analyze

```{r}
samdf <- data.frame(ps@sam_data)
```


## Read depth

```{r}
range(sample_data(ps)$reads_food)
```

```{r}
# Have I appropriately preserved failed samples in the phyloseq object?
# TODO: Check this in ONR and CHOMP runs
any(sample_sums(ps) == 0)
```

```{r}
median(sample_data(ps)$reads_food)
```

```{r}
sd(sample_data(ps)$reads_food)
```

### Read count histograms

```{r}
hist(sample_data(ps)$reads_food)
```

```{r}
# Faceted by study
ggplot(samdf, aes(x = reads_food, fill = study)) +
     geom_histogram(binwidth = 10000, boundary = 0) +
     facet_wrap(~study) +
     scale_fill_manual(values = c('gray80', 'gray40')) +
     # xlim(0, 50000) +
     labs(x = 'Read count (food taxa)', y = 'Samples (n)') +
     theme_bw() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1),
           legend.title = element_blank(),
           legend.position = 'none')
```

### Rarefaction curves

#### Stats

```{r}
# Number of dropped samples
sum(sample_sums(ps) < 10500)
```

```{r}
# Percentage
# Number of dropped samples
sum(sample_sums(ps) < 10500)/nsamples(ps) * 100
```

## Richness by participant

## By subject 

```{r}
# Refactor subject IDs in order of descending median taxa observations
alpha.stats <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(med = median(obs_rare_10500, na.rm = TRUE), 
               sd = sd(obs_rare_10500, na.rm = TRUE)) %>% 
     arrange(desc(med))

alpha$subj <- factor(alpha$subj, levels = alpha.stats$subj)
```

```{r}
alpha %>% 
     ggplot(aes(x = subj, y = obs_rare_10500)) +
     geom_boxplot() +
     facet_wrap(~study, scales = 'free_x') +
     labs(x = 'Participant') +
     theme_bw() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(), 
          panel.grid.major.x = element_blank()
     )
```


```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'miniseq', 'Combined', 'alpha diversity',
            paste0(plotID, '_Alpha diversity by study, raw.pdf')),
       height = 3, width = 5)
```

```{r}
alpha %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     ggplot(aes(x = subj, y = obs_rare_10500, fill = study)) +
     geom_boxplot() +
     # facet_wrap(~study, scales = 'free_x') +
     labs(x = 'Participant', y = 'trnL richness') +
     scale_fill_manual(values = c('gray80', 'gray40')) +
     theme_bw() +
     theme(
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(), 
          panel.grid.major.x = element_blank(),
          legend.position = c(0.9,0.8),
          legend.background = element_rect(color = 'black'),
          legend.title = element_blank()
     )
```

```{r}
adult.stats <- 
     alpha %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     group_by(subj) %>% 
     summarize(med = median(obs_rare_10500,
                            na.rm = TRUE),
               sd = sd(obs_rare_10500,
                       na.rm = TRUE))

range(adult.stats$med, na.rm = T)
range(adult.stats$sd, na.rm = T)
```


```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 'manuscript', '1',
            paste0(plotID, '_B_Richness by subject, 5,000 read rarefaction.pdf')),
       height = 4, width = 6)
```

## Comparison to HEI



### Adult-1

### Adult-2