---
title: "Figure 3"
output:
  html_document:
    df_print: paged
    fig_width: 5
    fig_height: 3.5
    toc: true
    toc_float: true
---

```{r libraries, include=FALSE, echo=FALSE}
library(fitdistrplus)
library(patchwork)
library(phyloseq)
library(tidyverse)

select <- dplyr::select
```

```{r include=FALSE, echo=FALSE}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data
```{r}
# Full trnL phyloseq object
ps <- readRDS('data/phyloseq_trnL.rds')

# Alpha diversity estimates
alpha <- read_csv('data/alpha.csv')
```

# Pre-process

```{r}
# Filter trnL data to successfully sequenced Adolescent samples
ps <- subset_samples(ps, study == 'Adolescent') 
ps <- prune_samples(sample_sums(ps) > 0, ps)

# Move forward with pMR only
alpha <- 
     alpha |> 
     filter(grepl(name, pattern = 'POMMS')) |> 
     filter(measure == 'Richness' & rarefied == 'Raw') |> 
     select(-rarefied)
```

```{r}
# Adolescent demographic data
samdf <- 
     ps@sam_data |> 
     data.frame() |> 
     rownames_to_column(var = 'row') |> 
     filter(study == 'Adolescent')
```

### Completeness

Check completeness of model covariates.  

```{r}
# Do all subjects have an entry timepoint?
n_distinct(samdf$subj)

samdf |> 
     filter(timepoint == 'Entry') |> 
     nrow()
```

Almost, but not everyone.  So will group by subject below rather than just filtering to one timepoint. 

```{r}
# Check only one measure per subject
samdf |>
     group_by(subj, ethnicity) |> 
     count() |> 
     ungroup() |> 
     group_by(ethnicity) |> 
     count()
```

```{r}
samdf |> 
     group_by(subj, sex) |> 
     count() |> 
     ungroup() |> 
     group_by(sex) |> 
     count()
```

```{r}
samdf |> 
     group_by(subj, race) |> 
     count() |> 
     ungroup() |> 
     group_by(race) |> 
     count()
```

```{r}
# Recode these to reduce categories with small number
# Mixed-race responses as "Multiple"
samdf$race[grep(' ', samdf$race)] <- 'Multiple'

samdf |> 
     group_by(subj, race) |> 
     count() |> 
     ungroup() |> 
     group_by(race) |> 
     count()
```

```{r}
samdf |> 
     group_by(subj, casecontrol) |> 
     count() |> 
     ungroup() |> 
     group_by(casecontrol) |> 
     count()
```

#### SES indicators

```{r}
samdf |> 
     group_by(subj, ses) |> 
     count() |> 
     ungroup() |> 
     group_by(ses) |> 
     count()
```

```{r}
# Make "any" category for worry/actual insecurity
samdf <- 
     mutate(
          samdf,
          across(.cols = c(worry, actual),
                 .fns = ~ ifelse(.x != 'Never true',
                                 yes = 'Yes',
                                 no = 'No'),
                 .names = '{.col}_ever'
                 )
     )
```

```{r}
samdf |> 
     group_by(subj, actual_ever) |> 
     count() |> 
     ungroup() |> 
     group_by(actual_ever) |> 
     count()
```

```{r}
# Recode NA as Unknown so these aren't discarded from the model
samdf$ses[is.na(samdf$ses)] <- 'Unknown'
samdf$worry_ever[is.na(samdf$worry_ever)] <- 'Unknown'
samdf$actual_ever[is.na(samdf$actual_ever)] <- 'Unknown'
```

#### Age

```{r}
# Age into quartiles by year, as in POMMS group analysis
samdf <- 
     samdf |> 
     mutate(age_years = floor(age_days/365),
            age_q = cut(age_years,
                        breaks = 4))

samdf |> 
     group_by(subj, age_q) |> 
     count() |> 
     ungroup() |> 
     group_by(age_q) |> 
     count()
```

If age not available from one of the samples, take it from another:

```{r}
correction <- 
     samdf |> 
     group_by(subj) |> 
     summarize(n_age_q = n_distinct(age_q)) |> 
     filter(n_age_q == 2)

samdf |> 
     filter(subj %in% correction$subj) |> 
     select(subj, starts_with('age')) |> 
     arrange(subj)
```
Okay, so some of these are appropriately when the participant's age has changed.  We want to just keep those that are NA. 

```{r}
# Get a list of subjects for which we'll replace one value with the other
correction <- 
     samdf |> 
     filter(subj %in% correction$subj) |> 
     select(subj, starts_with('age')) |> 
     group_by(subj) |> 
     summarize(na.present = any(is.na(age))) |> 
     filter(na.present == TRUE) |> 
     pull(subj)

length(correction)
```

Awesome!! This is exactly the same as the number of missing values above.  

```{r}
samdf |> 
     filter(subj %in% correction) |> 
     select(subj, timepoint, starts_with('age')) |> 
     arrange(subj, desc(timepoint))
```
Manually inspected and for all of these it's the 6 month timepoint missing. So add half a year (183 d) to their age_days variable, then convert.

```{r}
correction.df <- 
     samdf |> 
     filter(subj %in% correction) |> 
     filter(timepoint == 'Entry') |> 
     select(subj, timepoint, age_days) |> 
     mutate(timepoint = '6 months',
            age_days_corr = age_days + 183) |> 
     select(subj, timepoint, ends_with('corr'))

correction.df
```
```{r}
# Bind to sample data and replace values if needed
samdf <- 
     samdf |> 
     select(-age) |> 
     left_join(correction.df) |> 
     mutate(age_days = 
                 ifelse(is.na(age_days),
                        yes = age_days_corr,
                        no = age_days)) 

# Peek at results
samdf |> 
     filter(subj %in% correction.df$subj) |> 
     select(subj, starts_with('age')) |> 
     arrange(subj)

# Now, can drop corrected column, recalculate 
samdf <- 
     samdf |> 
     select(-age_days_corr) |> 
          mutate(age_years = floor(age_days/365),
            age_q = cut(age_years,
                        breaks = 4))
```

```{r}
# Any missing now? 
samdf |> 
     group_by(subj, age_q) |> 
     count() |> 
     ungroup() |> 
     group_by(age_q) |> 
     count()
```

```{r}
# Convert to 10s of years for better visualization
samdf$age_decades <- samdf$age_days/365/10 
```

#### Set reference

```{r}
class(samdf$sex)

samdf$sex <- factor(samdf$sex,
                          levels = c('M', 'F'))

levels(samdf$sex)
```

```{r}
class(samdf$race)

samdf$race <- factor(samdf$race,
                     levels = c('WH', 
                                'BL', 
                                'AS', 
                                'Multiple'))

levels(samdf$race)
```

```{r}
class(samdf$ethnicity)

samdf$ethnicity <- factor(samdf$ethnicity,
                          levels = c('NH', 'HI'))

levels(samdf$ethnicity)
```

```{r}
samdf$age_q <- as.ordered(samdf$age_q)
levels(samdf$age_q)
```

```{r}
class(samdf$casecontrol)

samdf$casecontrol <- factor(samdf$casecontrol,
                          levels = c('Control', 'Case'))

levels(samdf$casecontrol)
```

```{r}
class(samdf$ses)

samdf$ses <- 
     factor(samdf$ses,
            levels = c('<24,999',
                       '25,000-49,999',
                       '50,000-74,999',
                       '75,000-99,999',
                       '>100,000',
                       'Unknown')) 
levels(samdf$ses)
```

```{r}
class(samdf$actual_ever)

samdf$actual_ever <- 
     factor(samdf$actual_ever,
            levels = c('No',
                       'Yes',
                       'Unknown')) 

levels(samdf$actual_ever)
```

## Join

```{r}
# Pull from sample data, and join to results
alpha <- 
     samdf |> 
     select(row, 
            subj,
            timepoint,
            casecontrol,
            age_decades, 
            sex,
            race,
            ethnicity,
            ses,
            worry_ever,
            actual_ever) |> 
     right_join(alpha,
                by = c('row'='name'))

alpha
```

# Analyze

```{r}
# Per subject median, MAD
alpha |> 
     group_by(subj) |> 
     summarize(value = mean(value)) |> 
     ungroup() |> 
     summarize(med = median(value),
               mad = mad(value))
```

## Model: pMR response

```{r}
# Filter to Entry timepoint
dim(alpha)
alpha.subset <- filter(alpha, timepoint == 'Entry') 
dim(alpha.subset)
```

```{r}
# Check distribution of response variable
descdist(alpha.subset$value, discrete = TRUE, boot = 500)
```

```{r}
# Check distribution
p.pmr.hist <- 
     ggplot(alpha.subset, aes(x = value)) +
     geom_histogram(binwidth = 2,
                    boundary = 0) +
     labs(x = 'pMR\n', y = 'Frequency') +
     theme(panel.grid.minor = element_blank())

p.pmr.hist
```
```{r}
# QQ plot
qqnorm(alpha.subset$value, pch = 1, frame = FALSE)
qqline(alpha.subset$value, col = "steelblue", lwd = 2)
```

```{r}
f = "value ~ age_decades + sex + race + ethnicity + casecontrol + ses + actual_ever"

model <- lm(f,
            data = alpha.subset)

summary(model)
```

### Plot betas

```{r}
# Get confidence intervals
cis <- 
     confint(model) |> 
     data.frame() |> 
     rownames_to_column(var = 'covar') 

# Get beta hat and p from model
betas <- 
     summary(model)$coeff[, c(1, 4)] |> 
     data.frame() |> 
     rownames_to_column(var = 'covar') 

betas <- 
     full_join(betas, cis) |> 
     rename(p = Pr...t..,
            CI2.5 = X2.5..,
            CI97.5 = X97.5..)

betas
rm(cis)
```

```{r}
# Assign covariate grouping
betas$group <- NULL
betas$group[grepl(pattern = 'age|race|sex|ethnicity',
                  betas$covar)] <- 'Demographic'
betas$group[grepl(pattern = 'case',
                  betas$covar)] <- 'Clinical'
betas$group[grepl(pattern = 'ses|ever',
                  betas$covar)] <- 'Socioeconomic'

betas$group <- factor(betas$group,
                      levels = c('Socioeconomic',
                                 'Clinical',
                                 'Demographic'))
```

```{r}
# Refactor by values of Estimate
betas <- 
     betas |> 
     arrange(group, covar)
```

```{r}
# Recode names for more interpretable plot
betas$covar <- 
     factor(betas$covar,
            levels = c('actual_everUnknown',
                       'actual_everYes',
                       'sesUnknown',
                       'ses>100,000',
                       'ses75,000-99,999',
                       'ses50,000-74,999',
                       'ses25,000-49,999',
                       'casecontrolCase',
                       'ethnicityHI',
                       'raceMultiple',
                       'raceAS',
                       'raceBL',
                       'sexF',
                       'age_decades',
                       '(Intercept)'),
            labels = c('foodinsecUnknown',
                       'foodinsecYes',
                       'incomeUnknown',
                       'income>100K',
                       'income75-99K',
                       'income50-74K',
                       'income25-49K',
                       'casecontrolCase',
                       'ethnicityHispanic',
                       'raceMultiple',
                       'raceAsian',
                       'raceBlack',
                       'sexFemale',
                       'ageDecades',
                       '(Intercept)'))
```

```{r}
# Recode p value as significant, trend, n.s.
betas$p_cat <- 
     cut(betas$p,
         breaks = c(0, 0.05, 0.10, 1),
         labels = c('Significant',
                    'Trend',
                    'Not significant')) 
```

```{r}
# Plots in style of Yap et al
p.betas <- 
     betas |> 
     filter(covar != "(Intercept)") |> # Exclude intercept
     # Exclude Unknown covariate categories (missing data)
     filter(!grepl(pattern = 'Unknown', covar)) |> 
     ggplot(aes(x = covar, y = Estimate, color = p_cat)) +
     geom_hline(yintercept = 0,
                linetype = 'dotted',
                color = '#E15759') +
     geom_point(size = 2) +
     geom_linerange(aes(ymin = CI2.5, 
                       ymax = CI97.5,
                       color = p_cat),
                    size = 1) +
     scale_color_manual(values = c('#e03531',
                                   '#f0bd27',
                                   'gray40')) +
     labs(x = 'Covariate', y = 'Regression coefficient') +
     coord_flip() +
     theme(legend.position = 'none')

p.betas
```

### Visualize raw relationships

#### Age

```{r fig.width = 2.5, fig.height = 3}
p.age <- 
     ggplot(alpha.subset, aes(x = age_decades*10,
                         y = value)) +
     geom_point(alpha = 0.2) +
     geom_smooth(method = 'lm',
                 color = '#e03531') +
     scale_x_continuous(breaks = seq(10, 18, by = 2)) +
     labs(title = 'Age (years)', y = 'pMR') + 
     ylim(0, 35) +
     theme(axis.title.x =  element_blank())

p.age
```

#### Income categories

```{r}
# Clean up household income labels for plotting
levels(alpha.subset$ses)
```

```{r}
alpha.subset$ses <- 
     factor(alpha.subset$ses,
            labels = c('<$25K',
                       '$25-$49K',
                       '$50-$74K',
                       '$75K-$99K',
                       '>$100K',
                       'Unknown'))
```

```{r fig.height=3.5, fig.width=2.5}
# As dotplot
p.income <- 
     alpha.subset |> 
     filter(ses != 'Unknown') |> # Remove Unknown category
     ggplot(aes(x = ses, y = value)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Income', y = 'pMR')  +
     theme(axis.text.x = element_text(angle = 30, hjust = 1),
           axis.title = element_blank()) 

p.income
```

#### Case status

```{r}
alpha.subset$casecontrol <- 
     factor(alpha.subset$casecontrol,
            labels = c('Healthy\n(control)',
                       'Obesity\n(case)'))
```

```{r fig.height=4, fig.width=3}
p.case <- 
     alpha.subset |> 
     ggplot(aes(x = casecontrol, y = value)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Weight status', y = 'pMR') +
     theme(axis.title = element_blank())

p.case
```

#### Food insecurity

```{r fig.height=3, fig.width=2.5}
p.insec <- 
     alpha.subset |> 
     filter(actual_ever != 'Unknown') |> 
     ggplot(aes(x = actual_ever, y = value)) +
     geom_dotplot(binaxis = 'y',
                  binwidth = 1,
                  stackdir = 'center',
                  stackratio = 0.75,
                  alpha = 0.2) +
     stat_summary(fun= median,
                  geom = 'crossbar',
                  width = 0.25,
                  color = '#e03531') +
     stat_summary(fun= median,
                  fun.min = function(x){median(x) - mad(x)},
                  fun.max = function(x){median(x) + mad(x)},
                  geom = 'errorbar',
                  width = 0.1,
                  color = '#e03531') +
     ylim(0, 35) +
     labs(title = 'Food insecurity', y = 'pMR') +
     theme(axis.title = element_blank())

p.insec
```

Now, combine all of the above with patchwork:

```{r fig.height=5, fig.width=7.5}
(( p.pmr.hist ) | (p.betas)) /
(p.age | p.income | p.case | p.insec)
```

```{r}
# ggsave('Fig 3.pdf', device = cairo_pdf, height = 5, width = 7.5)
```
