---
title: "Figure S3"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r include=FALSE, echo=FALSE}
library(eulerr)
library(phyloseq)
library(readxl)
library(tidyverse)
```

```{r include=FALSE, echo=FALSE}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

```{r}
# trnL phyloseq
ps.mb <- readRDS('data/phyloseq_trnL.rds')

# Weight Loss menu phyloseq
ps.menu <- readRDS('data/phyloseq_WeightLoss_menu.rds')

# Mapping file to match taxon names between trnL, menu
name.sync <- read_excel('data/trnL to menu taxon mapping.xlsx')

# Controlled Feeding per-food performance
performance.cf <- read_csv('data/Controlled Feeding per-food performance.csv')
```
# Pre-process

```{r}
# Subset to trnL to Weight Loss cohort only
ps.mb <-
     ps.mb |> 
     subset_samples(study == 'Weight Loss') %>%
     prune_taxa(taxa_sums(.) > 0, .) %>%
     prune_samples(sample_sums(.) > 0, .)
```

```{r}
# Rename metabarcoding data to be consistent with menu
sample_names(ps.mb) <- 
     paste(sample_data(ps.mb)$subj,
           sample_data(ps.mb)$date,
           sep = '_')

head(sample_names(ps.menu))
head(sample_names(ps.mb))
```
```{r}
# Remove non-plant entries from menu
ps.menu <- subset_taxa(ps.menu,
                       kingdom == 'Viridiplantae')
ps.menu
```
```{r}
# Make a variable for 1-2 day lag menu from sample dates
sample_data(ps.mb)$delta1 <- 
     paste(sample_data(ps.mb)$subj,
           sample_data(ps.mb)$date - 1,
           sep = '_')
sample_data(ps.mb)$delta2 <- 
     paste(sample_data(ps.mb)$subj,
           sample_data(ps.mb)$date - 2,
           sep = '_')
```

```{r}
samdf.mb <- 
     ps.mb |> 
     sample_data() |> 
     data.frame() |> 
     rownames_to_column(var = 'row')
```

```{r}
# Merge 1-2 menu days prior and match to samples
# Note: Can't use phyloseq's merge_samples here because some samples make up >1 merge

# Do manually on ASV table
asvtab.menu <- otu_table(ps.menu)@.Data
asvtab.menu.12 <- data.frame()

for (sample in seq(dim(samdf.mb)[1])){
     # Pull menu day -1
     one.before <- asvtab.menu[samdf.mb$delta1[sample], ]
     # Pull menu day -2
     two.before <- asvtab.menu[samdf.mb$delta2[sample], ]
     # Merge and place in new, aggregated OTU table
     asvtab.menu.12 <- rbind(asvtab.menu.12,
                             one.before + two.before)
     # Update food names (only has to be done once)
     if (sample == 1){
          names(asvtab.menu.12) <- names(one.before + two.before)}
     # Update sample name
     row.names(asvtab.menu.12)[sample] <- samdf.mb$row[sample]
}

# Now rebuild a subsetted phyloseq object
ps.menu.12 <- phyloseq(otu_table(asvtab.menu.12, taxa_are_rows = FALSE),
                       sample_data(ps.mb), # Now this matches
                       tax_table(ps.menu)) # Menu-specific taxonomy

# Remove any taxa that aren't present any longer
ps.menu.12 <- prune_taxa(taxa_sums(ps.menu.12) > 0, ps.menu.12)
ps.menu.12
```
```{r}
# Helper function for getting lowest named taxonomic level
lowest_level <- function(taxtab){
     # Update taxa names from ASV sequence to identified taxon at the most
     # precise phylogenetic level possible

     # This gets the right-most, non-NA value
     lowest.index <- max.col(!is.na(taxtab), 'last')
     taxtab$name <- taxtab[cbind(seq_along(lowest.index),
                                         lowest.index)]

     taxtab
}
```

```{r}
# Match taxa names between the two datasets
taxtab.mb <- 
     ps.mb@tax_table@.Data |> 
     data.frame() |> 
     lowest_level()

# Rename with lowest-level name
taxa_names(ps.mb) <- tax_table(ps.mb)@.Data[, 'name']

# Link that to common name
tax_table(ps.mb) <- 
     name.sync |> 
     select(mb_taxa, common_name) |> 
     right_join(data.frame(ps.mb@tax_table@.Data),
               by = c('mb_taxa' = 'name')) |>
     distinct() |> 
     column_to_rownames(var = 'mb_taxa') |> 
     as.matrix()

# Rename with linked common name
taxa_names(ps.mb) <- tax_table(ps.mb)@.Data[, 'common_name']
```

```{r}
# Synchronize menu to metabarcoding names
# Add name column
tax_table(ps.menu.12) <- 
     lowest_level(data.frame(ps.menu.12@tax_table@.Data)) |> 
     as.matrix()

taxa_names(ps.menu.12) <- tax_table(ps.menu.12)@.Data[, 'name']

# Link that to common name
tax_table(ps.menu.12) <- 
     name.sync |> 
     select(menu_taxa, common_name) |> 
     left_join(x = data.frame(ps.menu.12@tax_table@.Data),
               by = c('name' = 'menu_taxa')) |> 
     select(common_name, everything()) |> 
     distinct() |> 
     column_to_rownames(var = 'name') |> 
     as.matrix()

# Renaming more complicated, see below
```

Some menu items now have the same label and should be merged. A straightforward merge on common name labels doesn't work using `tax_glom` because they differ at a higher rank.  Iterate over the tips instead, merging one at a time:

```{r}
# Get a list of these duplicates
dups <- 
     ps.menu.12@tax_table@.Data |> 
     data.frame() |> 
     filter(duplicated(common_name)) |> 
     pull(common_name) |> 
     unique()

dups
```
```{r}
ps.menu.12
merges = list()
for (taxon in dups){
     # Get the list of equivalent taxa at the species level for that food
     eq_taxa <- 
          ps.menu.12@tax_table@.Data |> 
          data.frame() |> 
          filter(common_name == taxon) |> 
          select(-common_name) |> 
          MButils::lowest_level() |> 
          pull(name)
     
     merges <- append(merges, list(eq_taxa))
}

if (length(merges) == 1){
     # Just do one merge
      ps.menu.id <- merge_taxa(ps.menu.12, merges[[1]])
} else {
     # Merge once to make new phyloseq
     ps.menu.id <- merge_taxa(ps.menu.12, merges[[1]])
     # Then loop over remainder
     for (i in 2:length(merges)){
          ps.menu.id <- merge_taxa(ps.menu.id, merges[[i]])
     }
}

taxa_names(ps.menu.id) <- tax_table(ps.menu.id)[, 'common_name']
ps.menu.id
```
# Analysis

## A

```{r}
# Plants
# Get union of taxa detected in the two analyses
mb.taxa <- taxa_names(ps.mb)
menu.taxa <- taxa_names(ps.menu.id)

all.plants <- 
     union(mb.taxa,
           menu.taxa) 

length(all.plants)
```

```{r}
# Make dataframe for plot input
# Plants
plant.df <- 
     data.frame(taxon = all.plants) |> 
     mutate(dna = taxon %in% mb.taxa,
            menu = taxon %in% menu.taxa)
```

```{r fig.height=3, fig.width=3}
# Version with labels for reference
euler(plant.df[,c('dna', 'menu')]) |> 
     plot(
          fills = c('#849db1', 'white'),
          labels = TRUE,
          shape = "ellipse", 
          quantities = TRUE
     )
```

## B

```{r}
# Extract OTU tables
asvtab.mb <- otu_table(ps.mb)@.Data
asvtab.menu.12 <- otu_table(ps.menu.id)@.Data

# Synchronize names
dim(asvtab.mb)
dim(asvtab.menu.12)
length(union(colnames(asvtab.menu.12), colnames(asvtab.mb)))
```

Want to make these tables the same same size (taking the union of their columns, and populating the new additions to each with zeros since the food wasn't detected in that dataset).
```{r}
# Helper function to pad out columns
pad_columns <- function(x, y, fill = 0){
     # Takes two matrices and synchronizes columns across them, filling added 
     # cols with a set value
     # Returns a list of the two updated matrices
     
     # Find missing columns in both matrices
     missing.x <- setdiff(colnames(y), colnames(x))
     missing.y <- setdiff(colnames(x), colnames(y))
     
     # Pad out columns of x
     fill.x <- matrix(fill, 
                      nrow = dim(x)[1],
                      ncol = length(missing.x))
     colnames(fill.x) <- missing.x 
     x <- cbind(x, fill.x)
     
     # Pad out columns of y
     fill.y <- matrix(fill, 
                      nrow = dim(y)[1],
                      ncol = length(missing.y))
     colnames(fill.y) <- missing.y
     y <- cbind(y, fill.y)
     
     # Arrange the columns so they appear in identical order
     u <- sort(union(colnames(x), colnames(y)))
     x <- x[, u]
     y <- y[, u]
     
     list(x, y)
}
```

```{r}
padded <- pad_columns(asvtab.mb, asvtab.menu.12)
asvtab.mb <- padded[[1]]
asvtab.menu.12 <- padded[[2]]

rm(padded)
```

```{r}
# Confirm row and column names are equal before proceeding
all(row.names(asvtab.mb) == row.names(asvtab.menu.12))
all(colnames(asvtab.mb) == colnames(asvtab.menu.12))
```

```{r}
# Label predictions 
mb.pos <- asvtab.mb > 0
menu.pos <- asvtab.menu.12 > 0

tp <- mb.pos & menu.pos
tn <- !mb.pos & !menu.pos
fp <- mb.pos & !menu.pos
fn <- !mb.pos & menu.pos
```

```{r}
# Summarize predictions
pred <- bind_rows(colSums(tp), 
                  colSums(tn), 
                  colSums(fp), 
                  colSums(fn))

pred$prediction <- c('tp', 'tn', 'fp', 'fn')
```

```{r}
# Reformat to long for visualization
pred.long <- pivot_longer(pred,
                          cols = -prediction,
                          names_to = 'common_name',
                          values_to = 'count')

# Factor data for visualization
# Calculate F-measure to be used downstream
# For this ranking, combine results from both cohorts
pred.summary <- 
     pred.long |> 
     group_by(prediction, common_name) |> 
     pivot_wider(names_from = prediction, values_from = count) |> 
     mutate(recall = tp/(tp + fn),
            precision = tp/(tp + fp),
            accuracy = (tp + tn)/(tp + tn + fp + fn),
            f_measure = (2 * recall * precision)/(recall + precision)) |> 
     arrange(desc(f_measure), fn, fp)

pred.summary
```

```{r}
# Now that joins complete, add factor levels
pred.long <- 
     pred.long |> 
     mutate(
          common_name = factor(common_name, 
                               levels = pred.summary$common_name),
          prediction = factor(prediction, levels = c('tp',
                                                     'tn',
                                                     'fp',
                                                     'fn'),
                              labels = c('True positive',
                                         'True negative',
                                         'False positive',
                                         'False negative')))
```

```{r}
# Add facets to go across multiple panels
pred.long <- 
     pred.long |> 
     arrange(common_name) |> 
     mutate(facet = sort(rep(1:4, 96))) # Tall: 4 rows
```

```{r fig.height=12, fig.width=6}
pred.long |>
     ggplot(aes(x = common_name, y = count, fill = prediction)) +
     geom_bar(position = 'stack', stat = 'identity') +
     scale_fill_manual(values = c('#309343', # TP
                                  '#8AC373', # TN
                                  '#D7191C', # FP
                                  '#FDAE61')) + # FN
     facet_wrap(~facet,
                nrow = 4,
                scales = 'free_x') +
     theme_classic() +
     scale_y_continuous(breaks = c(0, 10, 20, 30, 40)) +
     theme(axis.text.x = ggtext::element_markdown(angle = 45, 
                                                  hjust = 1,
                                                  size = 12,
                                                  face = 'bold'),
           axis.text.y = element_text(size = 12,
                                      face = 'bold'),
           axis.ticks.x = element_blank(),
           axis.line.x = element_blank(),
           axis.title = element_blank(),
           legend.position = 'none',
           legend.title = element_blank(), 
           plot.margin = margin(0, 0, 0, 0.75, 'in'),
           strip.background = element_blank(),
           strip.text = element_blank()) 
```
```{r}
# ggsave('Fig S3b.pdf'), height = 18, width = 9)
```

## C

Compare to performance in Controlled Feeding cohort
```{r}
pred.summary.wl <- pred.summary
pred.summary.cf <- performance.cf
```

```{r}
# Need to sync common name labels between these
# Happen due to slight differences in detected taxa or manual naming in each cohort

# Updates to Weight Loss
pred.summary.wl$common_name[pred.summary.wl$common_name == 'grains'] <- 
     'wheat'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'scallion, chive, leek'] <-
     'scallion'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'citrus family'] <- 
     'citrus'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'oats'] <- 
     'oat'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'potato, tomatillo'] <- 
     'potato'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'peas'] <- 
     'pea'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'apple, pear'] <- 
     'apple'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'chard, beet'] <- 
     'sugar beet'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'laurel family'] <- 
     'cinnamon'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'mustard family'] <- 
     'mustards, rapeseed'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'onion, shallot'] <- 
     'onion'
pred.summary.wl$common_name[pred.summary.wl$common_name == 'stone fruits'] <- 
     'peach'

# Change ABC
pred.summary.cf$common_name[pred.summary.cf$common_name == 'herbs'] <- 
     'oregano, thyme'

pred.summary.cf$common_name[pred.summary.cf$common_name == 'herbs'] <- 
     'oregano, thyme'
```

```{r}
pred.summary.cf$study <- 'Controlled Feeding'
pred.summary.wl$study <- 'Weight Loss'
```

```{r}
# Simplify: all other metrics can be recalculated from TP, TN, FP, FN
pred.summary.cf <- 
     pred.summary.cf |> 
     select(study, common_name, tp:fn)

pred.summary.wl <- 
     pred.summary.wl |> 
     select(study, common_name, tp:fn)
```

```{r}
# Join together
pred.summary <- 
     bind_rows(
          pred.summary.cf,
          pred.summary.wl
          )

rm(pred.summary.wl, pred.summary.cf)
```

```{r}
# Catch naming differences and correct:
pred.summary |> 
     group_by(common_name) |> 
     filter(length(common_name) == 1)
```
Screened these and added above. 

```{r}
# Filter to only taxa within the two
n_distinct(pred.summary$common_name)
pred.summary <- 
     pred.summary |> 
     group_by(common_name) |> 
     filter(length(common_name) == 2) |> 
     ungroup()

n_distinct(pred.summary$common_name)
```

```{r}
# Calculate FPR, FNR
pred.summary <- 
     pred.summary %>% 
     mutate(fnr = fn/(tp+fn),
            fpr = fp/(tn+fp))
```

```{r}
# Organize labels for plot
pred.metrics <- 
     pred.summary %>% 
     select(-(tp:fn)) %>% 
     pivot_longer(cols = c(fpr, fnr),
                  names_to = 'metric',
                  values_to = 'value') 

pred.metrics
```
```{r fig.height=5, fig.width=3}
pred.metrics %>% 
     pivot_wider(names_from = 'study',
                 values_from = 'value') %>% 
     mutate(metric = factor(metric,
                            levels = c('fpr', 'fnr'),
                            labels = c('FPR', 'FNR'))) %>% 
     ggplot(aes(x = `Weight Loss`, y = `Controlled Feeding`)) +
     geom_point(size = 3, 
                alpha = 0.7) +
     scale_x_continuous(breaks = seq(0, 1, 0.25),
                        labels = c(0, 0.25, 0.5, 0.75, 1)) +
     scale_y_continuous(breaks = seq(0, 1, 0.25),
                        labels = c(0, 0.25, 0.5, 0.75, 1)) +
     facet_wrap(~metric,
                nrow = 2)
```
```{r}
# ggsave('Fig S3c.pdf', height=5, width=3, device= cairo_pdf)
```

```{r}
pred.metrics %>% 
     pivot_wider(names_from = 'study',
                 values_from = 'value') %>% 
     group_by(metric) %>% 
     summarize(
          rho = cor.test(~ `Controlled Feeding` + `Weight Loss`,
                         method = 'spearman',
                         exact = FALSE)$estimate,
          p = cor.test(~ `Controlled Feeding` + `Weight Loss`,
                       method = 'spearman',
                       exact = FALSE)$p.value)
```
