---
title: "Figure 3"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

# Read in data

```{r}
ps = readRDS("../data/raw/20210612_updated_ps.rds")

sample_names(ps) %>%
  length()

sample_data(ps)$subj = as.double(sample_data(ps)$subj)
```

```{r}
data = read_csv("../data/metadata/20210805_pomms_clean_metadata_estimate.csv")

data
```

# Merging metadata and phyloseq object

```{r}
row_df = as_tibble(sample_data(ps), rownames) %>%
  select(R24_id, subj, visit)

full_data = left_join(row_df, data) %>%
  arrange(subj)
  
```

```{r}
sam = as_tibble(sample_data(ps), rownames= "row_names") %>%
    select(row_names, everything())

complete_data = left_join(sam, full_data) %>%
    column_to_rownames("row_names")

sample_data(ps) = sample_data(complete_data)

#sample_data(ps)
```
```{r}
rm(sam)
rm(complete_data)
```

# Alpha Diversity

```{r}
ps_foods = subset_taxa(ps, !is.na(kingdom))
#ps_foods = ps

```

```{r}
# Update read counts in phyloseq object
sample_data(ps_foods)$reads = sample_sums(ps_foods)
```

```{r}
# What percentage of reads does this retain (overall)?
sum(sample_sums(ps_foods))/sum(sample_sums(ps))
```

```{r}
# What percentage of reads does this retain (per sample)?
(sample_sums(ps_foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     ggplot(., aes(x = percentage)) +
     geom_histogram(binwidth = 0.05, boundary = 0) +
     theme_bw()
```
Idea here would be to follow up on samples that have <90% of their reads preserved after removing those not identified to plant.


```{r}
(sample_sums(ps_foods)/sample_sums(ps)) %>% 
     data.frame(percentage = . ) %>% 
     filter(is.na(percentage) | percentage < 0.9)
```

## Read counts
```{r}
# Remove samples that do not have any food reads (8 in all)
ps_foods = 
     ps_foods %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)


ps_foods
```

## Estimate alpha diversity:

```{r}
alpha = estimate_richness(ps_foods, measures = c('Observed', 'Shannon')) %>%
     rownames_to_column(var = 'name') %>%
  separate(name, into= c("x", "name"), sep = "X") %>%
  select(!x) %>%
  rename_with(tolower)
  

alpha
```

```{r}
# Join to read counts
samdf = 
     data.frame(ps_foods@sam_data) %>% 
     rownames_to_column(var = 'name')

alpha = 
     samdf %>% 
     #select(name, reads, subj) %>% 
     right_join(alpha)

alpha
```

```{r}
alpha = alpha %>%
  group_by(subj) %>%
  mutate(samples= n(),
         mean_shannon = mean(shannon),
         mean_observed = mean(observed))%>%
  ungroup() %>%
  rename_with(tolower)  %>%
  mutate(paired = rep(1:(n()/2),each=2),
         visit=factor(visit)) 

alpha
```

```{r}
# Paired samples
paired = alpha %>%
  filter(samples == 2) %>%
  arrange(subj, visit)

paired

#266 rows total

266/2

#133 paired samples
```

```{r}
#Glancing at change in alpha diversity by visit

alpha %>%
  filter(samples == 2) %>%
  arrange(subj, visit) %>%
  select(subj, visit, observed, shannon)

```

```{r}
alpha_long = pivot_longer(alpha, cols = c(observed, shannon,mean_observed, mean_shannon),
                           names_to = 'measure', values_to = 'value')
alpha_long$visit =  as.character(alpha_long$visit)

alpha_long
```

# Change in alpha diversity

## How does alpha diversity change between visits, seperated by intervention group:
```{r}
# This is conducted considering only subjects that had two samples: i.e. pairs.

alpha_long %>%
  #filter(is.na(Food.Insecurity.Scrn.Complete) == FALSE) %>%
  filter(samples == 2) %>%
  #select(subj, Visit.Age, Samples, visit, Food.Insecurity.Scrn.Complete, Food1, Food_run_out, Food2, Food_ran_out, Observed, Shannon, Mean_Observed,Mean_Shannon) %>%
  #pivot_longer(Observed:Mean_Shannon, names_to = "Measure", values_to = "Value") %>%
  filter(measure == "observed" | measure == "shannon") %>%
  ggplot(aes(x= group, y = value, colour= visit))+
  geom_jitter(alpha = 0.65, size=1, width= 0.2)+
  geom_boxplot(alpha = 0.5)+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust= 1))
  

ggsave("change_alpha_diversity_group.png", width = 10)
```

## Were subjects who increased in alpha diversity the same as those who lost weight?

```{r}
# more_div = alpha %>%
#   select(subj, visit, Delta_BMI_P95, Observed, Shannon) %>%
#   pivot_wider(names_from = visit, values_from= c(Observed, Shannon))%>%
#   filter(Observed_5 > Observed_1) %>%
#   pull(subj)
# 
# less_weight = alpha %>%
#   select(subj, visit, Delta_BMI_P95, Observed, Shannon) %>%
#   pivot_wider(names_from = visit, values_from= c(Observed, Shannon))%>%
#   filter(Delta_BMI_P95 < 0) %>%
#   pull(subj)
#   
# (intersect(less_weight, more_div) %>%
#   length())/(more_div %>%
#   length())*100
# 
# (intersect(less_weight, more_div) %>%
#   length())/(less_weight %>%
#   length())*100

```
No, more than half of those who lost weight are not the same as those who increased in diversity.

# By Read Depth

```{r}
ggplot(alpha_long, aes(x = reads, y = value)) +
     geom_point() +
     labs(x = 'Read depth', y = 'Alpha diversity') +
     facet_wrap(~measure, scales = 'free_y') +
     theme_bw()
```

# Plots

## Treating all points as separate
```{r}
alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  ggplot(aes(x= age, y = value, group = age))+
  geom_boxplot()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()
```
```{r}
alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  ggplot(aes(x= age_days, y = value))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()
```

## Considering averages

```{r}
alpha_long %>%
  filter(measure == "mean_observed" | measure == "mean_shannon") %>%
  ggplot(aes(x= age, y = value, group = age))+
  geom_boxplot()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()

#ggsave("figures/alpha_age_average_boxplot.png", width = 9)

alpha_long %>%
  filter(measure == "mean_observed" | measure == "mean_shannon") %>%
  ggplot(aes(x= age_days, y = value))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()

#ggsave("figures/alpha_age_average_scatter.png", width = 9)

```
## Individual points separated by visit

```{r}
alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  filter(visit == 1) %>%
  ggplot(aes(x= age, y = value, group = age))+
  geom_boxplot()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()+
  labs(title = "Visit 1")

#ggsave("figures/alpha_age_visit1_boxplot.png", width = 9)


alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  filter(visit == 5) %>%
  ggplot(aes(x= age, y = value, group = age))+
  geom_boxplot()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()+
  labs(title = "Visit 5")

#ggsave("figures/alpha_age_visit5_boxplot.png", width = 9)


```

```{r}
alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  filter(visit == 1) %>%
  ggplot(aes(x= age_days, y = value))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()+
  labs(title = "Visit 1")

#ggsave("figures/alpha_age_visit1_scatter.png", width = 9)

alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
  filter(visit == 5) %>%
  ggplot(aes(x= age_days, y = value))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()+
  labs(title = "Visit 5")

#ggsave("figures/alpha_age_visit5_scatter.png", width = 9)

```

## Different plots:

```{r}
alpha_long%>%
  filter(measure == "observed" | measure == "shannon") %>%
  #filter(group == "Lifestyle") %>%
  ggplot(aes(x= visit, y = value, colour = group))+
  geom_jitter() +
  geom_line(aes(group=paired)) +
  facet_wrap(~measure, scales = "free_y")+
  theme_bw()
  #theme(legend.position = "none")
```


# Correlation testing

--> Try fitting a linear regression too maybe?

## By Age_Days:

```{r}
alpha$age_days %>%
  unique()
```
### V1 subset
```{r}
v1= alpha %>%
  filter(visit == 1)
```

#### Spearman

```{r}
cor.test(x= v1$age_days, y= v1$observed, method= "spearman", exact = FALSE) %>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "observed")
```

```{r}
cor.test(x= v1$age_days, y= v1$shannon, method= "spearman", exact = FALSE) %>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "shannon")

```

#### Pearson
```{r}
cor.test(x= v1$age_days, y= v1$observed, method= "pearson", exact = FALSE) %>%
  glance()%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "observed")
```

```{r}
cor.test(x= v1$age_days, y= v1$shannon, method= "pearson", exact = FALSE)%>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "shannon")

```
#### regression

```{r}
glance(lm(v1$observed ~ v1$age_days))
```
```{r}
glance(lm(v1$shannon ~ v1$age_days))
```


### V5 subset

```{r}
v5= alpha %>%
  filter(visit == 5)
```

#### Spearman
```{r}
cor.test(x= v5$age_days, y= v5$observed, method= "spearman", exact = FALSE) %>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "observed")
```

```{r}
cor.test(x= v5$age_days, y= v5$shannon, method= "spearman", exact = FALSE) %>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "shannon")

```
#### Pearson


```{r}
cor.test(x= v5$age_days, y= v5$observed, method= "pearson", exact = FALSE) %>%
  glance()%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "observed")
```

```{r}
cor.test(x= v5$age_days, y= v5$shannon, method= "pearson", exact = FALSE)%>%
  glance() %>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "shannon")

```
#### regression

```{r}
glance(lm(v5$observed ~ v5$age_days))
```

```{r}
glance(lm(v5$shannon ~ v5$age_days))
```

### By averages 

#### Spearman

```{r}
glance(cor.test(x = alpha$age_days, y = alpha$mean_observed, method = "spearman", exact = FALSE))%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "mean_observed")
```

```{r}
glance(cor.test(x = alpha$age_days, y = alpha$mean_shannon, method = "spearman", exact = FALSE))%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "mean_shannon")
```
#### Pearson

```{r}
glance(cor.test(x = alpha$age_days, y = alpha$mean_observed, method = "pearson"))%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "mean_observed")
  
```

```{r}
glance(cor.test(x = alpha$age_days, y = alpha$mean_shannon, method = "pearson"))%>%
  select(estimate, statistic, p.value, method) %>%
  mutate(x = "age_days", y = "mean_shannon")

```
#### regression

```{r}
glance(lm(alpha$mean_observed ~ alpha$age_days))
```

```{r}
glance(lm(alpha$mean_shannon ~ alpha$age_days))
```


## Alpha diversity grant figure:

```{r}

annotation = data.frame(value = c("Observed", "Shannon"), label = c("rho = -0.1, p = 0.054", "rho = -0.15, p = 0.004"))


 alpha_long%>%
  filter(measure == "observed" | measure == "shannon") %>%
ggplot( aes(x = age_days, y = value)) +
     geom_point(color= "grey28", size= 2.5, alpha= 0.5) +
  #geom_smooth(method = "lm")+
     labs(x = expression(paste("Age in days")), y = expression(paste("trnl ", alpha, " diversity"))) +
     facet_wrap(~measure, scales = 'free_y')+
    #geom_text(x = 4500, y = 0, aes(label = label), data = annotation)+
     theme_bw()+
  theme(axis.title.y = element_text(colour= "grey41", face= "bold", size = 20),
        axis.title.x = element_text(colour= "grey41", face= "bold", size = 20),
        strip.text.x = element_text(size = 15, face = "bold"),
        axis.text.x = element_text(face="bold", colour = "black"),
        axis.text.y = element_text(face="bold", colour = "black"))

#ggsave("C:/Users/Ammara Aqeel/Box/project_davidlab/LAD_LAB_Personnel/Ammara_A/Grants/Brianna_analysis/grant_alpha_diversity_figure.png", width= 6, height = 4)
```

```{r}
alpha_long %>%
  filter(measure == "observed" | measure == "shannon") %>%
ggplot(aes(x = age, y = value, group = age)) +
     geom_boxplot(color= "grey28", alpha= 0.5, size= 1.05) +
     labs(x = expression(paste("Age")), y = expression(paste("trnl ", alpha, " diversity"))) +
     facet_wrap(~measure, scales = 'free_y')+
     theme_bw()+
  theme(axis.title.y = element_text(colour= "grey41", face= "bold", size = 20),
        axis.title.x = element_text(colour= "grey41", face= "bold", size = 20),
        strip.text.x = element_text(size = 15, face = "bold"),
        axis.text.x = element_text(face="bold", colour = "black", size= 15),
        axis.text.y = element_text(face="bold", colour = "black", size = 15))

#ggsave("C:/Users/Ammara Aqeel/Box/project_davidlab/LAD_LAB_Personnel/Ammara_A/Grants/Brianna_analysis/grant_alpha_diversity_boxplot.png", width= 8, height = 5)
```
# Rough Stuff, don't run
## By Age:

```{r}
alpha$age %>%
  unique()
```

```{r}
glance(kruskal.test(age ~ observed , data = alpha))%>%
  mutate(x = "age", y = "observed")
  
```

```{r}
glance(kruskal.test(age ~ shannon , data = alpha)) %>%
  mutate(x = "age", y = "shannon")
```

