---
title: "Joint phyloseq QC"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
# library(iNEXT)
library(phyloseq)
library(scales) # For label = comma
library(tidyverse)
library(vegan)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    legend.text = element_text(size = 11))
)
```

# Read in data

```{r}
# Load last saved phyloseq
ps <-
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>%
     readRDS()

ps
```

# Summary analyses

```{r}
# These will require sample data
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row')
```

## Success rate

Samples with negative Qubit readings:
```{r}
samdf %>% 
     filter(qubit < 0) %>% 
     select(study, subj, qubit, reads_tot) %>% 
     arrange(desc(reads_tot))
```

```{r}
samdf %>% 
     filter(reads_tot < 1000) %>% 
     select(study, subj, qubit, reads_tot) %>% 
     arrange(desc(reads_tot))
```
```{r}
# Joint condition
failed <- 
     samdf %>% 
     filter(qubit < 0 | reads_tot < 1000) %>% 
     group_by(study) %>% 
     count(name = 'n_failed')

failed
```

```{r}
# Calculate failure rate by study
failed <- 
     samdf %>% 
     group_by(study) %>% 
     count(name = 'n_total') %>% 
     left_join(failed) %>% 
     mutate(n_failed = ifelse(is.na(n_failed),
                                    yes = 0,
                                    no = n_failed),
            n_success = n_total - n_failed,
            pct = n_success/n_total)

failed
```

```{r}
# Overall failure rate
sum(failed$n_failed)/sum(failed$n_total)
```

```{r}
# Stats on study-specific failure rate
chisq <- 
     failed %>% 
     select(-n_total) %>% 
     column_to_rownames(var = 'study') %>% 
     as.matrix() %>% 
     chisq.test(x = .)

chisq
```

Remove these failed samples
Unclear how this will affect Qubit results below, so creating a separate object

```{r}
ps %>% 
     subset_samples(qubit > 0 & reads_tot > 1000) %>% 
     saveRDS(here('data',
                  'processed',
                  'phyloseq',
                  'trnL_combined_ps_filter.rds'))
```

## Descriptive stats
```{r}
# Number of subjects
samdf %>% 
     group_by(study) %>% 
     summarize(n = n_distinct(subj))
```
```{r}
# Total subjects
n_distinct(samdf$subj)
```

```{r}
# Samples per subject (mean and standard deviation)
samdf %>% 
     group_by(study, subj) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(study) %>% 
     summarize(mean = mean(n), sd = sd(n),
               med = median(n), mad = mad(n))
```
Median and median absolute deviations are better measures when data is non-normal:

```{r}
samdf %>% 
     group_by(study, subj) %>% 
     count() %>% 
     ungroup() %>% 
     ggplot(aes(x = n)) +
     geom_histogram(binwidth = 1,
                    boundary = 0) +
     facet_wrap(~study,
                scales = 'free_y')
```
Ok, so think best to report median and median absolute deviation.

## Taxa

### Number

```{r}
# Total
ntaxa(ps)
```

```{r}
# Per subject
# Get each subject's mean alpha diversity
alpha <-
     ps %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     otu_table() %>% 
     data.frame() %>% 
     apply(1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'row') %>% 
     left_join(samdf)

# alpha <- 
#      alpha %>% 
#      group_by(study, subj) %>% 
#      summarize(obs_mean = mean(obs))
```

```{r}
median(alpha$obs)
hist(alpha$obs)
```


### Most common

```{r}
asvtab <- 
     ps@otu_table@.Data %>% 
     data.frame()
```

```{r}
# Consumed in more than one individual
more_than_one <- 
     apply(asvtab, 2, function(x){sum(x > 0) > 1})

sum(more_than_one)
```

```{r}
# Consumed in >50% of individuals
multiple <- 
     apply(asvtab, 2, function(x){sum(x > 0) > (0.5*length(x))})

sum(multiple)
```

```{r}
# What are these taxa?
taxtab <- 
     ps@tax_table@.Data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'asv')

taxtab %>% 
     filter(asv %in% names(which(multiple)))
```

```{r}
# How many samples have chocolate?
choc <- 
     prune_taxa('ATCCTATTATTTTATTATTTTACGAAACTAAACAAAGGTTCAGCAAGCGAGAATAATAAAAAAAG', ps) %>% 
     sample_sums()

sum(choc > 0)/(1001-24) # 24 samples w/no reads (failed)
```

```{r}
# How many people have chocolate?
choc <- 
     ps %>% 
     psmelt %>% 
     filter(OTU == 'ATCCTATTATTTTATTATTTTACGAAACTAAACAAAGGTTCAGCAAGCGAGAATAATAAAAAAAG') %>% 
     group_by(subj, OTU) %>% 
     summarize(detected = any(Abundance > 0))

table(choc$detected)
```

### Reads mapped

```{r}
# % reads mapping to food
sum(samdf$reads_food)/sum(samdf$reads_tot)
```

```{r}
# % unmapped
1 - sum(samdf$reads_food)/sum(samdf$reads_tot)
```

```{r}
# Range across samples
# Include only those with >1,000 reads as otherwise this gets funky
# (e.g. 2 reads but one not identified)
samdf %>% 
     select(row, starts_with('reads')) %>% 
     filter(reads_tot >= 1000) %>% 
     mutate(pct_no_id = 1 - (reads_food/reads_tot)) %>% 
     arrange(desc(pct_no_id))
```

```{r}
samdf %>% 
     select(row, starts_with('reads')) %>% 
     filter(reads_tot >= 1000) %>% 
     mutate(pct_no_id = 1 - (reads_food/reads_tot)) %>% 
     pull(pct_no_id) %>% 
     range()
```

Ooh-- definitely need to follow up on this.

```{r}
samdf %>% 
     select(study, row, starts_with('reads')) %>% 
     # filter(reads_tot >= 1000) %>% 
     mutate(pct_no_id = 1 - (reads_food/reads_tot)) %>% 
     # Color those higher than dataset average 
     ggplot(aes(x = reads_tot, y = reads_food)) +
     geom_abline(slope = 1, 
                 intercept = 0, 
                 color = '#e15759',
                 linetype = 'dashed') +
     geom_point(alpha = 0.5) +
     coord_equal()+
     scale_x_continuous(label = comma) + scale_y_continuous(label = comma) +
     # scale_x_log10(label = comma) + scale_y_log10(label = comma) +
     facet_wrap(~study, nrow = 1) +
     labs(y = 'Mapped reads', x = 'Total reads') +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Unassigned reads by study (linear scale).pdf')),
       device = cairo_pdf,
       height = 3, width = 6)
```

```{r}
samdf %>% 
     select(study, row, starts_with('reads')) %>% 
     mutate(pct_no_id = 1 - (reads_food/reads_tot)) %>% 
     summarize(median = median(pct_no_id,
                               na.rm = TRUE),
               mad = mad(pct_no_id,
                         na.rm = TRUE))
```


## Read count threshold

### Distribution

```{r}
# Faceted by study
ggplot(samdf, aes(x = reads_tot, fill = study)) +
     geom_histogram(bins = 10, boundary = 0) +
     facet_wrap(~study, scales = 'free_y') +
     scale_fill_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     # xlim(0, 50000) +
     labs(x = 'Read count', y = 'Samples (n)') +
     theme_bw() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1),
           legend.title = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Read depth of samples by study, facet.pdf')),
#        height = 4, width = 6)
```

```{r}
# Range of values in each study
samdf %>% 
     group_by(study) %>% 
     summarize(min = min(reads_tot),
               max = max(reads_tot))
```

```{r}
# What about after removing zeros?
samdf %>% 
     filter(reads_food > 0) %>% 
     group_by(study) %>% 
     summarize(min = min(reads_tot),
               max = max(reads_tot),
               mean = mean(reads_tot),
               sd = sd(reads_tot))
```

### Relationship to input

#### DNA added

```{r}
ggplot(samdf, aes(x = ng_added, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date, 
                nrow = 2) +
     labs(y = 'Reads', x = 'DNA added to pool (ng)') + 
     scale_y_continuous(label = comma) +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Read depth v ng added.pdf')),
       height = 4, width = 6)
```

```{r}
# Concerning PCR batch is #1 from 20210322: examine this more closely
samdf %>% 
     filter(seq_date == '20210322') %>% 
     mutate(pcr_batch = factor(pcr_batch)) %>% 
     ggplot(aes(x = ng_added, y = reads_tot, color = pcr_batch)) +
     geom_point(alpha = 0.8) +
     labs(y = 'Reads', x = 'DNA added to pool (ng)', color = 'Batch') + 
     scale_y_continuous(label = comma) +
     theme_bw()
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_20210322 Batch 1.pdf')),
       height = 3, width = 4)
```

```{r}
ggplot(samdf, aes(x = qubit, y = ng_added, color = study)) +
     geom_point(alpha = 0.5) +
     scale_color_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     facet_wrap(~seq_date, nrow = 2) +
          labs(x = 'Qubit measurement',
          y = 'DNA added to pool (ng)') +
     theme(legend.position = 'bottom',
           legend.title = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_DNA added v. Qubit.pdf')),
       height = 4, width = 5)
```

```{r}
ggplot(samdf, aes(x = qubit, y = ng_added, color = reads_tot > 10000)) +
     geom_point(alpha = 0.2) +
     scale_color_manual(values = c('red', 'gray20')) +
     facet_wrap(~seq_date, nrow = 2) +
     labs(x = 'Qubit measurement',
          y = 'DNA added to pool (ng)',
          color = '> 10,000 reads') +
     theme(legend.position = 'bottom')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_DNA added v. Qubit.pdf')),
       height = 4, width = 5)
```

#### Qubit

Histogram of Qubit values

```{r}
ggplot(samdf, aes(x = qubit, fill = study)) +
     geom_histogram(stat = 'bin',
                    bins = 10,
                    boundary = 0) +
     scale_fill_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     facet_wrap(~study, 
                nrow = 2,
                scales = 'free_y') +
     labs(y = 'Samples (n)', x = 'Qubit reading') + 
     theme_bw() +
     theme(legend.title = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Qubit histogram by study.pdf')),
#        height = 4, width = 6)
```

```{r}
ggplot(samdf, aes(x = qubit, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date, nrow = 2) +
     labs(y = 'Reads', x = 'Qubit reading') + 
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Read depth v Qubit.pdf')),
#        height = 4, width = 6)
```

Add custom threshold line based on "ng added" above-- above what measure where we able to max out?

```{r}
qubit.threshold <- 
     samdf %>% 
     group_by(seq_date) %>% 
     arrange(desc(ng_added), qubit) %>% # Lowest Qubit at which max added
     summarize(threshold = first(qubit))
```

```{r}
# Add this to sample data, make new over/under threshold variable
samdf <- 
     samdf %>% 
     left_join(qubit.threshold) %>% 
     mutate(over_threshold = qubit >= threshold)
```

```{r}
ggplot(samdf, aes(x = qubit, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     geom_vline(aes(xintercept = threshold), 
                color = 'red',
                linetype = 'dashed') +
     facet_wrap(~seq_date, nrow = 2) +
     labs(y = 'Reads', x = 'Qubit reading') + 
     scale_y_continuous(label = comma)
```
```{r}
# Make custom batch variable and curate for plotting
samdf.batch <- 
     samdf %>% 
     filter(!over_threshold) %>%
     mutate(
          # Missing data for DFC
          pcr_batch = ifelse(seq_date == '20191206',
                             yes = '1',
                             no = pcr_batch),
          # Make combined batch name
          batch = paste(seq_date, pcr_batch, sep = '_'))

# How many samples in each?
samdf.batch %>% 
     group_by(batch) %>% 
     count()
```

```{r}
format_x <- function(x){
     format(round(x, 1), nsmall=1)
}

# Maybe try only those with >10 samples?
samdf %>% 
     filter(!over_threshold) %>% 
     mutate(
          # Missing data for DFC
          pcr_batch = ifelse(seq_date == '20191206',
                             yes = '1',
                             no = pcr_batch),
          seq_date = factor(seq_date, 
                              labels = 0:7)) %>% 
     group_by(seq_date) %>% 
     mutate(size = length(seq_date),
            pcr_batch = factor(pcr_batch)) %>% 
     filter(size >= 10) %>%
     ggplot(aes(x = qubit, y = reads_tot)) +
     geom_smooth(method = lm,
                 aes(color = seq_date != 6)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date, nrow = 2,
                scales = 'free') +
     labs(y = 'Reads', x = 'Post-PCR\nconcentration (ng/ul)',
          color = 'Significant\nlinear fit') + 
     scale_color_manual(values = c('#e15759', '#59a14f')) +
     scale_y_continuous(labels = comma) +
     scale_x_continuous(labels = format_x) +
     theme(axis.text.x = element_text(size = 8),
           axis.text.y = element_text(size = 8),
           legend.position = c(1,0),
           legend.justification = c(1,0))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Read depth v Qubit below pooling threshold.pdf')),
       device = cairo_pdf,
       height = 4, width = 8)
```

```{r}
# Maybe try only those with >10 samples?
samdf %>% 
     mutate(seq_date = factor(seq_date, 
                              labels = 1:8)) %>% 
     group_by(seq_date) %>% 
     mutate(size = length(seq_date),
            pcr_batch = factor(pcr_batch)) %>% 
     # filter(size >= 10) %>% 
     ggplot(aes(x = qubit, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date, nrow = 2,
                scales = 'free') +
     labs(y = 'Reads', x = 'Qubit reading') + 
     scale_y_continuous(labels = comma) +
     scale_x_continuous(labels = format_x) +
     theme(axis.text.x = element_text(size = 8),
           axis.text.y = element_text(size = 8))
```

```{r}
samdf %>% 
     filter(!over_threshold) %>% 
     group_by(seq_date) %>% 
     summarize(
          rho = cor.test(~qubit + reads_tot)$estimate,
          p = cor.test(~qubit + reads_tot)$p.value
     ) %>% 
     mutate(p_bh = p.adjust(p, method = 'BH'))
```
```{r}
samdf %>% 
     filter(over_threshold) %>% 
     group_by(seq_date) %>% 
     summarize(
          rho = cor.test(~qubit + reads_tot)$estimate,
          p = cor.test(~qubit + reads_tot)$p.value
     ) %>% 
     mutate(p_bh = p.adjust(p, method = 'BH'))
```

```{r}
# What about with a linear model?
results.df <- 
     data.frame(seq_date = NULL,
                threshold = NULL,
                r2 = NULL,
                intercept_b = NULL,
                intercept_p = NULL,
                qubit_b = NULL,
                qubit_p = NULL)

for (d in unique(samdf$seq_date)){
     # Model: above threshold
     above <- 
          samdf %>% 
          filter(over_threshold) %>% 
          filter(seq_date == d) %>% 
          lm(reads_tot ~ qubit, data = .)
     
     # Model: below threshold
     below <- 
          samdf %>% 
          filter(!over_threshold) %>% 
          filter(seq_date == d) %>% 
          lm(reads_tot ~ qubit, data = .)
     
     above.df <- 
          data.frame(seq_date = d,
                     threshold = 'Above',
                     r2 = summary(above)$r.squared,
                     intercept_b = summary(above)$coefficients[1,1],
                     intercept_p = summary(above)$coefficients[1,4],
                     qubit_b = summary(above)$coefficients[2,1],
                     qubit_p = summary(above)$coefficients[2,4])
     
     below.df <- 
          data.frame(seq_date = d,
                     threshold = 'Below',
                     r2 = summary(below)$r.squared,
                     intercept_b = summary(below)$coefficients[1,1],
                     intercept_p = summary(below)$coefficients[1,4],
                     qubit_b = summary(below)$coefficients[2,1],
                     qubit_p = summary(below)$coefficients[2,4])
     
     results.df <- 
          bind_rows(results.df,
                    above.df,
                    below.df)
}
```

```{r}
# Recode significance level
results.df <- 
     results.df %>% 
     mutate(across(.cols = ends_with('_p'),
                   .fns = ~ifelse(.x <= 0.05,
                                   yes = TRUE,
                                   no = FALSE)))

# Pivot longer
results.df <-
     results.df %>%
     pivot_longer(cols = c('intercept_b',
                           'qubit_b'),
                  names_to = 'term',
                  values_to = 'coeff') %>% 
     mutate(term = gsub(term, 
                        pattern = '_b',
                        replacement = ''),
            signif = ifelse(term == 'intercept',
                            yes = intercept_p,
                            no = qubit_p)) %>% 
     select(-ends_with('_p'))
```

```{r}
# Visualize
results.df %>% 
     filter(signif) %>% 
ggplot(aes(x = seq_date, y = coeff, fill = threshold)) +
     geom_col(position = 'dodge') +
     facet_wrap(~term) +
     labs(x = 'Sequencing batch',
          y = 'Coefficient value') +
     theme(axis.text.x = element_text(angle = 45,
                                      hjust = 1))
```

```{r}
ggplot(results.df, aes(x = seq_date, y = r2, fill = threshold)) +
     geom_col(position = 'dodge') +
     labs(x = 'Sequencing batch',
          y = 'Coefficient value') +
     theme(axis.text.x = element_text(angle = 45,
                                      hjust = 1))
```

### Relationship to diet surveys

For Adult-1 and Adult-2, check against FVS, hPDI scores

```{r}
# Read in this data
ffq <- 
     here('data', 
          'processed', 
          'DHQ3',
          'Adult-1 and Adult-2 dietary indices.csv') %>% 
     read_csv()
```

```{r}
ffq <- 
     samdf %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     select(study, subj, reads_tot, over_threshold) %>% 
     right_join(ffq) %>% 
     mutate(over_1k = ifelse(reads_tot >= 1000,
                             yes = 'Over\n1,000\nreads',
                             no = 'Under\n1,000\nreads'),
            over_1k = factor(over_1k,
                             levels = c('Under\n1,000\nreads',
                                        'Over\n1,000\nreads')))
```

Really neat:
- HEI plant significant, HEI non-plant not
- PDI, uPDI significant and opposite directions, hPDI not
- FVS plant residuals significant, animal residuals unrelated

```{r}
ggplot(ffq, aes(x = over_1k, y = hei_plant_all)) +
     geom_boxplot() +
     labs(y = 'HEI-2015\n(plant component)') +
     ylim(7, 35) +
     theme(axis.title.x = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_HEI over-under 1,000 reads.pdf')),
       height = 4, width = 2.5)
```

```{r}
wilcox.test(hei_plant_all ~ over_1k, ffq,
            conf.int = TRUE)
```

```{r}
wilcox.test(hei_non_plant ~ over_1k, ffq,
       conf.int = TRUE)
```

```{r}
ggplot(ffq, aes(x = over_1k, y = PDI)) +
     geom_boxplot() +
     labs(y = '\nPDI') +
     ylim(40, 80) +
     theme(axis.title.x = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_PDI over-under 1,000 reads.pdf')),
       height = 4, width = 2.5)
```

```{r}
wilcox.test(PDI ~ over_1k, ffq,
            conf.int = TRUE)
```

```{r}
wilcox.test(hPDI ~ over_1k, ffq,
       conf.int = TRUE)
```

```{r}
wilcox.test(uPDI ~ over_1k, ffq,
            conf.int = TRUE)
```

```{r}
ggplot(ffq, aes(x = over_1k, y = fvs_plant_residual)) +
     geom_boxplot() +
     labs(y = 'FVS score residuals\n(plant items)') +
     ylim(-13, 33) +
     theme(axis.title.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_FVS plant residual over-under 1,000 reads.pdf')),
       height = 4, width = 2.5)
```

```{r}
wilcox.test(fvs_plant_residual ~ over_threshold, ffq,
            conf.int = TRUE)
```

```{r}
wilcox.test(fvs_animal_residual ~ over_threshold, ffq,
            conf.int = TRUE)
```

### Effect on richness

```{r}
melt <- psmelt(ps)
```

```{r}
melt <- 
     melt %>% 
     group_by(study, seq_date, Sample, reads_tot, qubit) %>% 
     filter(Abundance > 0) %>% 
     count()
```

```{r}
ggplot(melt, aes(x = reads_tot, y = n)) + 
     geom_point(alpha = 0.5) +
     facet_wrap(~study) 
```

```{r}
ggplot(melt, aes(x = reads_tot, y = n/reads_tot, color = study)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date,
                nrow = 2,
                scales = 'free_x') +
     scale_y_log10() +
     scale_color_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     labs(y = '*trnL* taxa per read', x = 'Total reads') +
     theme(axis.text.x = element_text(angle = 20,
                                      hjust = 1), 
           axis.title.y = ggtext::element_markdown(),
           legend.title = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Observations per read v Qubit.pdf')),
       height = 4, width = 9)
```

## Rarefaction

### Necessary?

Use iNEXT package here to assess.

```{r}
# Function needs input x, dataframe of species x sites
x <- 
     ps@otu_table@.Data %>% 
     t() %>% 
     data.frame()

head(x)
```
For input datatype, this is one of
- Individual-based abundance data (datatype = "abundance")
- Sampling-unit-based incidence frequencies data (datatype = "incidence_freq")
- Species by sampling-units incidence matrix (datatype = "incidence_raw")-- this has to be 0/1

```{r}
sampling_units <- 
     apply(x, 2, sum) %>% 
     matrix(ncol = dim(x)[2]) %>% 
     data.frame()

names(sampling_units) <- names(x)  

x <- bind_rows(sampling_units, x)
row.names(x)[1] <- 'sampling_units'

head(x)
```
```{r}
# Remove entries with 0 reads
missing <- (x['sampling_units', ] == 0)
sum(missing)

dim(x)
x <- x[, !missing]
dim(x)
```

```{r}
re <- 
     iNEXT(x,
           q = 0, # Richness
           datatype = 'incidence_freq')
```

```{r}
ggiNEXT(re, type = 1) +
     labs(x = 'Sequencing reads', y = 'pMR') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_R-E, richness x sequencing reads.pdf')),
       height = 3, width = 4)
```

```{r}
ggiNEXT(re, type = 2) +
     labs(x = 'Sequencing reads', y = 'Sample coverage') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_R-E, coverage x sequencing reads.pdf')),
       height = 3, width = 4)
```

Now make same plots, faceted by sample
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 
         'manuscript', 
         'supplemental',
         'rarefaction-extrapolation',
         paste0(plotID, '_R-E, richness x sequencing reads, faceted.pdf')),
       height = 200, width = 25)
p <- ggiNEXT(re, type = 1, facet = 'site') +
     facet_wrap(~site, ncol = 10,
                scales = 'free') + 
     labs(x = 'Sequencing reads', y = 'pMR') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 
         'manuscript', 
         'supplemental',
         'rarefaction-extrapolation',
         paste0(plotID, '_R-E, sample coverage x sequencing reads, faceted.pdf')),
       height = 200, width = 25)
p <- ggiNEXT(re, type = 2, facet = 'site') +
     facet_wrap(~site, ncol = 10,
                scales = 'free') + 
     labs(x = 'Sequencing reads', y = 'Sample coverage') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

### Iterative PERMANOVA

Want to do this with Jaccard distance, since ideally analyses here are based on number of observed species. 

```{r}
# Try 500-read increments
cat('The maximum number of reads assigned to any sample in the dataset is',  
    max(sample_sums(ps)))
cutoffs <- seq(0, 100000, by = 500)
```

```{r}
# Now iterate over PERMANOVA
# The idea would be to threshold by a particular cutoff
# And then test to see if cutoff still applies
set.seed(10101)
outputs <- list()

for (i in seq_along(cutoffs)){
        # Make function using that cutoff variable
        cutoff <- cutoffs[i]
        
        # Subset phyloseq object to only samples above that have read counts
        # greater than cutoff
        # ps.temp <- subset_samples(ps.clr,
        ps.temp <- subset_samples(ps,
                                  reads_food > cutoff)
        
        # Update sample data
        samdf.temp <- as(sample_data(ps.temp), 'data.frame')
        
        output <-
             vegan::adonis(
                  distance(
                       ps.temp,
                       method = 'jaccard', binary = TRUE
                  ) ~ reads_food,
                  data = samdf.temp,
                  strata = samdf.temp$subj)
        
        outputs[[i]] <- output$aov.tab
        names(outputs)[i] <- cutoff
}

# Tabulate values into dataframe

cutoffs <- names(outputs)
r2s <- lapply(outputs, '[[', 5) %>% 
        lapply('[[', 1) %>% 
        unlist()
pvals <- lapply(outputs, '[[', 6) %>% 
        lapply('[[', 1) %>% 
        unlist()

readcounts.permanova <- data.frame(cutoff = names(outputs),
                                   R2 = r2s,
                                   p = pvals) 

readcounts.permanova$cutoff <- as.numeric(readcounts.permanova$cutoff)

readcounts.permanova <- pivot_longer(readcounts.permanova, 
                                     cols = -cutoff,
                                     names_to = 'fit',
                                     values_to = 'value')
```

Note: often getting error here that "Set of permutations < 'minperm'. Generating entire set. My guess is this happens because so few samples are left after filtering that they can't be permuted up to nperm times.  TODO: Figure out if this should change interpretation of results?

```{r}
# Plot results
ggplot(readcounts.permanova,
       aes(x = cutoff, y = value, color = fit, group = fit, by = fit)) +
     geom_line(size = 1) +
     geom_hline(aes(yintercept = 0.05), linetype = 'dashed', 
                color = '#F8766D') +
     scale_color_manual(values = c('gray40', 
                                   'gray80')) +
     labs(x = 'Read count cutoff', 
          y = 'PERMANOVA result', 
          color = 'Fit parameter', 
          title = 'Jaccard (with strata)',
          caption = 'Significance threshold crossed at 10,500 reads') +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 'manuscript', 'supplemental',
#             paste0(plotID, '_Iterative PERMANOVA (Jaccard with strata).pdf')),
#        height = 4, width = 6)
```

```{r}
# When is threshold crossed?
readcounts.permanova %>% 
     filter(fit == 'p' & value > 0.05)
```

### Vegan

#### Inspect curves

```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- otu_table(ps)@.Data
dim(asvtab)
```

```{r}
# View curves
# Randomly choose 20 CHOMP rows
rarecurve(asvtab[1:20, ], col = "blue")
```

```{r}
# Contrast with 20 ONR rows
n <- sample(grep('ONR-', row.names(asvtab)), 20)
rarecurve(asvtab[n, ], col = "blue")
```

More important than the curves may be conclusions we can draw from the slope: `rareslope` calculates the slope of `rarecurve` at `sample` size (can be a vector).
```{r}
slopes <- rareslope(asvtab, seq(100, 20000, 100))
```

```{r}
slopes.long <- 
     slopes %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     pivot_longer(cols = -row,
                  names_to = 'sample',
                  values_to = 'slope') %>% 
     mutate(sample = as.numeric(gsub('N', '', sample)),
            study = gsub('-\\d+$', '', row))

slopes.long
```

```{r}
ggplot(slopes.long, aes(x = sample, y = slope, group = sample)) +
     geom_point(alpha = 0.5) 
```

```{r}
means <- colSums(slopes)/dim(slopes)[1]
```

```{r}
# What does slope *mean*? For each additional read, we see that many additional species in the sample. Taking the inverse asks: How many reads do we need to see one additional species?
1/means
```

Ok, so let's consider then a 5000-read threshold, which would mean we'd need 14,626 additional reads to get one more species. Could perhaps go even lower based on inflection point of graph above.

```{r}
raremin <- c(1000, 5000, 10000) 

# Rarefy to each of these thresholds
asvtab <- ps@otu_table@.Data
asvtab.rare <- rarefy(asvtab, sample = raremin, se = TRUE)

head(asvtab.rare)
```

```{r}
# Re-organize results
rarefaction.df <- 
     asvtab.rare %>% 
     data.frame() %>% 
     rownames_to_column(var = 'name') %>% 
     separate(name, into = c('name', 'measure'), sep = '\\.') %>% 
     mutate(measure = ifelse(measure == 'S',
                             yes = 'obs_rare',
                             no = 'se')) 

rarefaction.df
```

```{r}
# Arrange for plotting
rarefaction.df <- 
     rarefaction.df %>% 
     pivot_longer(cols = matches('N\\d+'),
                  names_to = 'threshold',
                  values_to = 'rarefied') %>% 
     mutate(threshold = gsub(threshold, 
                             pattern = '^N', 
                             replacement = '')) %>% 
     mutate(threshold = factor(threshold,
                               levels = c(1000, 5000, 10000))) %>% 
     pivot_wider(names_from = 'measure',
                 values_from = 'rarefied')

rarefaction.df
```
```{r}
# Join to actual observations
rarefaction.df <- 
     apply(asvtab, 1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name') %>% 
     right_join(rarefaction.df) %>% 
     select(name, obs, everything())

rarefaction.df
```

Make sure-- are these being correctly populated with NA when the sample is below threshold?  Don't think so:

```{r}
any(is.na(rarefaction.df$obs_rare))
```

```{r}
# Join to sample data
rarefaction.df <- 
     select(samdf, row, reads_food) %>% 
     right_join(rarefaction.df, by = c('row' = 'name')) %>% 
     select(name = row, obs, reads_food, everything())

rarefaction.df
```
```{r}
# Replace values below threshold with NA

# Go back to numeric; need to compare
rarefaction.df$threshold <- 
     as.character(rarefaction.df$threshold) %>%
     as.numeric()

rarefaction.df <- 
     mutate(rarefaction.df,
            across(.cols = c(obs_rare, se),
                   ~ifelse(reads_food < threshold,
                           yes = NA,
                           no = .x)))
```

How to these estimates compare to raw data?
```{r}
# Could maybe add standard error on top of this?
ggplot(rarefaction.df, aes(x = obs, y = obs_rare)) + 
     geom_abline(slope = 1, intercept = 0, color = 'red') +
     geom_errorbar(aes(ymin = obs_rare - se, 
                       ymax = obs_rare + se), 
                   width = 0.2) +
     geom_point(alpha = 0.5) +
     facet_wrap(~threshold) + 
     coord_equal() +
     labs(x = 'pMR (raw)', y = 'pMR (rarefied)')
```
This is maybe a visualization that appears better than reality, since I'm not showing the points that are removed from the dataset? 
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'rarefaction-extrapolation',
#             paste0(plotID, '_Rarefied x raw pMR.pdf')),
#        height = 4, width = 4)
```

```{r}
# Save rarefied data
saveRDS(rarefaction.df,
        here('data', 
             'processed',
             'rarefaction',
             'Rarefied trnL counts.rds'))
```

Now, try repeating read count plots and alpha diversity with rarefied data. 

# Comparing cohorts

## Raw richness

Do on a per-subject basis, so one subject having more samples doesn't weight the data unfairly.

```{r}
# Get each subject's mean alpha diversity
alpha <- 
     ps@otu_table@.Data %>% 
     apply(1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'row') %>% 
     left_join(samdf)

alpha <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(obs_mean = mean(obs))
```

```{r}
# Factor levels for plotting
alpha$study <- 
     factor(alpha$study,
            levels = c('Weight Loss',
                       'Adult-1', 
                       'Adult-2',
                       'Adolescent'))
```

```{r}
# Plot
ggplot(alpha, aes(x = study, y = obs_mean, group = study)) +
     geom_boxplot() +
     labs(y = 'pMR') +
     theme(axis.title.x = element_blank(),
           axis.text = element_text(angle = 20, hjust = 1))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript', 
            '5',
            paste0(plotID, 'Raw pMR by study.pdf')),
       height = 4, width = 2.5)
```


```{r}
result.anova <- aov(obs_mean ~ study,
                    data = alpha)

summary(result.anova)
```

```{r}
TukeyHSD(result.anova)
```

## Rarefied richness

Do on a per-subject basis, so one subject having more samples doesn't weight the data unfairly.

```{r}
# Normalize withineach subject
stats.rare <- 
     samdf %>% 
     select(name = row, study, subj) %>% 
     right_join(rarefaction.df) %>% 
     group_by(study, subj, threshold) %>% 
     summarize(obs_rare_mean = mean(obs_rare, na.rm = TRUE)) %>% 
     filter(obs_rare_mean != 'NaN')
```

```{r}
# Factor levels for plotting
stats.rare$study <- 
     factor(stats.rare$study,
            levels = c('Weight Loss',
                       'Adult-1', 
                       'Adult-2',
                       'Adolescent'))
```

```{r}
# Plot
ggplot(stats.rare, aes(x = study, y = obs_rare_mean, group = study)) +
     geom_boxplot() +
     labs(y = 'Rarefied pMR') +
     facet_wrap(~threshold) +
     theme(axis.title.x = element_blank(),
           axis.text.x = element_text(angle = 20, hjust = 1))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript', 
            '5',
            paste0(plotID, '_Rarefied pMR by study.pdf')),
       height = 4, width = 6)
```


```{r}
result.anova <- 
     stats.rare %>% 
     filter(threshold == 10000) %>% 
     aov(obs_rare_mean ~ study,
                    data = .)

summary(result.anova)
```

```{r}
TukeyHSD(result.anova)
```

