---
title: "Joint phyloseq QC"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(iNEXT)
library(phyloseq)
library(tidyverse)
library(vegan)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

```{r}
# Load last saved phyloseq
ps <-
     here('data', 
          'processed', 
          'phyloseq', 
          'trnL_combined_ps.rds') %>%
     readRDS()

ps
```

# Summary analyses

```{r}
# These will require sample data
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row')
```

## Success rate

Samples with negative Qubit readings:
```{r}
samdf %>% 
     filter(qubit < 0) %>% 
     select(study, subj, qubit, reads_tot) %>% 
     arrange(desc(reads_tot))
```

```{r}
samdf %>% 
     filter(reads_tot < 1000) %>% 
     select(study, subj, qubit, reads_tot) %>% 
     arrange(desc(reads_tot))
```
```{r}
# Joint condition
failed <- 
     samdf %>% 
     filter(qubit < 0 | reads_tot < 1000) %>% 
     group_by(study) %>% 
     count(name = 'n_failed')

failed
```

```{r}
# Calculate failure rate by study
failed <- 
     samdf %>% 
     group_by(study) %>% 
     count(name = 'n_total') %>% 
     left_join(failed) %>% 
     mutate(n_failed = ifelse(is.na(n_failed),
                                    yes = 0,
                                    no = n_failed),
            n_success = n_total - n_failed)

failed
```

```{r}
# Overall failure rate
sum(failed$n_failed)/sum(failed$n_total)
```

```{r}
# Stats on study-specific failure rate
chisq <- 
     failed %>% 
     select(-n_total) %>% 
     column_to_rownames(var = 'study') %>% 
     as.matrix() %>% 
     chisq.test(x = .)

chisq
```

Remove these failed samples
Unclear how this will affect Qubit results below, so creating a separate object

```{r}
ps %>% 
     subset_samples(qubit > 0 & reads_tot > 1000) %>% 
     saveRDS(here('data',
                  'processed',
                  'phyloseq',
                  'trnL_combined_ps_filter.rds'))
```

## Descriptive stats
```{r}
# Number of subjects
samdf %>% 
     group_by(study) %>% 
     summarize(n = n_distinct(subj))
```
```{r}
# Total subjects
n_distinct(samdf$subj)
```

```{r}
# Samples per subject (mean and standard deviation)
samdf %>% 
     group_by(study, subj) %>% 
     count() %>% 
     ungroup() %>% 
     group_by(study) %>% 
     summarize(mean = mean(n), sd = sd(n))
```

## Number of taxa

```{r}
# Total
ntaxa(ps)
```

```{r}
# Consumed in >1 individual
asvtab <- 
     ps@otu_table@.Data %>% 
     data.frame()

multiple <- 
     apply(asvtab, 2, function(x){sum(x > 0) > (0.5*length(x))})
```

## Reads mapping to food

```{r}
# % reads mapping to food
sum(samdf$reads_food)/sum(samdf$reads_tot)
```

```{r}
# Range across samples
# Include only those with >1,000 reads
samdf %>% 
     select(row, starts_with('reads')) %>% 
     filter(reads_tot >= 1000) %>% 
     mutate(pct_no_id = 1 - (reads_food/reads_tot)) %>% 
     arrange(desc(pct_no_id))
```
Ooh-- definitely need to follow up on this.

## Read count threshold

### Distribution

```{r}
# Faceted by study
ggplot(samdf, aes(x = reads_tot, fill = study)) +
     geom_histogram(bins = 10, boundary = 0) +
     facet_wrap(~study, scales = 'free_y') +
     scale_fill_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     # xlim(0, 50000) +
     labs(x = 'Read count', y = 'Samples (n)') +
     theme_bw() +
     theme(axis.text.x = element_text(angle = 20, hjust = 1),
           legend.title = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Read depth of samples by study, facet.pdf')),
#        height = 4, width = 6)
```

```{r}
ggplot(samdf, aes(x = reads_food, y = ..density.., 
                  color = study, fill = study)) +
     geom_density(aes(color = study),
                  alpha = 0.5,
                  size = 1.5) +
     scale_color_manual(values = c('darkgreen', 
                                   'gray80', 
                                   'gray40', 
                                   'gold2')) +
     scale_fill_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     labs(x = 'Read count (food taxa)', y = 'Samples (n)') +
     theme_bw() +
     theme(legend.title = element_blank())
```

Need to hold off on this for now-- not informative I think unless scaled to actual size of the cohorts. 

### Relationship to input

#### DNA added

```{r}
ggplot(samdf, aes(x = ng_added, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     # geom_hline(yintercept = 10000,
     #            color = 'red',
     #            linetype = 'dashed') +
     # geom_vline(xintercept = 10,
     #            color = 'red',
     #            linetype = 'dashed') +
     facet_wrap(~seq_date, 
                nrow = 2) +
     labs(y = 'Reads', x = 'DNA added to pool (ng)') + 
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Read depth v ng added.pdf')),
       height = 4, width = 6)
```

```{r}
ggplot(samdf, aes(x = qubit, y = ng_added, color = study)) +
     geom_point(alpha = 0.5) +
     scale_color_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     facet_wrap(~seq_date, nrow = 2) +
          labs(x = 'Qubit measurement',
          y = 'DNA added to pool (ng)') +
     theme(legend.position = 'bottom',
           legend.title = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_DNA added v. Qubit.pdf')),
       height = 4, width = 5)
```

```{r}
ggplot(samdf, aes(x = qubit, y = ng_added, color = reads_tot > 10000)) +
     geom_point(alpha = 0.2) +
     scale_color_manual(values = c('red', 'gray20')) +
     facet_wrap(~seq_date, nrow = 2) +
     labs(x = 'Qubit measurement',
          y = 'DNA added to pool (ng)',
          color = '> 10,000 reads') +
     theme(legend.position = 'bottom')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_DNA added v. Qubit.pdf')),
       height = 4, width = 5)
```

#### Qubit

Histogram of Qubit values

```{r}
ggplot(samdf, aes(x = qubit, fill = study)) +
     geom_histogram(stat = 'bin',
                    bins = 10,
                    boundary = 0) +
     scale_fill_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     facet_wrap(~study, 
                nrow = 2,
                scales = 'free_y') +
     labs(y = 'Samples (n)', x = 'Qubit reading') + 
     theme_bw() +
     theme(legend.title = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Qubit histogram by study.pdf')),
#        height = 4, width = 6)
```

```{r}
ggplot(samdf, aes(x = qubit, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date, nrow = 2) +
     labs(y = 'Reads', x = 'Qubit reading') + 
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'manuscript', 
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Read depth v Qubit.pdf')),
#        height = 4, width = 6)
```

Add custom threshold line based on "ng added" above-- above what measure where we able to max out?

```{r}
qubit.threshold <- 
     samdf %>% 
     group_by(seq_date) %>% 
     arrange(desc(ng_added), qubit) %>% # Lowest Qubit at which max added
     summarize(threshold = first(qubit))
```

```{r}
# Add this to sample data, make new over/under threshold variable
samdf <- 
     samdf %>% 
     left_join(qubit.threshold) %>% 
     mutate(over_threshold = qubit >= threshold)
```

```{r}
ggplot(samdf, aes(x = qubit, y = reads_tot)) +
     geom_point(alpha = 0.5) +
     geom_vline(aes(xintercept = threshold), 
                color = 'red',
                linetype = 'dashed') +
     facet_wrap(~seq_date, nrow = 2) +
     labs(y = 'Reads', x = 'Qubit reading') + 
     theme_bw()
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'read depth',
#             paste0(plotID, '_Read depth v Qubit, with threshold.pdf')),
#        height = 4, width = 6)
```

Now, can filter out points to the right of this line and test for relationship.

```{r}
table(samdf$over_threshold)
```

```{r}
samdf %>% 
     filter(over_threshold == FALSE) %>% 
     group_by(seq_date) %>% 
     summarize(
          rho = cor.test(~qubit + reads_tot)$estimate,
          p = cor.test(~qubit + reads_tot)$p.value
     ) %>% 
     mutate(p_bh = p.adjust(p, method = 'BH'))
```
```{r}
samdf %>% 
     filter(over_threshold == TRUE) %>% 
     group_by(seq_date) %>% 
     summarize(
          rho = cor.test(~qubit + reads_tot)$estimate,
          p = cor.test(~qubit + reads_tot)$p.value
     ) %>% 
     mutate(p_bh = p.adjust(p, method = 'BH'))
```

### Effect on richness

```{r}
melt <- psmelt(ps)
```

```{r}
melt <- 
     melt %>% 
     group_by(study, seq_date, Sample, reads_tot, qubit) %>% 
     filter(Abundance > 0) %>% 
     count()
```

```{r}
ggplot(melt, aes(x = reads_tot, y = n/reads_tot, color = study)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~seq_date,
                nrow = 2,
                scales = 'free_x') +
     scale_y_log10() +
     scale_color_manual(values = c('darkgreen', 
                                  'gray80', 
                                  'gray40', 
                                  'gold2')) +
     labs(y = '*trnL* taxa per read', x = 'Total reads') +
     theme(axis.text.x = element_text(angle = 20,
                                      hjust = 1), 
           axis.title.y = ggtext::element_markdown(),
           legend.title = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'read depth',
            paste0(plotID, '_Observations per read v Qubit.pdf')),
       height = 4, width = 9)
```

## Rarefaction

### Necessary?

Use iNEXT package here to assess.

```{r}
# Function needs input x, dataframe of species x sites
x <- 
     ps@otu_table@.Data %>% 
     t() %>% 
     data.frame()

head(x)
```
For input datatype, this is one of
- Individual-based abundance data (datatype = "abundance")
- Sampling-unit-based incidence frequencies data (datatype = "incidence_freq")
- Species by sampling-units incidence matrix (datatype = "incidence_raw")-- this has to be 0/1

```{r}
sampling_units <- 
     apply(x, 2, sum) %>% 
     matrix(ncol = dim(x)[2]) %>% 
     data.frame()

names(sampling_units) <- names(x)  

x <- bind_rows(sampling_units, x)
row.names(x)[1] <- 'sampling_units'

head(x)
```
```{r}
# Remove entries with 0 reads
missing <- (x['sampling_units', ] == 0)
sum(missing)

dim(x)
x <- x[, !missing]
dim(x)
```

```{r}
re <- 
     iNEXT(x,
           q = 0, # Richness
           datatype = 'incidence_freq')
```

```{r}
ggiNEXT(re, type = 1) +
     labs(x = 'Sequencing reads', y = 'pMR') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_R-E, richness x sequencing reads.pdf')),
       height = 3, width = 4)
```

```{r}
ggiNEXT(re, type = 2) +
     labs(x = 'Sequencing reads', y = 'Sample coverage') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript', 
            'supplemental',
            'rarefaction-extrapolation',
            paste0(plotID, '_R-E, coverage x sequencing reads.pdf')),
       height = 3, width = 4)
```

Now make same plots, faceted by sample
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 
         'manuscript', 
         'supplemental',
         'rarefaction-extrapolation',
         paste0(plotID, '_R-E, richness x sequencing reads, faceted.pdf')),
       height = 200, width = 25)
p <- ggiNEXT(re, type = 1, facet = 'site') +
     facet_wrap(~site, ncol = 10,
                scales = 'free') + 
     labs(x = 'Sequencing reads', y = 'pMR') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 
         'manuscript', 
         'supplemental',
         'rarefaction-extrapolation',
         paste0(plotID, '_R-E, sample coverage x sequencing reads, faceted.pdf')),
       height = 200, width = 25)
p <- ggiNEXT(re, type = 2, facet = 'site') +
     facet_wrap(~site, ncol = 10,
                scales = 'free') + 
     labs(x = 'Sequencing reads', y = 'Sample coverage') +
     theme_bw() +
     theme(legend.position = 'none',
           axis.text = element_text(size = 12),
           axis.title = element_text(size = 14,
                                     face = 'bold'),
           strip.text = element_text(size = 12,
                                     face = 'bold'))
```

### Iterative PERMANOVA

Want to do this with Jaccard distance, since ideally analyses here are based on number of observed species. 

```{r}
# Try 500-read increments
cat('The maximum number of reads assigned to any sample in the dataset is',  
    max(sample_sums(ps)))
cutoffs <- seq(0, 100000, by = 500)
```

```{r}
# Now iterate over PERMANOVA
# The idea would be to threshold by a particular cutoff
# And then test to see if cutoff still applies
set.seed(10101)
outputs <- list()

for (i in seq_along(cutoffs)){
        # Make function using that cutoff variable
        cutoff <- cutoffs[i]
        
        # Subset phyloseq object to only samples above that have read counts
        # greater than cutoff
        # ps.temp <- subset_samples(ps.clr,
        ps.temp <- subset_samples(ps,
                                  reads_food > cutoff)
        
        # Update sample data
        samdf.temp <- as(sample_data(ps.temp), 'data.frame')
        
        output <-
             vegan::adonis(
                  distance(
                       ps.temp,
                       method = 'jaccard', binary = TRUE
                  ) ~ reads_food,
                  data = samdf.temp,
                  strata = samdf.temp$subj)
        
        outputs[[i]] <- output$aov.tab
        names(outputs)[i] <- cutoff
}

# Tabulate values into dataframe

cutoffs <- names(outputs)
r2s <- lapply(outputs, '[[', 5) %>% 
        lapply('[[', 1) %>% 
        unlist()
pvals <- lapply(outputs, '[[', 6) %>% 
        lapply('[[', 1) %>% 
        unlist()

readcounts.permanova <- data.frame(cutoff = names(outputs),
                                   R2 = r2s,
                                   p = pvals) 

readcounts.permanova$cutoff <- as.numeric(readcounts.permanova$cutoff)

readcounts.permanova <- pivot_longer(readcounts.permanova, 
                                     cols = -cutoff,
                                     names_to = 'fit',
                                     values_to = 'value')
```

Note: often getting error here that "Set of permutations < 'minperm'. Generating entire set. My guess is this happens because so few samples are left after filtering that they can't be permuted up to nperm times.  TODO: Figure out if this should change interpretation of results?

```{r}
# Plot results
ggplot(readcounts.permanova,
       aes(x = cutoff, y = value, color = fit, group = fit, by = fit)) +
     geom_line(size = 1) +
     geom_hline(aes(yintercept = 0.05), linetype = 'dashed', 
                color = '#F8766D') +
     scale_color_manual(values = c('gray40', 
                                   'gray80')) +
     labs(x = 'Read count cutoff', 
          y = 'PERMANOVA result', 
          color = 'Fit parameter', 
          title = 'Jaccard (with strata)',
          caption = 'Significance threshold crossed at 10,500 reads') +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 'manuscript', 'supplemental',
#             paste0(plotID, '_Iterative PERMANOVA (Jaccard with strata).pdf')),
#        height = 4, width = 6)
```

```{r}
# When is threshold crossed?
readcounts.permanova %>% 
     filter(fit == 'p' & value > 0.05)
```

### iNEXT



### Vegan

#### Inspect curves

```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- otu_table(ps)@.Data
dim(asvtab)
```

```{r}
# View curves
# Randomly choose 20 CHOMP rows
rarecurve(asvtab[1:20, ], col = "blue")
```

```{r}
# Contrast with 20 ONR rows
n <- sample(grep('ONR-', row.names(asvtab)), 20)
rarecurve(asvtab[n, ], col = "blue")
```

More important than the curves may be conclusions we can draw from the slope: `rareslope` calculates the slope of `rarecurve` at `sample` size (can be a vector).
```{r}
slopes <- rareslope(asvtab, seq(100, 20000, 100))
```

```{r}
slopes.long <- 
     slopes %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     pivot_longer(cols = -row,
                  names_to = 'sample',
                  values_to = 'slope') %>% 
     mutate(sample = as.numeric(gsub('N', '', sample)),
            study = gsub('-\\d+$', '', row))

slopes.long
```

```{r}
ggplot(slopes.long, aes(x = sample, y = slope, group = sample)) +
     geom_point(alpha = 0.5) 
```

```{r}
means <- colSums(slopes)/dim(slopes)[1]
```

```{r}
# What does slope *mean*? For each additional read, we see that many additional species in the sample. Taking the inverse asks: How many reads do we need to see one additional species?
1/means
```

Ok, so let's consider then a 5000-read threshold, which would mean we'd need 14,626 additional reads to get one more species. Could perhaps go even lower based on inflection point of graph above.

```{r}
raremin <- c(1000, 5000, 10000) 

# Rarefy to each of these thresholds
asvtab <- ps@otu_table@.Data
asvtab.rare <- rarefy(asvtab, sample = raremin, se = TRUE)

head(asvtab.rare)
```

```{r}
# Re-organize results
rarefaction.df <- 
     asvtab.rare %>% 
     data.frame() %>% 
     rownames_to_column(var = 'name') %>% 
     separate(name, into = c('name', 'measure'), sep = '\\.') %>% 
     mutate(measure = ifelse(measure == 'S',
                             yes = 'obs_rare',
                             no = 'se')) 

rarefaction.df
```

```{r}
# Arrange for plotting
rarefaction.df <- 
     rarefaction.df %>% 
     pivot_longer(cols = matches('N\\d+'),
                  names_to = 'threshold',
                  values_to = 'rarefied') %>% 
     mutate(threshold = gsub(threshold, 
                             pattern = '^N', 
                             replacement = '')) %>% 
     mutate(threshold = factor(threshold,
                               levels = c(1000, 5000, 10000))) %>% 
     pivot_wider(names_from = 'measure',
                 values_from = 'rarefied')

rarefaction.df
```
```{r}
# Join to actual observations
rarefaction.df <- 
     apply(asvtab, 1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name') %>% 
     right_join(rarefaction.df) %>% 
     select(name, obs, everything())

rarefaction.df
```

Make sure-- are these being correctly populated with NA when the sample is below threshold?  Don't think so:

```{r}
any(is.na(rarefaction.df$obs_rare))
```

```{r}
# Join to sample data
rarefaction.df <- 
     select(samdf, row, reads_food) %>% 
     right_join(rarefaction.df, by = c('row' = 'name')) %>% 
     select(name = row, obs, reads_food, everything())

rarefaction.df
```
```{r}
# Replace values below threshold with NA

# Go back to numeric; need to compare
rarefaction.df$threshold <- 
     as.character(rarefaction.df$threshold) %>%
     as.numeric()

rarefaction.df <- 
     mutate(rarefaction.df,
            across(.cols = c(obs_rare, se),
                   ~ifelse(reads_food < threshold,
                           yes = NA,
                           no = .x)))
```

How to these estimates compare to raw data?
```{r}
# Could maybe add standard error on top of this?
ggplot(rarefaction.df, aes(x = obs, y = obs_rare)) + 
     geom_abline(slope = 1, intercept = 0, color = 'red') +
     geom_errorbar(aes(ymin = obs_rare - se, 
                       ymax = obs_rare + se), 
                   width = 0.2) +
     geom_point(alpha = 0.5) +
     facet_wrap(~threshold) + 
     coord_equal() +
     labs(x = 'pMR (raw)', y = 'pMR (rarefied)')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results',
#             'manuscript',
#             'supplemental',
#             'rarefaction-extrapolation',
#             paste0(plotID, '_Rarefied x raw pMR.pdf')),
#        height = 4, width = 4)
```

```{r}
# Save rarefied data
# saveRDS(rarefaction.df,
#         here('data', 
#              'processed', 
#              'rarefaction',
#              'Rarefied trnL counts.rds'))
```

Now, try repeating read count plots and alpha diversity with rarefied data. 

# Comparing cohorts

## Raw richness

Do on a per-subject basis, so one subject having more samples doesn't weight the data unfairly.

```{r}
# Get each subject's mean alpha diversity
alpha <- 
     ps@otu_table@.Data %>% 
     apply(1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'row') %>% 
     left_join(samdf)

alpha <- 
     alpha %>% 
     group_by(study, subj) %>% 
     summarize(obs_mean = mean(obs))
```

```{r}
# Factor levels for plotting
alpha$study <- 
     factor(alpha$study,
            levels = c('Weight Loss',
                       'Adult-1', 
                       'Adult-2',
                       'Adolescent'))
```

```{r}
# Plot
ggplot(alpha, aes(x = study, y = obs_mean, group = study)) +
     geom_boxplot() +
     theme_classic() +
     labs(y = '*trnL* richness') +
     theme(axis.title.x = element_blank(),
           axis.title.y = ggtext::element_markdown(size = 14,
                                                   face = 'bold'),
           axis.text = element_text(size = 12))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript', 
            '5',
            paste0(plotID, '_A_Raw trnL richness by study.pdf')),
       height = 4, width = 4)
```


```{r}
result.anova <- aov(obs_mean ~ study,
                    data = alpha)

summary(result.anova)
```

```{r}
TukeyHSD(result.anova)
```

## Rarefied richness

Do on a per-subject basis, so one subject having more samples doesn't weight the data unfairly.

```{r}
# Normalize within each subject
stats.rare <- 
     rarefaction.df %>% 
     group_by(study, subj) %>% 
     summarize(obs_rare_mean = mean(obs_rare))
```

```{r}
# Factor levels for plotting
stats.rare$study <- 
     factor(stats.rare$study,
            levels = c('Weight Loss',
                       'Adult-1', 
                       'Adult-2',
                       'Adolescent'))
```

```{r}
# Plot
ggplot(stats.rare, aes(x = study, y = obs_rare_mean, group = study)) +
     geom_boxplot() +
     theme_classic() +
     labs(y = 'Rarefied *trnL* richness') +
     theme(axis.title.x = element_blank(),
           axis.title.y = ggtext::element_markdown(size = 14,
                                                   face = 'bold'),
           axis.text.x = element_text(size = 12))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript', 
            '5',
            paste0(plotID, '_A_Rarefied trnL richness by study.pdf')),
       height = 4, width = 4)
```


```{r}
result.anova <- aov(obs_rare_mean ~ study,
                    data = stats.rare)

summary(result.anova)
```

```{r}
TukeyHSD(result.anova)
```

