---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp <- 
     here('data', 'processed', 'phyloseq', '20210714_ps_trnL_CHOMP.rds') %>% 
     readRDS()

ps.chomp
```

### ONR

```{r}
ps.onr <- 
     here('data', 'processed', 'phyloseq', '20210617_ps_trnL_ONR.rds') %>% 
     readRDS()

ps.onr
```

### DFC

```{r}
ps.dfc <- 
     here('data', 'processed', 'phyloseq', '20210623_ps_trnL_DFC.rds') %>% 
     readRDS()

ps.dfc
```

### POMMS

```{r}
ps.pomms <- 
     here('data', 'processed', 'phyloseq', '20210612_ps_trnL_POMMS.rds') %>% 
     readRDS()

ps.pomms
```

### Tidy

#### Reduce sample data

To begin with, let's consider only study and other broad metadata variables:
* Age (pediatric/adolescent/adult)
* Diet (free-eating/weight loss/controlled feeding)

```{r}
# Populate all study variables
# (Tried to do before loading, but names not consistent, and ONR has some NAs)

sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.dfc)$study <- 'DFC'
sample_data(ps.onr)$study <- 'ONR'
sample_data(ps.pomms)$study <- 'POMMS'
```

```{r}
# Add demographics, as above
# Age
sample_data(ps.chomp)$age <- 'Adult'
sample_data(ps.dfc)$age <- 'Adult'
sample_data(ps.onr)$age <- 'Adult'
sample_data(ps.pomms)$age <- 'Adolescent'

# Diet
sample_data(ps.chomp)$diet <- 'Free-eating'
sample_data(ps.dfc)$diet <- 'Weight loss'
sample_data(ps.onr)$diet <- 'Free-eating'
sample_data(ps.pomms)$diet <- 'Free-eating'
```

#### Unique sample names

```{r}
# Otherwise, with well-based labeling, some might be combined in merge
sample_names(ps.chomp) <- paste(sample_data(ps.chomp)$study, 
                              seq(nsamples(ps.chomp)), sep = '-')

sample_names(ps.dfc) <- paste(sample_data(ps.dfc)$study, 
                              seq(nsamples(ps.dfc)), sep = '-')

sample_names(ps.onr) <- paste(sample_data(ps.onr)$study, 
                              seq(nsamples(ps.onr)), sep = '-')

sample_names(ps.pomms) <- paste(sample_data(ps.pomms)$study, 
                              seq(nsamples(ps.pomms)), sep = '-')
```

### Merge

Note that initially when I did this merge I had 1035 samples, and now I have 20 more.  I think this is due to the fact that I now preserve samples with 0 reads in the phyloseq objects, even if they fell out earlier in the pipeline.

```{r}
ps <- merge_phyloseq(ps.chomp, ps.dfc, ps.onr, ps.pomms)
ps
```

```{r}
# Adjust sample variables en masse for subsequent plots
sample_data(ps)$diet <- factor(sample_data(ps)$diet, 
                               levels = c('Free-eating', 'Weight loss'))

sample_data(ps)$reads <- sample_sums(ps)

sample_data(ps)$reads_binned <- cut(sample_data(ps)$reads, 
                                    breaks = c(-1, 1001, 5001, 10001, 30001,
                                               max(sample_data(ps)$reads) + 1), 
                                    labels = c('<1,000 reads', '1,000-5,000', 
                                               '5,000-10,000', '10,000-30,000',
                                               '>30,000'))

sample_data(ps)$study <- 
     factor(sample_data(ps)$study, 
            levels = c('POMMS', 'CHOMP', 'ONR', 'DFC'), 
            labels = c('Adolescent', 'Adult-1', 'Adult-2', 'Weight Loss'))
```

Note-- also might be interesting to try this ordination not just on ASVs identified to a food. Because of two features:
* ASVs not identified make up the minority of the total reads in the dataset, so they shouldn't affect the ordination much (at least not for PCA on CLR-transformed counts)
* However, ASVs that do map to a food, but aren't yet labeled as such, will be shared across cohorts and contribute meaningfully to the ordination, which is what will ultimately happen when I label them.

So I think reasonable to at least try an ordination on the full dataset (~1000 ASVs) for comparison to the one on the food-only dataset (~10% of that number). 

### Filter

#### Read counts

```{r}
# Remove samples with 0 read counts
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps
```

#### Plant taxa

Here, decide whether to proceed with the data as-is (all identified SVs), or analyze only those labeled to a food species.

##### Foods only

What percentage of reads in the dataset are labeled with a food taxon name?

```{r}
taxtab <- data.frame(ps@tax_table)
asvtab <- data.frame(ps@otu_table)

unassigned <- 
     taxtab %>% 
     filter(is.na(superkingdom)) %>% 
     row.names()

sum(asvtab[, unassigned])/sum(asvtab)
```

```{r}
# Consider foods only
ps <- subset_taxa(ps, !is.na(superkingdom))
ps
```

After this step, need to check again for samples with 0 read counts, as their only reads may have been unlabeled.
```{r}
# Remove samples with 0 read counts
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps
```

```{r}
# How many identifications at each phylogenetic level?
ps@tax_table@.Data %>% 
     apply(2, function(x){sum(!is.na(x))})
```


```{r}
# Clarify sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rename(reads_tot = reads,
            reads_tot_binned = reads_binned)

# Replace in object
sample_data(ps) <- samdf

# Calculate food-only reads
sample_data(ps)$reads_food <- sample_sums(ps)
```

##### Glom names

```{r}
taxtab <- 
     data.frame(ps@tax_table) %>% 
     lowest_level() # Get lowest named level
```

```{r}
# How many duplicates?
taxtab$name[duplicated(taxtab$name)]
```

This is a bit fraught at higher levels, because depending on exact ASVs these families may actually map to different species.  Ideally go back and do this by exact ASV identity, but for now, try merging everything.

```{r}
# Replace taxonomy table in phyloseq object
tax_table(ps) <- as.matrix(taxtab)

# Glom at this new level
ps <- tax_glom(ps, taxrank = 'name')
ps
```

### Save

```{r}
# saveRDS(ps,
#         here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds'))
```

```{r}
# Load last saved phyloseq
ps <-
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>%
     readRDS()

ps
```

# Summary analyses

```{r}
# These will require sample data
samdf <- 
     ps@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row')
```

## % reads mapping to food

```{r}
sum(samdf$reads_food)/sum(samdf$reads_tot)
```

## Test for read count threshold

### Distribution

```{r}
# Faceted by study
ggplot(samdf, aes(x = reads_food)) +
     geom_histogram(binwidth = 10000, boundary = 0) +
     facet_wrap(~study, scales = 'free_y') +
     # xlim(0, 50000) +
     labs(x = 'Read count', y = 'Count') +
     theme_bw()
```

```{r}
ggplot(samdf, aes(x = reads_food, color = study)) +
     geom_histogram(binwidth = 10000,
                    boundary = 0) +
     geom_density(aes(color = study), 
                  alpha = 0.5,
                  size = 2,
                  stat = 'bin',
                  binwidth = 10000) +
     labs(x = 'Read count (food taxa)', y = 'Count') +
     theme_bw() +
     theme(legend.title = element_blank())
```


```{r}
# Overall histogram
ggplot(samdf, aes(x = reads_food)) +
     geom_histogram(binwidth = 1000, boundary = 0) +
     # facet_wrap(~study, scales = 'free_y') +
     xlim(0, 50000) +
     labs(x = 'Read count', y = 'Count') +
     theme_bw()
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 'manuscript', 'supplemental',
#             paste0(plotID, '_Read depth of all samples, zoom.png')),
#        height = 4, width = 6)
```

```{r}
# Overall histogram (zoom)
ggplot(samdf, aes(x = reads_food)) +
     geom_histogram(binwidth = 1000, boundary = 0) +
     # facet_wrap(~study, scales = 'free_y') +
     xlim(0, 50000) +
     labs(x = 'Read count', y = 'Count') +
     theme_bw()
```

## Iterative PERMANOVA

Want to do this with Jaccard distance, since ideally analyses here are based on 
number of observed species. 

```{r}
# Try 500-read increments
cat('The maximum number of reads assigned to any sample in the dataset is',  
    max(sample_sums(ps)))
cutoffs <- seq(0, 100000, by = 500)
```

```{r}
# Now iterate over PERMANOVA
# The idea would be to threshold by a particular cutoff
# And then test to see if cutoff still applies
set.seed(10101)
outputs <- list()

for (i in seq_along(cutoffs)){
        # Make function using that cutoff variable
        cutoff <- cutoffs[i]
        
        # Subset phyloseq object to only samples above that have read counts
        # greater than cutoff
        # ps.temp <- subset_samples(ps.clr,
        ps.temp <- subset_samples(ps,
                                  reads_food > cutoff)
        
        # Update sample data
        samdf.temp <- as(sample_data(ps.temp), 'data.frame')
        
        output <-
             vegan::adonis(
                  distance(
                       ps.temp,
                       method = 'jaccard', binary = TRUE
                  ) ~ reads_food,
                  data = samdf.temp,
                  strata = samdf.temp$subj)
        
        outputs[[i]] <- output$aov.tab
        names(outputs)[i] <- cutoff
}

# Tabulate values into dataframe

cutoffs <- names(outputs)
r2s <- lapply(outputs, '[[', 5) %>% 
        lapply('[[', 1) %>% 
        unlist()
pvals <- lapply(outputs, '[[', 6) %>% 
        lapply('[[', 1) %>% 
        unlist()

readcounts.permanova <- data.frame(cutoff = names(outputs),
                                   R2 = r2s,
                                   p = pvals) 

readcounts.permanova$cutoff <- as.numeric(readcounts.permanova$cutoff)

readcounts.permanova <- pivot_longer(readcounts.permanova, 
                                     cols = -cutoff,
                                     names_to = 'fit',
                                     values_to = 'value')
```
Note: often getting error here that "Set of permutations < 'minperm'. Generating entire set. My guess is this happens because so few samples are left after filtering that they can't be permuted up to nperm times.  TODO: Figure out if this should change interpretation of results?

```{r}
# Plot results
ggplot(readcounts.permanova,
       aes(x = cutoff, y = value, color = fit, group = fit, by = fit)) +
     geom_line(size = 1) +
     geom_hline(aes(yintercept = 0.05), linetype = 'dashed', 
                color = '#F8766D') +
     scale_color_manual(values = c('gray40', 'gray80')) +
     labs(x = 'Read count cutoff', 
          y = 'PERMANOVA result', 
          color = 'Fit parameter', 
          title = 'Jaccard (with strata)',
          caption = 'Significance threshold crossed at 10,500 reads') +
     theme_bw()
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 'manuscript', 'supplemental',
            paste0(plotID, '_Iterative PERMANOVA (Jaccard with strata).pdf')),
       height = 4, width = 6)
```

```{r}
# When is threshold crossed?
readcounts.permanova %>% 
     filter(fit == 'p' & value > 0.05)
```
## Rarefaction

#### Inspect curves

```{r}
# Remove samples with counts below that threshold, and get ASV table
asvtab <- otu_table(ps)@.Data
dim(asvtab)
```

```{r}
# View curves
# Randomly choose 20 CHOMP rows
rarecurve(asvtab[1:20, ], col = "blue")
```

More important than the curves may be conclusions we can draw from the slope: `rareslope` calculates the slope of `rarecurve` at `sample` size (can be a vector).
```{r}
slopes <- rareslope(asvtab, seq(100, 20000, 100))
```

```{r}
slopes.long <- 
     slopes %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     pivot_longer(cols = -row,
                  names_to = 'sample',
                  values_to = 'slope') %>% 
     mutate(sample = as.numeric(gsub('N', '', sample)),
            study = gsub('-\\d+$', '', row))

slopes.long
```

```{r}
ggplot(slopes.long, aes(x = sample, y = slope, group = sample)) +
     geom_point() 
```

```{r}
means <- colSums(slopes)/dim(slopes)[1]
```

```{r}
# What does slope *mean*? For each additional read, we see that many additional species in the sample. Taking the inverse asks: How many reads do we need to see one additional species?
1/means
```

Ok, so let's consider then a 5000-read threshold, which would mean we'd need 14,626 additional reads to get one more species. Could perhaps go even lower based on inflection point of graph above.

From iterative PERMANOVA on Jaccard distances, may want threshold of 10,500.

```{r}
raremin <- 5000

# Discard samples with less than this read count
asvtab <- 
     ps %>% 
     prune_samples(sample_sums(.) > raremin, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     otu_table() %>% 
     getElement('.Data')

asvtab.rare <- rarefy(asvtab, sample = raremin, se = TRUE)
```

```{r}
# How does this compare to actual number passed to analysis below?
rarefaction.df <- 
     data.frame(obs_rare = asvtab.rare[1, ],
                obs_rare_se = asvtab.rare[2, ]) %>% 
     rownames_to_column(var = 'name')

# Add actual # observed taxa
rarefaction.df <- 
     rowSums(asvtab > 0) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name') %>% 
     full_join(rarefaction.df)
```

How well do they agree with one another?
```{r}
# Could maybe add standard error on top of this?
ggplot(rarefaction.df, aes(x = obs, y = obs_rare)) + 
     geom_point() +
     coord_equal() +
     theme_bw()
```

Now, try repeating read count plots and alpha diversity with rarefied data. 
