---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20220214_ps_trnL_CHOMP.rds') %>% 
     readRDS()

ps.chomp
```

### ONR

```{r}
ps.onr <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20220214_ps_trnL_ONR.rds') %>% 
     readRDS()

ps.onr
```

### DFC

```{r}
ps.dfc <- 
     here('data', 
          'processed',
          'phyloseq', 
          '20220126_ps_trnL_DFC.rds') %>% 
     readRDS()

ps.dfc
```

### POMMS

```{r}
ps.pomms <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20211221_ps_trnL_POMMS.rds') %>% 
     readRDS()

ps.pomms
```

### Tidy

#### Add variables

To begin with, let's consider only study and other broad metadata variables:
* Age (pediatric/adolescent/adult)
* Diet (free-eating/weight loss/controlled feeding)

```{r}
# Populate all study variables
# (Tried to do before loading, but names not consistent, and ONR has some NAs)

sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.dfc)$study <- 'DFC'
sample_data(ps.onr)$study <- 'ONR'
sample_data(ps.pomms)$study <- 'POMMS'
```

#### Unique sample names

```{r}
# Otherwise, with well-based labeling, some might be combined in merge
sample_names(ps.chomp) <- paste(sample_data(ps.chomp)$study, 
                              seq(nsamples(ps.chomp)), sep = '-')

sample_names(ps.dfc) <- paste(sample_data(ps.dfc)$study, 
                              seq(nsamples(ps.dfc)), sep = '-')

sample_names(ps.onr) <- paste(sample_data(ps.onr)$study, 
                              seq(nsamples(ps.onr)), sep = '-')

sample_names(ps.pomms) <- paste(sample_data(ps.pomms)$study, 
                              seq(nsamples(ps.pomms)), sep = '-')
```

### Merge

Note that initially when I did this merge I had 1035 samples, and now I have 20 more.  I think this is due to the fact that I now preserve samples with 0 reads in the phyloseq objects, even if they fell out earlier in the pipeline.

```{r}
ps <- merge_phyloseq(ps.chomp, ps.dfc, ps.onr, ps.pomms)
ps
```

```{r}
# Adjust sample variables en masse for subsequent plots
sample_data(ps)$reads <- sample_sums(ps)

sample_data(ps)$study <- 
     factor(sample_data(ps)$study, 
            levels = c('DFC', 'CHOMP', 'ONR', 'POMMS'), 
            labels = c('Weight Loss', 'Adult-1', 'Adult-2', 'Adolescent'))
```

### Tidy

```{r}
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row')
```


##### BMI

Uppercase in ONR, lower case in others

```{r}
samdf <- 
     samdf %>% 
     unite(col = bmi,
           bmi, BMI,
           na.rm = TRUE) %>% 
     mutate(bmi = ifelse(bmi == '', # Replace now empty values with NA
                         yes = NA,
                         no = bmi))
```


### Add variables

##### Cycle correction

PCR B cycle number only added once changed from 8 to 10 cycles (early 2021). Populate remaining values with 8 cycles.

```{r}
samdf %>% 
     group_by(seq_date, pcrb_cycles) %>% 
     count()
```

```{r}
sample_data(ps)$pcrb_cycles[is.na(sample_data(ps)$pcrb_cycles)] <- 8
```

##### Study day

Most accurate would be study *date*, but that would take more work. 

```{r}
# First, have to split ONR_day into week and weekday
samdf <- 
     samdf %>% 
     mutate(ONR_week = gsub(pattern = 'T|F',
                            replacement = '',
                            ONR_day),
            ONR_weekday = gsub(pattern = '\\d',
                               replacement = '',
                               ONR_day),
            ONR_weekday = ifelse(ONR_weekday == 'T',
                                 yes = 'Tuesday',
                                 no = 'Friday')) %>% 
     select(-ONR_day)
```

```{r}
# Calculate study integer day, assuming Monday of week 1 is day 1
samdf$CH_weekday_i <- 
     samdf$CH_weekday %>% 
     factor(levels = c("Monday", "Tuesday", "Wednesday", 
                       "Thursday", "Friday", "Saturday", "Sunday"),
            ordered = TRUE) %>% 
     as.integer()

samdf$ONR_weekday_i <- 
     samdf$ONR_weekday %>% 
     factor(levels = c("Monday", "Tuesday", "Wednesday", 
                       "Thursday", "Friday", "Saturday", "Sunday"),
            ordered = TRUE) %>% 
     as.integer()

# Convert types to numeric
samdf$ONR_week <- as.numeric(samdf$ONR_week)

# Calculate overall study day
samdf <- 
     samdf %>% 
     mutate(CH_day = (7*(CH_week - 1) + CH_weekday_i),
            ONR_day = (7*(ONR_week - 1) + ONR_weekday_i))
```

```{r}
# Replace in phyloseq object
sample_data(ps) <- column_to_rownames(samdf, var = 'row')

rm(samdf)
```

### Filter

#### Read counts

Keep samples with 0 counts for subsequent QC.

```{r}
# How many?
ps %>% 
     prune_samples(sample_sums(.) == 0, .) %>% 
     nsamples()
```

#### Plant taxa

Here, decide whether to proceed with the data as-is (all identified SVs), or analyze only those labeled to a food species.

##### Foods only

What percentage of reads in the dataset are labeled with a food taxon name?

```{r}
taxtab <- data.frame(ps@tax_table)
asvtab <- data.frame(ps@otu_table)

unassigned <- 
     taxtab %>% 
     filter(is.na(superkingdom)) %>% 
     row.names()

sum(asvtab[, unassigned])/sum(asvtab)
```

###### OPTIONAL: Update taxonomic assignment

```{r}
ref <- 
     here('..',
          'food-dbs',
          'data',
          'processed',
          'dada2-compatible',
          'trnL',
          '20210825_trnL_prior_format.fasta')
          # 'trnLGH.fasta')

taxtab.species <- MButils::assignSpecies_mod(taxa_names(ps),
                                             ref,
                                             tryRC = TRUE)
```

How many ASVs unassigned?
```{r}
unassigned <- taxtab.species$asv[is.na(taxtab.species$Species)]

# Percentage of sequence variants
length(unassigned)/dim(asvtab)[2]
```

```{r}
# Percentage of reads mapping to these unassigned species
sum(asvtab[, unassigned])/sum(asvtab)
```

Try implementing functions from taxa
Now want to build out whole tree
```{r}
# Separate ID from species name for querying
taxtab.species <- 
     taxtab.species %>%
     separate(col = Species,
              into = c('index', 'label'), sep = '\\s',
              extra = 'merge')

taxmap <- taxa::lookup_tax_data(taxtab.species, 
                                type = 'taxon_name', 
                                column = 'label')

taxonomy <- taxa::taxonomy_table(taxmap, 
                                 use_ranks = c('superkingdom', 
                                               'kingdom', 
                                               'phylum', 
                                               'order',
                                               'family',
                                               'genus',
                                               'species', 
                                               'subspecies',
                                               'varietas'))
```

Join to results
Need to do separate joins: One for subspecies-level designations, and then the remainder at the species level.
```{r}
# Variety
taxonomy.var <- filter(taxonomy, !is.na(varietas))
taxtab.var <- 
     taxtab.species %>% 
     left_join(taxonomy.var, by = c('label' = 'varietas')) %>% 
     filter(!is.na(species)) %>% 
     select(asv, superkingdom, kingdom, phylum, order, family,
            genus, species, varietas = label)

# Subspecies
taxonomy.sub <- filter(taxonomy, !is.na(subspecies))
taxtab.sub <- 
     taxtab.species %>% 
     left_join(taxonomy.sub, by = c('label' = 'subspecies')) %>% 
     filter(!is.na(species)) %>% 
     select(asv, superkingdom, kingdom, phylum, order, family,
            genus, species, subspecies = label)

taxtab.species <- 
     taxtab.species %>% 
     filter(!(asv %in% c(taxtab.var$asv,
                         taxtab.sub$asv))) %>% 
     left_join(taxonomy, by = c('label' = 'species')) %>%
     select(asv, 
            superkingdom, 
            kingdom, 
            phylum, 
            order, 
            family,
            genus, 
            species = label)

# Missing columns will be filled with NA
taxtab.species <- bind_rows(taxtab.species, 
                            taxtab.sub,
                            taxtab.var)
```

Find last common ancestor
```{r}
# Group by ASV
lca <- 
     taxtab.species %>%
     group_by(asv) %>%
     summarize_all(n_distinct) %>%
     column_to_rownames(var = 'asv')

# Now, relabel all those with >1 name at a particular level as NA
# As a placeholder, keep only the first species, knowing it will be overwritten
taxtab.species.lca <- 
     taxtab.species %>%
     group_by(asv) %>%
     summarize_all(first) %>%
     column_to_rownames(var = 'asv')
```

Now do relabeling 
```{r}
# Confirm ordering is okay
all(rownames(lca) == rownames(taxtab.species.lca))

taxtab.species.lca[lca > 1] = NA
taxtab.species.lca <- 
     rownames_to_column(taxtab.species.lca, var = 'asv')
```

```{r}
colSums(!is.na(taxtab.species.lca))/nrow(taxtab.species.lca)
```

```{r}
# blast <- 
#      taxtab.species %>%
#      filter(is.na(species)) %>%
#      pull(asv)
# 
# names(blast) <- as.character(seq_along(blast))

# blast <- Biostrings::DNAStringSet(blast)
# Biostrings::writeXStringSet(blast,
# here('data', 'processed', 'miniseq', '20201201',
#      '20210608_no_tax_assignment.fasta'))
```

Replace in phyloseq object

```{r}
taxdf <- 
     taxtab.species.lca %>%
     column_to_rownames(var = 'asv') %>%
     as.matrix()

ps.update <- ps

tax_table(ps.update) <- tax_table(taxdf)
```

###### OPTIONAL: Compare to prior

Does this substantially change the number of observed taxa per sample?
```{r}
# All ASVs
asvtab <- ps@otu_table@.Data
asvtab.update <- ps.update@otu_table@.Data

dim(asvtab)
dim(asvtab.update)
```

```{r}
# Food ASVs
asvtab <- 
     ps %>% 
     subset_taxa(!is.na(superkingdom)) %>% 
     otu_table()
asvtab <- asvtab@.Data

asvtab.update <- 
     ps.update %>% 
     subset_taxa(!is.na(superkingdom)) %>% 
     otu_table()
asvtab.update <- asvtab.update@.Data

dim(asvtab)
dim(asvtab.update)
```

```{r}
# Get observations per sample
obs.df <- 
     apply(asvtab, 1, function(x){sum(x > 0)}) %>% 
     data.frame(summer21 = .) %>% 
     rownames_to_column(var = 'row')

obs.df <- 
     apply(asvtab.update, 1, function(x){sum(x > 0)}) %>% 
     data.frame(aug21 = .) %>% 
     rownames_to_column(var = 'row') %>% 
     full_join(obs.df)
     
obs.df
```

```{r}
ggplot(obs.df, aes(x = summer21, y = aug21)) +
     geom_point(alpha = 0.5) +
     coord_equal() +
     labs(x = 'Summer 2021\n(mixed references)',
          y = 'August 2021\n(full set of manual additions)')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('..',
            'food-dbs',
            'results',
            paste0(plotID, '_Database update (Summer21 to Aug21).pdf')),
       height = 4, width = 4)
```


###### Subset

2021/02: -5 taxa from prior

```{r}
# Consider foods only
ps <- subset_taxa(ps, !is.na(superkingdom))
ps
```

After this step, need to check again for samples with 0 read counts, as their only reads may have been unlabeled.
```{r}
# Now, how many?
ps %>% 
     prune_samples(sample_sums(.) == 0, .) %>% 
     nsamples()
```

```{r}
# How many identifications at each phylogenetic level?
ps@tax_table@.Data %>% 
     apply(2, function(x){sum(!is.na(x))})
```


```{r}
# Clarify sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rename(reads_tot = reads)

# Replace in object
sample_data(ps) <- samdf

# Calculate food-only reads
sample_data(ps)$reads_food <- sample_sums(ps)
```

##### Glom names

```{r}
taxtab <- 
     data.frame(ps@tax_table) %>% 
     lowest_level() # Get lowest named level
```

```{r}
# How many duplicates?
taxtab$name[duplicated(taxtab$name)] %>% 
     unique() %>% 
     sort()
```

This is a bit fraught at higher levels, because depending on exact ASVs these families may actually map to different species. 

```{r}
# Update names at higher levels to reflect distinction

# Asteraceae (NEW <- ONLY)
taxtab['ATCACGTTTTCCGAAAACAAACAAAGGTTCAGAAAGCGAAAATAAAAAAG', 
       'name'] <- 
     'Asteraceae (sunflower, sunchoke, dandelion)'
taxtab['AAATCACGTTTTCCGAAAACAAACAAAGGTTCAGAAAGCGAAAAGAAAAAAAG', 
       'name'] <- 
     'Asteraceae (sunflower, sunchoke, dandelion)'

# Cucurbitaceae 
# These are subsequences; label by shortest one

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAGG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAGGG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

# Fabaceae (NEW REF ONLY)
# Subsequences
taxtab['ATCCTGTTTTCCGAAAACAAAGAAAAGTTCATAAAGTGATAATAAAAAAGG',
       'name'] <-
     'Fabaceae (soybean, guar)'

taxtab['AAATCCTGTTTTCCGAAAACAAAGAAAAGTTCATAAAGTGATAATAAAAAAGG',
       'name'] <-
     'Fabaceae (soybean, guar)'

# Distinct
taxtab['ATCCTGTTTTCTGAAAACAAAGAAAAATTCAGAAAGTTATAATAAAAAAGG', 
       'name'] <- 
     'Fabaceae (hyacinth bean, adzuki bean, moth bean, rice bean, black gram bean)'

# Lamiaceae (NEW REF ONLY)
taxtab['ATCCTGTTTTCTCAAAACAAAGGTTCAAAAAACGAAAAAAAAAG', 'name'] <- 
     'Lamiaceae (lemon balm, rosemary)'

taxtab['ATCCTGTTTTCTCAAAACAAAAGTTCAAAAAACGAAAAAAAAAAAAAG', 'name'] <- 
     'Lamiaceae (basil, catnip)'

taxtab['ATCCTGTTTTCTCAAAACAAAGGTTCAAAAAACGAAAAAAAAAAG', 'name'] <- 
     'Lamiaceae (korean mint, african basil, rosemary)'

# # Lauraceae (OLD REF ONLY)
# # Also subsequences
# taxtab['ATCCTGTTTTCAGAAAACAAGGGTTCAGAAAGCGAGAACCAAAAAAAGGATAG', 
#        'name'] <- 'Lauraceae (cinnamon, bay leaf, avocado)'
# 
# taxtab['ATCCTGTTTTCAGAAAACAAGGGTTCAGAAAGCGAGAACCAAAAAAAG', 
#        'name'] <- 'Lauraceae (cinnamon, bay leaf, avocado)'

# # Poaceae
# # Also subsequences, but should be labeled
# 
# taxtab['ATCCGTGTTTTGAGAAAACAAGGGGTTCTCGAACTAGAATACAAAGGAAAAG', 
#        'name'] <- 'Poaceae (rye, wheat)'
# 
# taxtab['AATCCGTGTTTTGAGAAAACAAGGGGTTCTCGAACTAGAATACAAAGGAAAAG', 
#        'name'] <- 'Poaceae (rye, wheat)'

# Rosaceae
# Not listed, but needs to be manually labeled because of pear cultivar
# that can't be looked up: propagates NAs
taxtab['ATCCTGTTTTATGAAAATAAACAAGGGTTTCATAAACCGAAAATAAAAAAG', 
       'name'] <- 'Rosaceae (apples, pears)'

taxtab['ATCCCGTTTTATGAAAACAAACAAGGGTTTCAGAAAGCGAGAATAAATAAAG', 
       'name'] <- 'Rosaceae (strawberry, brambles, rose)'

# Replace higher levels in taxonomy (need to fix this upstream)
rosaceae <- c('ATCCTGTTTTATGAAAATAAACAAGGGTTTCATAAACCGAAAATAAAAAAG',
                'ATCCCGTTTTATGAAAACAAACAAGGGTTTCAGAAAGCGAGAATAAATAAAG')

taxtab <- 
     taxtab %>% 
     rownames_to_column(var = 'asv') %>% 
     mutate(superkingdom = ifelse(asv %in% rosaceae),
                            yes = 'Eukaryota', no = superkingdom,
            kingdom = ifelse(asv %in% rosaceae),
                            yes = 'Viridiplantae', no = kingdom,
            superkingdom = ifelse(asv %in% rosaceae),
                            yes = 'Eukaryota', no = superkingdom,
            superkingdom = ifelse(asv %in% rosaceae),
                            yes = 'Eukaryota', no = superkingdom,
            superkingdom = ifelse(asv %in% rosaceae),
                            yes = 'Eukaryota', no = superkingdom,
                         
                 family = ifelse(asv %in% rosaceae,
                            yes = '))
```

```{r}
# Brassica
taxtab %>% filter(name == 'Brassica') %>% row.names()

taxtab['ATCCTGGGTTACGCGAACAAACCAGAGTTTAGAAAGCGG', 
       'name'] <- 'Brassica (juncea, rapa)'
taxtab['ATCTTGGGTTACGCGAACAAACCAGAGTTTAGAAAGCGG', 
       'name'] <- 'Brassica (napus, oleracea)'

# Carya-- substrings of each other
taxtab %>% filter(name == 'Carya') %>% row.names()

# Citrus-- substrings
taxtab %>% filter(name == 'Citrus') %>% row.names()
```

```{r}
# Replace taxonomy table in phyloseq object
tax_table(ps) <- as.matrix(taxtab)

# Now, these and all others can be merged by name
ps <- tax_glom(ps, taxrank = 'name')
ps
```

### Save

```{r}
saveRDS(ps,
        here('data', 
             'processed', 
             'phyloseq', 
             'trnL_combined_ps.rds'))
```
