---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp.trnL <- 
     here('data', 'processed', 'phyloseq', '20210222_ps_trnL_chomp.rds') %>% 
     readRDS()

ps.chomp.joint <- 
     here('data', 'processed', 'phyloseq', '20210428_ps_joint_chomp.rds') %>% 
     readRDS()
```

### ONR
```{r}
ps.onr <- 
     here('data', 'processed', 'phyloseq', '20210427_ps_trnL_onr.rds') %>% 
     readRDS()
```

# Pre-process data

## trnL

### Filter

```{r}
# Currently working only with trnL
# Keep only plant ASVs (the joint object has the metadata we need)

asvs.plant <- taxa_names(ps.chomp.trnL)
ps.chomp <- prune_taxa(asvs.plant, ps.chomp.joint)
rm(ps.chomp.trnL, ps.chomp.joint)

# Remove controls
ps.chomp <- subset_samples(ps.chomp, type == 'sample')
```

CHOMP phyloseq object contains Wednesday and Thursday samples sequenced on two separate runs; keep only the second of these (February 2021). 

```{r}
# Select only MiniSeq experiments of interest:
# Substituted Feb 2021 run in place of Mar 2020
ps.chomp <- 
    ps.chomp %>% 
    subset_samples(seq_date != '20200304') %>% 
    prune_taxa(taxa_sums(.) > 0, .)
```

### Synchronize metadata

```{r}
# Add study column for each
sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.onr)$study <- 'ONR'
```

```{r}
intersect(sample_variables(ps.chomp), sample_variables(ps.onr))
```

```{r}
# In CHOMP, but not in ONR
setdiff(sample_variables(ps.chomp), sample_variables(ps.onr))
```

```{r}
# In ONR, but not in CHOMP
setdiff(sample_variables(ps.onr), sample_variables(ps.chomp))
```

```{r}
# Reconcile differences thoughtfully :P
# TODO
```

### Merge

```{r}
ps <- merge_phyloseq(ps.chomp, ps.onr)
```

### Convert variable types

Weird, couldn't figure out how to do dates prior to merge_phyloseq step; was getting an error that origin wasn't specified, though I thought I had when initializing both objects.

```{r}
# Date type
sample_data(ps)$seq_date <- 
     as.Date(as.character(sample_data(ps)$seq_date),
             format = '%Y%m%d', origin = '1900-01-01')

# Factor type
sample_data(ps)$study <- factor(sample_data(ps)$study)
```

```{r}
saveRDS(ps, here('data', 'processed', 'phyloseq',
                 '20210408_ps_trnL.rds'))
```

