---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp <- 
     here('data', 'processed', 'phyloseq', '20210714_ps_trnL_CHOMP.rds') %>% 
     readRDS()

ps.chomp
```

### ONR

```{r}
ps.onr <- 
     here('data', 'processed', 'phyloseq', '20210617_ps_trnL_ONR.rds') %>% 
     readRDS()

ps.onr
```

### DFC

```{r}
ps.dfc <- 
     here('data', 'processed', 'phyloseq', '20210623_ps_trnL_DFC.rds') %>% 
     readRDS()

ps.dfc
```

### POMMS

```{r}
ps.pomms <- 
     here('data', 'processed', 'phyloseq', '20210816_ps_trnL_POMMS.rds') %>% 
     readRDS()

ps.pomms
```

### Tidy

#### Reduce sample data

To begin with, let's consider only study and other broad metadata variables:
* Age (pediatric/adolescent/adult)
* Diet (free-eating/weight loss/controlled feeding)

```{r}
# Populate all study variables
# (Tried to do before loading, but names not consistent, and ONR has some NAs)

sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.dfc)$study <- 'DFC'
sample_data(ps.onr)$study <- 'ONR'
sample_data(ps.pomms)$study <- 'POMMS'
```

```{r}
# Add demographics, as above
# Age
sample_data(ps.chomp)$age <- 'Adult'
sample_data(ps.dfc)$age <- 'Adult'
sample_data(ps.onr)$age <- 'Adult'
sample_data(ps.pomms)$age <- 'Adolescent'

# Diet
sample_data(ps.chomp)$diet <- 'Free-eating'
sample_data(ps.dfc)$diet <- 'Weight loss'
sample_data(ps.onr)$diet <- 'Free-eating'
sample_data(ps.pomms)$diet <- 'Free-eating'
```

#### Unique sample names

```{r}
# Otherwise, with well-based labeling, some might be combined in merge
sample_names(ps.chomp) <- paste(sample_data(ps.chomp)$study, 
                              seq(nsamples(ps.chomp)), sep = '-')

sample_names(ps.dfc) <- paste(sample_data(ps.dfc)$study, 
                              seq(nsamples(ps.dfc)), sep = '-')

sample_names(ps.onr) <- paste(sample_data(ps.onr)$study, 
                              seq(nsamples(ps.onr)), sep = '-')

sample_names(ps.pomms) <- paste(sample_data(ps.pomms)$study, 
                              seq(nsamples(ps.pomms)), sep = '-')
```

### Merge

Note that initially when I did this merge I had 1035 samples, and now I have 20 more.  I think this is due to the fact that I now preserve samples with 0 reads in the phyloseq objects, even if they fell out earlier in the pipeline.

```{r}
ps <- merge_phyloseq(ps.chomp, ps.dfc, ps.onr, ps.pomms)
ps
```

```{r}
# Adjust sample variables en masse for subsequent plots
sample_data(ps)$diet <- factor(sample_data(ps)$diet, 
                               levels = c('Free-eating', 'Weight loss'))

sample_data(ps)$reads <- sample_sums(ps)

sample_data(ps)$study <- 
     factor(sample_data(ps)$study, 
            levels = c('POMMS', 'CHOMP', 'ONR', 'DFC'), 
            labels = c('Adolescent', 'Adult-1', 'Adult-2', 'Weight Loss'))
```

Note-- also might be interesting to try this ordination not just on ASVs identified to a food. Because of two features:
* ASVs not identified make up the minority of the total reads in the dataset, so they shouldn't affect the ordination much (at least not for PCA on CLR-transformed counts)
* However, ASVs that do map to a food, but aren't yet labeled as such, will be shared across cohorts and contribute meaningfully to the ordination, which is what will ultimately happen when I label them.

So I think reasonable to at least try an ordination on the full dataset (~1000 ASVs) for comparison to the one on the food-only dataset (~10% of that number). 

### Filter

#### Read counts

```{r}
# Remove samples with 0 read counts
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps
```

#### Plant taxa

Here, decide whether to proceed with the data as-is (all identified SVs), or analyze only those labeled to a food species.

##### Foods only

What percentage of reads in the dataset are labeled with a food taxon name?

```{r}
taxtab <- data.frame(ps@tax_table)
asvtab <- data.frame(ps@otu_table)

unassigned <- 
     taxtab %>% 
     filter(is.na(superkingdom)) %>% 
     row.names()

sum(asvtab[, unassigned])/sum(asvtab)
```

```{r}
# Consider foods only
ps <- subset_taxa(ps, !is.na(superkingdom))
ps
```

After this step, need to check again for samples with 0 read counts, as their only reads may have been unlabeled.
```{r}
# Remove samples with 0 read counts
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps
```

```{r}
# How many identifications at each phylogenetic level?
ps@tax_table@.Data %>% 
     apply(2, function(x){sum(!is.na(x))})
```


```{r}
# Clarify sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rename(reads_tot = reads)

# Replace in object
sample_data(ps) <- samdf

# Calculate food-only reads
sample_data(ps)$reads_food <- sample_sums(ps)
```

##### Glom names

```{r}
taxtab <- 
     data.frame(ps@tax_table) %>% 
     lowest_level() # Get lowest named level
```

```{r}
# How many duplicates?
taxtab$name[duplicated(taxtab$name)]
```

This is a bit fraught at higher levels, because depending on exact ASVs these families may actually map to different species.  Ideally go back and do this by exact ASV identity, but for now, try merging everything.

```{r}
# Replace taxonomy table in phyloseq object
tax_table(ps) <- as.matrix(taxtab)

# Glom at this new level
ps <- tax_glom(ps, taxrank = 'name')
ps
```

### Save

```{r}
# saveRDS(ps,
#         here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds'))
```
