---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20210714_ps_trnL_CHOMP.rds') %>% 
     readRDS()

ps.chomp
```

### ONR

```{r}
ps.onr <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20210617_ps_trnL_ONR.rds') %>% 
     readRDS()

ps.onr
```

### DFC

```{r}
ps.dfc <- 
     here('data', 
          'processed',
          'phyloseq', 
          '20211116_ps_trnL_DFC.rds') %>% 
     readRDS()

ps.dfc
```

### POMMS

```{r}
ps.pomms <- 
     here('data', 
          'processed', 
          'phyloseq', 
          '20211221_ps_trnL_POMMS.rds') %>% 
     readRDS()

ps.pomms
```

### Tidy

#### Add variables

To begin with, let's consider only study and other broad metadata variables:
* Age (pediatric/adolescent/adult)
* Diet (free-eating/weight loss/controlled feeding)

```{r}
# Populate all study variables
# (Tried to do before loading, but names not consistent, and ONR has some NAs)

sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.dfc)$study <- 'DFC'
sample_data(ps.onr)$study <- 'ONR'
sample_data(ps.pomms)$study <- 'POMMS'
```

#### Unique sample names

```{r}
# Otherwise, with well-based labeling, some might be combined in merge
sample_names(ps.chomp) <- paste(sample_data(ps.chomp)$study, 
                              seq(nsamples(ps.chomp)), sep = '-')

sample_names(ps.dfc) <- paste(sample_data(ps.dfc)$study, 
                              seq(nsamples(ps.dfc)), sep = '-')

sample_names(ps.onr) <- paste(sample_data(ps.onr)$study, 
                              seq(nsamples(ps.onr)), sep = '-')

sample_names(ps.pomms) <- paste(sample_data(ps.pomms)$study, 
                              seq(nsamples(ps.pomms)), sep = '-')
```

### Merge

Note that initially when I did this merge I had 1035 samples, and now I have 20 more.  I think this is due to the fact that I now preserve samples with 0 reads in the phyloseq objects, even if they fell out earlier in the pipeline.

```{r}
ps <- merge_phyloseq(ps.chomp, ps.dfc, ps.onr, ps.pomms)
ps
```

```{r}
# Adjust sample variables en masse for subsequent plots
sample_data(ps)$reads <- sample_sums(ps)

sample_data(ps)$study <- 
     factor(sample_data(ps)$study, 
            levels = c('DFC', 'CHOMP', 'ONR', 'POMMS'), 
            labels = c('Weight Loss', 'Adult-1', 'Adult-2', 'Adolescent'))
```

### Tidy

#### Add variables

PCR B cycle number only added once changed from 8 to 10 cycles (early 2021). Populate remaining values with 8 cycles.

```{r}
samdf <- data.frame(ps@sam_data)

samdf %>% 
     group_by(seq_date, pcrb_cycles) %>% 
     count()
```

```{r}
sample_data(ps)$pcrb_cycles[is.na(sample_data(ps)$pcrb_cycles)] <- 8
```

### Filter

#### Read counts

Keep samples with 0 counts for subsequent QC.

```{r}
# How many?
ps %>% 
     prune_samples(sample_sums(.) == 0, .) %>% 
     nsamples()
```

#### Plant taxa

Here, decide whether to proceed with the data as-is (all identified SVs), or analyze only those labeled to a food species.

##### Foods only

What percentage of reads in the dataset are labeled with a food taxon name?

```{r}
taxtab <- data.frame(ps@tax_table)
asvtab <- data.frame(ps@otu_table)

unassigned <- 
     taxtab %>% 
     filter(is.na(superkingdom)) %>% 
     row.names()

sum(asvtab[, unassigned])/sum(asvtab)
```

```{r}
# Consider foods only
ps <- subset_taxa(ps, !is.na(superkingdom))
ps
```

After this step, need to check again for samples with 0 read counts, as their only reads may have been unlabeled.
```{r}
# Now, how many?
ps %>% 
     prune_samples(sample_sums(.) == 0, .) %>% 
     nsamples()
```

```{r}
# How many identifications at each phylogenetic level?
ps@tax_table@.Data %>% 
     apply(2, function(x){sum(!is.na(x))})
```


```{r}
# Clarify sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rename(reads_tot = reads)

# Replace in object
sample_data(ps) <- samdf

# Calculate food-only reads
sample_data(ps)$reads_food <- sample_sums(ps)
```

##### Glom names

```{r}
taxtab <- 
     data.frame(ps@tax_table) %>% 
     lowest_level() # Get lowest named level
```

```{r}
# How many duplicates?
taxtab$name[duplicated(taxtab$name)]
```

This is a bit fraught at higher levels, because depending on exact ASVs these families may actually map to different species. 

```{r}
# Update names at higher levels to reflect distinction
# Cucurbitaceae 
# These are subsequences; label by shortest one

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAGG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

taxtab['ATCCTTTTTCCGAAAACAAAAAAAAGGG', 
       'name'] <- 'Cucurbitaceae (pumpkin, squash, gourd, Chinese okra)'

# Lauraceae
# Also subsequences
taxtab['ATCCTGTTTTCAGAAAACAAGGGTTCAGAAAGCGAGAACCAAAAAAAGGATAG', 
       'name'] <- 'Lauraceae (cinnamon, bay leaf, avocado)'

taxtab['ATCCTGTTTTCAGAAAACAAGGGTTCAGAAAGCGAGAACCAAAAAAAG', 
       'name'] <- 'Lauraceae (cinnamon, bay leaf, avocado)'

# Poaceae
# Also subsequences, but should be labeled

taxtab['ATCCGTGTTTTGAGAAAACAAGGGGTTCTCGAACTAGAATACAAAGGAAAAG', 
       'name'] <- 'Poaceae (rye, wheat)'

taxtab['AATCCGTGTTTTGAGAAAACAAGGGGTTCTCGAACTAGAATACAAAGGAAAAG', 
       'name'] <- 'Poaceae (rye, wheat)'

# Rosaceae
taxtab['ATCCTGTTTTATGAAAATAAACAAGGGTTTCATAAACCGAAAATAAAAAAG', 
       'name'] <- 'Rosaceae (apples, pears)'

taxtab['ATCCCGTTTTATGAAAACAAACAAGGGTTTCAGAAAGCGAGAATAAATAAAG', 
       'name'] <- 'Rosaceae (strawberry, brambles, rose)'

```

```{r}
# Replace taxonomy table in phyloseq object
tax_table(ps) <- as.matrix(taxtab)

# Now, these and all others can be merged by name
ps <- tax_glom(ps, taxrank = 'name')
ps
```

### Save

```{r}
saveRDS(ps,
        here('data', 
             'processed', 
             'phyloseq', 
             'trnL_combined_ps.rds'))
```
