---
title: "Joint phyloseq"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(MButils) # for lowest_level naming
library(phyloseq)
library(tidyverse)
```

# Load input data

## trnL metabarcoding

### CHOMP

```{r}
ps.chomp <- 
     here('data', 'processed', 'phyloseq', '20210714_ps_trnL_CHOMP.rds') %>% 
     readRDS()

ps.chomp
```

### ONR

```{r}
ps.onr <- 
     here('data', 'processed', 'phyloseq', '20210617_ps_trnL_ONR.rds') %>% 
     readRDS()

ps.onr
```

### DFC

```{r}
ps.dfc <- 
     here('data', 'processed', 'phyloseq', '20210623_ps_trnL_DFC.rds') %>% 
     readRDS()

ps.dfc
```

### POMMS

```{r}
ps.pomms <- 
     here('data', 'processed', 'phyloseq', '20210612_ps_trnL_POMMS.rds') %>% 
     readRDS()

ps.pomms
```

### Tidy

#### Reduce sample data

To begin with, let's consider only study and other broad metadata variables:
* Age (pediatric/adolescent/adult)
* Diet (free-eating/weight loss/controlled feeding)

```{r}
# Populate all study variables
# (Tried to do before loading, but names not consistent, and ONR has some NAs)

sample_data(ps.chomp)$study <- 'CHOMP'
sample_data(ps.dfc)$study <- 'DFC'
sample_data(ps.onr)$study <- 'ONR'
sample_data(ps.pomms)$study <- 'POMMS'
```

```{r}
# Add demographics, as above
# Age
sample_data(ps.chomp)$age <- 'Adult'
sample_data(ps.dfc)$age <- 'Adult'
sample_data(ps.onr)$age <- 'Adult'
sample_data(ps.pomms)$age <- 'Adolescent'

# Diet
sample_data(ps.chomp)$diet <- 'Free-eating'
sample_data(ps.dfc)$diet <- 'Weight loss'
sample_data(ps.onr)$diet <- 'Free-eating'
sample_data(ps.pomms)$diet <- 'Free-eating'
```

#### Unique sample names

```{r}
# Otherwise, with well-based labeling, some might be combined in merge
sample_names(ps.chomp) <- paste(sample_data(ps.chomp)$study, 
                              seq(nsamples(ps.chomp)), sep = '-')

sample_names(ps.dfc) <- paste(sample_data(ps.dfc)$study, 
                              seq(nsamples(ps.dfc)), sep = '-')

sample_names(ps.onr) <- paste(sample_data(ps.onr)$study, 
                              seq(nsamples(ps.onr)), sep = '-')

sample_names(ps.pomms) <- paste(sample_data(ps.pomms)$study, 
                              seq(nsamples(ps.pomms)), sep = '-')
```

### Merge

Note that initially when I did this merge I had 1035 samples, and now I have 20 more.  I think this is due to the fact that I now preserve samples with 0 reads in the phyloseq objects, even if they fell out earlier in the pipeline.

```{r}
ps <- merge_phyloseq(ps.chomp, ps.dfc, ps.onr, ps.pomms)
ps
```

```{r}
# Adjust sample variables en masse for subsequent plots
sample_data(ps)$diet <- factor(sample_data(ps)$diet, 
                               levels = c('Free-eating', 'Weight loss'))

sample_data(ps)$reads <- sample_sums(ps)

sample_data(ps)$reads_binned <- cut(sample_data(ps)$reads, 
                                    breaks = c(-1, 1001, 5001, 10001, 30001,
                                               max(sample_data(ps)$reads) + 1), 
                                    labels = c('<1,000 reads', '1,000-5,000', 
                                               '5,000-10,000', '10,000-30,000',
                                               '>30,000'))
```

### Save

```{r}
# saveRDS(ps,
#         here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds'))
```


