---
title: "Mantel tests"
output: html_notebook
---

There are a few different ways these comparisons could be done.  List them out:
- DHQ3 (1 observation per participant)
     + Compare each DHQ3 to averaged trnL compositions
     - Replicate each DHQ3 up to match the number of trnL samples from a given participant
- ASA24 (2-3 observations per participant)
     + Average all ASA24s, and compare them to averaged trnL compositions
     - Try to match ASA24s to stool samples

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(tidyverse)
library(vegan)
```

# Load input data

## trnL metabarcoding

```{r}
# Joint phyloseq object
ps <- 
     here('data', 'processed', 'phyloseq', '20210428_ps_trnL.rds') %>% 
     readRDS()

ps
```

## DHQ3

### CHOMP

#### FNDDS nutrients

```{r}
# All FNDDS nutrients (64)
dhq3.nut.chomp <- 
     here('data', 'processed', 'DHQ3',
          '20201022_DHQ3 kcal-normalized FNDDS nutrients_chomp.rds') %>% 
     readRDS()
```

#### Food groups

```{r} 
# All food groups (41)
dhq3.foodgroups.chomp <- 
     here('data', 'processed', 'DHQ3', 
          '20201022_DHQ3 kcal-normalized food groups_chomp.rds') %>% 
     readRDS()

# Only plant-derived food groups (but NOT including oils, solid fats, sweeteners, or alcohols) (22)
dhq3.plantgroups.chomp <- 
     here('data', 'processed', 'DHQ3', 
          '20201022_DHQ3 kcal-normalized plant food groups_chomp.rds') %>% 
     readRDS()
```

### ONR

#### FNDDS nutrients

```{r}
# All FNDDS nutrients (64)
dhq3.nut.onr <- 
     here('data', 'processed', 'DHQ3',
          '20210305_DHQ3 kcal-normalized FNDDS nutrients_onr.rds') %>% 
     readRDS()
```

#### Food groups

```{r}
# All food groups (41)
dhq3.foodgroups.onr <- 
     here('data', 'processed', 'DHQ3', 
          '20210305_DHQ3 kcal-normalized food groups_onr.rds') %>% 
     readRDS()

# Only plant-derived food groups (but NOT including oils, solid fats, sweeteners, or alcohols) (22)
dhq3.plantgroups.onr <-
     here('data', 'processed', 'DHQ3', 
          '20210305_DHQ3 kcal-normalized plant food groups_onr.rds') %>% 
     readRDS()
```

## ASA24 

### CHOMP

#### Nutrients

```{r}
asa24.nut.chomp <- 
     here('data', 'processed', 'ASA24', 'all records',
          '20210405_Nutrients, kcal-normalized, all records_chomp.rds') %>% 
     readRDS()

asa24.nut.1d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 1 day prior_CHOMP.rds') %>% 
     readRDS()

asa24.nut.2d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 2 day prior_CHOMP.rds') %>% 
     readRDS()

asa24.nut.3d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 3 day prior_CHOMP.rds') %>% 
     readRDS()
```

#### Food groups

```{r}
# All
asa24.foodgroups.chomp <- 
     here('data', 'processed', 'ASA24', 'all records',
          '20210405_Food groups, kcal-normalized, all records_chomp.rds') %>% 
     readRDS()

asa24.foodgroups.1d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records', 
          'Food groups, kcal-normalized, 1 day prior_CHOMP.rds') %>% 
     readRDS()

asa24.foodgroups.2d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records', 
          'Food groups, kcal-normalized, 2 day prior_CHOMP.rds') %>% 
     readRDS()

asa24.foodgroups.3d.chomp <- 
     here('data', 'processed', 'ASA24', 'paired records', 
          'Food groups, kcal-normalized, 3 day prior_CHOMP.rds') %>% 
     readRDS()
```

### ONR

#### Nutrients

```{r}
asa24.nut.onr <-
     here('data', 'processed', 'ASA24', 'all records',
          '20210405_Nutrients, kcal-normalized, all records_onr.rds') %>%
     readRDS()

asa24.nut.1d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 1 day prior_ONR.rds') %>% 
     readRDS()

asa24.nut.2d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 2 day prior_ONR.rds') %>% 
     readRDS()

asa24.nut.3d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Nutrients, kcal-normalized, 3 day prior_ONR.rds') %>% 
     readRDS()
```

#### Food groups

```{r}
# All
asa24.foodgroups.onr <- 
     here('data', 'processed', 'ASA24', 'all records',
          '20210405_Food groups, kcal-normalized, all records_onr.rds') %>% 
     readRDS()

asa24.foodgroups.1d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Food groups, kcal-normalized, 1 day prior_ONR.rds') %>% 
     readRDS()

asa24.foodgroups.2d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records',
          'Food groups, kcal-normalized, 2 day prior_ONR.rds') %>% 
     readRDS()

asa24.foodgroups.3d.onr <- 
     here('data', 'processed', 'ASA24', 'paired records', 
          'Food groups, kcal-normalized, 3 day prior_ONR.rds') %>% 
     readRDS()
```

# Prepare

## trnL

```{r}
# Filter out any samples with zero counts
# Necessary before converting to relative abundances
ps <- 
     ps %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)
```

### Merge

```{r}
# Add pseudocount
otu_table(ps) <- otu_table(ps) + 1
```

### Convert to relative abundance

```{r}
# Convert to relative abundance
ps.ra <- transform_sample_counts(ps, function(x){x/sum(x)}) 
```

### Merge within each subject

```{r}
ps.ra.merged <- merge_samples(ps.ra, 'subj')

# Now, need to normalize by total number of samples for each subject to return sums to 1
ps.ra.merged <- transform_sample_counts(ps.ra.merged,
                                        function(x){x/sum(x)})
```

### CLR transform

```{r}
# CLR
asvtab.merged <- otu_table(ps.ra.merged)@.Data
asvtab.merged.clr <- driver::clr(asvtab.merged)

# Replace names, but remember these aren't interpretable as strict ASVs any longer
colnames(asvtab.merged.clr) <- colnames(asvtab.merged)
```

### Rename 

CHOMP already has duplicates remove, so can rename as-is.  However, ONR has 13 samples without a confirmed study day, that I remove here since I can't be certain when comparing to ASA24 study day. Note that these are preserved in the trnL "averages" created for the Mantel against DHQ3 data. 

```{r}
ps
ps <- subset_samples(ps, !is.na(study_day))
ps
```

```{r}
# Rename samples with SUBJ_STUDY DAY to match naming pattern of the "# days prior" files
head(sample_names(ps))
sample_names(ps) <- paste(sample_data(ps)$subj, sample_data(ps)$study_day, 
                          sep = '_')
head(sample_names(ps))
```

```{r}
# CLR-transform now that these samples removed
asvtab <- otu_table(ps)@.Data
asvtab.clr <- driver::clr(asvtab)

# Replace names, but remember these aren't interpretable as strict ASVs any longer
colnames(asvtab.clr) <- colnames(asvtab)
```

## DHQ3

### Tidy

```{r}
# Concatenate dataframes
dhq3.nut <- bind_rows(dhq3.nut.chomp, dhq3.nut.onr)
dhq3.foodgroups <- bind_rows(dhq3.foodgroups.chomp, dhq3.foodgroups.onr)
dhq3.plantgroups <- bind_rows(dhq3.plantgroups.chomp, dhq3.plantgroups.onr)

# Remove original variables
rm(list = ls()[grep('.chomp$|.onr$', ls())])
```

```{r}
# Check completeness
any(is.na(dhq3.nut))
any(is.na(dhq3.foodgroups))
any(is.na(dhq3.plantgroups))
```

Ok, DHQ3 appears to have missing data:

```{r}
missing <- apply(dhq3.nut, 1, function(x){any(is.na(x))})
which(missing)
```

```{r}
# Remove
dhq3.nut <- dhq3.nut[!missing, ]
dhq3.foodgroups <- dhq3.foodgroups[!missing, ]
dhq3.plantgroups <- dhq3.plantgroups[!missing, ]
```

```{r}
# Confirm
any(is.na(dhq3.nut))
any(is.na(dhq3.foodgroups))
any(is.na(dhq3.plantgroups))
```

## ASA24 

### Tidy

```{r}
# Concatenate dataframes
asa24.nut <- bind_rows(asa24.nut.chomp, asa24.nut.onr)
asa24.foodgroups <- bind_rows(asa24.foodgroups.chomp, asa24.foodgroups.onr)

asa24.nut.1d <- bind_rows(asa24.nut.1d.chomp, asa24.nut.1d.onr)
asa24.foodgroups.1d <- bind_rows(asa24.foodgroups.1d.chomp, asa24.foodgroups.1d.onr)

asa24.nut.2d <- bind_rows(asa24.nut.2d.chomp, asa24.nut.2d.onr)
asa24.foodgroups.2d <- bind_rows(asa24.foodgroups.2d.chomp, asa24.foodgroups.2d.onr)

asa24.nut.3d <- bind_rows(asa24.nut.3d.chomp, asa24.nut.3d.onr)
asa24.foodgroups.3d <- bind_rows(asa24.foodgroups.3d.chomp, asa24.foodgroups.3d.onr)

# Remove original variables
rm(list = ls()[grep('.chomp$|.onr$', ls())])
```

```{r}
# Check completeness
any(is.na(asa24.nut))
any(is.na(asa24.foodgroups))

any(is.na(asa24.nut.1d))
any(is.na(asa24.foodgroups.1d))

any(is.na(asa24.nut.2d))
any(is.na(asa24.foodgroups.2d))

any(is.na(asa24.nut.3d))
any(is.na(asa24.foodgroups.3d))
```

### Average

Create an averaged version where each participants' ASAs are grouped into one.

```{r}
# Helper function for averaging multiple recalls from the same subject
average_asa <- function(input){
     # Assumes rownames are of the format SUBJ_RECALL
     input %>% 
     rownames_to_column(var = 'asa') %>% 
     separate(asa, into = c('subj', 'recall'), sep = '_') %>% 
     select(-recall) %>% 
     group_by(subj) %>% 
     summarize(across(everything(), mean)) %>% 
     column_to_rownames(var = 'subj')
}
```

```{r}
# Make averaged, centered, and scaled versions of the data
asa24.nut.avg <- 
     average_asa(asa24.nut) %>% 
     scale()
asa24.foodgroups.avg <- 
     average_asa(asa24.foodgroups) %>% 
     scale()
```

# Mantel

## ASA24 x DHQ3

```{r}
# Subset to shared samples
shared <- intersect(row.names(asa24.nut.avg),
                    row.names(dhq3.nut))

length(shared)
```

```{r}
# Calculate distance 
dist.asa24.nut <- 
     dist(asa24.nut.avg[shared, ], method = 'euclidean') 
dist.asa24.foodgroups <- 
     dist(asa24.foodgroups.avg[shared, ], method = 'euclidean') 

dist.dhq3.nut <- 
     dist(dhq3.nut[shared, ], method = 'euclidean') 
dist.dhq3.foodgroups <- 
     dist(dhq3.foodgroups[shared, ], method = 'euclidean') 
```

```{r}
# Run Mantel
mantel(dist.asa24.nut, dist.dhq3.nut)
```

```{r}
# Run Mantel
mantel(dist.asa24.foodgroups, dist.dhq3.foodgroups)
```


## trnL x ASA24

### Averaged ASA
```{r}
# Subset to shared samples
shared <- intersect(row.names(asa24.nut.avg),
                    sample_names(ps.ra.merged))

length(shared)
```

```{r}
# Euclidean distance on CLR-transformed counts
dist.trnL <- dist(asvtab.merged.clr[shared, ], method = 'euclidean')
```

Note that I haven't centered or scaled the ASA24 data.  Would that be important to do?? I think scaling would be important so differences in one variable don't drive all others.

```{r}
# Define plant and animal variables
plant <- 
     asa24.foodgroups %>% 
     select(matches('^F_|G_|PF_LEGUMES|PF_NUTSDS|PF_SOY|V_')) %>% 
     names()

animal <- 
     asa24.foodgroups %>% 
     select(matches('^D_|PF_CUREDMEAT|PF_EGGS|PF_MEAT|PF_ORGAN|PF_POULT|PF_SEAFD|SOLID_FATS')) %>% 
     names()
```


```{r}
# Within-ASA24
dist.asa24.nut <- dist(asa24.nut.avg[shared, ], method = 'euclidean')
dist.asa24.plant <- dist(asa24.foodgroups.avg[shared, plant], method = 'euclidean') 
dist.asa24.animal <- dist(asa24.foodgroups.avg[shared, animal], method = 'euclidean') 
```

```{r}
# Run Mantel
mantel(dist.asa24.nut, dist.trnL)
```

```{r}
mantel(dist.asa24.plant, dist.trnL)
```

```{r}
mantel(dist.asa24.animal, dist.trnL)
```

### Paired ASA-stool samples

#### 1 day lag
```{r}
# Subset to shared samples
shared <- intersect(row.names(asa24.nut.1d),
                    sample_names(ps))

length(shared)
```

```{r}
# Calculate distances
# Euclidean distance on CLR-transformed counts
dist.trnL <- dist(asvtab.clr[shared, ], method = 'euclidean')

# Within-ASA24
dist.asa24.nut <- 
     dist(scale(asa24.nut.1d[shared, ]), method = 'euclidean')
dist.asa24.plant <- 
     dist(scale(asa24.foodgroups.1d[shared, plant]), method = 'euclidean') 
dist.asa24.animal <- 
     dist(scale(asa24.foodgroups.1d[shared, animal]), method = 'euclidean') 
```

```{r}
# Run Mantel
mantel(dist.asa24.nut, dist.trnL)
mantel(dist.asa24.plant, dist.trnL)
mantel(dist.asa24.animal, dist.trnL)
```

#### 2 days lag
```{r}
# Subset to shared samples
shared <- intersect(row.names(asa24.nut.2d),
                    sample_names(ps))

length(shared)
```

```{r}
# Calculate distances
# Euclidean distance on CLR-transformed counts
dist.trnL <- dist(asvtab.clr[shared, ], method = 'euclidean')

# Within-ASA24
dist.asa24.nut <- 
     dist(scale(asa24.nut.2d[shared, ]), method = 'euclidean')
dist.asa24.plant <- 
     dist(scale(asa24.foodgroups.2d[shared, plant]), method = 'euclidean') 
dist.asa24.animal <- 
     dist(scale(asa24.foodgroups.2d[shared, animal]), method = 'euclidean') 
```

```{r}
# Run Mantel
mantel(dist.asa24.nut, dist.trnL)
mantel(dist.asa24.plant, dist.trnL)
mantel(dist.asa24.animal, dist.trnL)
```

#### 3 days lag
```{r}
# Subset to shared samples
shared <- intersect(row.names(asa24.nut.3d),
                    sample_names(ps))

length(shared)
```

```{r}
# Calculate distances
# Euclidean distance on CLR-transformed counts
dist.trnL <- dist(asvtab.clr[shared, ], method = 'euclidean')

# Within-ASA24
dist.asa24.nut <- 
     dist(scale(asa24.nut.3d[shared, ]), method = 'euclidean')
dist.asa24.plant <- 
     dist(scale(asa24.foodgroups.3d[shared, plant]), method = 'euclidean') 
dist.asa24.animal <- 
     dist(scale(asa24.foodgroups.3d[shared, animal]), method = 'euclidean') 
```

```{r}
# Run Mantel
mantel(dist.asa24.nut, dist.trnL)
mantel(dist.asa24.plant, dist.trnL)
mantel(dist.asa24.animal, dist.trnL)
```

## trnL x DHQ3

## Match samples 
```{r}
# Subset to shared samples
shared <- intersect(row.names(dhq3.nut),
                    sample_names(ps.ra.merged))

length(shared)
```

## Calculate distance

### Within-trnL

```{r}
# Euclidean distance on CLR-transformed counts
dist.trnL <- dist(asvtab.clr[shared, ], method = 'euclidean')
```

### Within-DHQ3

```{r}
dist.dhq3.nut <- dist(dhq3.nut[shared, ], method = 'euclidean')
dist.dhq3.plantgroups <- dist(dhq3.plantgroups[shared, ], method = 'euclidean') 
```

## Run

```{r}
mantel(dist.trnL, dist.dhq3.nut)
```

```{r}
mantel(dist.trnL, dist.dhq3.plantgroups)
```

```{r}
# Off the cuff: animal food groups
animal <- c(19:28, 34:41) # Columns corresponding to animal foods
colnames(dhq3.foodgroups[, animal])
dist.dhq3.animal <- dist(dhq3.foodgroups[shared, animal], method = 'euclidean') 

mantel(dist.trnL, dist.dhq3.animal)
```

