---
title: "Plant taxa"
output: html_notebook
---

Used to make Figure 1B.

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(ape) # For phylogenetic trees
library(Biostrings)
library(ggtree)
library(ggtreeExtra)
library(here)
library(phyloseq)
library(MButils) # For lowest level naming
library(metacoder)
library(Taxonstand) # For synchronizing plant names
library(tidyverse)
library(vegan)
library(UpSetR)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed',
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

## Reference database

```{r}
db <- 
     here('..', 
          'food-dbs', 
          'data', 
          'processed',
          'dada2-compatible',
          'trnL',
          'trnLGH.fasta') %>% 
     readDNAStringSet()
```

#### PhyloFood

This follows on tree presented by Ruben Milla in "Crop Origins and Phylo Food: A database and a phylogenetic tree to stimulate comparative analyses on the origins of food crops."

From README:
This folder contains five files:
A readme file (readme_ phylo_food_live.txt);
The R code used to build the tree (phylo_food_live_code.txt);
the literature sources cited in the code and used to resolve the tree, in BibTeX format (sources_used_to_resolve_phylo_food_live.txt);
the time-calibrated and fully resolved phylogenetic tree of crop species (phylo_food_live.tre);
the non-fully resolved phylogenetic tree of vascular plants, including the 866 species of crops (phylo_GBOTB_with_crops_live.tre).

```{r}
crop.dir <- 
     here('data',
          'processed',
          'reference',
          'Crop_Origins_Phylo',
          'Crop_Origins_Phylo_v_live')

metadata <- 
     here(crop.dir,
          'crop_origins_v_live',
          'crop_origins_live_db.csv') %>% 
     read_delim(delim = ';')

tree <- 
     here(crop.dir,
          'phylo_food_v_live',
          'phylo_food_live.tre') %>% 
     ape::read.tree()
```

## FAO name mapping

from Martin et al., PLOS ONE, 2019

```{r}
fao.names <- 
     here('..',
          'food-dbs',
          'data',
          'processed',
          'parsed-refs',
          'Martin_PLOSONE_2019_FAOSTAT to taxon mapping.csv') %>% 
     read_csv()
```

```{r}
dim(fao.names)
```

# Pre-process

## trnL

### Detected/not detected

```{r}
melt <- psmelt(ps)
```

```{r}
detections <- 
     melt %>% 
     group_by(superkingdom, 
              kingdom, 
              phylum,
              order,
              family, 
              genus,
              species) %>% # Only go to species, since this is what's in tree
     summarize(abundance = sum(Abundance)) %>% 
     mutate(in_data = abundance > 0)

detections
```
## Tree data

### CropOrigins

Here, likely want to subset to only species used as food.

```{r}
names(metadata)
```

```{r}
# Get entries for use categories
metadata %>% 
     select(starts_with('use')) %>% 
     apply(2, unique)
```

```{r}
metadata %>% 
     group_by(use_food) %>% 
     count()
```
```{r}
# Join to FAO names
# Note that some of the corresponding species are duplicated: just pick the first entry for simplicity
dim(fao.names)
n_distinct(fao.names$Species)

fao.names <- 
     fao.names %>% 
     group_by(Species) %>% 
     summarize(across(everything(),
                      first))

# Now can join without introducing duplicates
metadata <- 
     left_join(metadata, fao.names,
               by = c('species_name' = 'Species'))
```

### PhyloFood

```{r}
str(tree)
```

Interesting that internal nodes aren't labeled here.  This would be nice because some metabarcoding entries will terminate internally in the tree. Or could I just display species, without collapsing up to last common ancestor?

```{r}
# Filter to only those used as food, get names in same format as tree
crops <- 
     metadata %>% 
     filter(use_food == 1) %>% 
     pull(species_name) %>% 
     gsub(pattern = ' ', replacement = '_')

head(crops)
```

```{r}
# Subset tree to these taxa only
tree <- keep.tip(tree, crops)

tree
rm(crops)
```

#### Taxonomy

Taxonomic information up to order level already present!! Clean up structure:

```{r}
# Filter to only those used as food
crops <- 
     metadata %>% 
     filter(use_food == 1) %>% 
     select(id, 
            common_name_crop,
            species_name,
            species_name_,
            order, 
            family,
            genus,
            sps,
            `FAO group code`) %>% 
     mutate(species = paste(genus, sps)) %>% 
     select(-sps)
```

##### Manual corrections

```{r}
# Noticed in joins below
# Chenopodiaceae labeled as Amaranthaceae for quinoa
crops$family[crops$genus == 'Chenopodium'] <- 'Chenopodiaceae'
```

```{r}
# How many major crops? (Lines with an FAO group code entry?)
crops %>% 
     filter(!is.na(`FAO group code`)) %>% 
     dim()
```

This was lower than I expected so I also tested as a join from raw FAO data with the same result (177).  So now, FAO group code is only present if the listed species is a major crop.

### Reference

Organize in a taxonomy-table style.

```{r}
db
```

```{r}
db.df <- 
     data.frame(name = names(db),
                seq = as.character(db))

db.df
```

```{r}
# Separate name and assign ranks
db.df <- 
     db.df %>%
     separate(name, 
              into = c( "accession",
                        "name" ),
              sep = '\\s',
              extra = 'merge') 

# Add ASV info alongside
db.df$seq <- as.character(db)
```

#### Taxonomy

```{r}
# Look up accession taxonomy with taxa package functions
taxmap <- metacoder::lookup_tax_data(db.df$accession, 
                                type = 'seq_id')
```

```{r}
# Make taxonomy table
taxonomy <- 
     metacoder::taxonomy_table(taxmap, 
                          use_ranks = c('superkingdom', 
                                        'kingdom',
                                        'phylum', 
                                        'order', 
                                        'family',
                                        'genus', 
                                        'species',
                                        'subspecies',
                                        'varietas'),
                          add_id_col = TRUE) %>% 
     data.frame()
```

Note that not all taxids are terminally listed here:
```{r}
dim(taxonomy)[1]
length(unique(taxmap$data$query_data))
```

This is because some are internal nodes. Manually add their taxonomy too:

```{r }
add <- setdiff(unique(names(taxmap$data$query_data)),
               taxonomy$taxon_id)

length(add)
```

Interesting.  Thinking that differences here (compared to 858-525 above) are due to repeated taxa?

```{r}
# Manually inspect
taxmap$data$tax_data %>%
     filter(taxon_id %in% add) %>% 
     View()

# Write full table, subset these columns out
# write_csv(taxonomy,
#           here('data', 
#                'processed',
#                'reference',
#                '20220216_Internal taxonomy table nodes.csv'))

add.rows <-
     here('data',
          'processed',
          'reference',
          '20220216_Internal taxonomy table nodes.csv') %>%
     read_csv(col_types = cols(.default = 'c'))
```

```{r }
# Bind together
taxonomy <-
     taxonomy %>%
     bind_rows(add.rows) %>%
     as.data.frame() # This is required in order for lowest level step to work-- so weird-- the code below doesn't seem to work on a tibble
```

```{r}
# Now, connect this back to the database and make sure that everything has an entry

# Link from accession to taxon ID
dim(db.df)
db.df <- 
     data.frame(accession = taxmap$data$query_data,
                taxon_id = names(taxmap$data$query_data)) %>% 
     left_join(db.df, .) %>% 
     distinct()
dim(db.df)
```

```{r}
# Link from taxon ID to taxonomy
db.df <- 
     left_join(db.df, taxonomy)

# Any missing a species-level designation?
any(is.na(db.df$species))
```

```{r}
# This is one accession for which look-up failed:
# filter(db.df, is.na(species))
# 
# # Remove it, duplicate other Theobroma row and replace accession
# row <- filter(db.df, species == 'Theobroma cacao')
# row$accession <- 'KR864758.1'
# 
# db.df <- 
#      db.df %>% 
#      filter(!is.na(species)) %>% 
#      bind_rows(row)
```

### Integrate

```{r}
names(db.df)
```

```{r}
names(crops)
```

```{r}
names(detections)
```


```{r}
intersect(names(db.df),
          names(crops)) %>% 
     intersect(names(detections))
```

```{r}
# Simplify the database input so we only have one line per included taxon
nrow(db.df)

db.df.join <- 
     db.df %>% 
     select(superkingdom:species) %>% # Leave off subspecies and variety information
     mutate(in_ref = TRUE) %>% 
     distinct() 

nrow(db.df.join)
```

```{r}
# Pre-join dimensions
nrow(db.df.join)
nrow(crops)
nrow(detections) # Though note duplications here due to NAs at genus or species level
```

```{r}
# Join on shared taxonomy columns
combined.df <- 
     full_join(db.df.join, crops) %>% 
     # Add a label for membership in tree and major crop status
     mutate(in_tree = !is.na(id),
            major_crop = !is.na(`FAO group code`)) %>% 
     select(-id) %>% 
     full_join(detections)

dim(combined.df)
```

#### Major crops in reference

#### Add major/minor crop label

Interestingly, from Milla et al. 2021 methods: "Before matching, the FAO Commodity List was checked for taxonomic accuracy and validity against The Plant List using function TPL of the Taxonstand package in R v.3.6.2. Commodity groups composed of a single crop species were automatically matched but most groups included several taxa. However, all groups but wheat were composed of a major contributor to agricultural production, together with congeneric minor crops. Therefore, production data were assigned to the major crop of each multispecific group. For wheat, we assigned 93% of production to bread wheat and 7% to durum. *The 112 crop species with a tracked record in FAOSTAT* were operationally coded as major crops and the remaining 754 crop species that were included in the Crop Origins database, but which lacked a FAOSTAT record, were considered as minor crops."

```{r}
# Make a simplified version of this for major crop tally in visualization:
major.crop <- 
     metadata %>% 
     filter(use_food == 1) %>% 
     select(species_name_, # Tip labels in tree
            `FAO group code`) %>%  # From FAO labeling
     mutate(major_crop = ifelse(!is.na(`FAO group code`),
                                yes = TRUE,
                                no = FALSE))
```

```{r}
# Now, add if the taxon is in the reference
# First, need to align naming with tip names to do lookup
db.df <- 
     db.df %>% 
     mutate(species_name_ = 
                 gsub(name,
                      pattern = ' var\\. .*$', 
                      replacement = ''),
            species_name_ =
                 gsub(species_name_,
                      pattern = ' subsp\\. .*$', 
                      replacement = ''))
```

```{r}
# Find those more than 3 words long
ex <- '\\w+\\s\\w+\\s'
db.df %>% 
     filter(grepl(species_name_, pattern = ex))
```
Ok, checking against major crops, now know to remove "x" designation-- this will get us a few more matches (except Glycyrrhiza and Juglans). Can safely ignore these because each component species is in the reference and will match. 

```{r}
db.df <- 
     db.df %>% 
     mutate(species_name_ = 
                 gsub(species_name_,
                      pattern = ' f\\. .*$', 
                      replacement = ''),
            species_name_ =
                 gsub(species_name_,
                      pattern = ' x ', 
                      replacement = ' '), 
            # Final mutate to add underscore
            species_name_ = 
                 gsub(species_name_,
                      pattern = ' ',
                      replacement = '_')
            )
```

Thinking I can add this to the tree as a heatmap-style layer (function gheatmap-- following example in Fig. 4.17 [here](https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html)).

```{r}
# Now, add as column to major crop dataframe:
head(major.crop)
```
```{r}
major.crop$in_ref <- 
     major.crop$species_name_ %in% db.df$species_name_

sum(major.crop$in_ref)
```

```{r}
# Now, add subtler agreement categories (if I want to color-code)
major.crop <- 
     major.crop %>% 
     mutate(agree = major_crop & in_ref,
            yes_crop_no_ref = major_crop & !in_ref,
            no_crop_yes_ref = !major_crop & in_ref)
```

```{r}
# And recode more subtle categories as a numeric scale for coloring plot

# In both crop and reference: 1, dark green
major.crop$in_ref[major.crop$agree] <- 1

# In reference, but not a major crop: 2, light green
major.crop$in_ref[major.crop$no_crop_yes_ref] <- 2

# Major crop, but not in reference: 3, yellow/orange
# Not a major crop, not in reference: 0, white
major.crop$in_ref[major.crop$yes_crop_no_ref] <- 3

# Convert to factor
major.crop <- mutate(major.crop,
                     in_ref = factor(in_ref))
```

```{r}
# Name rows so compatible with plotting function
major.crop <- 
     column_to_rownames(major.crop, var = 'species_name_')
```

# Statistics

How many distinct taxa?

```{r}
db.df %>% 
     select(name) %>% 
     n_distinct()
```

How many distinct species?

```{r}
n_distinct(db.df$species)
```

How many total sequences? Had previously reported unique, but I don't think this makes sense.

```{r}
length(db.df$seq)
```

Range of sequence lengths

```{r}
# Subtract length of primer sequences
range(width(db)) - nchar('GGGCAATCCTGAGCCAA') - nchar('CCATTGAGTCTCTGCACCTATC')
```

## % tree present in ref

```{r}
combined.df %>% 
     select(family, 
            in_ref,
            in_tree) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) 
```
```{r}
combined.df %>% 
     select(family, 
            in_ref,
            in_tree) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) %>% 
     filter(in_tree == TRUE) %>% 
     summarize(across(starts_with('in_'),
                      ~sum(.x, na.rm = TRUE)))
```

```{r}
72/115
```

What about only major crop families?

```{r}
combined.df %>% 
     filter(!is.na(`FAO group code`)) %>% 
     select(family, 
            in_ref,
            in_tree) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) %>% 
     filter(in_tree == TRUE) %>% 
     summarize(across(starts_with('in_'),
                      ~sum(.x, na.rm = TRUE)))
```
```{r}
45/54
```

## % ref detected from data

```{r}
combined.df %>% 
     select(family, 
            in_data,
            in_ref) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) %>% 
     filter(in_ref == TRUE) %>%
     summarize(across(starts_with('in_'),
                      ~sum(.x, na.rm = TRUE)))
```
```{r}
54/80
```

```{r}
combined.df %>% 
     filter(!is.na(`FAO group code`)) %>% 
     select(family, 
            in_data,
            in_ref) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) %>% 
     filter(in_ref == TRUE) %>%
     summarize(across(starts_with('in_'),
                      ~sum(.x, na.rm = TRUE)))
```
```{r}
33/45
```

```{r}
combined.df %>% 
     filter(!is.na(`FAO group code`) & in_ref) %>% 
     select(family, 
            in_data,
            in_ref) %>% 
     group_by(family) %>% 
     # Is that family represented by any sequence in the reference,
     # or any branch on the tree?
     summarize(across(starts_with('in_'),
                   ~any(.x, na.rm = TRUE))) %>% 
     filter(in_ref == TRUE) %>%
     summarize(across(starts_with('in_'),
                      ~sum(.x, na.rm = TRUE)))
```
## Major crops missing from reference

```{r}
missing <- 
     major.crop %>% 
     filter(yes_crop_no_ref) %>% 
     # Join to metadata common name for better interpretation
     rownames_to_column(var = 'species_name_') %>% 
     select(species_name_, 
            `FAO group code`) %>% 
     left_join(metadata)

missing
```


# Visualize taxa

## Overlay on Phylo Food

First, try visualizing *just* Phylo Food tree
```{r}
t <- ggtree(tree,
            layout = 'circular')
     # geom_tiplab(size = 0.5) +
     # geom_text2(aes(subset=!isTip,
     #                label = node),
     #            color = 'red',
     #            hjust = -0.3,
     #            size = 0.5)

t
```

Now, try overlaying species that are in the reference. From above, just by simple text matching, this is 

```{r}
sum(taxa.ref %in% taxa.tree)
```

groupOTU accepts a vector of OTUs (taxa name) or a list of OTUs. groupOTU will trace back from OTUs to their most recent common ancestor and cluster them together,

### Color by +- reference

```{r}
# Create a grouping based on inclusion/exclusion from reference
tree.overlay <- 
     groupOTU(tree,
              .node = unique(db.df$species_name_))
```

```{r}
t <- 
     ggtree(tree.overlay,
            aes(color = group),
            layout = 'circular') +
     # geom_tiplab(aes(subset = (label %in% taxa.ref), size = 0.5),
     geom_tiplab(size = 0.8) +
     scale_color_manual(values = c('gray80', 'black')) +
     theme(legend.position = 'none')

t
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree.eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

### Color by Milla categs

Wonder if I can label by consecutive leaf names?
Or leaves corresponding to particular internal nodes that I can then back out?

```{r}
# Do tip names go in order? Think yes:
tree$tip.label %>% head(50)
```

Saved these as CSV, then manually labeled with species groupings.  This is so they can be colored using groupOTU, as groupClade was missing a few lingering branches.  
```{r}
# Read in annotations
groups <- 
     here('data',
          'processed',
          'reference',
          'Phylo Food tip groups.csv') %>% 
     read_csv()

# Reformat for groupOTU()
# split(x, f) divides the data in the vector x into the groups defined by f
x <- split(groups$species_name_, groups$group)

tree.overlay <- 
     groupOTU(tree,
              x)

overlay.colors <- 
     c('#7e756d', # Root branches
       '#849db1', # Basal magnolids
       '#fbb04e', # Campanulidae
       '#6e8b66', # Fabidae
       '#b66353', # Lamiidae
       '#4f6980', # Malvidae
       '#a2ceaa', # Non-Poales monocots
       '#bfbb60', # Other eudicots
       '#d7ce9f') # Poales
```

```{r}
t <- 
     ggtree(tree.overlay,
            aes(color = group),
            # layout = 'circular') +
            layout = 'fan',
            open.angle = 20) +
     # geom_tiplab(size = 0.5) +
     scale_color_manual(values = overlay.colors)

t
```
```{r}
# Now, try adding major/minor crop data in outer ring
t.heatmap <- 
     gheatmap(p = t,
              data = select(major.crop,
                            in_ref),
         offset = 1,
         width = 0.05,
         colnames = FALSE,
         color = NULL
     ) +
     # Traffic color scheme-- a bit busy
     scale_fill_manual(values = c('white',
                                  'black', 
                                  '#989ca3', # 
                                  '#ff684c')) + 
     theme(legend.position = 'none') 

t.heatmap
```

```{r}
open_tree(t.heatmap, 20) %>% 
     rotate_tree(-20)
```


```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree.eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

```{r}
# Do a Chi squared test
# Are we more or less likely to include a crop in the reference, given its
# major/minor status?

# Convert data back to binary and tally
m <- 
     major.crop %>% 
     select(major_crop, in_ref) %>% 
     mutate(across(everything(), ~.x != 0)) %>% 
     group_by(major_crop, in_ref) %>% 
     count()

m
```

```{r}
# Pivot to wide
m <- 
     pivot_wider(m, 
                 names_from = in_ref,
                 names_prefix = 'in_ref_',
                 values_from = n)

m
```
```{r}
test <- chisq.test(m)
test
```

```{r}
test$expected
```

## UpSet plot

```{r}
taxtab <- 
     ps@tax_table@.Data %>% 
     data.frame()

head(taxa_names(ps))
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
146 - length(weightloss.taxa)
146 - length(adult1.taxa)
146 - length(adult2.taxa)
146 - length(adolescent.taxa)
```

```{r}
# Try dataframe structure for UpSet plot input
input.df <- 
     data.frame(taxon = taxa_names(ps)) %>% 
     mutate(`Weight Loss` = taxon %in% weightloss.taxa,
            `Adult-1` = taxon %in% adult1.taxa,
            `Adult-2` = taxon %in% adult2.taxa,
            `Adolescent` = taxon %in% adolescent.taxa
            ) %>% 
     mutate(across(.cols = -taxon,
                   as.numeric))

input.df
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results',
         'manuscript',
         'supplemental',
         paste0(plotID,
                '_UpSet plot of trnL taxa.pdf')),
    onefile = FALSE, # Remove blank page
    height = 4, width = 6)

upset(input.df,
      sets = c('Adolescent', 'Adult-2', 'Adult-1', 'Weight Loss'),
      keep.order = TRUE,
      order.by = 'freq',
      text.scale = 1.5,
      mb.ratio = c(0.55, 0.45),
      mainbar.y.label = 'Shared plants',
      sets.x.label = 'Total plants') 
```


```{r}
# Summarize phyloseq by cohort
listInput <- list(`Weight Loss` = weightloss.taxa,
                  `Adult-1` = adult1.taxa,
                  `Adult-2` = adult2.taxa,
                  `Adolescent` = adolescent.taxa)

upset(fromList(listInput), order.by = 'freq')
```
