---
title: "Plant taxa"
output: html_notebook
---

Used to make Figure 1B.

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(ape) # For phylogenetic trees
library(Biostrings)
library(ggtree)
library(ggtreeExtra)
library(here)
library(phyloseq)
library(MButils) # For lowest level naming
library(metacoder)
library(Taxonstand) # For synchronizing plant names
library(tidyverse)
library(vegan)
library(UpSetR)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed',
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

## Reference database

```{r}
db <- 
     here('..', 
          'food-dbs', 
          'data', 
          'processed',
          'dada2-compatible',
          'trnL',
          'trnLGH.fasta') %>% 
     readDNAStringSet()
```

#### PhyloFood

This follows on tree presented by Ruben Milla in "Crop Origins and Phylo Food: A database and a phylogenetic tree to stimulate comparative analyses on the origins of food crops."

From README:
This folder contains five files:
A readme file (readme_ phylo_food_live.txt);
The R code used to build the tree (phylo_food_live_code.txt);
the literature sources cited in the code and used to resolve the tree, in BibTeX format (sources_used_to_resolve_phylo_food_live.txt);
the time-calibrated and fully resolved phylogenetic tree of crop species (phylo_food_live.tre);
the non-fully resolved phylogenetic tree of vascular plants, including the 866 species of crops (phylo_GBOTB_with_crops_live.tre).

```{r}
crop.dir <- 
     here('data',
          'processed',
          'reference',
          'Crop_Origins_Phylo',
          'Crop_Origins_Phylo_v_live')

metadata <- 
     here(crop.dir,
          'crop_origins_v_live',
          'crop_origins_live_db.csv') %>% 
     read_delim(delim = ';')

tree <- 
     here(crop.dir,
          'phylo_food_v_live',
          'phylo_food_live.tre') %>% 
     ape::read.tree()
```

## FAO name mapping

from Martin et al., PLOS ONE, 2019

```{r}
fao.names <- 
     here('..',
          'food-dbs',
          'data',
          'processed',
          'parsed-refs',
          'Martin_PLOSONE_2019_FAOSTAT to taxon mapping.csv') %>% 
     read_csv()
```

```{r}
dim(fao.names)
```

# Pre-process

## trnL

```{r}
# First, remove name column from taxonomy, as not needed here
taxtab <- 
     ps@tax_table %>% 
     data.frame() %>% 
     select(-name)

tax_table(ps) <- tax_table(as.matrix(taxtab))
```

## Tree data

### CropOrigins

Here, likely want to subset to only species used as food.

```{r}
names(metadata)
```

```{r}
# Get entries for use categories
metadata %>% 
     select(starts_with('use')) %>% 
     apply(2, unique)
```

```{r}
metadata %>% 
     group_by(use_food) %>% 
     count()
```
```{r}
# Tip labels are already encoded in the variable 'species_name_'
# Add major/minor crop data
# First, format compatibly with tip labels
fao.names <- 
     fao.names %>% 
     # Remove infraspecific labels
     mutate(tip = gsub(pattern = 'var.*',
                       replacement = '', 
                       Species),
            tip = gsub(pattern = 'subsp.*',
                       replacement = '', 
                       tip),
            # Replace space with underscore
            tip = gsub(tip, 
                       pattern = ' ',
                       replacement = '_')) 
```

```{r}
# Join to metadata
metadata <- 
     left_join(metadata, fao.names,
               by = c('species_name_' = 'tip'))
```

```{r}
# Make a simplified version of this for major crop tally in visualization:
major.crop <- 
     metadata %>% 
     select(species_name_, # Tip labels in tree
            Species) %>%  # From FAO labeling
     distinct() %>% 
     mutate(major_crop = ifelse(!is.na(Species),
                                yes = TRUE,
                                no = FALSE))
```

### PhyloFood

```{r}
str(tree)
```

Interesting that internal nodes aren't labeled here.  This would be nice because some metabarcoding entries will terminate internally in the tree. Or could I just display species, without collapsing up to last common ancestor?

```{r}
# Filter to only those used as food, get names in same format as tree
crops <- 
     metadata %>% 
     filter(use_food == 1) %>% 
     pull(species_name) %>% 
     gsub(pattern = ' ', replacement = '_')

head(crops)
```

```{r}
# Subset tree to these taxa only
tree <- keep.tip(tree, crops)

tree
```

### Reference

Organize in a taxonomy-table style.

```{r}
db
```

```{r}
db.df <- 
     data.frame(name = names(db),
                seq = as.character(db))

db.df
```

```{r}
# Separate name and assign ranks
db.df <- 
     db.df %>%
     separate(name, 
              into = c( "accession",
                        "name" ),
              sep = '\\s',
              extra = 'merge') 

# Add ASV info alongside
db.df$seq <- as.character(db)
```

Steps for taxonomy: DON'T RUN RIGHT NOW

```{r}
# Look up accession taxonomy with taxa package functions
taxmap <- metacoder::lookup_tax_data(db.df$accession, 
                                type = 'seq_id')
```

```{r}
# Make taxonomy table
# Note may need to make changes to ranks below based on animal/plant db
taxonomy <- 
     metacoder::taxonomy_table(taxmap, 
                          use_ranks = c('superkingdom', 
                                        'kingdom',
                                        'phylum', 
                                        'order', 
                                        'family',
                                        'genus', 
                                        'species',
                                        'subspecies',
                                        'varietas'),
                          add_id_col = TRUE) %>% 
     data.frame()
```

Note that not all taxids are terminally listed here:
```{r}
dim(taxonomy)[1]
length(unique(taxmap$data$query_data))
```

This is because some are internal nodes. Manually add their taxonomy too:

```{r }
add <- setdiff(unique(names(taxmap$data$query_data)),
               taxonomy$taxon_id)

length(add)
```

Interesting.  Thinking that differences here (compared to 858-525 above) are due to repeated taxa?

```{r}
# Manually inspect
taxmap$data$tax_data %>%
     filter(taxon_id %in% add) %>% 
     View()

# Write full table, subset these columns out
# write_csv(taxonomy,
#           here('data', 
#                'processed',
#                'reference',
#                '20220216_Internal taxonomy table nodes.csv'))

add.rows <-
     here('data',
          'processed',
          'reference',
          '20220216_Internal taxonomy table nodes.csv') %>%
     read_csv(col_types = cols(.default = 'c'))
```


```{r }
# Bind together
taxonomy <-
     taxonomy %>%
     bind_rows(add.rows) %>%
     as.data.frame() # This is required in order for lowest level step to work-- so weird-- the code below doesn't seem to work on a tibble
```

```{r}
# Now, connect this back to the database and make sure that everything has an entry

# Link from accession to taxon ID
dim(db.df)
db.df <- 
     data.frame(accession = taxmap$data$query_data,
                taxon_id = names(taxmap$data$query_data)) %>% 
     left_join(db.df, .) %>% 
     distinct()
dim(db.df)
```

```{r}
# Link from taxon ID to taxonomy
db.df <- 
     left_join(db.df, taxonomy)

# Any missing a species-level designation?
any(is.na(db.df$species))
```

### Integrate

How many names are present in tree labels? 592 in reference:

```{r}
taxa.ref <- 
     db.df %>% 
     pull(name) %>% 
     unique()

taxa.tree <- 
     tree$tip.label %>% 
     gsub(pattern = '_', replacement = ' ')

head(taxa.ref)
head(taxa.tree)
```

Interesting-- note that every entry appears to be at the species level.  Don't worry about extending for now; instead, drop subspecies and variety info.

```{r}
taxa.ref <- 
     taxa.ref %>% 
     gsub(pattern = ' var\\. .*$', replacement = '') %>% 
     gsub(pattern = ' subsp\\. .*$', replacement = '')

sum(taxa.ref %in% taxa.tree)/length(taxa.ref)
```

```{r}
taxa.ref[!(taxa.ref %in% taxa.tree)] %>% head(20)
```

```{r}
taxa.tree[!(taxa.tree %in% taxa.ref)] %>% head(20)
```

#### Relabel internal nodes

https://bioconnector.github.io/workshops/r-ggtree.html#internal_node_number

Another idea would be to just numerically label the internal nodes, then manually note the ones I want to highlight. 

```{r}
head(tree$edge)
dim(tree$edge)
```

"The matrix edge contains the beginning and ending node number for all the nodes and tips in the tree. By convention, the tips of the tree are numbered 1 through n for n tips; and the nodes are numbered n + 1 through n + m for m nodes. m = n - 1 for a fully bifurcating tree. This is just to keep track of which nodes are internal and which are leaves."

```{r}
# First, organize and put nodes in order
labels.df <- data.frame(tree$edge)

labels.df <- 
     labels.df %>% 
     rownames_to_column(var = 'edge') %>% 
     group_by(edge) %>% 
     mutate(from = max(X1, X2),
            to = min(X1, X2)) %>% 
     ungroup()
```

```{r}
# Get tip names
tips <- 
     data.frame(name = tree$tip.label,
                number = 1:866) 

# Join to overall labels
labels.df <- left_join(labels.df, tips, by = c('to'='number'))
```

```{r}
# Now get genus labels from those:
genera <- 
     labels.df %>% 
     filter(!is.na(name)) %>% 
     select(from, name) %>% 
     mutate(genus = gsub('_.*', '', name)) %>% 
     select(from, genus) %>% 
     distinct()

dim(genera)
```

```{r}
# Note that some are duplicated
# For that, just want to choose the largest (most internal?? node and label that one)
genera <- 
     genera %>% 
     group_by(genus) %>% 
     summarize(number = max(from))

# Now label these in the node list
# Just label in the 'to' column
labels.df <- 
     labels.df %>% 
     left_join(., genera, by = c('to'='number')) %>% 
     mutate(name = ifelse(is.na(name), 
                          yes = genus,
                          no = name)) %>% 
     select(-genus)
```

```{r}
# Relabel in tree
head(tree$node.label)
length(tree$node.label)
```

```{r}
# Subtract 866 from index
# tree$node.label <- 
#      labels.df %>% 
#      select(to, name) %>% 
#      filter(to > 866) %>% 
#      mutate(to = to - 866) %>% 
#      pull(name)
```

Or try using ape::makeNodeLabel

```{r}
nodeList <- unique(db.df$genus)
names(nodeList) <- nodeList

# This ran for ~5 mins with no results
# ape::makeNodeLabel(tree,
#                    method = 'user',
#                    nodeList = nodeList)
```

#### Taxonstand

Try synchronizing names to see how this impacts result.

```{r}
# Getting error here-- need to follow up
# taxonstand <- Taxonstand::TPL(db.df$name)
```

## Add major/minor crop label

Interestingly, from Milla et al. 2021 methods: "Before matching, the FAO Commodity List was checked for taxonomic accuracy and validity against The Plant List using function TPL of the Taxonstand package in R v.3.6.2. Commodity groups composed of a single crop species were automatically matched but most groups included several taxa. However, all groups but wheat were composed of a major contributor to agricultural production, together with congeneric minor crops. Therefore, production data were assigned to the major crop of each multispecific group. For wheat, we assigned 93% of production to bread wheat and 7% to durum. *The 112 crop species with a tracked record in FAOSTAT* were operationally coded as major crops and the remaining 754 crop species that were included in the Crop Origins database, but which lacked a FAOSTAT record, were considered as minor crops."

However, more than 112 listed:
```{r}
dim(fao.names)
n_distinct(fao.names$Species)
```

Do only 112 turn up in PhyloFood taxa?

```{r}
intersect(tree$tip.label,
          fao.names$tip) %>% 
     length()
```

Maybe this further reduction comes after naming corrections above, or after reducing to the "major crop of each multispecific group"? (They had to do this in order to map FAOSTAT production data onto food species, but we don't need to).  

# Statistics

How many distinct taxa?

```{r}
db.df %>% 
     select(name) %>% 
     n_distinct()
```

How many total sequences? Had previously reported unique, but I don't think this makes sense.

```{r}
length(db.df$seq)
```

Range of sequence lengths

```{r}
# Subtract length of primer sequences
range(width(db)) - nchar('GGGCAATCCTGAGCCAA') - nchar('CCATTGAGTCTCTGCACCTATC')
```

# Visualize taxa

## Overlay on Phylo Food

First, try visualizing *just* Phylo Food tree
```{r}
t <- ggtree(tree,
            layout = 'circular') +
     geom_tiplab(size = 0.5) +
     geom_text2(aes(subset=!isTip,
                    label = node),
                color = 'red',
                hjust = -0.3,
                size = 0.5)

t
```

Now, try overlaying species that are in the reference. From above, just by simple text matching, this is 

```{r}
sum(taxa.ref %in% taxa.tree)
```

Name as in tree style so I can directly reference tip labels.
```{r}
db.df <- 
     db.df %>% 
     # Remove subspecies and variety info
     mutate(tree_name = gsub(name,
                             pattern = ' var\\. .*$', 
                             replacement = ''),
            tree_name = gsub(tree_name,
                             pattern = ' subsp\\. .*$', 
                             replacement = '')) %>% 
     # Replace space with underscore
     mutate(tree_name = gsub(pattern = ' ',
                             replacement = '_',
                             tree_name))

taxa.ref <- 
     db.df %>% 
     pull(tree_name) %>% 
     unique()

head(taxa.ref)
```

```{r}
length(setdiff(taxa.ref, tree$tip.label))
length(setdiff(tree$tip.label, taxa.ref))
```

groupOTU accepts a vector of OTUs (taxa name) or a list of OTUs. groupOTU will trace back from OTUs to their most recent common ancestor and cluster them together,

```{r}
tree.overlay <- 
     groupOTU(tree,
              .node = taxa.ref)
```

```{r}
t <- 
     ggtree(tree.overlay,
            aes(color = group),
            layout = 'circular') +
     # Basal magnolids
     geom_hilight(node = 1547,
                  fill = '#7e756d', 
                  alpha = 0.5,
                  size = 0) +
     # Non-Poales monocots
     # Fabidae
     geom_hilight(node = 1025,
                  fill = '#849db1',
                  alpha = 0.8,
                  size = 0) +
     # Campanulidae
     geom_hilight(node = 800,
                  fill = '#ffda66',
                  alpha = 0.8,
                  size = 0) +
     # Lamiidae
     geom_hilight(node = 853,
                  fill = '#fbb04e',
                  alpha = 0.8,
                  size = 0) +
     # Malvidae
     geom_hilight(node = 1269,
                  fill = '#4f6980',
                  alpha = 0.8,
                  size = 0) +
     # Poaceae
     geom_hilight(node = 1427,
                  fill = '#b66353',
                  alpha = 0.8,
                  size = 0) +
     # # Non-Poales monocots
     geom_hilight(node = 1516,
                  fill = '#ff9da7',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = 1532,
                  fill = '#ff9da7',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = 1476,
                  fill = '#ff9da7',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = c(1407, 1512),
                  fill = '#ff9da7',
                  alpha = 0.8,
                  size = 0) +
     # Other Eudicots (add in installments 2/2 tree structure)
     geom_hilight(node = 964,
                  fill = '#638b66',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = 932,
                  fill = '#638b66',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = 1392,
                  fill = '#638b66',
                  alpha = 0.8,
                  size = 0) +
     geom_hilight(node = 1394,
                  fill = '#638b66',
                  alpha = 0.8,
                  size = 0) +
     # geom_tiplab(aes(subset = (label %in% taxa.ref), size = 0.5),
     # geom_tiplab(size = 0.8) +
     scale_color_manual(values = c('gray80', 'black')) +
     theme(legend.position = 'none')

t
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree (with colors).eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

```{r}
# Now, try adding major/minor crop data in outer ring
t + 
     geom_fruit(
          data = major.crop,
          geom = geom_bar,
          mapping = aes(y = species_name_, 
                        x = 1,
                        fill = major_crop),
          stat = 'identity',
          orientation = 'y') +
     # Fill values for major/minor crop
     scale_fill_manual(values = c('white', 'black'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree (major crop label and types).eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

## Metacoder tree

### Reference

#### Full tree

This doesn't take into account the fact that some of our detections won't match nodes because of sequence duplication.

```{r}
# Consider every taxon in the reference
# Make a dummy phyloseq object to hold
taxtab <- 
     taxonomy %>% 
     column_to_rownames(var = 'taxon_id') %>% 
     as.matrix()

asvtab <- matrix(data = 1,
                 nrow = 1,
                 ncol = dim(taxtab)[1],
                 dimnames = list('all', row.names(taxtab)))

ps.ref <- 
     phyloseq(tax_table(taxtab),
              otu_table(asvtab, taxa_are_rows = FALSE))

ps.ref
```

```{r}
x.ref <- parse_phyloseq(ps.ref)
```

377 taxa, and 117 terminal nodes
Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x.ref$data$tax_abund <- calc_taxon_abund(x.ref, 
                                         'otu_table')

dim(x.ref$data$tax_abund)

# Convert to presence-absence
x.ref$data$tax_abund <- counts_to_presence(x.ref, 'tax_abund')
x.ref$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '1', 
          paste0(plotID, '_Full reference.pdf')),
         height = 10, width = 10)

heat_tree(
     x.ref,
     node_label = taxon_names,
     # node_color = n_samples,
     # node_color_range = pal[1:7],
     # node_size = n_samples,
     # node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

#### Realistic tree

This is with last common ancestor labeling done as would be in object.

- Group by sequence
     + Are any subsequences?
- Do last common ancestor
- Get "full" taxonomy
- Extend phyloseq object

```{r}
# Are any sequences substrings of one another?
seqs <- db.df$seq

substring <- function(s){
     # Remove s from the current list of strings to be considered
     seqs.temp <- seqs[seqs!=s]
     # Now check this list for membership
     any(str_detect(pattern = s,
                string = seqs.temp))
}

db.df$ss <- 
     lapply(db.df$seq,
            substring) %>% 
     unlist()

db.df %>% 
     filter(ss) %>% View()

```

#### Add possible taxa

Here, need to add empty rows to phyloseq object and corresponding taxonomy for everything that could have potentially been detected by the reference.  

```{r}
asvtab <- ps@otu_table@.Data
taxtab <- ps@tax_table@.Data
```

```{r}
# Pull only sequences that don't have an ASV table sequence as a substring

add <- c()

for (s in db.df$seq){
     ss <- 
          lapply(colnames(asvtab),
                  function(x){str_detect(pattern = x,
                                         string = s)}) %>% 
          unlist() %>% 
          any()
     
     # If no ASV as a substring, add it to list
     if(!ss){
          add <- c(add, s)
     }
}

length(add)
length(unique(add))
```

```{r}
# Pull rows to join to taxonomy table
add.tax <- 
     db.df %>% 
     filter(seq %in% add) 
```

Need to label duplicated sequences with their last common ancestor to get this working properly.

```{r eval = FALSE}
# Group by ASV
lca <- 
     add.tax %>%
     group_by(seq) %>%
     summarize_all(n_distinct) %>%
     column_to_rownames(var = 'seq')

# Now, relabel all those with >1 name at a particular level as NA
# As a placeholder, keep only the first species, knowing it will be overwritten
add.tax.lca <- 
     add.tax %>%
     group_by(seq) %>%
     slice_head(n = 1) %>%
     column_to_rownames(var = 'seq')
```

Now do relabeling 
```{r eval = FALSE}
# Confirm ordering is okay
all(rownames(lca) == rownames(add.tax.lca))

add.tax.lca[lca > 1] = NA
```

```{r}
# Combine rows
taxtab <- 
     rbind(taxtab,
           add.tax.lca) %>% 
     distinct() # Some of the taxa were already detected
```

Now can update ASV table

```{r}
# Make empty columns for ASV table
add.asv <- 
     matrix(0, 
            nrow = dim(asvtab)[1],
            ncol = dim(add.tax.lca)[1])

colnames(add.asv) <- rownames(add.tax.lca)
rownames(add.asv) <- rownames(asvtab)

asvtab <- cbind(asvtab, add.asv)
```

Need to make one dummy sample to contain all these taxa, so that the zeros are then labeled in the plot below.

```{r}
# Add to ASV table
row <- matrix(1, ncol = dim(asvtab)[2])
rownames(row) <- 'Reference'
colnames(row) <- colnames(asvtab)

asvtab <- rbind(asvtab, row)
```

```{r}
# Add to sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row')

# Add a distinguishing variable
samdf$type <- 'sample' # All existing rows are samples

row <- matrix(NA, ncol = dim(samdf)[2])
colnames(row) <- names(samdf)
row <- 
     data.frame(row) %>% 
     mutate(row = 'Reference',
            type = 'reference')

samdf <- 
     bind_rows(samdf, row) %>% 
     column_to_rownames(var = 'row') %>% 
     select(-sample) # Need to toss this, otherwise it's looked up below
``` 

```{r}
dim(asvtab)
dim(samdf)
dim(taxtab)
```

```{r}
# Update phyloseq object
ps

ps <- phyloseq(otu_table(asvtab, taxa_are_rows = FALSE),
               sample_data(samdf),
               tax_table(as.matrix(taxtab)))

ps
```

```{r}

```


### All samples

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table')

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Samples detected, size.pdf')),
         height = 10, width = 10)

pal <- brewer.pal(9, 'Greens')

heat_tree(
     x,
     node_label = taxon_names,
     node_color = n_samples,
     node_color_range = pal[1:7],
     node_size = n_samples,
     node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

Now, try with group color
Need to re-make data structure.

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$study)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Detected (Adolescent).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(Adolescent, pal[7], "grey"))
```
Color argument "accepts a vector, the same length taxon_id or a factor of its length."
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results',
         'manuscript',
         '5', 
          paste0(plotID, '_Detected (Adult-1).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, pal[7], "grey"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, pal[7], "grey"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_Detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, pal[7], "grey"))
```

#### Visualize

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$type)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript',
         '1', 
          paste0(plotID, '_B.pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(sample, "darkgreen", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adolescent).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "grey", "red"),
          node_label = taxon_names)
```

## UpSet plot

```{r}
# Update taxa names to their lowest specified level
all(taxa_names(ps) == row.names(taxtab)) 
taxa_names(ps) <- lowest_level(taxtab)$name
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
146 - length(weightloss.taxa)
146 - length(adult1.taxa)
146 - length(adult2.taxa)
146 - length(adolescent.taxa)
```

```{r}
# Summarize phyloseq by cohort
listInput <- list(`Weight Loss` = weightloss.taxa,
                  `Adult-1` = adult1.taxa,
                  `Adult-2` = adult2.taxa,
                  `Adolescent` = adolescent.taxa)

upset(fromList(listInput), order.by = 'freq')
```
