---
title: "Figure 1A"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(MButils) # For lowest level naming
library(metacoder)
library(RColorBrewer)
library(tidyverse)
library(vegan)
library(UpSetR)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

# Tidy

```{r}
# First, remove name column from taxonomy, as not needed here
taxtab <- 
     ps@tax_table %>% 
     data.frame() %>% 
     select(-name)

tax_table(ps) <- tax_table(as.matrix(taxtab))
```

## Filter

```{r}
# # Remove any taxon seen with 3 or fewer counts in a sample
# asvtab <- ps@otu_table@.Data
# asvtab[asvtab <= 3] <- 0
# 
# # Replace in phyloseq object
# otu_table(ps) <- otu_table(asvtab,
#                            taxa_are_rows = FALSE)
```

# Count taxa

```{r}
taxtab <- data.frame(ps@tax_table)
```

```{r}
# Number of taxa detected
dim(taxtab)[1]
```

```{r}
# Number of taxa per phylogenetic category
apply(taxtab, 2, function(x){n_distinct(x)})
```


# Visualize taxa

## Metacoder tree

### All samples
```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

377 taxa, and 117 terminal nodes
Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table')

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_n samples detected, size.pdf')),
         height = 10, width = 10)

pal <- brewer.pal(9, 'Greens')

heat_tree(
     x,
     node_label = taxon_names,
     node_color = n_samples,
     node_color_range = pal[1:7],
     node_size = n_samples,
     node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

Now, try with group color
Need to re-make data structure.

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$study)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adolescent).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "red", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adolescent).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "grey", "red"))
```
Color argument "accepts a vector, the same length taxon_id or a factor of its length."
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adult-1).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, "red", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adult-1).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, "grey", "red"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, "red", "grey"))

# Not detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, "grey", "red"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, "red", "grey"))

# Not detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, "grey", "red"))
```

## UpSet plot

```{r}
# Update taxa names to their lowest specified level
all(taxa_names(ps) == row.names(taxtab)) 
taxa_names(ps) <- lowest_level(taxtab)$name
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
118 - length(weightloss.taxa)
118 - length(adult1.taxa)
118 - length(adult2.taxa)
118 - length(adolescent.taxa)
```

```{r}
# Summarize phyloseq by cohort
listInput <- list(`Weight Loss` = weightloss.taxa,
                  `Adult-1` = adult1.taxa,
                  `Adult-2` = adult2.taxa,
                  `Adolescent` = adolescent.taxa)

upset(fromList(listInput), order.by = 'freq')
```
