---
title: "Plant taxa"
output: html_notebook
---

Used to make Figure 1B.

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(ape) # For phylogenetic trees
library(Biostrings)
library(ggtree)
library(ggtreeExtra)
library(here)
library(phyloseq)
library(MButils) # For lowest level naming
library(metacoder)
library(Taxonstand) # For synchronizing plant names
library(tidyverse)
library(vegan)
library(UpSetR)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 
          'processed',
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

## Reference database

```{r}
db <- 
     here('..', 
          'food-dbs', 
          'data', 
          'processed',
          'dada2-compatible',
          'trnL',
          'trnLGH.fasta') %>% 
     readDNAStringSet()
```

#### PhyloFood

This follows on tree presented by Ruben Milla in "Crop Origins and Phylo Food: A database and a phylogenetic tree to stimulate comparative analyses on the origins of food crops."

From README:
This folder contains five files:
A readme file (readme_ phylo_food_live.txt);
The R code used to build the tree (phylo_food_live_code.txt);
the literature sources cited in the code and used to resolve the tree, in BibTeX format (sources_used_to_resolve_phylo_food_live.txt);
the time-calibrated and fully resolved phylogenetic tree of crop species (phylo_food_live.tre);
the non-fully resolved phylogenetic tree of vascular plants, including the 866 species of crops (phylo_GBOTB_with_crops_live.tre).

```{r}
crop.dir <- 
     here('data',
          'processed',
          'reference',
          'Crop_Origins_Phylo',
          'Crop_Origins_Phylo_v_live')

metadata <- 
     here(crop.dir,
          'crop_origins_v_live',
          'crop_origins_live_db.csv') %>% 
     read_delim(delim = ';')

tree <- 
     here(crop.dir,
          'phylo_food_v_live',
          'phylo_food_live.tre') %>% 
     ape::read.tree()
```

## FAO name mapping

from Martin et al., PLOS ONE, 2019

```{r}
fao.names <- 
     here('..',
          'food-dbs',
          'data',
          'processed',
          'parsed-refs',
          'Martin_PLOSONE_2019_FAOSTAT to taxon mapping.csv') %>% 
     read_csv()
```

```{r}
dim(fao.names)
```

# Pre-process

## trnL

```{r}
# First, remove name column from taxonomy, as not needed here
taxtab <- 
     ps@tax_table %>% 
     data.frame() %>% 
     select(-name)

tax_table(ps) <- tax_table(as.matrix(taxtab))
```

## Tree data

### CropOrigins

Here, likely want to subset to only species used as food.

```{r}
names(metadata)
```

```{r}
# Get entries for use categories
metadata %>% 
     select(starts_with('use')) %>% 
     apply(2, unique)
```

```{r}
metadata %>% 
     group_by(use_food) %>% 
     count()
```
### PhyloFood

```{r}
str(tree)
```

Interesting that internal nodes aren't labeled here.  This would be nice because some metabarcoding entries will terminate internally in the tree. Or could I just display species, without collapsing up to last common ancestor?

```{r}
# Filter to only those used as food, get names in same format as tree
crops <- 
     metadata %>% 
     filter(use_food == 1) %>% 
     pull(species_name) %>% 
     gsub(pattern = ' ', replacement = '_')

head(crops)
```

```{r}
# Subset tree to these taxa only
tree <- keep.tip(tree, crops)

tree
rm(crops)
```

### Reference

Organize in a taxonomy-table style.

```{r}
db
```

```{r}
db.df <- 
     data.frame(name = names(db),
                seq = as.character(db))

db.df
```

```{r}
# Separate name and assign ranks
db.df <- 
     db.df %>%
     separate(name, 
              into = c( "accession",
                        "name" ),
              sep = '\\s',
              extra = 'merge') 

# Add ASV info alongside
db.df$seq <- as.character(db)
```

Steps for taxonomy: DON'T RUN RIGHT NOW

```{r}
# Look up accession taxonomy with taxa package functions
taxmap <- metacoder::lookup_tax_data(db.df$accession, 
                                type = 'seq_id')
```

```{r}
# Make taxonomy table
# Note may need to make changes to ranks below based on animal/plant db
taxonomy <- 
     metacoder::taxonomy_table(taxmap, 
                          use_ranks = c('superkingdom', 
                                        'kingdom',
                                        'phylum', 
                                        'order', 
                                        'family',
                                        'genus', 
                                        'species',
                                        'subspecies',
                                        'varietas'),
                          add_id_col = TRUE) %>% 
     data.frame()
```

Note that not all taxids are terminally listed here:
```{r}
dim(taxonomy)[1]
length(unique(taxmap$data$query_data))
```

This is because some are internal nodes. Manually add their taxonomy too:

```{r }
add <- setdiff(unique(names(taxmap$data$query_data)),
               taxonomy$taxon_id)

length(add)
```

Interesting.  Thinking that differences here (compared to 858-525 above) are due to repeated taxa?

```{r}
# Manually inspect
taxmap$data$tax_data %>%
     filter(taxon_id %in% add) %>% 
     View()

# Write full table, subset these columns out
# write_csv(taxonomy,
#           here('data', 
#                'processed',
#                'reference',
#                '20220216_Internal taxonomy table nodes.csv'))

add.rows <-
     here('data',
          'processed',
          'reference',
          '20220216_Internal taxonomy table nodes.csv') %>%
     read_csv(col_types = cols(.default = 'c'))
```


```{r }
# Bind together
taxonomy <-
     taxonomy %>%
     bind_rows(add.rows) %>%
     as.data.frame() # This is required in order for lowest level step to work-- so weird-- the code below doesn't seem to work on a tibble
```

```{r}
# Now, connect this back to the database and make sure that everything has an entry

# Link from accession to taxon ID
dim(db.df)
db.df <- 
     data.frame(accession = taxmap$data$query_data,
                taxon_id = names(taxmap$data$query_data)) %>% 
     left_join(db.df, .) %>% 
     distinct()
dim(db.df)
```

```{r}
# Link from taxon ID to taxonomy
db.df <- 
     left_join(db.df, taxonomy)

# Any missing a species-level designation?
any(is.na(db.df$species))
```

### Integrate

#### Major crops in reference

#### Add major/minor crop label

Interestingly, from Milla et al. 2021 methods: "Before matching, the FAO Commodity List was checked for taxonomic accuracy and validity against The Plant List using function TPL of the Taxonstand package in R v.3.6.2. Commodity groups composed of a single crop species were automatically matched but most groups included several taxa. However, all groups but wheat were composed of a major contributor to agricultural production, together with congeneric minor crops. Therefore, production data were assigned to the major crop of each multispecific group. For wheat, we assigned 93% of production to bread wheat and 7% to durum. *The 112 crop species with a tracked record in FAOSTAT* were operationally coded as major crops and the remaining 754 crop species that were included in the Crop Origins database, but which lacked a FAOSTAT record, were considered as minor crops."

```{r}
# Tip labels are already encoded in the variable 'species_name_'
# Add major/minor crop data
# First, format compatibly with tip labels
fao.names <- 
     fao.names %>% 
     # Remove infraspecific labels
     mutate(tip = gsub(pattern = 'var.*',
                       replacement = '', 
                       Species),
            tip = gsub(pattern = 'subsp.*',
                       replacement = '', 
                       tip),
            # Replace space with underscore
            tip = gsub(tip, 
                       pattern = ' ',
                       replacement = '_')) 
```

```{r}
# Join to metadata
metadata <- 
     left_join(metadata, fao.names,
               by = c('species_name_' = 'tip'))
```

```{r}
# Make a simplified version of this for major crop tally in visualization:
major.crop <- 
     metadata %>% 
     select(species_name_, # Tip labels in tree
            Species) %>%  # From FAO labeling
     distinct() %>% 
     mutate(major_crop = ifelse(!is.na(Species),
                                yes = TRUE,
                                no = FALSE))
```

```{r}
# Now, add if the taxon is in the reference
# First, need to align naming with tip names to do lookup
db.df <- 
     db.df %>% 
     mutate(species_name_ = 
                 gsub(name,
                      pattern = ' var\\. .*$', 
                      replacement = ''),
            species_name_ =
                 gsub(species_name_,
                      pattern = ' subsp\\. .*$', 
                      replacement = ''))
```

```{r}
# Find those more than 3 words long
ex <- '\\w+\\s\\w+\\s'
db.df %>% 
     filter(grepl(species_name_, pattern = ex))
```
Ok, checking against major crops, now know to remove "x" designation-- this will get us a few more matches (except Glycyrrhiza and Juglans). Can safely ignore these because each component species is in the reference and will match. 

```{r}
db.df <- 
     db.df %>% 
     mutate(species_name_ = 
                 gsub(species_name_,
                      pattern = ' f\\. .*$', 
                      replacement = ''),
            species_name_ =
                 gsub(species_name_,
                      pattern = ' x ', 
                      replacement = ' '), 
            # Final mutate to add underscore
            species_name_ = 
                 gsub(species_name_,
                      pattern = ' ',
                      replacement = '_')
            )
```

Thinking I can add this to the tree as a heatmap-style layer (function gheatmap-- following example in Fig. 4.17 [here](https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html)).

```{r}
# Now, add as column to major crop dataframe:
head(major.crop)
```
```{r}
major.crop$in_ref <- 
     major.crop$species_name_ %in% db.df$species_name_

sum(major.crop$in_ref)
```

```{r}
# Now, add subtler agreement categories (if I want to color-code)
major.crop <- 
     major.crop %>% 
     mutate(agree = major_crop & in_ref,
            yes_crop_no_ref = major_crop & !in_ref,
            no_crop_yes_ref = !major_crop & in_ref)
```

# Statistics

How many distinct taxa?

```{r}
db.df %>% 
     select(name) %>% 
     n_distinct()
```

How many total sequences? Had previously reported unique, but I don't think this makes sense.

```{r}
length(db.df$seq)
```

Range of sequence lengths

```{r}
# Subtract length of primer sequences
range(width(db)) - nchar('GGGCAATCCTGAGCCAA') - nchar('CCATTGAGTCTCTGCACCTATC')
```

# Visualize taxa

## Overlay on Phylo Food

First, try visualizing *just* Phylo Food tree
```{r}
t <- ggtree(tree,
            layout = 'circular')
     # geom_tiplab(size = 0.5) +
     # geom_text2(aes(subset=!isTip,
     #                label = node),
     #            color = 'red',
     #            hjust = -0.3,
     #            size = 0.5)

t
```

Now, try overlaying species that are in the reference. From above, just by simple text matching, this is 

```{r}
sum(taxa.ref %in% taxa.tree)
```

groupOTU accepts a vector of OTUs (taxa name) or a list of OTUs. groupOTU will trace back from OTUs to their most recent common ancestor and cluster them together,

### Color by +- reference

```{r}
# Create a grouping based on inclusion/exclusion from reference
tree.overlay <- 
     groupOTU(tree,
              .node = unique(db.df$species_name_))
```

```{r}
t <- 
     ggtree(tree.overlay,
            aes(color = group),
            layout = 'circular') +
     # geom_tiplab(aes(subset = (label %in% taxa.ref), size = 0.5),
     geom_tiplab(size = 0.8) +
     scale_color_manual(values = c('gray80', 'black')) +
     theme(legend.position = 'none')

t
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree.eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

### Color by Milla categs

Wonder if I can label by consecutive leaf names?
Or leaves corresponding to particular internal nodes that I can then back out?

```{r}
# Do tip names go in order? Think yes:
tree$tip.label %>% head(50)
```

Saved these as CSV, then manually labeled with species groupings.  This is so they can be colored using groupOTU, as groupClade was missing a few lingering branches.  
```{r}
# Read in annotations
groups <- 
     here('data',
          'processed',
          'reference',
          'Phylo Food tip groups.csv') %>% 
     read_csv()

# Reformat for groupOTU()
# split(x, f) divides the data in the vector x into the groups defined by f
x <- split(groups$species_name_, groups$group)

tree.overlay <- 
     groupOTU(tree,
              x)

overlay.colors <- 
     c('#7e756d', # Root branches
       '#849db1', # Basal magnolids
       '#fbb04e', # Campanulidae
       '#6e8b66', # Fabidae
       '#b66353', # Lamiidae
       '#4f6980', # Malvidae
       '#d7ce9f', # Non-Poales monocots
       '#bfbb60', # Other eudicots
       '#a2ceaa') # Poales
```

```{r}
t <- 
     ggtree(tree.overlay,
            aes(color = group),
            layout = 'circular') +
     scale_color_manual(values = overlay.colors) +
     geom_tiplab(size = 0.8) +
     theme(legend.position = 'none') 

t
```


```{r}
# Now, try adding major/minor crop data in outer ring
# Format a bit more precisely
major.crop <- 
     column_to_rownames(major.crop, var = 'species_name_')

gheatmap(p = t,
              data = select(major.crop,
                            in_ref,
                            major_crop),
         offset = 1,
         width = 0.1,
         colnames = FALSE,
         color = NULL
     ) +
     scale_fill_manual(values = c('white', 'black')) +
     theme(legend.position = 'none')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            '1',
            paste0(plotID, '_Overlay on Phylo Food tree (types, with labels).eps')),
       # Try this to preserve background transparency
       device = cairo_ps,
       height = 10, width = 10)
```

## Metacoder tree

### Reference

#### Full tree

This doesn't take into account the fact that some of our detections won't match nodes because of sequence duplication.

```{r}
# Consider every taxon in the reference
# Make a dummy phyloseq object to hold
taxtab <- 
     taxonomy %>% 
     column_to_rownames(var = 'taxon_id') %>% 
     as.matrix()

asvtab <- matrix(data = 1,
                 nrow = 1,
                 ncol = dim(taxtab)[1],
                 dimnames = list('all', row.names(taxtab)))

ps.ref <- 
     phyloseq(tax_table(taxtab),
              otu_table(asvtab, taxa_are_rows = FALSE))

ps.ref
```

```{r}
x.ref <- parse_phyloseq(ps.ref)
```

377 taxa, and 117 terminal nodes
Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x.ref$data$tax_abund <- calc_taxon_abund(x.ref, 
                                         'otu_table')

dim(x.ref$data$tax_abund)

# Convert to presence-absence
x.ref$data$tax_abund <- counts_to_presence(x.ref, 'tax_abund')
x.ref$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '1', 
          paste0(plotID, '_Full reference.pdf')),
         height = 10, width = 10)

heat_tree(
     x.ref,
     node_label = taxon_names,
     # node_color = n_samples,
     # node_color_range = pal[1:7],
     # node_size = n_samples,
     # node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

#### Realistic tree

This is with last common ancestor labeling done as would be in object.

- Group by sequence
     + Are any subsequences?
- Do last common ancestor
- Get "full" taxonomy
- Extend phyloseq object

```{r}
# Are any sequences substrings of one another?
seqs <- db.df$seq

substring <- function(s){
     # Remove s from the current list of strings to be considered
     seqs.temp <- seqs[seqs!=s]
     # Now check this list for membership
     any(str_detect(pattern = s,
                string = seqs.temp))
}

db.df$ss <- 
     lapply(db.df$seq,
            substring) %>% 
     unlist()

db.df %>% 
     filter(ss) %>% View()

```

#### Add possible taxa

Here, need to add empty rows to phyloseq object and corresponding taxonomy for everything that could have potentially been detected by the reference.  

```{r}
asvtab <- ps@otu_table@.Data
taxtab <- ps@tax_table@.Data
```

```{r}
# Pull only sequences that don't have an ASV table sequence as a substring

add <- c()

for (s in db.df$seq){
     ss <- 
          lapply(colnames(asvtab),
                  function(x){str_detect(pattern = x,
                                         string = s)}) %>% 
          unlist() %>% 
          any()
     
     # If no ASV as a substring, add it to list
     if(!ss){
          add <- c(add, s)
     }
}

length(add)
length(unique(add))
```

```{r}
# Pull rows to join to taxonomy table
add.tax <- 
     db.df %>% 
     filter(seq %in% add) 
```

Need to label duplicated sequences with their last common ancestor to get this working properly.

```{r eval = FALSE}
# Group by ASV
lca <- 
     add.tax %>%
     group_by(seq) %>%
     summarize_all(n_distinct) %>%
     column_to_rownames(var = 'seq')

# Now, relabel all those with >1 name at a particular level as NA
# As a placeholder, keep only the first species, knowing it will be overwritten
add.tax.lca <- 
     add.tax %>%
     group_by(seq) %>%
     slice_head(n = 1) %>%
     column_to_rownames(var = 'seq')
```

Now do relabeling 
```{r eval = FALSE}
# Confirm ordering is okay
all(rownames(lca) == rownames(add.tax.lca))

add.tax.lca[lca > 1] = NA
```

```{r}
# Combine rows
taxtab <- 
     rbind(taxtab,
           add.tax.lca) %>% 
     distinct() # Some of the taxa were already detected
```

Now can update ASV table

```{r}
# Make empty columns for ASV table
add.asv <- 
     matrix(0, 
            nrow = dim(asvtab)[1],
            ncol = dim(add.tax.lca)[1])

colnames(add.asv) <- rownames(add.tax.lca)
rownames(add.asv) <- rownames(asvtab)

asvtab <- cbind(asvtab, add.asv)
```

Need to make one dummy sample to contain all these taxa, so that the zeros are then labeled in the plot below.

```{r}
# Add to ASV table
row <- matrix(1, ncol = dim(asvtab)[2])
rownames(row) <- 'Reference'
colnames(row) <- colnames(asvtab)

asvtab <- rbind(asvtab, row)
```

```{r}
# Add to sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row')

# Add a distinguishing variable
samdf$type <- 'sample' # All existing rows are samples

row <- matrix(NA, ncol = dim(samdf)[2])
colnames(row) <- names(samdf)
row <- 
     data.frame(row) %>% 
     mutate(row = 'Reference',
            type = 'reference')

samdf <- 
     bind_rows(samdf, row) %>% 
     column_to_rownames(var = 'row') %>% 
     select(-sample) # Need to toss this, otherwise it's looked up below
``` 

```{r}
dim(asvtab)
dim(samdf)
dim(taxtab)
```

```{r}
# Update phyloseq object
ps

ps <- phyloseq(otu_table(asvtab, taxa_are_rows = FALSE),
               sample_data(samdf),
               tax_table(as.matrix(taxtab)))

ps
```

```{r}

```


### All samples

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table')

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Samples detected, size.pdf')),
         height = 10, width = 10)

pal <- brewer.pal(9, 'Greens')

heat_tree(
     x,
     node_label = taxon_names,
     node_color = n_samples,
     node_color_range = pal[1:7],
     node_size = n_samples,
     node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

Now, try with group color
Need to re-make data structure.

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$study)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Detected (Adolescent).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(Adolescent, pal[7], "grey"))
```
Color argument "accepts a vector, the same length taxon_id or a factor of its length."
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results',
         'manuscript',
         '5', 
          paste0(plotID, '_Detected (Adult-1).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, pal[7], "grey"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 
         'manuscript', 
         '5', 
         paste0(plotID, '_Detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, pal[7], "grey"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_Detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.005, 0.025),
          node_size = n_samples,
          node_legend_title = element_blank(),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, pal[7], "grey"))
```

#### Visualize

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$type)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript',
         '1', 
          paste0(plotID, '_B.pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(sample, "darkgreen", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adolescent).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "grey", "red"),
          node_label = taxon_names)
```

## UpSet plot

```{r}
# Update taxa names to their lowest specified level
all(taxa_names(ps) == row.names(taxtab)) 
taxa_names(ps) <- lowest_level(taxtab)$name
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
146 - length(weightloss.taxa)
146 - length(adult1.taxa)
146 - length(adult2.taxa)
146 - length(adolescent.taxa)
```

```{r}
# Summarize phyloseq by cohort
listInput <- list(`Weight Loss` = weightloss.taxa,
                  `Adult-1` = adult1.taxa,
                  `Adult-2` = adult2.taxa,
                  `Adolescent` = adolescent.taxa)

upset(fromList(listInput), order.by = 'freq')
```
