---
title: "Figure 1A"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(Biostrings)
library(here)
library(phyloseq)
library(MButils) # For lowest level naming
library(metacoder)
library(RColorBrewer)
library(tidyverse)
library(vegan)
library(UpSetR)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

## Reference database

```{r}
db <- 
     here('..', 
          'food-dbs', 
          'data', 
          'processed',
          'dada2-compatible',
          'trnL',
          '20210825_trnL_prior_format (taxonomy).fasta') %>% 
     readDNAStringSet()
```

# Pre-process

## trnL
```{r}
# First, remove name column from taxonomy, as not needed here
taxtab <- 
     ps@tax_table %>% 
     data.frame() %>% 
     select(-name)

tax_table(ps) <- tax_table(as.matrix(taxtab))
```

### Filter

```{r}
# # Remove any taxon seen with 3 or fewer counts in a sample
# # In current data, this is 60 entries
# asvtab <- ps@otu_table@.Data
# asvtab[asvtab <= 3] <- 0
# 
# # Replace in phyloseq object
# otu_table(ps) <- otu_table(asvtab,
#                            taxa_are_rows = FALSE)
```

## Reference

Organize in a taxonomy-table style.

```{r}
db
```

```{r}
db.df <- 
     data.frame(name = names(db),
                seq = as.character(db))

db.df
```
```{r}
# Separate name and assign ranks
db.df <- 
     db.df %>% 
     # Trim terminal semicolon, not needed here
     mutate(name = gsub(name, pattern = ';$', replacement = '')) %>% 
     separate(name, 
              into = c( "superkingdom",
                        "kingdom",
                        "phylum",
                        "order",
                        "family",
                        "genus",
                        "species",
                        "subspecies or varietas"),
              sep = ';') # Based on current structure any sub-species-level designation will land here.
```

Note that this has subspecies as a column, whereas phyloseq object does not.  Likely need to go back and correct this.  But for clearer merge, just drop subspecies column for now. 

```{r}
dim(db.df)
db.df <- 
     db.df %>% 
     filter(!(grepl(`subspecies or varietas`,
                    pattern = 'subsp\\.')))
dim(db.df)
```

```{r}
# Now can rename to varietas column
names(db.df)[names(db.df) == 'subspecies or varietas'] <- 'varietas'
```

Also need to manually correct cyanobacterial by shifting all entries right
```{r}
cyano <- 
     db.df %>% 
     filter(superkingdom != 'Eukaryota')

# Remove these rows from the full reference
db.df <- 
     db.df %>% 
     filter(superkingdom == 'Eukaryota')

cyano
```
```{r}
# Correct
cyano <- 
     cyano %>% 
     select(-species) %>% # This column is empty
     select(species = genus,
            genus = family,
            family = order,
            order = phylum,
            phylum = kingdom,
            everything()) %>% 
     mutate(kingdom = NA)

cyano
```

```{r}
db.df <- bind_rows(db.df, cyano)
```

# Visualize taxa

## Metacoder tree

### All samples
```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

377 taxa, and 117 terminal nodes
Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table')

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data (base plot -- no group color)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_n samples detected, size.pdf')),
         height = 10, width = 10)

pal <- brewer.pal(9, 'Greens')

heat_tree(
     x,
     node_label = taxon_names,
     node_color = n_samples,
     node_color_range = pal[1:7],
     node_size = n_samples,
     node_size_range = c(0.005, 0.025),
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)'
)
```

Now, try with group color
Need to re-make data structure.

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$study)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adolescent).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "red", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adolescent).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "grey", "red"),
          node_label = taxon_names)
```
Color argument "accepts a vector, the same length taxon_id or a factor of its length."
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adult-1).pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, "red", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adult-1).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-1`, "grey", "red"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, "red", "grey"))

# Not detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adult-2).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Adult-2`, "grey", "red"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# Detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, "red", "grey"))

# Not detected
pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Weight Loss).pdf')),
         height = 10, width = 10)

heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(x$data$tax_abund$`Weight Loss`, "grey", "red"))
```

### Possible v actual detection

#### Add possible taxa

Here, need to add empty rows to phyloseq object and corresponding taxonomy for everything that could have potentially been detected by the reference.  

```{r}
asvtab <- ps@otu_table@.Data
taxtab <- ps@tax_table@.Data
```

```{r}
add <- db.df$seq[!(db.df$seq %in% colnames(asvtab))]
```

```{r}
# Pull rows to join to taxonomy table
add.tax <- 
     db.df %>% 
     filter(seq %in% add) 
```

Need to label duplicated sequences with their last common ancestor to get this working properly.

```{r eval = FALSE}
# Group by ASV
lca <- 
     add.tax %>%
     group_by(seq) %>%
     summarize_all(n_distinct) %>%
     column_to_rownames(var = 'seq')

# Now, relabel all those with >1 name at a particular level as NA
# As a placeholder, keep only the first species, knowing it will be overwritten
add.tax.lca <- 
     add.tax %>%
     group_by(seq) %>%
     slice_head(n = 1) %>%
     column_to_rownames(var = 'seq')
```

Now do relabeling 
```{r eval = FALSE}
# Confirm ordering is okay
all(rownames(lca) == rownames(add.tax.lca))

add.tax.lca[lca > 1] = NA
```

```{r}
# Combine rows
taxtab <- rbind(taxtab,
                add.tax.lca)
```

Now can update ASV table

```{r}
# Make empty columns for ASV table
add.asv <- 
     matrix(0, 
            nrow = dim(asvtab)[1],
            ncol = dim(add.tax.lca)[1])

colnames(add.asv) <- rownames(add.tax.lca)
rownames(add.asv) <- rownames(asvtab)

asvtab <- cbind(asvtab, add.asv)
```

Need to make one dummy sample to contain all these taxa, so that the zeros are then labeled in the plot below.

```{r}
# Add to ASV table
row <- matrix(1, ncol = dim(asvtab)[2])
rownames(row) <- 'Reference'
colnames(row) <- colnames(asvtab)

asvtab <- rbind(asvtab, row)
```

```{r}
# Add to sample data
samdf <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row')

# Add a distinguishing variable
samdf$type <- 'sample' # All existing rows are samples

row <- matrix(NA, ncol = dim(samdf)[2])
colnames(row) <- names(samdf)
row <- 
     data.frame(row) %>% 
     mutate(row = 'Reference',
            type = 'reference')

samdf <- 
     bind_rows(samdf, row) %>% 
     column_to_rownames(var = 'row') %>% 
     select(-sample) # Need to toss this, otherwise it's looked up below
``` 

```{r}
dim(asvtab)
dim(samdf)
dim(taxtab)
```

```{r}
# Update phyloseq object
ps

ps <- phyloseq(otu_table(asvtab, taxa_are_rows = FALSE),
               sample_data(samdf),
               tax_table(as.matrix(taxtab)))

ps
```

#### Visualize

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 
                                     'otu_table',
                                     groups = x$data$sample_data$type)

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

pdf(here('results', 
         'manuscript',
         '1', 
          paste0(plotID, '_B.pdf')),
         height = 10, width = 10)

# Detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(sample, "darkgreen", "grey"))

pdf(here('results', 'manuscript', '5', 
          paste0(plotID, '_A_not detected (Adolescent).pdf')),
         height = 10, width = 10)

# Not detected
heat_tree(x, 
          node_size_range = c(0.01, 0.01),
          node_color = ifelse(Adolescent, "grey", "red"),
          node_label = taxon_names)
```

## UpSet plot

```{r}
# Update taxa names to their lowest specified level
all(taxa_names(ps) == row.names(taxtab)) 
taxa_names(ps) <- lowest_level(taxtab)$name
```

```{r}
adult1.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-1') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adult2.taxa <- 
     ps %>% 
     subset_samples(study == 'Adult-2') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

weightloss.taxa <- 
     ps %>% 
     subset_samples(study == 'Weight Loss') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()

adolescent.taxa <- 
     ps %>% 
     subset_samples(study == 'Adolescent') %>% 
     prune_taxa(taxa_sums(.) > 0, .) %>% 
     taxa_names()
```

```{r}
# How many detected in each?
length(weightloss.taxa)
length(adult1.taxa)
length(adult2.taxa)
length(adolescent.taxa)
```

```{r}
# Check names equal count above
union(weightloss.taxa, adult1.taxa) %>% 
     union(adult2.taxa) %>% 
     union(adolescent.taxa) %>% 
     length()
```

```{r}
# How many *not* detected in each?
118 - length(weightloss.taxa)
118 - length(adult1.taxa)
118 - length(adult2.taxa)
118 - length(adolescent.taxa)
```

```{r}
# Summarize phyloseq by cohort
listInput <- list(`Weight Loss` = weightloss.taxa,
                  `Adult-1` = adult1.taxa,
                  `Adult-2` = adult2.taxa,
                  `Adolescent` = adolescent.taxa)

upset(fromList(listInput), order.by = 'freq')
```
