---
title: "Figure 1A"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(metacoder)
library(RColorBrewer)
library(tidyverse)
library(vegan)
```

# Read in data

## trnL
Note this is already trimmed to only those taxa identified to a food species, with duplicated taxa collapsed.
```{r}
# Read in last saved phyloseq
ps <- 
     here('data', 'processed', 'phyloseq', 'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

# Tidy

```{r}
# First, remove name column from taxonomy, as not needed here
taxtab <- 
     ps@tax_table %>% 
     data.frame() %>% 
     select(-name)

tax_table(ps) <- tax_table(as.matrix(taxtab))
```

## Filter

```{r}
# Remove any taxon seen with 3 or fewer counts in a sample
asvtab <- ps@otu_table@.Data
asvtab[asvtab <= 3] <- 0

# Replace in phyloseq object
otu_table(ps) <- otu_table(asvtab,
                           taxa_are_rows = FALSE)
```

# Visualize taxa

## Metacoder tree

```{r}
# Convert phyloseq object to taxmap
x <- parse_phyloseq(ps)
```

Helpful functions for accessing data
```{r}
taxon_names(x) %>% head() # Taxa names
taxon_ranks(x) %>% head() # Taxa ranks
```

377 taxa, and 117 terminal nodes
Okay!  So calc_taxon_abund converts "per-observation counts (e.g. OTU counts) to per-taxon counts", where the per-taxon counts span the full tree
```{r}
# Following tutorial https://github.com/grunwaldlab/metacoder/issues/267: 
# 1. calc_taxon_abund
# 2. counts_to_presence

x$data$tax_abund <- calc_taxon_abund(x, 'otu_table')

dim(x$data$tax_abund)

# Convert to presence-absence
x$data$tax_abund <- counts_to_presence(x, 'tax_abund')
x$data$tax_abund
```
```{r}
# Convert to numeric, in order to...
x$data$tax_abund <- 
     x$data$tax_abund %>% 
     mutate(across(-taxon_id, as.numeric))

# ...total counts per taxon in this presence-absence scheme
x$data$tax_occ <- 
     calc_n_samples(x, 'tax_abund')
```

```{r}
# Plot heat tree based on presence-absence data
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

gradient <- brewer.pal(9, 'YlOrRd')

heat_tree(
     x,
     node_label = taxon_names,
     node_size = n_samples,
     node_color = n_samples,
     node_legend_title = element_blank(),
     node_color_axis_label = 'Samples (n)',
     node_color_range = c(first(gradient), last(gradient)),
     output_file = here('results', 'manuscript', '1', 
          paste0(plotID, '_A_n samples detected, size and color.pdf'))
)
```

