---
title: "Figure 1"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(car) # For vif()
library(fitdistrplus)
library(here)
library(lme4) # For glmer()
library(phyloseq)
library(tidyverse)

# Make dplyr select the default
select <- dplyr::select
```

# Load data

## trnL richness (+/- rarefaction)

```{r}
alpha <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS()
```

## Weight Loss menu data

```{r}
# Menu phyloseq object (taxa consumed)
ps.menu <- 
     here('data', 
          'processed', 
          'Weight Loss', 
          '20210220_ps_menu.rds') %>% 
     readRDS()
```

```{r}
# Menu summaries (nutrients consumed)
nutrients <- 
     here('data',
          'processed',
          'Weight Loss',
          'Nutrients per day.csv') %>% 
     read_csv()
```

# Pre-process

## Metabarcoding data

```{r}
# Subset to Weight Loss samples and metadata only
alpha <- 
     alpha %>% 
     filter(study == 'Weight Loss') %>% 
     # Drop columns that are all NA
     select(
          where(function(x){ 
               !all(is.na(x))
          })
     ) %>% 
      # Re-order remainder
     select(
          subj,
          date,
          obs, 
          threshold, 
          starts_with('obs_rare')
          ) 

alpha
```

```{r}
# Add link to menu day at varying lag

# Label each stool sample with a week of preceding days as SUBJ_DATE
for (delta in 1:7){
     varname <- paste0('delta', delta)
     alpha <- mutate(alpha, 
                     !!varname := date - delta,)
}

alpha
```
## Menu data

```{r}
head(sample_names(ps.menu))
```

```{r}
# Filter to plants only
ntaxa(ps.menu)
ps.menu <- subset_taxa(ps.menu,
                       kingdom == 'Viridiplantae')
ntaxa(ps.menu)
```

### Normalize by kcal

```{r}
macros <- c('carbohydrates (g)',
            'fat (g)',
            'protein (g)',
            'dietary fiber (g)')

nutrients <- 
     nutrients %>% 
     mutate(across(all_of(macros), 
                   ~ . / calories,
                   .names = '{.col}_norm'))
```

### % cal by nutrient

```{r}
# 4 kcal/gram for carbs, protein
nutrients <- 
     nutrients %>% 
     mutate(across(c(`carbohydrates (g)`, `protein (g)`, ), 
                   ~ . * 4/ calories,
                   .names = '{.col}_pct_tot'))

# 9 kcal/g for fat
nutrients$`fat (g)_pct_tot` <- nutrients$`fat (g)`*9/nutrients$calories
```

```{r}
# Sanity check
nutrients %>% select(ends_with('pct_tot')) %>% rowSums() %>% round()
```

Ok, good, all within rounding error. 

### Calculate richness

```{r}
# Get intake table
asvtab.menu <- ps.menu@otu_table@.Data 
# Calculate number of non-zero entries
alpha.menu <- apply(asvtab.menu,
                    1, 
                    function(x){sum(x > 0)})  

# Reformat as dataframe
alpha.menu <- 
     data.frame(obs_menu = alpha.menu) %>% 
     rownames_to_column(var = 'row') %>% 
     separate(row, 
              into = c('subj', 'date_menu'), 
              sep = '_')
```

```{r}
# Add weekday
alpha.menu <- 
     alpha.menu %>% 
     mutate(date_menu = as.Date(date_menu, format = '%Y-%m-%d'),
            weekday_menu = weekdays(date_menu)) %>% 
     select(subj, date_menu, weekday_menu, obs_menu)

alpha.menu
```
```{r}
# Join to nutrient intakes
alpha.menu <- 
     left_join(alpha.menu, 
               nutrients,
               by = c('subj', 
                      'date_menu' = 'date', 
                      'weekday_menu' = 'weekday'))

rm(nutrients)
```

# Analyze

## Range of taxa 

Overall dataset

```{r}
range(alpha.menu$obs_menu)
```

What if I filter out weekends (lower quality?)
```{r}
alpha.menu %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     pull(obs_menu) %>% 
     range()
```

Median within each subject

```{r}
# Including weekends 
alpha.menu %>% 
     group_by(subj) %>% 
     summarize(median = median(obs_menu),
               sd = sd(obs_menu))
```

```{r}
# Excluding weekends
alpha.menu %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     group_by(subj) %>% 
     summarize(median = median(obs_menu),
               sd = sd(obs_menu))
```

## Correlation with trnL richness

Have two dataframes:

* `alpha`
* `alpha.menu`

First, might want to understand day-to-day variability in trnL richness compared to menu richness.

### Raw richness

```{r}
# 1 day of lag
lag <- 'delta1'

# Make joint object
alpha.joint <- 
     alpha %>%
     select(subj,
            date,
            date_menu = all_of(lag), 
            starts_with('obs')) %>% 
     # mutate(delta = gsub('delta', '', lag)) %>% 
     left_join(alpha.menu) %>% 
     select(subj,
            starts_with('date'),
            weekday_menu,
            starts_with('obs'),
            everything())

head(alpha.joint)
```
Range of menu taxa compared to range of trnL taxa
```{r}
# Pivot data to long form by method
alpha.range <- 
     alpha.joint %>% 
     select(subj, obs, obs_menu) %>% 
     pivot_longer(cols = -subj,
                  names_to = 'measure',
                  values_to = 'richness') %>% 
     mutate(measure = ifelse(measure == 'obs',
                             yes = 'Metabarcoding',
                             no = 'Menu'))

alpha.range
```
```{r}
# Factor subjects
alpha.range$subj <- 
     factor(alpha.range$subj,
            levels = c('JN55', 'AK65', 'PC74', 'ML48'),
            labels = 1:4)
```

```{r}
ggplot(alpha.range, aes(x = subj, y = richness, color = measure)) +
     geom_boxplot() +
     scale_color_manual(values = c('black', 'darkgreen')) +
     labs(x = 'Participant', 
          y = 'Plant taxon richness', 
          color = 'Measure') +
     theme_classic() +
     theme(axis.title = element_text(size = 14),
           legend.position = c(1, 0.8),
           legend.justification = 'right')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'manuscript', '1',
            paste0(plotID, '_A_trnL richness x menu richness.pdf')),
       height = 3, width = 4)
```

```{r}
ggplot(alpha.joint, aes(x = obs_menu, y = obs)) +
     geom_point(size = 2, alpha = 0.8) +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown()) +
     theme(axis.title = element_text(size = 14))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript', 
#             '1',
#             paste0(plotID, '_B_trnL richness x menu richness.pdf')),
#        height = 3, width = 3)
```

```{r}
cor.test(alpha.joint$obs,
         alpha.joint$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

### Rarefied richness
```{r}
cor.test(alpha.joint$obs_rare,
         alpha.joint$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

```{r}
ggplot(alpha.joint, aes(x = obs_menu, y = obs_rare)) +
     geom_point(size = 2, alpha = 0.8) +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown()) +
     theme(axis.title = element_text(size = 14))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 'manuscript', '1',
            paste0(plotID, '_B_trnL richness x menu richness.pdf')),
       height = 3, width = 3)
```


Should I test other days, or leave it at that?

What about excluding weekends?

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>%
     cor.test(~ obs_rare + obs_menu, 
              data = .,
              method = 'spearman',
              alternative = 'greater')
```

How about Sundays, Mondays only

```{r}
alpha.joint %>% 
     filter(weekday_menu %in% c('Saturday', 'Sunday')) %>%
     ggplot(aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```


```{r}
alpha.joint %>% 
     filter(weekday_menu %in% c('Saturday', 'Sunday')) %>%
     cor.test(~ obs_rare + obs_menu, 
              data = .,
              method = 'spearman')
```

How about in comparison to the day after today?

```{r}
alpha.next <- 
     alpha %>% 
     select(subj, date, obs_rare) %>% 
     mutate(date_menu = date + 1) %>% 
     left_join(alpha.menu) %>% 
     select(subj,
            starts_with('date'),
            weekday_menu,
            starts_with('obs'),
            everything())

alpha.next
```

```{r}
ggplot(alpha.next, aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```
```{r}
cor.test(alpha.next$obs_rare,
         alpha.next$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

How about with a random day (within each subject)?

```{r}
# Make joint object
alpha.rand <- 
     alpha %>%
     select(subj, date, obs_rare)

# Randomly choose a comparator day (within each subject)
sample_subj <- function(s, n){
     alpha.menu %>% 
          filter(subj == s) %>% 
          pull(row) %>% 
          sample(size = n, replace = FALSE)
}
```

```{r}
# How many entries per subject do we need?
alpha.rand %>% 
     group_by(subj) %>% 
     count()
```

How about with a random day (any day)?

```{r}
alpha.rand <- 
     alpha %>%
     mutate(sample = paste(subj, date, sep = '_'))  %>% 
     select(subj, sample, obs_rare)

dim(alpha.rand)
```

```{r}
set.seed(1234)
alpha.rand <- 
     bind_cols(alpha.rand, 
               row = sample(alpha.menu$row, 32))

alpha.rand <- 
     alpha.rand %>% 
     select(-subj) %>% 
     left_join(alpha.menu)
```

```{r}
ggplot(alpha.rand, aes(x = obs_menu, y = obs_rare)) +
     geom_point() +
     lims(x = c(7, 46), y = c(7, 46)) +
     labs(x = 'Plant taxa (menu)', y = 'Plant taxa (*trnL*)') +
     coord_equal() +
     theme_bw() +
     theme(axis.title.y = ggtext::element_markdown())
```
```{r}
cor.test(alpha.rand$obs_rare,
         alpha.rand$obs_menu,
         method = 'spearman',
         alternative = 'greater')
```

## Relationship to macronutrient intake

Start with just one day's comparison. Can use joint object made above.

### Calories
```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = calories, y = obs_rare, color = subj)) +
     geom_point()
```

### Carbohydrates

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `carbohydrates (g)`, y = obs_rare, color = subj)) +
     geom_point()
```

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `carbohydrates (g)_pct_tot`, y = obs_rare, color = subj)) +
     geom_point()
```

### Fat

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `fat (g)`, y = obs_rare, color = subj)) +
     geom_point()
```

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `fat (g)_pct_tot`, y = obs_rare, color = subj)) +
     geom_point()
```

### Protein

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `protein (g)`, y = obs_rare, color = subj)) +
     geom_point()
```

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `protein (g)_pct_tot`, y = obs_rare, color = subj)) +
     geom_point()
```


### Fiber

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `dietary fiber (g)`, y = obs_rare, color = subj)) +
     geom_point()
```

```{r}
alpha.joint %>% 
     filter(!(weekday_menu %in% c('Saturday', 'Sunday'))) %>% 
     ggplot(aes(x = `dietary fiber (g)_norm`, y = obs_rare, color = subj)) +
     geom_point()
```

## Fitted model

```{r}
# First, want to remove observations following weekends.
alpha.joint.weekday <- 
     filter(
          alpha.joint,
          !(weekday_menu %in% c('Saturday', 'Sunday'))
     )
```

```{r}
# How many observations does this remove?
dim(alpha.joint)[1]
dim(alpha.joint.weekday)[1]
```

Think multicollinearity may be an issue here.

```{r}
# Try as a linear model? 
# kcal-normalized macronutrients
fit1 <- glm(obs_rare ~ 
                `fat (g)_norm` +
                `carbohydrates (g)_norm` +
                `protein (g)_norm` +
                 calories,
           family = Gamma(link = 'identity'),
           data = alpha.joint.weekday)

summary(fit1)
```

```{r}
vif(fit1)
```
Wow, these are super high.  Apparently VIF shouldn't be much greater than 5. 

```{r}
predictors <- c('calories',
                'fat (g)_norm',
                'carbohydrates (g)_norm',
                'protein (g)_norm')

select(alpha.joint.weekday, one_of(predictors)) %>% 
     apply(2, range, na.rm = TRUE)
```

```{r}
# Re-scale and center
alpha.joint.weekday <- 
     alpha.joint.weekday %>% 
     mutate(across(one_of(predictors),
                   scale,
                   .names = '{.col}_scale'))
```

```{r}
fit2 <- glmer(obs_rare ~ 
                   # `fat (g)_norm_scale` +
                   `carbohydrates (g)_norm_scale` +
                   `protein (g)_norm_scale` +
                   calories_scale +
                   (1 | subj),
           family = Gamma(link = 'identity'),
           data = alpha.joint.weekday)

summary(fit2)
```

Seems like there could be multiple reasons for singular fit warning, from [this thread](https://stackoverflow.com/questions/54597496/how-to-cope-with-a-singular-fit-in-a-linear-mixed-model-lme4):
- "...collinearity in fixed effects, as in any other linear model. That would need you to revise your model by removing terms."
- "But in lmer, that (or a "boundary (singular) fit" warning) can also be also triggered in quite simple models when a random effect variance is estimated very near zero and (very loosely) the data is not sufficiently informative to drag the estimate away from the zero starting value."

```{r}
isSingular(fit2)
```

The random effect variance here *is* quite small, and there's also collinearity. 

```{r}
anova(fit1, fit2, test="Chisq")
```

