---
title: "Adult-1, Adult-2 nutrient relationships"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(fitdistrplus)
library(here)
library(lme4) # For glmer()
library(phyloseq)
library(tidyverse)

# Make dplyr select the default
select <- dplyr::select
```

# Load data

## trnL richness (+/- rarefaction)

```{r}
alpha <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (10,500 reads).rds') %>% 
     readRDS()
```

## FFQ data

This is processed in a separate notebook ("Adult-1, Adult-2 diet survey data").  All nutrient and food group calories are present both in raw form and in an amount per 1,000-kcal in the diet.

```{r}
ffq <- 
     here('data', 
          'processed',
          'DHQ3', 
          'Adult-1 and Adult-2 FFQ with per-1000-kcal calculated nutrients.csv'
          ) %>% 
     read_csv()
```

# Pre-process

## Metabarcoding data

```{r}
# Subset to Weight Loss samples and metadata only
alpha <- 
     alpha %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     # Drop columns that are all NA
     select(
          where(function(x){ 
               !all(is.na(x))
          })
     ) %>% 
      # Re-order remainder
     select(
          study,
          subj,
          obs, 
          threshold, 
          starts_with('obs_rare')
          ) %>% 
     # Convert subject to factor
     mutate(subj = factor(subj)) %>% 
     arrange(study, subj)

alpha
```
## Diet survey data

```{r}
# Crude filter: refine later
ffq <- 
     filter(ffq, 
            !(`Energy (kcal)` < 500 | `Energy (kcal)` > 3500))
```

This has a lot of columns: subset to only those we'll use downstream: calories, macronutrients, and fiber.

```{r}
ffq <- 
     select(ffq,
            subj,
            `Energy (kcal)`,
            starts_with('Carbohydrate (g)'),
            starts_with('Total fat (g)'),
            starts_with('Protein (g)'),
            starts_with('Dietary fiber (g)'),
            starts_with('Energy from'))
```

# Analyze

## Relationship to macronutrient intake

### Each sample independently

```{r}
trnL.ffq.individual <- 
     left_join(alpha, ffq)
```

```{r}
# Create a skewness-kurtosis plot to find the best distrubution for your data
descdist(trnL.ffq.individual$obs_rare, discrete = FALSE, boot = 500)
```

```{r}
##fit your data to a variety of distributions to doublecheck the above results. 
##Make determination based on Q-Q plot
fit.norm <- fitdist(trnL.ffq.individual$obs_rare, "norm")
fit.gamma <- fitdist(trnL.ffq.individual$obs_rare, "gamma")
fit.weibull <- fitdist(trnL.ffq.individual$obs_rare, "weibull")

##look at the plots
plot(fit.norm)
plot(fit.gamma)
plot(fit.weibull)
```

```{r}
# Try as a linear model? 
fit1 <- glm(obs_rare ~ 
                `Energy (kcal)` +
                `Total fat (g)_per_1000kcal` +
                `Carbohydrate (g)_per_1000kcal` +
                `Protein (g)_per_1000kcal`,
           family = Gamma(link = 'identity'),
           data = trnL.ffq.individual)

summary(fit1)
```

Also tried a log link: AIC 2447.7.  Deviance and residual deviance are the same.

```{r}
# Now, fit a model with subject as random effect
fit2 <- glmer(obs_rare ~ 
                `Energy (kcal)` +
                `Total fat (g)_per_1000kcal` +
                `Carbohydrate (g)_per_1000kcal` +
                `Protein (g)_per_1000kcal` + 
                 (1 | subj),
           family = Gamma(link = 'identity'),
           data = trnL.ffq.individual)

summary(fit2)
```

Ok, so predictors are very different scale:

```{r}
predictors <- c('Energy (kcal)',
                'Total fat (g)_per_1000kcal',
                'Carbohydrate (g)_per_1000kcal',
                'Protein (g)_per_1000kcal')

select(trnL.ffq.individual, one_of(predictors)) %>% 
     apply(2, range, na.rm = TRUE)
```

```{r}
# Re-scale and center
trnL.ffq.individual <- 
     trnL.ffq.individual %>% 
     mutate(across(one_of(predictors),
                   scale,
                   .names = '{.col}_scale'))
```

```{r}
# Now try again with these scaled variables
fit2 <- glmer(obs_rare ~ 
                `Energy (kcal)_scale` +
                `Total fat (g)_per_1000kcal_scale` +
                `Carbohydrate (g)_per_1000kcal_scale` +
                `Protein (g)_per_1000kcal_scale` + 
                 (1 | subj),
           family = Gamma(link = 'identity'),
           data = trnL.ffq.individual)

summary(fit2)
```

### Average richness

```{r}
# Combine richness estimates and FFQ data
trnL.ffq.average <- 
     alpha %>% 
     group_by(subj) %>% 
     summarize(obs_mean = mean(obs_rare)) %>% 
     left_join(ffq)
```

```{r}
# Try as a linear model? 
fit1 <- glm(obs_mean ~ 
                `Energy (kcal)` +
                `Total fat (g)_per_1000kcal` +
                `Carbohydrate (g)_per_1000kcal` +
                `Protein (g)_per_1000kcal`,
           family = Gamma(link = 'identity'),
           data = trnL.ffq.average)

summary(fit1)
```

### Visualize

```{r}
ggplot(trnL.ffq.average, aes(x = `Energy (kcal)`, y = obs_mean)) +
     geom_point()
```
```{r}
ggplot(trnL.ffq.average, aes(x = `Carbohydrate (g)_per_1000kcal`, y = obs_mean)) +
     geom_point()
```
```{r}
ggplot(trnL.ffq.average, aes(x = `Protein (g)_per_1000kcal`, y = obs_mean)) +
     geom_point()
```
```{r}
ggplot(trnL.ffq.average, 
       aes(x = `Dietary fiber (g)_per_1000kcal`, y = obs_mean)) +
     geom_point()
```

## Relationship to HEI

## Relationship to
