---
title: "Adult-1, Adult-2 diet survey data"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(here)
library(phyloseq)
library(scales)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

# Read in data

## DHQ3

### Summary responses

```{r}
# Adult-1
summary.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3_summary.csv') %>% 
     read_csv()

head(summary.adult1)
```

```{r}
# Adult-2
summary.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3_summary.csv') %>% 
     read_csv()

head(summary.adult2)
```

### Detailed responses

```{r}
# Adult-1
detail.adult1 <- 
     here('data', 
          'processed', 
          'Adult-1', 
          'DHQ3_detail.csv') %>% 
     read_csv()

head(detail.adult1)
```

```{r}
# Adult-2
detail.adult2 <- 
     here('data', 
          'processed', 
          'Adult-2', 
          'DHQ3_detail.csv') %>% 
     read_csv()

head(detail.adult2)
```

### Calculated PDI

This is done individually in CHOMP and ONR project notebooks, and then loaded here.

```{r}
# Adult-1
pdi.adult1 <-
     here('data', 
          'processed', 
          'Adult-1', 
          'PDI scores by subject.csv') %>% 
     read_csv()

head(pdi.adult1)
```

```{r}
# Adult-2
pdi.adult2 <-
     here('data', 
          'processed', 
          'Adult-2', 
          'PDI scores by subject.csv') %>% 
     read_csv()

head(pdi.adult2)
```

## Demographic data

For Harris-Benedict equation

```{r}
# In sample metadata of trnL phyloseq object
samdf <- 
     here('data',
          'processed',
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS() %>% 
     sample_data() %>% 
     data.frame()
```

# Pre-process

```{r}
# Add study variable
summary.adult1$study <- 'Adult-1'
summary.adult2$study <- 'Adult-2'

detail.adult1$study <- 'Adult-1'
detail.adult2$study <- 'Adult-2'

pdi.adult1$study <- 'Adult-1'
pdi.adult2$study <- 'Adult-2'
```


```{r}
# Join together, arrange columns
summary <- 
     bind_rows(summary.adult1, summary.adult2) %>% 
     select(study, everything())

detail <- 
     bind_rows(detail.adult1, detail.adult2) %>% 
     select(study, everything())

pdi <- 
     bind_rows(pdi.adult1, pdi.adult2) %>% 
     select(study, everything())
```

```{r}
# Remove raw data files
rm(summary.adult1, 
   summary.adult2,
   detail.adult1, 
   detail.adult2,
   pdi.adult1,
   pdi.adult2)
```

## Tidy

```{r}
# Remove un-needed columns (metadata, supplements)
summary <- select(summary, 
                  -c(`Record Number`,
                     starts_with('SUPP'))
)
```

```{r}
# Re-code sex
summary <- 
     summary %>% 
     rename(sex = `Sex (1=male; 2=female)`) %>% 
     mutate(sex = factor(sex,
                         labels = c('Male', 'Female')))

detail <- 
     detail %>% 
     rename(sex = `Sex (1=male; 2=female)`) %>% 
     mutate(sex = factor(sex,
                         labels = c('Male', 'Female')))
```


```{r}
# Make labels plot-friendly
names(pdi)[names(pdi) == 'pdi'] <- 'PDI'
names(pdi)[names(pdi) == 'hpdi'] <- 'hPDI'
names(pdi)[names(pdi) == 'updi'] <- 'uPDI'
```

## QC and filter

From DHQ3 website:
There are no data cleaning or exclusion programs built into the software. Researchers can decide, based on break offs, energy, or other considerations, which respondents to exclude from analyses. Different investigators clean data in different ways: some like calorie exclusions, some like exclusions based on percentile cutpoints of nutrients or missing data, some like to transform data and then exclude, others want to exclude those respondents who did not take the task seriously (however that is defined). By looking at output files, one can write a program to check issues related to whether an individual always checks the same frequency or portion size. It is our experience that it is difficult to discern these types of errors from energy exclusions. FFQ data can be messy for some respondents and prone to measurement error for all.

Filtering criteria from Asnicar:
"Submitted FFQs were excluded if more than 10 food items were left unanswered or if the total energy intake estimate derived from the FFQ as a ratio of the participant’s estimated basal metabolic rate (determined by the Harris–Benedict equation) was more than 2 s.d. outside the mean of this ratio (<0.52 or >2.58)."
This is for a Harvard FFQ.

A minimum 10 kcal filter removes 3 participants who it appears minimally completed the survey. (Is there a DHQ3-reported parameter that reads this out??)

Other ideas for more data cleaning-- check for survey completeness, change sample ID to CHOMP participant ID, quality/sanity checks on data, filter incomplete responses...

```{r}
# Get subject demographics to calculate Harris-Benedict
samdf <- 
     samdf %>% 
     filter(study %in% c('Adult-1', 'Adult-2')) %>% 
     select(study, subj,
            age, gender, kg, m) %>% 
     distinct()

samdf
```
```{r}
# Join to reported kcal intakes
samdf <- 
     summary %>% 
     select(subj,
            kcal = `Energy (kcal)`) %>% 
     right_join(samdf) %>% 
     select(study, subj, kcal, everything())

samdf
```

### Harris-Benedict equation

Following latest revised version [here](https://en.wikipedia.org/wiki/Harris–Benedict_equation).

```{r}
samdf <-
     samdf %>% 
     mutate(harris_benedict = 
                 ifelse(gender == 'Male',
                        yes = 10*kg + 6.25*(100*m) - 5*age + 5,
                        no = 10*kg + 6.25*(100*m) - 5*age + 161))
```

In Asnicar, FFQ data was excluded if "Submitted FFQs were excluded if more than 10 food items were left unanswered or if the total energy intake estimate derived from the FFQ as a ratio of the participant’s estimated basal metabolic rate (determined by the Harris–Benedict equation) was more than 2 s.d. outside the mean of this ratio (<0.52 or >2.58)."

```{r}
# Calculate ratio
samdf <- 
     samdf %>% 
     mutate(energy_ratio = harris_benedict/kcal) %>% 
     select(study, 
            subj, 
            kcal,
            harris_benedict,
            energy_ratio)

# What's mean and SD?
median(samdf$energy_ratio, na.rm = TRUE)
mad(samdf$energy_ratio, na.rm = TRUE)
```

```{r}
median(samdf$energy_ratio, na.rm = TRUE) + 2*mad(samdf$energy_ratio, na.rm = TRUE)
median(samdf$energy_ratio, na.rm = TRUE) - 2*mad(samdf$energy_ratio, na.rm = TRUE)
```

```{r}
ggplot(samdf, aes(x = energy_ratio)) +
     geom_histogram(binwidth = 0.2,
                    boundary = 0) + 
     geom_vline(xintercept = 1, color = 'green', linetype = 'dashed') +
     geom_vline(xintercept = 1.755, color = 'red', linetype = 'dashed') +
     geom_vline(xintercept = 0.217, color = 'red', linetype = 'dashed') +
     # facet_wrap(~study) +
     labs(x = 'BMR : reported intake\nkcal ratio',
          y = 'Frequency')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'diet surveys',
            'ffq',
            paste0(plotID, '_BMR to reported intake kcal ratios.pdf')),
       height = 2.5, width = 3)
```

```{r}
# Flag those that fall outside median +/- 2 median absolute deviations
samdf <- 
     samdf %>% 
     mutate(pass_qc = ifelse(energy_ratio <= 1.755695 & energy_ratio >= 0.2165082,
                             yes = TRUE,
                             no = FALSE))

samdf %>% 
     group_by(pass_qc) %>% 
     count()
```
Ok, so this brings us up to 12 people excluded from this analysis.  Does it differ by study?

```{r}
samdf %>% 
     group_by(study, pass_qc) %>% 
     count()
```
```{r}
samdf %>% 
     group_by(pass_qc) %>% 
     count()
```
```{r}
# Percent passing QC (excluding NAs)
1-7/55
```

# Calculate indices

### FVS

In order to calculate, need to do a few things:
- Remove entries that shouldn't be counted (i.e. water)
- Combine food codes duplicated across seasons or settings
- Filter to only those consumed with a frequency greater than once weekly

```{r}
# Simplify columns as we don't need nutrients
fvs <- select(detail, 
              study,
              subj, 
              q_id,
              food_id,
              food_name,
              daily_freq)

fvs
```
```{r}
# Remove water 

# Entries this term shows up in:
filter(fvs,
       grepl(pattern = '[Ww]ater',
             food_name)) %>% 
     pull(food_name) %>% 
     unique()

# Remove
fvs <- filter(fvs,
              !grepl(pattern = '[Ww]ater',
                     food_name))

dim(fvs)
```

```{r}
# How often might a code be repeated?
fvs %>% 
     group_by(subj, food_id, food_name) %>% 
     count() %>% 
     arrange(desc(n))
```
Spot-checked these and even values of up to 6 are reasonable: this would happen if someone added milk to every hot drink, for example. 

What's tricky here is for seasonality, I think I'd want to average, but for different types of "occasions", I'd want to sum. 

```{r}
# Sum
# This is most permissive, not accurate for all cases, but just try for now

fvs <- 
     fvs %>% 
     group_by(study, subj, food_id, food_name) %>% 
     summarize(daily_freq = sum(daily_freq))

fvs
```
Just for fun-- what's the range of daily frequencies?

```{r}
ggplot(fvs, aes(x = daily_freq)) +
     geom_histogram(binwidth = 0.10,
                    boundary = 0)
```
```{r}
# Filter
dim(fvs)
fvs <- filter(fvs,
              daily_freq >= 1/7) # Greater than once weekly
dim(fvs)
```

#### Label plant foods

```{r}
# Save for manual labeling

# fvs %>% 
#      ungroup %>% 
#      select(starts_with('food')) %>% 
#      distinct() %>% 
#      arrange(food_id) %>% 
#      write_csv(here('data',
#                     'processed',
#                     'DHQ3',
#                     'Food ID to plant mapping.csv'))
```

```{r}
# Read back in 
plant.items <- 
     here('data',
          'processed',
          'DHQ3',
          'Food ID to plant mapping.csv') %>% 
     read_csv()
```

```{r}
# Calculate FVS: All food items
fvs.tot <- 
     fvs %>% 
     ungroup() %>% 
     group_by(study, subj) %>% 
     count(name = 'fvs_tot')
```

```{r}
# Calculate FVS: Plant food items
fvs.plant <- 
     fvs %>% 
     left_join(plant.items) %>% 
     ungroup() %>% 
     filter(!is.na(plant)) %>% 
     group_by(study, subj) %>% 
     count(name = 'fvs_plant')

dim(fvs.plant)
```

```{r}
fvs <- 
     full_join(fvs.tot, fvs.plant) %>% 
     mutate(fvs_animal = fvs_tot - fvs_plant)

rm(fvs.tot, fvs.plant)

fvs
```
```{r}
# Pivot longer for plotting
fvs %>% 
     pivot_longer(cols = starts_with('fvs'),
                  names_to = 'score',
                  values_to = 'value') %>% 
     mutate(score = factor(score,
                           levels = c('fvs_tot', 
                                      'fvs_plant',
                                      'fvs_animal'))) %>% 
     ggplot(aes(x = study, y = value, color = score)) +
     geom_boxplot() +
     labs(y = 'Dietary index score', color = 'FVS') +
     # scale_color_manual(values = c('#57606c', '#59a14f'),
     #                    labels = c('Total items',
     #                               'Plant items')) +
     theme(axis.title.x = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             '2',
#             paste0(plotID, '_FVS scores by cohort.pdf')),
#        height = 2.5, width = 4)
```

Mann-Whitney U (Wilcoxon): This “does not estimate the difference in medians (a common misconception) but rather the median of the difference between a sample from x and a sample from y.”

```{r}
# Are the studies significantly different from each other?
wilcox.test(fvs_plant ~ study,
            data = fvs,
            conf.int = TRUE)
```

```{r}
# Are the two measures correlated?
ggplot(fvs, aes(x = fvs_tot, y = fvs_plant)) +
     geom_point() +
     facet_wrap(~study) +
     labs(y = 'FVS\n(plant foods)',
          x = 'FVS\n(all foods)')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results', 
#             'manuscript',
#             '2',
#             paste0(plotID, '_Total FVS x plant FVS.pdf')),
#        height = 2.5, width = 4)
```

Overall, each component score (plant or animal) is much more strongly correlated with the total than it is with the other component.

```{r}
cor.test(fvs$fvs_plant,
         fvs$fvs_tot,
         method = 'spearman') 
```

```{r}
cor.test(fvs$fvs_plant,
         fvs$fvs_animal,
         method = 'spearman') 
```

```{r}
cor.test(fvs$fvs_tot,
         fvs$fvs_animal,
         method = 'spearman') 
```

### DDS

### HEI-2015

```{r}
hei <- select(summary, 
              study, 
              subj,
              matches('HEI'))

names(hei)
```

```{r}
hei <- 
     hei %>% 
     rename(hei_tot = 
                 'Total HEI-2015 Score',
            hei_veg = 
                 'HEI-2015 - Total Vegetables - Component Score',
            hei_greensbeans = 
                 'HEI-2015 - Greens and Beans - Component Score',
            hei_fruit = 
                 'HEI-2015 - Total Fruits - Component Score',
            hei_wholefruits = 
                 'HEI-2015 - Whole Fruits - Component Score',
            hei_grain = 
                 'HEI-2015 - Whole Grains - Component Score',
            hei_dairy = 
                 'HEI-2015 - Dairy - Component Score',
            hei_tot_prot = 
                 'HEI-2015 - Total Protein Foods - Component Score',
            hei_seafood = 
                 'HEI-2015 - Seafood and Plant Proteins - Component Score',
            hei_fa = 
                 'HEI-2015 - Fatty Acids - Component Score',
            hei_sodium = 
                 'HEI-2015 - Sodium - Component Score',
            hei_addsugar = 
                 'HEI-2015 - Added Sugars - Component Score',
            hei_satfat = 
                 'HEI-2015 - Saturated Fats - Component Score') %>% 
     # Make custom plant and non-plant columns
     mutate(hei_plant_all = 
                 hei_veg + 
                 hei_greensbeans + 
                 hei_fruit + 
                 hei_wholefruits + 
                 hei_grain,
            hei_non_plant = 
                 hei_dairy +
                 hei_sodium +
                 hei_addsugar +
                 hei_satfat)

hei
```

### PDI

Nothing to do here: processing already done in external notebooks.

# Join together

```{r}
# Join together results from above
output <- 
     select(hei, 
            study,
            subj,
            hei_plant_all,
            hei_non_plant) %>% 
     full_join(fvs, by = c('subj', 'study')) %>% 
     full_join(select(pdi, study:uPDI), by = c('subj', 'study')) %>% 
     full_join(select(samdf, study, subj, pass_qc, kcal)) %>% 
     select(study, subj, pass_qc, kcal, everything())

output
```

## Try residuals

### FVS

```{r}
ggplot(output, aes(x = kcal, y = fvs_tot)) +
     geom_point() +
     labs(x = 'Energy intake (kcal)',
          y = 'FVS\n(All items)')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             '2',
#             paste0(plotID, '_FVS scores by kcal.pdf')),
#        height = 2.5, width = 3)
```

```{r}
# Linear model (total)
mod_tot <- lm(fvs_tot ~ kcal, data = output)
summary(mod_tot)

# Get input data, for joining back to full results
resids <- mod_tot$model

# Populate with residuals
resids$fvs_tot_residual <- resid(mod_tot)
output <- left_join(output, resids)
```

```{r}
# Linear model (plant items)
mod_plant <- lm(fvs_plant ~ kcal, data = output)
summary(mod_plant)

# Get input data, for joining back to full results
resids <- mod_plant$model

# Populate with residuals
resids$fvs_plant_residual <- resid(mod_plant)
output <- left_join(output, resids)
```

```{r}
# Linear model (plant items)
mod_animal <- lm(fvs_animal ~ kcal, data = output)
summary(mod_animal)

# Get input data, for joining back to full results
resids <- mod_animal$model

# Populate with residuals
resids$fvs_animal_residual <- resid(mod_animal)
output <- left_join(output, resids)
```

### HEI

```{r}
ggplot(output, aes(x = kcal, y = hei_plant_all + hei_non_plant)) +
     geom_point() +
     labs(x = 'Energy intake (kcal)',
          y = 'HEI')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'manuscript',
#             '2',
#             paste0(plotID, '_HEI scores by kcal.pdf')),
#        height = 2.5, width = 3)
```

```{r}
# Linear model (total)
mod_tot <- lm(hei_plant_all + hei_non_plant ~ kcal, data = output)

summary(mod_tot)
```

Also checked components individually, and these aren't significant either. 

### PDI

```{r}
output %>% 
     pivot_longer(cols = matches('pdi'),
                  names_to = 'index',
                  values_to = 'value') %>% 
     mutate(index = factor(index,
                           levels = c('PDI', 'hPDI', 'uPDI'))) %>% 
     ggplot(aes(x = kcal, y = value)) +
     geom_point(alpha = 0.5) +
     facet_wrap(~index) +
     labs(x = 'Energy intake (kcal)',
          y = 'Score') +
     scale_x_continuous(labels = comma) +
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'diet surveys',
            'ffq',
            paste0(plotID, '_PDI scores by kcal.pdf')),
       height = 2.5, width = 5)
```

```{r}
# Linear model (total)
mod_pdi <- lm(PDI ~ kcal, data = output)
summary(mod_pdi)
```

```{r}
# Linear model (plant items)
mod_hpdi <- lm(hPDI ~ kcal, data = output) 
summary(mod_hpdi)

# Get input data, for joining back to full results
resids <- mod_hpdi$model

# Populate with residuals
resids$hPDI_residual <- resid(mod_hpdi)
output <- left_join(output, resids)
```

```{r}
mod_updi <- lm(uPDI ~ kcal, data = output) 
summary(mod_updi)

# Get input data, for joining back to full results
resids <- mod_updi$model

# Populate with residuals
resids$uPDI_residual <- resid(mod_updi)
output <- left_join(output, resids)
```

# Save output

```{r}
write_csv(
     output,
     here('data', 
          'processed',
          'DHQ3', 
          'Adult-1 and Adult-2 dietary indices.csv'
          )
)
```

