---
title: "Co-inertia: ASA24 & DHQ3"
author: "Brianna Petrone"
output: html_notebook
---

# Setup
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, include=FALSE, echo=FALSE}
library(ade4) # For co-inertia functions
library(here)
library(tidyverse)

source(here('..', 'functions', 'top_loadings.R'))
source(here('..', 'functions', 'rand_axis_test.R'))
```

# Load input data

## ASA24

### CHOMP

```{r}
data.dir <- here('data', 'processed', 'ASA24')
list.files(data.dir)

# Nutrients
asa24.nut.chomp <- 
    readRDS(file.path(data.dir, "20200405_Nutrients, kcal-normalized, all records_chomp.rds"))

# Food groups
asa24.foodgroups.chomp <- 
    readRDS(file.path(data.dir, 
                      "20200405_Food groups, kcal-normalized, all records_chomp.rds"))

# Plant food groups
asa24.plantgroups.chomp <- 
    readRDS(file.path(data.dir, 
                      "20200405_Plant food groups, kcal-normalized, all records_chomp.rds"))
```

### ONR

```{r}
# Nutrients
asa24.nut.onr <- 
    readRDS(file.path(data.dir, "20200405_Nutrients, kcal-normalized, all records_onr.rds"))

# Food groups
asa24.foodgroups.onr <- 
    readRDS(file.path(data.dir, 
                      "20200405_Food groups, kcal-normalized, all records_onr.rds"))

# Plant food groups
asa24.plantgroups.onr <- 
    readRDS(file.path(data.dir, 
                      "20200405_Plant food groups, kcal-normalized, all records_onr.rds"))
```

## DHQ3

### CHOMP

#### FNDDS nutrients

```{r}
# All FNDDS nutrients (64)
dhq3.nut.chomp <- 
     here('data', 'processed', 'DHQ3',
          '20201022_DHQ3 kcal-normalized FNDDS nutrients_chomp.rds') %>% 
     readRDS()
```

#### Food groups

```{r} 
# All food groups (41)
dhq3.foodgroups.chomp <- 
     here('data', 'processed', 'DHQ3', 
          '20201022_DHQ3 kcal-normalized food groups_chomp.rds') %>% 
     readRDS()

# Only plant-derived food groups (but NOT including oils, solid fats, sweeteners, or alcohols) (22)
dhq3.plantgroups.chomp <- 
     here('data', 'processed', 'DHQ3', 
          '20201022_DHQ3 kcal-normalized plant food groups_chomp.rds') %>% 
     readRDS()
```

### ONR

#### FNDDS nutrients

```{r}
# All FNDDS nutrients (64)
dhq3.nut.onr <- 
     here('data', 'processed', 'DHQ3',
          '20210305_DHQ3 kcal-normalized FNDDS nutrients_onr.rds') %>% 
     readRDS()
```

#### Food groups

```{r}
# All food groups (41)
dhq3.foodgroups.onr <- 
     here('data', 'processed', 'DHQ3', 
          '20210305_DHQ3 kcal-normalized food groups_onr.rds') %>% 
     readRDS()

# Only plant-derived food groups (but NOT including oils, solid fats, sweeteners, or alcohols) (22)
dhq3.plantgroups.onr <-
     here('data', 'processed', 'DHQ3', 
          '20210305_DHQ3 kcal-normalized plant food groups_onr.rds') %>% 
     readRDS()
```

# Tidy

## Concatenate dataframes

```{r}
# ASA24
asa24.nut <- bind_rows(asa24.nut.chomp, asa24.nut.onr)
asa24.foodgroups <- bind_rows(asa24.foodgroups.chomp, asa24.foodgroups.onr)
asa24.plantgroups <- bind_rows(asa24.plantgroups.chomp, asa24.plantgroups.onr)
```

```{r}
# DHQ3
dhq3.nut <- bind_rows(dhq3.nut.chomp, dhq3.nut.onr)
dhq3.foodgroups <- bind_rows(dhq3.foodgroups.chomp, dhq3.foodgroups.onr)
dhq3.plantgroups <- bind_rows(dhq3.plantgroups.chomp, dhq3.plantgroups.onr)
```

```{r}
# Remove original variables
rm(list = ls()[grep('.chomp$|.onr$', ls())])
```

## Check completeness

```{r}
any(is.na(asa24.nut))
any(is.na(asa24.foodgroups))
any(is.na(asa24.plantgroups))

any(is.na(dhq3.nut))
any(is.na(dhq3.foodgroups))
any(is.na(dhq3.plantgroups))
```

Ok, DHQ3 appears to have missing data:

```{r}
missing <- apply(dhq3.nut, 1, function(x){any(is.na(x))})
which(missing)
```

```{r}
# Remove
dhq3.nut <- dhq3.nut[!missing, ]
dhq3.foodgroups <- dhq3.foodgroups[!missing, ]
dhq3.plantgroups <- dhq3.plantgroups[!missing, ]
```

```{r}
# Confirm
any(is.na(dhq3.nut))
any(is.na(dhq3.foodgroups))
any(is.na(dhq3.plantgroups))
```

## Combine ASA24

### Combine

```{r}
# Create function that averages ASAs within an individual together
asa_average <- function(x){
     # Takes dataframe of ASA24 values, with rownames of format [subj]_[recall],
     # and returns a dataframe with subjects as rows, and all column values 
     # averaged within a subject
     x %>% 
     rownames_to_column(var = 'ID') %>% 
     separate(ID, into = c('ID', 'recall'), sep = '_') %>% 
     group_by(ID) %>% 
     summarize_at(vars(-recall), mean) %>% 
     column_to_rownames(var = 'ID')
}
```

```{r}
# Apply to files
asa24.nut.avg <- asa_average(asa24.nut)
asa24.foodgroups.avg <- asa_average(asa24.foodgroups)
asa24.plantgroups.avg <- asa_average(asa24.plantgroups)
```

## Replicate DHQ3

```{r}
# TODO
```

# Co-inertia

## Subset to shared samples

```{r}
# Shared rows in averaged ASA24 and DHQ3 data
shared.avg <- intersect(row.names(asa24.nut.avg), row.names(dhq3.nut))

# Note missing samples from either set
cat('Subjects who completed an ASA24, but not a DHQ3:\n',
    setdiff(row.names(asa24.nut.avg), row.names(dhq3.nut)), '\n')
cat('Subjects who completed an ASA24, but not a DHQ3:\n',
    setdiff(row.names(dhq3.nut), row.names(asa24.nut.avg)), '\n')
```

```{r}
# Shared rows in ASA24 and replicated DHQ3 data
```

## PCA

Should think about-- is scaling actually appropriate here? 
Need to justify to myself that this is the case.

```{r}
# Ordinate ASA24 input data with PCA
dudi.asa24.nut.avg <- dudi.pca(asa24.nut.avg[shared.avg, ], 
                               center = TRUE, scale = TRUE,
                               scannf = FALSE, nf = 2)

dudi.asa24.foodgroups.avg <- dudi.pca(asa24.foodgroups.avg[shared.avg, ],
                                      center = TRUE, scale = TRUE,
                                      scannf = FALSE, nf = 2)
```

```{r}
# Cumulative relative variation of eigenvalues
cumsum(dudi.asa24.nut.avg$eig / sum(dudi.asa24.nut.avg$eig)) %>% 
     plot()
```

```{r}
# Ordinate DHQ3 input data with PCA
dudi.dhq3.nut <- dudi.pca(dhq3.nut[shared.avg, ], 
                          center = TRUE, scale = TRUE,
                          scannf = FALSE, nf = 2)

dudi.dhq3.foodgroups <- dudi.pca(dhq3.foodgroups[shared.avg, ],
                                 center = TRUE, scale = TRUE,
                                 scannf = FALSE, nf = 2)
```

```{r}
# Cumulative relative variation of eigenvalues 
cumsum(dudi.dhq3.nut$eig / sum(dudi.asa24.nut.avg$eig)) %>% 
     plot()
```
## Co-inertia

### Nutrients

```{r}
# Are the row weights equal?
all.equal(dudi.asa24.nut.avg$lw, dudi.dhq3.nut$lw)
```

```{r}
# Run
coia.nut <- coinertia(dudi.asa24.nut.avg, dudi.dhq3.nut, 
                      scannf = FALSE, nf = 2)

summary(coia.nut)
```

```{r}
# Relative variation on first eigenvalue 
coia.nut$eig[1] / sum(coia.nut$eig) 
```

```{r}
randtest(coia.nut, nrepet = 999) 
```

```{r}
# Plot global data

plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 'co-inertia',
         paste0(plotID,
                '_Global plot, ASA24 nutrients v. DHQ3 nutrients.pdf')),
    height = 20, width = 20)
plot(coia.nut)

# Relevant plotIDs:
# 2021-04-07.359: First pass
```

### Food groups

```{r}
# Are the row weights equal?
all.equal(dudi.asa24.foodgroups.avg$lw, dudi.dhq3.foodgroups$lw)
```

```{r}
# Run
coia.foodgroups <- coinertia(dudi.asa24.foodgroups.avg, dudi.dhq3.foodgroups, 
                             scannf = FALSE, nf = 2)

summary(coia.foodgroups)
```

```{r}
# Relative variation on first eigenvalue 
coia.foodgroups$eig[1] / sum(coia.foodgroups$eig) 
```

```{r}
randtest(coia.foodgroups, nrepet = 999) 
```

```{r}
# Plot global data

plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 'co-inertia',
         paste0(plotID,
                '_Global plot, ASA24 food groups v. DHQ3 food groups.pdf')),
    height = 20, width = 20)
plot(coia.foodgroups)

# Relevant plotIDs:
# 2021-04-07.313: First pass
```