---
title: "Relative abundance plots"
output: html_notebook
---

# Setup

```{r setup, include = FALSE, echo = FALSE}
require(knitr)
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 14,
                                              face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
                     )
)
```

```{r}
# Plotting colors
colors <- 
     here('..', 'food-dbs', 'data', 'processed', 'plotting', 
          'trnL colors.csv') %>% 
     read_csv() %>% 
     select(family, family_color) %>% 
     distinct()

colors.vec <- colors$family_color
names(colors.vec) <- colors$family
# Convert NA to string representation
names(colors.vec)[is.na(names(colors.vec))] <- 'NA'
```

# Read in data

```{r}
# Full trnL phyloseq object
ps <- 
     here('data', 
          'processed', 
          'phyloseq',
          'trnL_combined_ps.rds') %>% 
     readRDS()

ps
```

```{r}
# Rarefied counts
alpha <- 
     here('data', 
          'processed', 
          'rarefaction',
          'Rarefied trnL counts (1,000 reads).rds') %>% 
     readRDS()

dim(alpha)
```

# Pre-process

Extend richness measures to the full cohort, to include samples that were filtered out because of low read counts.

```{r}
# Pull ASV table
asvtab <- otu_table(ps)@.Data

# Calculate richness 
obs <- 
     apply(asvtab, 1, function(x){sum(x > 0)}) %>% 
     data.frame(obs = .) %>% 
     rownames_to_column(var = 'name')
```

```{r}
# Join to read count information
obs <- 
     data.frame(ps@sam_data) %>% 
     rownames_to_column(var = 'row') %>% 
     select(row, study, subj, reads_food) %>% 
     full_join(obs, by = c('row' = 'name')) %>% 
     rename(name = row)
```

```{r}
# Join to rarefied richness
alpha <- 
     alpha %>% 
     select(name, starts_with('obs')) %>% 
     right_join(obs) %>% 
     select(name, study, subj, everything())

rm(obs)
```

```{r}
# Bin read counts
alpha$reads_food_binned <- 
     cut(alpha$reads_food, 
         breaks = c(-1, 500, 1001, 5001, 10001, 30001,
                    max(alpha$reads_food) + 1), 
         labels = c('<500 reads', 
                    '500-1,000', 
                    '1,000-5,000', 
                    '5,000-10,000', 
                    '10,000-30,000',
                    '>30,000'))
```

# Analyze

## Mean trnL richness by bin

```{r}
ggplot(alpha, aes(x = reads_food_binned, y = obs)) +
     geom_boxplot() +
     labs(x = 'Read depth bin', y = 'pMR') +
     theme(axis.text.x = element_text(angle = 20,
                                      hjust = 1))
```

```{r}
ggplot(alpha, aes(x = reads_food_binned, y = obs_rare)) +
     geom_boxplot() +
     labs(x = 'Read depth bin', y = 'Rarefied pMR') +
     theme(axis.text.x = element_text(angle = 20,
                                      hjust = 1))
```

## Example visualization

Idea would be to show samples that come from the same sequencing depth, but vary widely by their trnL richness.

```{r}
samples <- 
     ps %>% 
     subset_samples(reads_tot < 107000 & 
                         reads_tot > 106000) %>% 
     sample_names()

samples
```

```{r}
# Range of richness we'll see
alpha %>% 
     filter(name %in% samples) %>% 
     select(name, starts_with('obs')) %>% 
     arrange(obs)
```

```{r}
# Get melted data for these samples
melt <- 
     ps %>% 
     prune_samples(samples, .) %>%
     psmelt() %>% 
     # Factor samples based on level above
     mutate(Sample = factor(Sample,
                            levels = c('POMMS-1',
                                       'POMMS-244',
                                       'POMMS-315',
                                       'CHOMP-180')))
```

```{r}
# This gets order manually coded below
# melt %>% 
#      filter(Sample %in% c('POMMS-1', 'CHOMP-180')) %>% 
#      filter(Abundance > 0) %>% 
#      arrange(desc(Abundance))
```

```{r}
# Convert variable types for cleaner plotting
melt %>% 
     filter(Sample %in% c('POMMS-1', 'CHOMP-180')) %>% 
     filter(Abundance > 0) %>% 
     # Relevel the 'name' variable
     mutate(name = factor(name,
                          levels = c('Theobroma cacao',
                                     'Rosaceae (strawberry, brambles, rose)',
                                     'Pistacia vera',
                                     'Lamiales',
                                     'Zea mays',
                                     'Capsicum annuum',
                                     'Cocos nucifera',
                                     'Avena sativa',
                                     'Oryza sativa',
                                     'Juglans regia',
                                     'Poaceae (rye, wheat)',
                                     'Coriandrum sativum',
                                     'Piper longum',
                                     'Vaccinium pallidum',
                                     'Rubus idaeus',
                                     'Apiaceae',
                                     'Corylus avellana',
                                     'Allium'),
                          labels = c('Chocolate',
                                     'Rosaceae (strawberry, brambles, rose)',
                                     'Pistachio',
                                     'Lamiales (basil, oregano, lemon balm, \nKorean mint, sage, sesame, thyme, rosemary)',
                                     'Corn',
                                     'Peppers (bell, paprika)',
                                     'Coconut',
                                     'Oat',
                                     'Rice',
                                     'Walnut',
                                     'Poaceae (rye, wheat)',
                                     'Coriander/cilantro',
                                     'Pepper (black)',
                                     'Blueberry',
                                     'Raspberry',
                                     'Apiaceae (carrot, parsnip, fennel,\nparsnip, parsley)',
                                     'Hazelnut', 
                                     'Allium (leek, scallion, chive)'))) %>% 
     ggplot(aes(x = Sample, y = Abundance, fill = name)) +
     geom_bar(stat = 'identity', 
              position = 'stack', 
              color = 'black',
              size = 0.25) +
     scale_fill_manual(values = c('#B66353',
                                  '#D37295',
                                  '#B9CA5D',
                                  '#55AD89',
                                  '#F0BD27',
                                  '#E15759',
                                  '#BAB0AC',
                                  '#FFBE7D',
                                  'white',
                                  '#9D7660',
                                  '#F1CE63',
                                  '#59A14F',
                                  '#79706E',
                                  '#4E79A7',
                                  '#FF9D9A',
                                  '#F28E2B',
                                  '#D7B5A6',
                                  '#8CD17D')) +
     theme(axis.text.x = element_blank(),
           axis.ticks.x = element_blank(),
           legend.position = 'right',
           legend.title = element_blank(),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'abundance plots', 
            paste0(plotID, '_Low and high trnL richness example.pdf')),
       height = 5, width = 6)
```

```{r}
# Convert variable types for cleaner plotting
melt %>% 
     filter(Sample %in% c('POMMS-1', 'CHOMP-180')) %>% 
     filter(Abundance > 0) %>% 
     ggplot(aes(x = Sample, y = Abundance, fill = name)) +
     geom_bar(stat = 'identity', position = 'stack',
              fill = 'white', color = 'black',
              size = 0.25) +
     theme(axis.text.x = element_blank(),
           axis.ticks.x = element_blank(),
           legend.position = 'right',
           legend.title = element_blank(),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'abundance plots', 
            paste0(plotID, '_Low and high trnL richness example (no color).pdf')),
       height = 5, width = 3)
```

## Extremes of intake

Given some results different on either side of a read count threshold, want to understand: Do samples *look* appreciably different at high/low richness?

### Relative abundance

```{r}
# Randomly sample 10 from each bin
set.seed(2021)
samples <- 
     alpha %>% 
     group_by(reads_food_binned) %>% 
     sample_n(size = 10)
```

```{r}
# Get melted data for these samples
melt <- 
     ps %>% 
     prune_samples(samples$name, .) %>%
     psmelt()

# Add binned reads
melt <- 
     alpha %>% 
     select(Sample = name, reads_food_binned) %>% 
     right_join(melt)
```

```{r}
# Convert variable types for cleaner plotting
ggplot(melt, aes(x = Sample, y = Abundance)) +
     geom_bar(stat = 'identity', position = 'fill', 
              color = 'black', fill = 'white') +
     facet_wrap(~reads_food_binned, 
                scales = 'free_x',
                nrow = 1) + 
     # scale_fill_manual(values = colors.vec) +
     theme_classic() + 
     theme(axis.text.x = element_blank(),
           axis.ticks.x = element_blank(),
           legend.position = 'top',
           plot.margin = margin(0.5, 0.5, 0.5, 0.5, "in"))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 'abundance plots',
         paste0(plotID, 
                '_Relative abundances, 10 samples per read count bin.pdf')),
    height = 5, 
    width = 20)

ggplot(melt, aes(x = Sample, y = Abundance)) +
     geom_bar(stat = 'identity', position = 'fill', 
              color = 'black', fill = 'white') +
     facet_wrap(~reads_food_binned, 
                scales = 'free_x',
                nrow = 1) + 
     # scale_fill_manual(values = colors.vec) +
     theme_classic() + 
     theme(axis.text.x = element_blank(),
           axis.ticks.x = element_blank(),
           legend.position = 'top',
           plot.margin = margin(0.5, 0.5, 0.5, 0.5, "in"))

```
Based on this, looks like we accumulate more low-proportion things at higher read depth (>10,000).

Might be nice to color by family to see if there are trends in what turns up over time.

